### 2018-10-03 misc/written_project_proposal.tex
Start project proposal

\documentclass{article}

\usepackage[letterpaper,margin=0.5in]{geometry}
\usepackage{parskip}

\usepackage[backend=bibtex,style=ieee]{biblatex}
\bibliography{written_project_proposal}

\title{Towards a Verified Bzip2: A Functional Specification for Burrows-Wheeler}
\author{Jake Waksbaum '20\\Advisor: Andrew Appel}
\date{}

\begin{document}
\maketitle

\section{Motivation and Goal}
File compression software is often used to reduce the amount of disk
space occupied by archived files and to more efficiently transfer
data. Many types of file compression software, such as bzip2, 7zip,
unzip, have been found vulnerable to security exploits that can be
used in denial-of-service attacks or to enable code execution. Often
these vulnerabilities are the result of programmer errors related to
the management of low-level resources: buffer-overflow,
use-after-free, double-free, etc. These types of errors are
well-addressed using formal verification methods that demand proof
that low-level resources are correctly managed.

The goal of my project is to begin the work of producing a formally
verified implementation of the popular and open-source bzip2
compression algorithm. A formally verified compression and
decompression program won't suffer from the types of security exploits
that are so common in existing implementations.

\section{Problem Background and Related Work}
Coq is a theorem proving assistant that allows programmers to write
machine-checked proofs: proofs are verified by Coq to be sound. It
also includes a functional programming language called Gallina that
can be used to produce functional specifications\cite{7536361} that
are useful for describing the desired behavior of some C program.

The CompCert
compiler\cite{Stewart:2015:CC:2676726.2676985,leroy:inria-00000963} is
an optimizing C compiler, that has been formally verified to preserve
semantics: this means that the semantics of the executable code with
resepect to the semantics of the assembly language is the same as the
semantics of the source code with respect to the semantics of the
source langage, C. This allows us to have confidence that the
properties that we prove about our C programs also hold on the
compiled executables.

The Verified Software
Toolchain\cite{Appel:2011:VST:1987211.1987212,Appel:2014:PLC:2670099}
includes also a program logic for C called Verifiable C that is proved
sound with respect to the operational semantics of Compcert C. This
gives us tools in Coq to reason about our C programs to prove
properties such as safety. It includes
VST-Floyd\cite{cao2018vst-floyd:}, which helps automate some parts of
the the process of building functional correctness proofs of C
programs.

These tools were used to formalize the functional specifications of
the mbedTLS HMAC-DRBG, to prove that the specification truly produces
pseuodrandom output, and then to produce an executable for which that
result also holds\cite{Ye:2017:VCS:3133956.3133974}.

\section{Approach}
I plan to work towards a formally verified implementation of bzip2 by
implementing functional specifications for the core compression and
decompression passes that make up the bzip2 stack: run-length
encoding, Burrows-Wheeler, Huffman encoding, etc. This means writing a
functional program in Gallina that implements those algorithms, and
proving useful properties about those algorithms such as that the
decoder inverts the encoder. I will focus on the functional
specification of the algorithims as opposed to the verification of C
implementations.

\section{Plan}
I will start by working on a functional specification for run-length
encoding, as it is simple and the first phase of compression. Then, I
will work on a functional specification for Burrows-Wheeler. Depending
on how long that takes, I can move on to the rest of the compression
stack, implementing and proving theorems about as much of it as I can.

\section{Evaluation}
My main goal is to produce a functional specification for
Burrows-Wheeler and a proof that the decoder inverts the encoder. A
stretch goal would be to verify the entire bzip2 compression stack,
and an even bigger stretch goal would be to produce an entire Coq
program that can operate on real bzip2 zip files.

\printbibliography{}

\end{document}

### 2018-10-03 project_proposal/written_project_proposal.tex
Finish written proposal

--- 2018-10-03 misc/written_project_proposal.tex
@@
 space occupied by archived files and to more efficiently transfer
-data. Many types of file compression software, such as bzip2, 7zip,
-unzip, have been found vulnerable to security exploits that can be
-used in denial-of-service attacks or to enable code execution. Often
-these vulnerabilities are the result of programmer errors related to
-the management of low-level resources: buffer-overflow,
-use-after-free, double-free, etc. These types of errors are
-well-addressed using formal verification methods that demand proof
-that low-level resources are correctly managed.
+data. Many types of file compression software, such as
+bzip2\cite{bzip2}, 7zip\cite{7zip}, unzip\cite{infozip}, and
+WinRAR\cite{winrar} have been found vulnerable to security exploits
+that can be used in denial-of-service attacks or to enable arbitrary
+code execution. Often these vulnerabilities are the result of
+programmer errors related to the management of low-level resources:
+buffer-overflow, use-after-free, double-free, etc. These types of
+errors are well-addressed using formal verification methods that
+demand proof that low-level resources are correctly managed.
 
@@
 decompression program won't suffer from the types of security exploits
-that are so common in existing implementations.
+common to existing implementations.
 
@@
 compiler\cite{Stewart:2015:CC:2676726.2676985,leroy:inria-00000963} is
-an optimizing C compiler, that has been formally verified to preserve
-semantics: this means that the semantics of the executable code with
-resepect to the semantics of the assembly language is the same as the
-semantics of the source code with respect to the semantics of the
-source langage, C. This allows us to have confidence that the
+an optimizing C compiler, that has been formally verified in Coq to
+preserve semantics: this means that the semantics of the executable
+code with respect to the semantics of the assembly language are the
+same as the semantics of the source code with respect to the semantics
+of the source language, C. This allows us to have confidence that the
 properties that we prove about our C programs also hold on the
@@
 includes also a program logic for C called Verifiable C that is proved
-sound with respect to the operational semantics of Compcert C. This
-gives us tools in Coq to reason about our C programs to prove
-properties such as safety. It includes
+sound with respect to the operational semantics of CompCert C. This
+gives us tools in Coq to reason about our C programs and to prove
+properties such as correctness and safety. It includes
 VST-Floyd\cite{cao2018vst-floyd:}, which helps automate some parts of
-the the process of building functional correctness proofs of C
-programs.
+the process of building functional correctness proofs of C programs.
 
@@
 decoder inverts the encoder. I will focus on the functional
-specification of the algorithims as opposed to the verification of C
+specification of the algorithms as opposed to the verification of C
 implementations.
@@
 on how long that takes, I can move on to the rest of the compression
-stack, implementing and proving theorems about as much of it as I can.
+stack, implementing and proving theorems about as much of it as I can
+in the time that I have. Each component of the stack is easy to treat
+separately and can be proved correct independent of the other
+components.
 
@@
 and an even bigger stretch goal would be to produce an entire Coq
-program that can operate on real bzip2 zip files.
+program that can operate on real bzip2 zip files. Evaluating whether
+or not I have completed the proof is pretty simple: Coq will reject
+any incorrect proofs of my theorems.
 

### 2019-02-08 thesis/thesis.tex
Write thesis draft

\documentclass[11pt]{article}

\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}

\usepackage{amsthm}
\usepackage{thmtools}
\declaretheorem[style=plain,numberwithin=section]{theorem}
\declaretheorem[style=plain,sibling=theorem]{proposition}
\declaretheorem[style=plain,sibling=theorem]{lemma}
\declaretheorem[style=plain,sibling=theorem]{corollary}

\declaretheorem[style=definition,numberwithin=section]{definition}
\declaretheorem[style=definition,numberwithin=section]{example}
\declaretheorem[style=definition,numberwithin=section]{remark}

\setlength{\parskip}{1em}
\setlength{\parindent}{0em}
\usepackage[letterpaper,margin=1in]{geometry}

\author{Jake Waksbaum}
\date{\today}
\title{Thesis}
\hypersetup{
 pdfauthor={Jake Waksbaum},
 pdftitle={Thesis},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 26.1 (Org mode 9.1.9)},
 pdflang={English}}

\begin{document}

\maketitle
\tableofcontents

\section{Introduction}
\label{sec:intro}

The Burrows-Wheeler transform (BWT) is a reversible transformation
that makes a string more amenable to compression by other
methods \cite{bw}. The forwards-transform is straightforward to
describe: for a string of length \(n\), form the \(n \times n\) matrix of
all cyclic shifts of the original string, sort it lexicographically,
and take the last column of the matrix. In addition, we note that
index where the original string appears in the sorted rotation matrix.
As shown in \autoref{fig:bw_ex}, applying the algorithm to the string
\verb|abracadabra!| produces the new string \verb|ard!rcaaabb| and the
index \verb|3|. The tendency of the transformation to group the same
character together is what makes the resulting restring easier to
compress. Although it is not at all obvious, the transformed string,
together with the index, provide enough information to recover the the
original string. BWT is discussed more in \autoref{sec:bwt}.

\begin{figure}
  \centering
  \begin{tabular}{rc}
    0  & abracadabra! \\
    1  & bracadabra!a \\
    2  & racadabra!ab \\
    3  & acadabra!abr \\
    4  & cadabra!abra \\
    5  & adabra!abrac \\
    6  & dabra!abraca \\
    7  & abra!abracad \\
    8  & bra!abracada \\
    9  & ra!abracadab \\
    10 & a!abracadabr \\
    11 & !abracadabra
  \end{tabular}
  $\rightarrow$
  \begin{tabular}{rc}
    0  & !abracadabr\textbf{a} \\
    1  & a!abracadab\textbf{r} \\
    2  & abra!abraca\textbf{d} \\
    \textit{3} & \textit{abracadabra\textbf{!}} \\
    4  & acadabra!ab\textbf{r} \\
    5  & adabra!abra\textbf{c} \\
    6  & bra!abracad\textbf{a} \\
    7  & bracadabra!\textbf{a} \\
    8  & cadabra!abr\textbf{a} \\
    9  & dabra!abrac\textbf{a} \\
    10 & ra!abracada\textbf{b} \\
    11 & racadabra!a\textbf{b}
  \end{tabular}
  \caption{BWT acting on \texttt{abracadabra!}}
  \label{fig:bw_ex}
\end{figure}

Bird and Mu \cite{birdmu,pearls} analyze the Burrows-Wheeler algorithm
in a functional setting by expressing the algorithm as a composition
of many smaller functions. Then, they derive an expression for the
inverse-transform from the specification that it invert the
forwards-transform. In addition to providing an intuitive derivation
of the inverse-transform, this provides the outline of a proof of the
correctness of the Burrows-Wheeler algorithm. We give our version of
their derivation in \autoref{sec:deriv}

Our contribution is a formalization in the Coq proof assistant of the
derivations in \cite{birdmu,pearls}. This is useful in its own right
as a functional specification for a potential verified implementation
of a program that relies on Burrows-Wheeler \cite{appel-func-spec}.
More generally, this provides evidence that Bird's approach of
program calculation, in which programs are derived from their
specifications via algebraic manipulations, is useful for proving the
correctness of algorithms in a machine-checked setting. Bird has
applied this approach to greedy algorithms, the Boyer-Moore algorithm,
the Knuth-Morris-Pratt algoirthm, Sudoku solvers, arithmetic coding,
and the Schorr-Waite algorithm, the Johnson-Trotter algorithm, and
many more problems throughout his career \cite{pearls}. Finally, as a
part of this proof we develop a theory of stable sorts, proving
equivalence between different definitions of sorted lists and stable
sorts, as well as the equivalence of all stable sort algorithms to
each other.

\section{Burrows-Wheeler}
\label{sec:bw}

The Burrows-Wheeler transform was originally introduced as a pass in a
compression algorithm. First the input is transformed with BWT, which
tends to group together characters. The intuition for why this happens
is that all adjacent characters in the output string are followed in
the input string by characters that are adjacent in the sorted first
column. If there is any correlation between a character and the
character preceeding it, bringing together the characters in the first
column will tend to bring together similar characters in the last
column.

The next pass is a Move-to-Front (MTF) encoding, in which a character
is represented by the number of different characters encountered since
the last occurence of this character. This creates many runs of zeroes
from the runs in the output of BWT. On English texts, often 50-60\% of
the values produced by the MTF pass are zeroes
\cite{fenwick2007,bw-analysis}. Often a Run-Length Encoding (RLE) pass
is then used to encode only the runs of zeroes with their lengths,
using a bijective base-2 encoding \cite{bw-analysis, tsai_2016}.
Finally, some sort of zeroth-order coder like a Huffman coder is used.

BWT is also at the heart of algorithms that exploit the relationship
between the rotation matrix in BWT and the suffix array data structure
in order to search for patterns in a compressed text
\cite{ferragina_index}.

\section{The Derivation}
\label{sec:deriv}
In this section, we give an overview of the derivation without delving
into the proofs of the lemmas. We only derive the simples version of
the algorithm, although the derivations in \cite{birdmu,pearls}
provide additional optimizations that bring the final implementation
closer to the imperative implementation in \cite{bw} and other texts.

\subsection{Forwards-Transform}
\label{subsec:ft}
In general, the Burrows-Wheeler transform can be applied to lists of
any type that can be sorted, so we define our functions to operate on
\verb|list A|. We will discuss the sortable constraint in
Section~\autoref{sec:sort_stable}.

First, we define left and right rotations of a list.
\begin{verbatim}
Definition lrot (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: tl => tl ++ [hd]
  end.

Definition rrot (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: tl => last l hd :: init l
  end.
\end{verbatim}

Now we can produce the rotation matrix by repeatedly applying \verb|lrot|.
\begin{verbatim}
Fixpoint iter (f : A -> A) (n : nat) (z : A) : list A :=
  match n with
  | O   => []
  | S m => z :: iter f m (f z)
  end.

Definition rots (l : list A) : list (list A) :=
  iter lrot (length l) l.
\end{verbatim}

Finally, we produce the transformed string by taking the last column
of the sorted rotation matrix. \verb|lexsort| sorts the rotation matrix
lexicographically; it's precise implementation is not important.
\begin{verbatim}
Definition bwp (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: _ => List.map (fun x => last x hd) (lexsort (rots l))
  end.
\end{verbatim}

We also need to find the index in the sorted rotation matrix where the
original string appears.
\begin{verbatim}
Fixpoint findIndex (x : A) (ls : list A) : nat :=
  match ls with
  | [] => 0
  | hd :: tl =>
    match x == hd with
    | left _ => 0
    | right _ => S (findIndex x tl)
    end
  end.
\end{verbatim}
\begin{verbatim}
Definition bwn (l : list A) : nat :=
  findIndex l (lexsort (rots l)).
\end{verbatim}

Without proof, we note that we should have for any nonempty
\verb|x : list A| and \verb|d : A|,
\begin{verbatim}
nth (bwn x) (lexsort (rots x)) d = x
\end{verbatim}

\subsection{Inverse-Transform}
\label{subsec:invt}

Now we have to construct a function \verb|unbwt| such that
\begin{verbatim}
unbwt (bwn xs) (bwp xs) = xs.
\end{verbatim}
If we could recreate the entire sorted rotation matrix from
\verb|bwp x|, it woiuld be simple to recover \verb|x| by indexing into
the matrix at \verb|bwn x|. That is, we would like a function
\verb|recreate| such that
\begin{verbatim}
recreate (bwp l) = lexsort (rots l).
\end{verbatim}

Then, we could define \verb|unbwt| as
\begin{verbatim}
Definition unbwt (i : nat) (l : list A) : list A :=
  nth i (recreate l) l.
\end{verbatim}
and we would have the correctness of \verb|unbwt| as a result of the
correctness of \verb|bwn|.

\subsubsection{\texttt{recreate}}
We will recreate the sorted rotation matrix column by column, so
\verb|recreate j| will recreate the first \verb|j| columns of the sorted
rotation matrix.
\begin{verbatim}
Definition cols (j : nat) (l : list A) : list A := map (firstn j) l.

Theorem recreate_correct_ind : forall j l,
    j <= length l ->
    recreate j (bwp l) = cols j (lexsort (rots l)).
\end{verbatim}

To derive an implementation of \verb|recreate| from this correctness
specification, we need an expression for \verb|recreate 0| and
\verb|recreate (S j)|. We can substitute for \verb|j| on the
right-hand side and manipulate that expression algebraically until it
is in the form \verb|_ (bwp l)|. Then we can define \verb|recreate 0|
and \verb|recreate (S j)| to be precisely the \verb|_|, and by
running our derivation in reverse, we obtain a proof for the base and
inductive cases that uses only rewrites.

We start with the base case, \verb|recreate 0|:
\begin{verbatim}
  cols 0 (lexsort (rots l))
= { by unfold cols }
  map (firstn 0) (lexsort (rots l))
= { by firstn_O}
  map (const []) (lexsort (rots l))
= { by map_const }
  repeat [] (length (lexsort (rots l)))
= { by bwp_length, rots_length, lexsort_length }
  repeat [] (length (bwp l))
= { by map_const }
  map (const []) (bwp l)
\end{verbatim}

From this, we can define \verb|recreate 0 l = map (const []) l| and
we already have the proof for the base case of the correctness
specification.

For the inductive case, we have the expression
\begin{verbatim}
  cols (S j) (lexsort (rots l))
\end{verbatim}
on the right-hand side. Our plan is to express \verb|cols (S j)|
in terms of \verb|cols j|, and then to use the inductive hypothesis
\begin{verbatim}
  cols j (lexsort (rots l)) = recreate j
\end{verbatim}
to produce a well-formed recursive definition.

One way to do this is to say that taking the first \verb|S j| columns
of a matrix is the same as moving the first column to the back, taking
the first \verb|j| columns, and then sticking the last column back on
the front:
\begin{verbatim}
  cols (S j) l = prepend_col (map last (map lrot l)) (cols j (map lrot l))
\end{verbatim}
If we substitute \verb|map rrot l| for \verb|l| we find a simpler
expression
\begin{verbatim}
  cols (S j) (map rrot l) = prepend_col (map last l) (cols j l)
\end{verbatim}

Now we would like to find a way of introducing a \verb|map rrot|
before the \verb|lexsort (rots l)|. The insight here is that we can
assume that \verb|lexsort| is actually a radix-sort. That is, that
\begin{verbatim}
lexsort xs = rep (length xs) (hdsort ∘ map rrot) xs
\end{verbatim}
where \verb|hdsort| sorts the matrix stably on the the first column
and \verb|rep| is defined as
\begin{verbatim}
Fixpoint rep (f : A -> A) (n : nat) (z : A) : A :=
  match n with
  | O => z
  | S m => f (rep f m z)
  end.
\end{verbatim}

We can do this because there is only one way to sort the matrix, so
assuming \verb|lexsort| and the radix-sort are correct, they must be
doing the same thing. We will specify precisely when and why this is
valid later.

Stepping back, it is important to say that this insight, whether
explicitly stated or not, is always the crucial step in inverting the
Burrows-Wheeler transform. Because radix-sort only needs one column at
a time, this transformation will allow us to reconstruct the sorted
rotation matrix one column at a time.

In seeking to introduce an extra \verb|map rrot| term, we can note
that rotating and sorting an extra time doesn't change anything when
we are sorting the rotation matrix. This works because rotating the
last column to the front produces a matrix that is sorted on its last
\verb|n - 1| characters, so sorting stably on the first column will
re-sort the matrix. Crucially though, rotating the last column
forwards also permutes the entire matrix, so in addition to having a
sorted list, we have a sorted list of the same elements.

More precisely, we can characterize how \verb|map rrot| permutes
\verb|rots|
\begin{verbatim}
map rrot (rots x) = rrot (rots x)
\end{verbatim}

Using that we can show
\begin{verbatim}
  lexsort (rots x)
= { because rrot permutes its elements }
  lexsort (rrot (rots x))
= { because hdsort permutes its elements }
  lexsort (hdsort (rrot (rots x)))
= { radix-sort }
  rep n (hdsort ∘ rrot) (hdsort (rrot (rots x)))
= { rep_absorb_r }
  rep (S n) (hdsort ∘ rrot) (rots x)
= { rep_split_l }
  hdsort (rrot (rep n (hdsort ∘ rrot) (rots x)))
= { radix-sort }
  hdsort (rrot (lexsort (rots x)))
\end{verbatim}

From here we can calculate
\begin{verbatim}
  cols (S j) (lexsort (rots l))
= cols (S j) (hdsort (map rrot (lexsort (rots l))))
= { hdsort commutes with cols (S j) }
  hdsort (cols (S j) (map rrot (lexsort (rots l))))
= { reduce cols (S j) }
  hdsort (prepend_col (map last (lexsort (rots l))) (cols j (lexsort (rots l)))
= { by IH and def of bwp }
  hdsort (prepend_col (bwp l) (recreate j (bwp l)))
= (fun x => hdsort (prepend_col x (recreate j x))) bwp
\end{verbatim}

We can use this to define \verb|recreate|:
\begin{verbatim}
Fixpoint recreate (j : nat) (l : list A) : list (list A) :=
  match j with
  | O    => map (const []) l
  | S j' => hdsort (prepend_col l (recreate j' l))
  end.
\end{verbatim}

\section{Proving the Lemmas}
\label{sec:lemmas}

In this section we prove the lemmas from the previous section:
\begin{enumerate}
\item
\begin{verbatim}
map rrot (rots x) = rrot (rots x)
\end{verbatim}

\item \verb|hdsort| commutes with \verb|cols (S j)|
  This proof is cool because in theory it can be proven via
  parametricity, but that's a metatheorem in Coq.

\item
\begin{verbatim}
cols (S j) (map rrot l) = prepend_col (map last l) (cols j l)
\end{verbatim}
\end{enumerate}

\section{Sorting \& Stability}
\label{sec:sort_stable}

\subsection{Sorting}
\label{subsec:sorting}
What do we mean by the constraint that \verb|A| is sortable? We say
that a type must be equppied with a total, decidable preorder in order
to be sorted.
\begin{verbatim}
Class Preord (A : Type):=
  { le : A -> A -> Prop;
    le_trans : forall x y z, le x y -> le y z -> le x z;
    le_total : forall x y, le x y \/ le y x;
    le_dec : forall x y, {le x y} + {~le x y};
  }.
\end{verbatim}
The type of a sorting algorithm will be
\begin{verbatim}
sort : forall (A : Type), Preord A -> list A -> list A
\end{verbatim}

A correct sorting algorithm should satisfy
\begin{verbatim}
forall x, Sorted (sort x) /\ Permutation x (sort x)
\end{verbatim}
where we define the \verb|Sorted| relation on lists:
\begin{verbatim}
Inductive Sorted: list A -> Prop :=
| Sorted_nil:
    Sorted nil
| Sorted_cons: forall hd tl,
    (forall x, In x tl -> le hd x) ->
    Sorted tl ->
    Sorted (hd :: tl).
\end{verbatim}

We can also have defined \verb|Sorted| like this
\begin{verbatim}
Inductive LocallySorted : list A -> Prop :=
| LSorted_nil : LocallySorted nil
| LSorted_cons1 a : LocallySorted (a :: nil)
| LSorted_consn a b l :
    LocallySorted (b :: l) -> le a b -> LocallySorted (a :: b :: l).
\end{verbatim}

\begin{theorem}[Sorted-LocallySorted Equivalent]
\begin{verbatim}
forall l, Sorted l <-> LocallySorted l.
\end{verbatim}
\end{theorem}

\begin{proof}
  First we prove the forwards direction by induction on the evidence
  \verb|Sorted l|. For the base case, we have \verb|Lsorted_nil|. In
  the inductive case, we consider when \verb|tl| is empty and when it
  is \verb|b :: tl|. If it is empty, we have \verb|Lsorted_cons1 hd|.
  If it is not, we have \verb|LocallySorted (b :: tl)| by the
  inductive hypothesis. We also have that
  \verb|forall x, In x (b :: tl) -> le hd x)|, and because clearly
  \verb|In b (b :: tl)| we have \verb|le hd b|. This means we can
  apply \verb|Lsorted_consn| and we are done.

  Now we prove the backwards direction by induction on the evidence
  \verb|LocallySorted l|. The nil case is given by \verb|Sorted_nil|,
  and the singleton case is given by the \verb|Sorted_cons| followed
  by \verb|Sorted_nil|, using the fact that \verb|In x []| is
  \verb|False|. In the inductive case we have \verb|Sorted (b :: tl)|
  by the inductive hypothesis, which implies \verb|Sorted tl|. It
  suffices to show \verb|forall x, In x (b :: tl) -> le a x|.
  \verb|Sorted (b :: tl)| implies that
  \verb|forall x, In x tl -> le b x|, and we have that \verb|le a b|,
  so by the transitive property of \verb|le| we are done.
\end{proof}

We could also define sorted based on indexes:
\begin{verbatim}
Definition IndexSorted (l : list A) := forall i j d,
    i <= j < length l ->
    le (nth i l d) (nth j l d).
\end{verbatim}

\begin{theorem}[Sorted-IndexSorted Equivalent]
\begin{verbatim}
forall l, Sorted l <-> IndexSorted l.
\end{verbatim}
\end{theorem}

\begin{theorem}[Permutation-IndexPermutation Equivalent]
  The built-in \verb|Permutation xs ys| relation is equivalent to saying
  that there exists some permutation function \verb|p : 'S_n| such that
\begin{verbatim}
forall i : 'I_n, nth i xs = nth (p i) ys
\end{verbatim}
\end{theorem}

\subsection{Stability}
Any preorder induces a decidable equivalence relation:
\begin{verbatim}
Definition eqv {A} `{Preord A} (x y : A) : Prop := le x y /\ le y x.
\end{verbatim}
A stable sort is a sort that preserves the relative order of
equivalent elements. We can state this naturally in terms of the
permutations. We say that a permutation \verb|p| is stable on a list
\verb|l| if
\begin{verbatim}
forall i j : 'I_n, i < j -> eqv (nth i l) (nth j l) -> p i < p j
\end{verbatim}
Then, a sort is stable if for all \verb|l|, there exists a stable
permutation between \verb|l| and \verb|sort l|.

We can also say that two lists are stably related if
\begin{verbatim}
forall x, filter (eqv_dec x) l = filter (eqv_dec x) l'
\end{verbatim}

\begin{theorem}[PermStable-FilterStable Equivalent]
\begin{verbatim}
forall l,
  PermStable l l' <-> (FilterStable l l' /\ Permutation l l')
\end{verbatim}
\end{theorem}

Stability is important because we can prove that there is only one way
to stably sort any list. This also means that any two correct stable
sorting algorithms must agree always, and can be substituted for one another.

\begin{theorem}[StableUnique]
\begin{verbatim}
forall l l',
  Sorted l' -> Permutation l l' -> Stable l l' -> l = l'
\end{verbatim}
\end{theorem}

If we have a antisymmetric total preorder, also known simply as a
total order, then we have the important result that all sorts are
stable sorts. Antisymmetricity means that
\begin{equation*}
  \forall x y, x \leq y \wedge y \leq x \implies x = y
\end{equation*}
or in terms of equivalence
\begin{equation*}
  \forall x y, x \equiv y \implies x = y
\end{equation*}

\begin{theorem}[OrderStable]
\begin{verbatim}
forall (O : Preorder A) (l l' : A),
  (forall x y, eqv x y -> x = y) ->
  Permutation l l' -> Sorted l -> Sorted l' -> Stable l l'
\end{verbatim}
\end{theorem}

We require that the type \verb|A| be equipped with a decidable total
order. If \verb|A| has a decidable total order, then the lexicographic
preorder on \verb|list A| is also an order. This means that any
sorting algorithm using lexicographic order is stable, including
radix-sort. This is what allows us to conclude that \verb|lexsort| is
equivalent to radix-sort.

\begin{theorem}[RadixSort Correct]
\begin{verbatim}
forall x,
  Sorted (radixsort x) /\ Permutation x (radixsort x)
\end{verbatim}
\end{theorem}

\nocite{*}

\bibliography{thesis}{}

\bibliographystyle{plain}

\end{document}

### 2019-03-02 written_project_proposal/written_project_proposal.tex
Reorganize build system to use nix

--- 2018-10-03 project_proposal/written_project_proposal.tex
@@
 
+% Paper size and margins
 \usepackage[letterpaper,margin=0.5in]{geometry}
+
+% Use spaces between paragraphs
 \usepackage{parskip}
 
-\usepackage[backend=bibtex,style=ieee]{biblatex}
+% No numbers in section titles
+\usepackage{titlesec}
+\titlelabel{}
+
+% Bibliography
+\usepackage[style=ieee]{biblatex}
 \bibliography{written_project_proposal}
@@
 Coq is a theorem proving assistant that allows programmers to write
-machine-checked proofs: proofs are verified by Coq to be sound. It
+machine-checked proofs: proofs are verified by Coq to be valid. It
 also includes a functional programming language called Gallina that

### 2019-05-01 draft_paper/draft_paper.tex
Reorganize build system to use nix
Start final paper

--- 2019-02-08 thesis/thesis.tex
@@
-\documentclass[11pt]{article}
+\documentclass[11pt]{thesis}
 
-\usepackage{graphicx}
-\usepackage{grffile}
-\usepackage{longtable}
-\usepackage{wrapfig}
-\usepackage{rotating}
-\usepackage[normalem]{ulem}
-\usepackage{amsmath}
-\usepackage{textcomp}
-\usepackage{amssymb}
-\usepackage{capt-of}
-\usepackage{hyperref}
-
-\usepackage{amsthm}
-\usepackage{thmtools}
-\declaretheorem[style=plain,numberwithin=section]{theorem}
-\declaretheorem[style=plain,sibling=theorem]{proposition}
-\declaretheorem[style=plain,sibling=theorem]{lemma}
-\declaretheorem[style=plain,sibling=theorem]{corollary}
-
-\declaretheorem[style=definition,numberwithin=section]{definition}
-\declaretheorem[style=definition,numberwithin=section]{example}
-\declaretheorem[style=definition,numberwithin=section]{remark}
-
-\setlength{\parskip}{1em}
-\setlength{\parindent}{0em}
-\usepackage[letterpaper,margin=1in]{geometry}
-
+\title{Towards a Verified Bzip2: A Functional Specification for Burrows-Wheeler}
 \author{Jake Waksbaum}
-\date{\today}
-\title{Thesis}
-\hypersetup{
- pdfauthor={Jake Waksbaum},
- pdftitle={Thesis},
- pdfkeywords={},
- pdfsubject={},
- pdfcreator={Emacs 26.1 (Org mode 9.1.9)},
- pdflang={English}}
+\date{May 6, 2019}
 
@@
 \maketitle
-\tableofcontents
 
@@
 
-The Burrows-Wheeler transform (BWT) is a reversible transformation
-that makes a string more amenable to compression by other
-methods \cite{bw}. The forwards-transform is straightforward to
-describe: for a string of length \(n\), form the \(n \times n\) matrix of
-all cyclic shifts of the original string, sort it lexicographically,
-and take the last column of the matrix. In addition, we note that
-index where the original string appears in the sorted rotation matrix.
-As shown in \autoref{fig:bw_ex}, applying the algorithm to the string
-\verb|abracadabra!| produces the new string \verb|ard!rcaaabb| and the
-index \verb|3|. The tendency of the transformation to group the same
-character together is what makes the resulting restring easier to
-compress. Although it is not at all obvious, the transformed string,
-together with the index, provide enough information to recover the the
-original string. BWT is discussed more in \autoref{sec:bwt}.
+The Burrows-Wheeler transform (BWT) is an invertible transformation
+that makes a string more amenable to compression by other methods
+\cite{bw}. The forwards transform is straightforward to describe: for
+a string of length \(n\), form the \(n \times n\) rotaion matrix of all
+cyclic shifts of the original string, sort it lexicographically, and
+take the last column of the matrix. In addition, we note that index
+where the original string appears in the sorted rotation matrix. As
+shown in \cref{fig:bw_ex}, applying the algorithm to the string
+\texttt{abracadabra!} produces the new string \texttt{ard!rcaaabb} and
+the index \texttt{3}. The transformation tends to produce runs like
+the three \texttt{a}s or the two \texttt{b}s, and it is this feature
+that makes the resulting string easier to compress. Although it is not
+at all obvious, the transformed string, together with the index,
+provide enough information to recover the the original string. BWT is
+discussed more in \cref{sec:bwt}.
 
@@
   \centering
+  \begin{tt}
   \begin{tabular}{rc}
@@
   \end{tabular}
+  \end{tt}
   \caption{BWT acting on \texttt{abracadabra!}}
@@
 Bird and Mu \cite{birdmu,pearls} analyze the Burrows-Wheeler algorithm
-in a functional setting by expressing the algorithm as a composition
-of many smaller functions. Then, they derive an expression for the
-inverse-transform from the specification that it invert the
-forwards-transform. In addition to providing an intuitive derivation
-of the inverse-transform, this provides the outline of a proof of the
+in a functional setting. They express the algorithm as the composition
+of many smaller functions, and then derive an expression for the
+inverse transform from the specification that it inverts the forwards
+transform. In addition to providing an intuitive derivation of the
+inverse transform, this provides the outline of a proof of the
 correctness of the Burrows-Wheeler algorithm. We give our version of
-their derivation in \autoref{sec:deriv}
+their derivation in \cref{sec:forwards_BWT,sec:inverse_BWT}.
 
-Our contribution is a formalization in the Coq proof assistant of the
-derivations in \cite{birdmu,pearls}. This is useful in its own right
-as a functional specification for a potential verified implementation
-of a program that relies on Burrows-Wheeler \cite{appel-func-spec}.
-More generally, this provides evidence that Bird's approach of
-program calculation, in which programs are derived from their
-specifications via algebraic manipulations, is useful for proving the
-correctness of algorithms in a machine-checked setting. Bird has
-applied this approach to greedy algorithms, the Boyer-Moore algorithm,
-the Knuth-Morris-Pratt algoirthm, Sudoku solvers, arithmetic coding,
-and the Schorr-Waite algorithm, the Johnson-Trotter algorithm, and
-many more problems throughout his career \cite{pearls}. Finally, as a
-part of this proof we develop a theory of stable sorts, proving
-equivalence between different definitions of sorted lists and stable
-sorts, as well as the equivalence of all stable sort algorithms to
-each other.
+Our main contribution is a formalization in the Coq proof assistant of
+the BWT, together with a machine-checked proof that our implementation
+is correct. Specifically, we implement the functions
+$\mathtt{bwp} : \mathtt{list}~A \to \mathtt{list}~A$ and
+$\mathtt{bwn} : \mathtt{list}~A \to \mathbb{N}$ to produce the string
+and index part of the BWT respectively, and the function
+$\mathtt{unbwt} : \mathbb{N} \to \mathtt{list}~A \to \mathtt{list}~A$ to
+reverse the transform, and we prove the following theorem:
+\begin{theorem}[unbwt\_correct]
+  \label{thm:unbwt_correct}
+  For all lists $l$,
+  \begin{equation*}
+    \mathtt{unbwt}~(\mathtt{bwn}~l)~(\mathtt{bwp}~l) = l
+  \end{equation*}
+\end{theorem}
+The implementation of the functions and the outline of the proof are
+based on Bird et~al.'s derivations.
+
+The implementation is useful in its own right as a functional
+specification for a potential verified implementation of a program
+that relies on Burrows-Wheeler \cite{appel-func-spec}. But more
+generally, this provides evidence that Bird's approach of program
+calculation, in which programs are derived from their specifications
+via algebraic manipulations, is useful for proving the correctness of
+algorithms in a machine-checked setting. Bird has applied this
+approach to greedy algorithms, the Boyer-Moore algorithm, the
+Knuth-Morris-Pratt algoirthm, Sudoku solvers, arithmetic coding, and
+the Schorr-Waite algorithm, the Johnson-Trotter algorithm, and many
+more problems \cite{pearls}.
+
+As a part of the proof of \Cref{thm:unbwt_correct}, we implemented
+radix-sort and proved its correctness,
+\begin{theorem}[radixsort\_correct]
+  \label{thm:radixsort_correct}%
+  For all lists $l$,
+  \begin{equation*}
+    \mathtt{Sorted}~(\mathtt{radixsort}~l) \land \mathtt{Permutation}~(\mathtt{radixsort}~l)~l
+  \end{equation*}
+\end{theorem}
+and proved that any two stable sorts are interchangable,
+\begin{theorem}[stable\_sort\_unique]
+  \label{thm:radixsort_correct}%
+  For all functions $f, g : \mathtt{list}~A \to \mathtt{list}~A$,
+  \begin{gather*}
+    (\forall l,~\mathtt{Sorted}~(f~l) \land \mathtt{StablePermutation}~(f~l)~l) \implies \\
+    (\forall l,~\mathtt{Sorted}~(g~l) \land \mathtt{StablePermutation}~(g~l)~l) \implies \\
+    \forall l,~f~l = g~l
+  \end{gather*}
+\end{theorem}
+This in turn required developing suitable definitions for
+\texttt{Sorted}, \texttt{Permutation}, and \texttt{StablePermutation}.
+We explored multiple definitions for each of these properties, and
+proved equivalence between them.
+
+The rest of this paper is organize as follows. \Cref{sec:bwt} details
+the history of the BWT and its applications. \Cref{sec:ord} defines
+what we need to know about a type $A$ to sort a $\mathtt{list}~A$ as a
+part of BWT. Following Bird et~al., \cref{sec:forwards_BWT} defines
+the implementation of the forwards BWT in Coq, and
+\cref{sec:inverse_BWT} derives an implementation of the inverse BWT
+via program calculation. \Cref{sec:sorted_perm_stable} provides
+multiple definitions of \texttt{Sorted}, \texttt{Permutation}, and
+\texttt{StablePermutation}, and proves equivalence between them.
+\Cref{sec:sort} implements insertion sort and radixsort, and proves
+their correctness. \Cref{sec:BWT_proof} more formally proves the
+theorems about the BWT mentioned in previous sections, including
+\Cref{thm:unbwt_correct}.
+
+The full source code can be found at
+\url{https://github.com/jbaum98/verifed_bzip2/tree/2019-05-06}.
 
 \section{Burrows-Wheeler}
-\label{sec:bw}
+\label{sec:bwt}
 
@@
 column. If there is any correlation between a character and the
-character preceeding it, bringing together the characters in the first
+character preceding it, bringing together the characters in the first
 column will tend to bring together similar characters in the last
@@
 
-\section{The Derivation}
-\label{sec:deriv}
-In this section, we give an overview of the derivation without delving
-into the proofs of the lemmas. We only derive the simples version of
-the algorithm, although the derivations in \cite{birdmu,pearls}
-provide additional optimizations that bring the final implementation
-closer to the imperative implementation in \cite{bw} and other texts.
+\section{Orders and Preorders}
+\label{sec:ord}
 
-\subsection{Forwards-Transform}
-\label{subsec:ft}
-In general, the Burrows-Wheeler transform can be applied to lists of
-any type that can be sorted, so we define our functions to operate on
-\verb|list A|. We will discuss the sortable constraint in
-Section~\autoref{sec:sort_stable}.
+Although it is commonly demonstrated on strings, the BWT can be
+applied to any list whose elements are comparable. To make that
+precise, if we have a type $A$ and a relation on that type $\le$, we say
+that this forms a preorder on $A$ when $\le$ is reflexive and transitive, that is
+$x \le x$ for all $xA$, and
+\begin{equation*}
+  x \le y \implies y \le z \implies x \le z
+\end{equation*}
+for all $x,y,z$.
+
+We say that we have a \textit{total} preorder if for any two elements
+$x$ and $y$, $x \le y \lor y \le x$.
+
+We say that we have a \textit{decidable} preorder if we have a
+decision procedure for determining whether $x \le y$ or $x \not\le y$.
+
+We represent a total, decidable preorder on type with the
+\texttt{Preord} typeclass in Coq:
+\lstinputlisting[firstline=9, lastline=14]{../theories/BWT/Sorting/Ord.v}
+We do not explicitly require reflexivity because it follows from totality.
+
+From $\le$, we can define $x \ge y \coloneqq y \le x$, $x < y \coloneqq
+\lnot (y \le x)$, and $x > y \coloneqq \lnot (x \le y)$, and prove
+transitivity and reflexivity where appropriate.
+
+We can also define $x \equiv y \coloneqq x \le y \land y \le x$. $\equiv$ is a
+decidable equivalence relation, which means it is reflexive,
+transitive, and symmetric. The builtin class \texttt{EqDec} is used to
+describe types with decidable equivalence relations, so for any
+\texttt{Preord} instance we have the \texttt{EqDec} instance
+\texttt{Preord\_EqDec} using \texttt{eqv\_dec}:
+\lstinputlisting[firstline=116, lastline=120,
+  widthgobble=2]{../theories/BWT/Sorting/Ord.v}
+In Coq, the notation \texttt{===} is used for the relation of an
+\texttt{EqDec} instance and \texttt{==} for the decision procedure of
+an \texttt{EqDec} instance.
+
+We say that we have a total, decidable \textit{order} when we have
+that for all $x$ and $y$,
+\begin{equation*}
+  x \equiv y \implies x = y
+\end{equation*}
+that is, the induced equivalence relation of the preorder ($\equiv$)
+implies Leibniz equality ($=$). We use the \texttt{Ord} class in Coq for this:
+\lstinputlisting[firstline=175, lastline=176,
+  widthgobble=2]{../theories/BWT/Sorting/Ord.v}
+For the BWT, we require an \texttt{Ord} instance for $A$, but for other parts of the proof, such as sorting, \texttt{Preord} will suffice.
+
+\section{The Forwards BWT}
+\label{sec:forwards_BWT}
 
 First, we define left and right rotations of a list.
-\begin{verbatim}
-Definition lrot (l: list A) : list A :=
-  match l with
-  | [] => []
-  | hd :: tl => tl ++ [hd]
-  end.
+\lstinputlisting[widthgobble=2, firstline=17, lastline=22]{../theories/BWT/Lib/List.v}
+\lstinputlisting[widthgobble=2, firstline=16, lastline=21]{../theories/BWT/Rotation/Rotation.v}
+\lstinputlisting[widthgobble=2, firstline=23, lastline=28]{../theories/BWT/Rotation/Rotation.v}
 
-Definition rrot (l: list A) : list A :=
-  match l with
-  | [] => []
-  | hd :: tl => last l hd :: init l
-  end.
-\end{verbatim}
-
-Now we can produce the rotation matrix by repeatedly applying \verb|lrot|.
-\begin{verbatim}
-Fixpoint iter (f : A -> A) (n : nat) (z : A) : list A :=
-  match n with
-  | O   => []
-  | S m => z :: iter f m (f z)
-  end.
-
-Definition rots (l : list A) : list (list A) :=
-  iter lrot (length l) l.
-\end{verbatim}
+Now we can produce the rotation matrix by repeatedly applying \texttt{lrot}.
+\lstinputlisting[widthgobble=2, firstline=11, lastline=18]{../theories/BWT/Lib/Iterate.v}
+\lstinputlisting[widthgobble=2, firstline=17, lastline=18]{../theories/BWT/Rotation/Rots.v}
 
 Finally, we produce the transformed string by taking the last column
-of the sorted rotation matrix. \verb|lexsort| sorts the rotation matrix
-lexicographically; it's precise implementation is not important.
-\begin{verbatim}
-Definition bwp (l: list A) : list A :=
-  match l with
-  | [] => []
-  | hd :: _ => List.map (fun x => last x hd) (lexsort (rots l))
-  end.
-\end{verbatim}
+of the sorted rotation matrix. \texttt{lexsort} sorts the rotation
+matrix lexicographically using the \texttt{Ord} instance; we will
+discuss it's implementation in \cref{sec:sort}.
+\lstinputlisting[firstline=31,
+  lastline=35]{../theories/BWT/BurrowsWheeler.v}
+Note that in order to ensure that $\mathtt{last}$ is total, it takes
+an extra argument to return in the case of an empty list.
 
@@
 original string appears.
-\begin{verbatim}
-Fixpoint findIndex (x : A) (ls : list A) : nat :=
-  match ls with
-  | [] => 0
-  | hd :: tl =>
-    match x == hd with
-    | left _ => 0
-    | right _ => S (findIndex x tl)
-    end
-  end.
-\end{verbatim}
-\begin{verbatim}
-Definition bwn (l : list A) : nat :=
-  findIndex l (lexsort (rots l)).
-\end{verbatim}
+\lstinputlisting[firstline=12, lastline=19]{../theories/BWT/Lib/FindIndex.v}
+\lstinputlisting[firstline=73, lastline=74]{../theories/BWT/BurrowsWheeler.v}
 
-Without proof, we note that we should have for any nonempty
-\verb|x : list A| and \verb|d : A|,
-\begin{verbatim}
-nth (bwn x) (lexsort (rots x)) d = x
-\end{verbatim}
+\section{The Inverse BWT}
+\label{sec:inverse_BWT}
 
-\subsection{Inverse-Transform}
-\label{subsec:invt}
+Now we have to construct a function \texttt{unbwt} such that for all $l$,
+\begin{equation}
+  \mathtt{unbwt}~(\mathtt{bwn}~l)~(\mathtt{bwp}~xs) = l
+  \label{eq:unbwt}
+\end{equation}
+If we could recreate the entire sorted rotation matrix from
+$\mathtt{bwp}~l$, it would be simple to recover $x$ by indexing into
+the matrix at $\mathtt{bwn}~x$.
 
-Now we have to construct a function \verb|unbwt| such that
-\begin{verbatim}
-unbwt (bwn xs) (bwp xs) = xs.
-\end{verbatim}
-If we could recreate the entire sorted rotation matrix from
-\verb|bwp x|, it woiuld be simple to recover \verb|x| by indexing into
-the matrix at \verb|bwn x|. That is, we would like a function
-\verb|recreate| such that
-\begin{verbatim}
-recreate (bwp l) = lexsort (rots l).
-\end{verbatim}
-
-Then, we could define \verb|unbwt| as
-\begin{verbatim}
+More precisely, suppose we had a function function \texttt{recreate}
+such that
+\begin{equation}
+  \mathtt{recreate}~(\mathtt{bwp}~l) = \mathtt{lexsort}~(\mathtt{rots}~l)
+  \label{eq:recreate}
+\end{equation}
+Then, we could define \texttt{unbwt} as
+\begin{lstlisting}
 Definition unbwt (i : nat) (l : list A) : list A :=
   nth i (recreate l) l.
-\end{verbatim}
-and we would have the correctness of \verb|unbwt| as a result of the
-correctness of \verb|bwn|.
+\end{lstlisting}
+and \Cref{eq:unbwt} would follow directly from the following theorem:
+\begin{theorem}[bwn\_correct]
+  For every non-empty list $l$,
+  \begin{equation*}
+    (\mathtt{lexsort}~(\mathtt{rots}~l))_{(\mathtt{bwn}~l)} = l
+  \end{equation*}
+\end{theorem}
 
-\subsubsection{\texttt{recreate}}
-We will recreate the sorted rotation matrix column by column, so
-\verb|recreate j| will recreate the first \verb|j| columns of the sorted
-rotation matrix.
+How, then, do we implement \texttt{recreate}? We will recreate the
+sorted rotation matrix column by column, so \verb|recreate j| will
+recreate the first \verb|j| columns of the sorted rotation matrix.
 \begin{verbatim}
@@
 
-\section{Proving the Lemmas}
-\label{sec:lemmas}
+\section{\texttt{Sorted}, \texttt{Permutation}, and \texttt{StablePermutation}}
+\label{sec:sorted_perm_stable}
 
@@
 
-\section{Sorting \& Stability}
-\label{sec:sort_stable}
+\section{Insertion Sort, Radixsort, and \texttt{lexsort}}
+\label{sec:sort}
 
-\subsection{Sorting}
-\label{subsec:sorting}
-What do we mean by the constraint that \verb|A| is sortable? We say
-that a type must be equppied with a total, decidable preorder in order
-to be sorted.
-\begin{verbatim}
-Class Preord (A : Type):=
-  { le : A -> A -> Prop;
-    le_trans : forall x y z, le x y -> le y z -> le x z;
-    le_total : forall x y, le x y \/ le y x;
-    le_dec : forall x y, {le x y} + {~le x y};
-  }.
-\end{verbatim}
-The type of a sorting algorithm will be
-\begin{verbatim}
-sort : forall (A : Type), Preord A -> list A -> list A
-\end{verbatim}
+\section{Proving BWT Correct}
+\label{sec:BWT_proof}
 
-A correct sorting algorithm should satisfy
-\begin{verbatim}
-forall x, Sorted (sort x) /\ Permutation x (sort x)
-\end{verbatim}
-where we define the \verb|Sorted| relation on lists:
-\begin{verbatim}
-Inductive Sorted: list A -> Prop :=
-| Sorted_nil:
-    Sorted nil
-| Sorted_cons: forall hd tl,
-    (forall x, In x tl -> le hd x) ->
-    Sorted tl ->
-    Sorted (hd :: tl).
-\end{verbatim}
-
-We can also have defined \verb|Sorted| like this
-\begin{verbatim}
-Inductive LocallySorted : list A -> Prop :=
-| LSorted_nil : LocallySorted nil
-| LSorted_cons1 a : LocallySorted (a :: nil)
-| LSorted_consn a b l :
-    LocallySorted (b :: l) -> le a b -> LocallySorted (a :: b :: l).
-\end{verbatim}
-
-\begin{theorem}[Sorted-LocallySorted Equivalent]
-\begin{verbatim}
-forall l, Sorted l <-> LocallySorted l.
-\end{verbatim}
-\end{theorem}
-
-\begin{proof}
-  First we prove the forwards direction by induction on the evidence
-  \verb|Sorted l|. For the base case, we have \verb|Lsorted_nil|. In
-  the inductive case, we consider when \verb|tl| is empty and when it
-  is \verb|b :: tl|. If it is empty, we have \verb|Lsorted_cons1 hd|.
-  If it is not, we have \verb|LocallySorted (b :: tl)| by the
-  inductive hypothesis. We also have that
-  \verb|forall x, In x (b :: tl) -> le hd x)|, and because clearly
-  \verb|In b (b :: tl)| we have \verb|le hd b|. This means we can
-  apply \verb|Lsorted_consn| and we are done.
-
-  Now we prove the backwards direction by induction on the evidence
-  \verb|LocallySorted l|. The nil case is given by \verb|Sorted_nil|,
-  and the singleton case is given by the \verb|Sorted_cons| followed
-  by \verb|Sorted_nil|, using the fact that \verb|In x []| is
-  \verb|False|. In the inductive case we have \verb|Sorted (b :: tl)|
-  by the inductive hypothesis, which implies \verb|Sorted tl|. It
-  suffices to show \verb|forall x, In x (b :: tl) -> le a x|.
-  \verb|Sorted (b :: tl)| implies that
-  \verb|forall x, In x tl -> le b x|, and we have that \verb|le a b|,
-  so by the transitive property of \verb|le| we are done.
-\end{proof}
-
-We could also define sorted based on indexes:
-\begin{verbatim}
-Definition IndexSorted (l : list A) := forall i j d,
-    i <= j < length l ->
-    le (nth i l d) (nth j l d).
-\end{verbatim}
-
-\begin{theorem}[Sorted-IndexSorted Equivalent]
-\begin{verbatim}
-forall l, Sorted l <-> IndexSorted l.
-\end{verbatim}
-\end{theorem}
-
-\begin{theorem}[Permutation-IndexPermutation Equivalent]
-  The built-in \verb|Permutation xs ys| relation is equivalent to saying
-  that there exists some permutation function \verb|p : 'S_n| such that
-\begin{verbatim}
-forall i : 'I_n, nth i xs = nth (p i) ys
-\end{verbatim}
-\end{theorem}
-
-\subsection{Stability}
-Any preorder induces a decidable equivalence relation:
-\begin{verbatim}
-Definition eqv {A} `{Preord A} (x y : A) : Prop := le x y /\ le y x.
-\end{verbatim}
-A stable sort is a sort that preserves the relative order of
-equivalent elements. We can state this naturally in terms of the
-permutations. We say that a permutation \verb|p| is stable on a list
-\verb|l| if
-\begin{verbatim}
-forall i j : 'I_n, i < j -> eqv (nth i l) (nth j l) -> p i < p j
-\end{verbatim}
-Then, a sort is stable if for all \verb|l|, there exists a stable
-permutation between \verb|l| and \verb|sort l|.
-
-We can also say that two lists are stably related if
-\begin{verbatim}
-forall x, filter (eqv_dec x) l = filter (eqv_dec x) l'
-\end{verbatim}
-
-\begin{theorem}[PermStable-FilterStable Equivalent]
-\begin{verbatim}
-forall l,
-  PermStable l l' <-> (FilterStable l l' /\ Permutation l l')
-\end{verbatim}
-\end{theorem}
-
-Stability is important because we can prove that there is only one way
-to stably sort any list. This also means that any two correct stable
-sorting algorithms must agree always, and can be substituted for one another.
-
-\begin{theorem}[StableUnique]
-\begin{verbatim}
-forall l l',
-  Sorted l' -> Permutation l l' -> Stable l l' -> l = l'
-\end{verbatim}
-\end{theorem}
-
-If we have a antisymmetric total preorder, also known simply as a
-total order, then we have the important result that all sorts are
-stable sorts. Antisymmetricity means that
-\begin{equation*}
-  \forall x y, x \leq y \wedge y \leq x \implies x = y
-\end{equation*}
-or in terms of equivalence
-\begin{equation*}
-  \forall x y, x \equiv y \implies x = y
-\end{equation*}
-
-\begin{theorem}[OrderStable]
-\begin{verbatim}
-forall (O : Preorder A) (l l' : A),
-  (forall x y, eqv x y -> x = y) ->
-  Permutation l l' -> Sorted l -> Sorted l' -> Stable l l'
-\end{verbatim}
-\end{theorem}
-
-We require that the type \verb|A| be equipped with a decidable total
-order. If \verb|A| has a decidable total order, then the lexicographic
-preorder on \verb|list A| is also an order. This means that any
-sorting algorithm using lexicographic order is stable, including
-radix-sort. This is what allows us to conclude that \verb|lexsort| is
-equivalent to radix-sort.
-
-\begin{theorem}[RadixSort Correct]
-\begin{verbatim}
-forall x,
-  Sorted (radixsort x) /\ Permutation x (radixsort x)
-\end{verbatim}
-\end{theorem}
-
-\nocite{*}
-
-\bibliography{thesis}{}
-
-\bibliographystyle{plain}
+\printbibliography{}
 
 \end{document}
+
+We prove this using a few lemmas about \texttt{findIndex},
+\texttt{lexsort}, and \texttt{rots}.
+
+\begin{lemma}[findIndex\_correct]
+  For all elements $x$, lists $l$,
+  \begin{equation*}
+    (\exists y \in l,~x \equiv y) \implies l_{(\mathtt{findIndex}~x~xs)} \equiv x
+  \end{equation*}
+  \begin{proof}
+    This follows by induction on $xs$: At every step, either the
+    current element is equivalent to $x$ and we return $0$, or there
+    must be such a $y$ in the rest of the list.
+  \end{proof}
+\end{lemma}
+
+\begin{lemma}[orig\_in\_sorted\_rots]
+  For every non-empty list $l$,
+  \begin{equation*}
+    \exists y \in \mathtt{lexsort}~(\mathtt{rots}~l),~x = y
+  \end{equation*}
+  \begin{proof}
+    Clearly $l$ is in $\mathtt{rots}~l$ at index 0, and because
+    \texttt{lexsort} produces a permutation of its input, $l$ must be
+    in $\mathtt{lexsort}~(\mathtt{rots}~l)$ as well.
+  \end{proof}
+\end{lemma}
+
+\begin{theorem}[bwn\_correct]
+  For every non-empty list $xs$,
+  \begin{equation*}
+    (\mathtt{lexsort}~(\mathtt{rots}~xs))_{(\mathtt{bwn}~xs)} = xs
+  \end{equation*}
+  \begin{proof}
+    This follows from \textsf{findIndex\_correct} if we can show that
+    $\exists y \in xs,~x = y$, which we have by \textsf{orig\_in\_sorted\_rots}.
+  \end{proof}
+\end{theorem}

### 2019-05-02 draft_paper/draft_paper.tex
Stash

@@
 cyclic shifts of the original string, sort it lexicographically, and
-take the last column of the matrix. In addition, we note that index
+take the last column of the matrix. In addition, we note the index
 where the original string appears in the sorted rotation matrix. As
-shown in \cref{fig:bw_ex}, applying the algorithm to the string
-\texttt{abracadabra!} produces the new string \texttt{ard!rcaaabb} and
-the index \texttt{3}. The transformation tends to produce runs like
-the three \texttt{a}s or the two \texttt{b}s, and it is this feature
-that makes the resulting string easier to compress. Although it is not
-at all obvious, the transformed string, together with the index,
-provide enough information to recover the the original string. BWT is
-discussed more in \cref{sec:bwt}.
+shown in \Cref{fig:bw_ex}, applying the algorithm to the string
+\var{abracadabra!} produces the new string \var{ard!rcaaabb} and
+the index \var{3}. The transformation tends to produce runs like
+the three \var{a}s or the two \var{b}s in \var{ard!rcaaabb},
+and it is this feature that makes the resulting string easier to
+compress. Although it is not obvious, the transformed string, together
+with the index, provide enough information to recover the the original
+string. The BWT is discussed more in \cref{sec:bwt}.
 
@@
   \end{tt}
-  \caption{BWT acting on \texttt{abracadabra!}}
+  \caption{BWT acting on \var{abracadabra!}}
   \label{fig:bw_ex}
@@
 the BWT, together with a machine-checked proof that our implementation
-is correct. Specifically, we implement the functions
-$\mathtt{bwp} : \mathtt{list}~A \to \mathtt{list}~A$ and
-$\mathtt{bwn} : \mathtt{list}~A \to \mathbb{N}$ to produce the string
-and index part of the BWT respectively, and the function
-$\mathtt{unbwt} : \mathbb{N} \to \mathtt{list}~A \to \mathtt{list}~A$ to
-reverse the transform, and we prove the following theorem:
+is correct. Specifically, we implement the functions $\var{bwp} :
+\var{list}~A \to \var{list}~A$ and $\var{bwn} : \var{list}~A
+\to \mathbb{N}$ that produce the string and index part of the BWT
+respectively, and the function $\var{unbwt} : \mathbb{N} \to
+\var{list}~A \to \var{list}~A$ that reverses the transform. We
+prove the following theorem:
 \begin{theorem}[unbwt\_correct]
@@
   \begin{equation*}
-    \mathtt{unbwt}~(\mathtt{bwn}~l)~(\mathtt{bwp}~l) = l
+    \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l) = l
   \end{equation*}
@@
   \begin{equation*}
-    \mathtt{Sorted}~(\mathtt{radixsort}~l) \land \mathtt{Permutation}~(\mathtt{radixsort}~l)~l
+    \var{Sorted}~(\var{radixsort}~l) \land \var{Permutation}~(\var{radixsort}~l)~l
   \end{equation*}
@@
   \label{thm:radixsort_correct}%
-  For all functions $f, g : \mathtt{list}~A \to \mathtt{list}~A$,
+  For all functions $f, g : \var{list}~A \to \var{list}~A$,
   \begin{gather*}
-    (\forall l,~\mathtt{Sorted}~(f~l) \land \mathtt{StablePermutation}~(f~l)~l) \implies \\
-    (\forall l,~\mathtt{Sorted}~(g~l) \land \mathtt{StablePermutation}~(g~l)~l) \implies \\
+    (\forall l,~\var{Sorted}~(f~l) \land \var{StablePermutation}~(f~l)~l) \implies \\
+    (\forall l,~\var{Sorted}~(g~l) \land \var{StablePermutation}~(g~l)~l) \implies \\
     \forall l,~f~l = g~l
@@
 This in turn required developing suitable definitions for
-\texttt{Sorted}, \texttt{Permutation}, and \texttt{StablePermutation}.
+\var{Sorted}, \var{Permutation}, and \var{StablePermutation}.
 We explored multiple definitions for each of these properties, and
@@
 the history of the BWT and its applications. \Cref{sec:ord} defines
-what we need to know about a type $A$ to sort a $\mathtt{list}~A$ as a
+what we need to know about a type $A$ to sort a $\var{list}~A$ as a
 part of BWT. Following Bird et~al., \cref{sec:forwards_BWT} defines
@@
 via program calculation. \Cref{sec:sorted_perm_stable} provides
-multiple definitions of \texttt{Sorted}, \texttt{Permutation}, and
-\texttt{StablePermutation}, and proves equivalence between them.
+multiple definitions of \var{Sorted}, \var{Permutation}, and
+\var{StablePermutation}, and proves equivalence between them.
 \Cref{sec:sort} implements insertion sort and radixsort, and proves
@@
 We represent a total, decidable preorder on type with the
-\texttt{Preord} typeclass in Coq:
+\var{Preord} typeclass in Coq:
 \lstinputlisting[firstline=9, lastline=14]{../theories/BWT/Sorting/Ord.v}
-We do not explicitly require reflexivity because it follows from totality.
+We do not explicitly require reflexivity because it follows from
+totality. Given a \var{Preord} instance for a type $A$, we will be
+able to sort $\var{list}~A$ as shown in \cref{sec:sort}.
 
@@
 decidable equivalence relation, which means it is reflexive,
-transitive, and symmetric. The builtin class \texttt{EqDec} is used to
-describe types with decidable equivalence relations, so for any
-\texttt{Preord} instance we have the \texttt{EqDec} instance
-\texttt{Preord\_EqDec} using \texttt{eqv\_dec}:
-\lstinputlisting[firstline=116, lastline=120,
-  widthgobble=2]{../theories/BWT/Sorting/Ord.v}
-In Coq, the notation \texttt{===} is used for the relation of an
-\texttt{EqDec} instance and \texttt{==} for the decision procedure of
-an \texttt{EqDec} instance.
+transitive, and symmetric, and we have a decision procedure for it.
+\lstinputlisting[
+  firstline=116, lastline=120, widthgobble=2
+]{../theories/BWT/Sorting/Ord.v}
+
+The builtin class \var{EqDec} is used to describe types with
+decidable equivalence relations, so for any \var{Preord} instance
+we have the \var{EqDec} instance \var{Preord\_EqDec} using
+\var{eqv\_dec}. In Coq, the notation \var{===} is used for the
+equivalence relation of an \var{EqDec} instance and \var{==} for
+its decision procedure.
 
@@
 that is, the induced equivalence relation of the preorder ($\equiv$)
-implies Leibniz equality ($=$). We use the \texttt{Ord} class in Coq for this:
-\lstinputlisting[firstline=175, lastline=176,
-  widthgobble=2]{../theories/BWT/Sorting/Ord.v}
-For the BWT, we require an \texttt{Ord} instance for $A$, but for other parts of the proof, such as sorting, \texttt{Preord} will suffice.
+implies Leibniz equality ($=$). We use the \var{Ord} class in Coq
+for this:
+\lstinputlisting[firstline=175, lastline=176, widthgobble=2]{../theories/BWT/Sorting/Ord.v}
+For the BWT, we require an \var{Ord} instance for $A$, but
+elsewhere, such as in sorting, \var{Preord} will suffice.
 
@@
 
-First, we define left and right rotations of a list.
-\lstinputlisting[widthgobble=2, firstline=17, lastline=22]{../theories/BWT/Lib/List.v}
-\lstinputlisting[widthgobble=2, firstline=16, lastline=21]{../theories/BWT/Rotation/Rotation.v}
-\lstinputlisting[widthgobble=2, firstline=23, lastline=28]{../theories/BWT/Rotation/Rotation.v}
-
-Now we can produce the rotation matrix by repeatedly applying \texttt{lrot}.
-\lstinputlisting[widthgobble=2, firstline=11, lastline=18]{../theories/BWT/Lib/Iterate.v}
-\lstinputlisting[widthgobble=2, firstline=17, lastline=18]{../theories/BWT/Rotation/Rots.v}
-
-Finally, we produce the transformed string by taking the last column
-of the sorted rotation matrix. \texttt{lexsort} sorts the rotation
-matrix lexicographically using the \texttt{Ord} instance; we will
-discuss it's implementation in \cref{sec:sort}.
-\lstinputlisting[firstline=31,
-  lastline=35]{../theories/BWT/BurrowsWheeler.v}
-Note that in order to ensure that $\mathtt{last}$ is total, it takes
+First, we define the left and right rotations of a list. \var{lrot}
+moves the first element of a list to the end.
+\lstinputlisting[widthgobble=2, firstline=24, lastline=28]{../theories/BWT/Rotation/Rotation.v}
+\var{rrot} moves the last element of a list to the front, and
+relies on \var{init}, which drops the last element of a list.
+\lstinputlisting[widthgobble=2, firstline=18, lastline=22]{../theories/BWT/Lib/List.v}
+\lstinputlisting[widthgobble=2, firstline=17, lastline=21]{../theories/BWT/Rotation/Rotation.v}
+Note that in order to ensure that $\var{last}$ is total, it takes
 an extra argument to return in the case of an empty list.
 
-We also need to find the index in the sorted rotation matrix where the
-original string appears.
-\lstinputlisting[firstline=12, lastline=19]{../theories/BWT/Lib/FindIndex.v}
-\lstinputlisting[firstline=73, lastline=74]{../theories/BWT/BurrowsWheeler.v}
+Now we would likel to produce the rotation matrix by repeatedly
+applying \var{lrot}. We will use \var{iter}, which takes a function
+$f : A \to A$, a number $n$, and an initial value $z$, and returns the
+list of intermediate values obtained by applying $f$ to $z$ $n$ times.
+Or,
+\begin{equation}
+  \var{iter}~f~n~z = [z,~f~z,~f~(f~z),~\ldots,~f^n z]
+  \label{eq:iter}
+\end{equation}
+\lstinputlisting[
+  widthgobble=2,
+  firstline=14,
+  lastline=18
+]{../theories/BWT/Lib/Iterate.v}
+
+We can prove that this definition satisfies \Cref{eq:iter} by defining
+the function \var{rep} for repeated function application.
+\lstinputlisting[
+  widthgobble=2,
+  firstline=8,
+  lastline=12
+]{../theories/BWT/Lib/Repeat.v}
+In proofs, we will still use the notation $f^i z$ for $\var{rep}~f~i~z$.
+
+\begin{theorem}[rep\_r]
+  For all $f$, $n$, and $z$,
+  \begin{equation}
+     f^i~(f~z) = f^{i+1}~z
+  \end{equation}
+\end{theorem}
+\begin{proof}
+  The theorem follows by induction on $n$.
+\end{proof}
+
+\begin{theorem}[iter\_length]
+  For all $f$, $n$, and $z$,
+  \begin{equation}
+    \var{length}~(\var{iter}~f~n~z) = n
+  \end{equation}
+\end{theorem}
+\begin{proof}
+  The theorem follows by induction on $n$.
+\end{proof}
+
+\begin{theorem}[iter\_nth]
+  For all $f$, $n$, $z$, and $i < n$,
+  \begin{equation}
+    (\var{iter}~f~n~z)_i = f^i~z
+  \end{equation}
+\end{theorem}
+\begin{proof}
+  The theorem follows by induction on $i$. The inductive case is
+  \begin{align*}
+       (\var{iter}~f~(n+1)~z)_{i+1}
+    &= (z :: \var{iter}~f~n~(f~z))_{i+1} &&\{\text{\var{iter} def.}\} \\
+    &= (\var{iter}~f~n~(f~z))_i          &&\{\text{by \var{nth} def.}\} \\
+    &= f^i~(f~z)                            &&\{\text{by \var{rep\_r}}\} \\
+    &= f^{i+1}~z && \qedhere
+  \end{align*}
+\end{proof}
+
+Now we can define \var{rots} to produce the rotation matrix.
+\lstinputlisting[
+  widthgobble=2,
+  firstline=17,
+  lastline=18,
+]{../theories/BWT/Rotation/Rots.v}
+
+\var{bwp} produces the transformed string by taking the last column
+of the sorted rotation matrix. \var{lexsort} sorts the rotation
+matrix lexicographically using the \var{Ord} instance; we will
+discuss it's implementation in \cref{sec:sort}.
+\lstinputlisting[
+  widthgobble=2,
+  firstline=31,
+  lastline=35
+]{../theories/BWT/BurrowsWheeler.v}
+
+\var{bwn} returns the index in the sorted rotation matrix where the
+original string appears. It relies on \var{findIndex}, which is
+defined with respect to some decidable equivalence relation.
+\lstinputlisting[
+  widthgobble=2,
+  firstline=12,
+  lastline=19,
+]{../theories/BWT/Lib/FindIndex.v}
+
+When we use \var{findIndex} in \var{bwn}, we are using the
+induced equivalence relation from the \var{Ord} instance, so it
+finds the row that is Leibniz equal to $l$.
+\lstinputlisting[
+  widthgobble=2,
+  firstline=73,
+  lastline=74,
+]{../theories/BWT/BurrowsWheeler.v}
+
+Given these definitions, we can quickly prove that \var{bwn}
+satisfies its specification, namely that for all non-empty $l$
+\begin{equation*}
+    (\var{lexsort}~(\var{rots}~l))_{(\var{bwn}~l)} = l
+\end{equation*}
+
+\begin{lemma}[findIndex\_correct]
+  For all elements $x$ and lists $l$,
+  \begin{equation*}
+    (\exists y \in l,~x \equiv y) \implies l_{(\var{findIndex}~x~xs)} \equiv x
+  \end{equation*}
+  \begin{proof}
+    This follows by induction on $l$: At every step, either the
+    current element is equivalent to $x$ and we return $0$, or there
+    must be such a $y$ in the rest of the list.
+  \end{proof}
+\end{lemma}
+
+\begin{lemma}[orig\_in\_sorted\_rots]
+  For every non-empty list $l$,
+  \begin{equation*}
+    \exists y \in \var{lexsort}~(\var{rots}~l),~x = y
+  \end{equation*}
+  \begin{proof}
+    Clearly $l$ is in $\var{rots}~l$ at index 0, and because
+    \var{lexsort} produces a permutation of its input, $l$ must be
+    in $\var{lexsort}~(\var{rots}~l)$ as well.
+  \end{proof}
+\end{lemma}
+
+\begin{theorem}[bwn\_correct]
+  For every non-empty list $xs$,
+  \begin{equation*}
+    (\var{lexsort}~(\var{rots}~xs))_{(\var{bwn}~xs)} = xs
+  \end{equation*}
+  \begin{proof}
+    This follows from \var{findIndex\_correct} if we can show that
+    $\exists y \in xs,~x = y$, which we have by \var{orig\_in\_sorted\_rots}.
+  \end{proof}
+\end{theorem}
 
@@
 
-Now we have to construct a function \texttt{unbwt} such that for all $l$,
+Now we want to construct a function \var{unbwt} such that for all $l$,
 \begin{equation}
-  \mathtt{unbwt}~(\mathtt{bwn}~l)~(\mathtt{bwp}~xs) = l
+  \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l) = l
   \label{eq:unbwt}
@@
 If we could recreate the entire sorted rotation matrix from
-$\mathtt{bwp}~l$, it would be simple to recover $x$ by indexing into
-the matrix at $\mathtt{bwn}~x$.
+$\var{bwp}~l$, it would be simple to recover $l$ by indexing into
+the matrix at $\var{bwn}~l$.
 
-More precisely, suppose we had a function function \texttt{recreate}
+More precisely, suppose we had a function function \var{recreate}
 such that
 \begin{equation}
-  \mathtt{recreate}~(\mathtt{bwp}~l) = \mathtt{lexsort}~(\mathtt{rots}~l)
+  \var{recreate}~(\var{bwp}~l) = \var{lexsort}~(\var{rots}~l)
   \label{eq:recreate}
 \end{equation}
-Then, we could define \texttt{unbwt} as
+Then, we could define \var{unbwt} as
 \begin{lstlisting}
@@
 \end{lstlisting}
-and \Cref{eq:unbwt} would follow directly from the following theorem:
-\begin{theorem}[bwn\_correct]
-  For every non-empty list $l$,
+and \Cref{eq:unbwt} would follow directly from the \var{bwn\_correct}:
+\begin{align}
+     \label{eq:unbwt_correct_start}
+     \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l)
+  &= (\var{recreate}~(\var{bwp}~l))_{(\var{bwn})~l} &&\\
+  &= (\var{lexsort}~(\var{rots}~l))_{(\var{bwn})~l}
+     &&\text{\{by \Cref{eq:recreate}\}}\\
+  &= l &&\text{\{\var{recreate\_correct}\}}
+     \label{eq:unbwt_correct_end}
+\end{align}
+
+How, then, do we implement \var{recreate} such that
+\Cref{eq:recreate} holds? One idea is to recreate the sorted rotation
+matrix column by column. We define $\var{cols}~j$ to take the first
+$j$ columns of a matrix:
+\lstinputlisting[
+  widthgobble=2,
+  firstline=19,
+  lastline=20,
+]{../theories/BWT/Columns.v}
+Then we introduce a parameter $j$ to \var{recreate}, and our new
+specification is that $\var{recreate}~j$ recreates the first $j$
+columns of the sorted rotation matrix:
+\begin{theorem}[recreate\_correct\_ind]
+  For all lists $l$ and $j \le \var{length}~l$,
+  \begin{equation}
+    \var{recreate}~j~(\var{bwp}~l) =
+    \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
+    \label{eq:recreate_correct_ind}
+  \end{equation}
+\end{theorem}
+
+Now we can redefine \var{unbwt} as
+\lstinputlisting[
+  widthgobble=2,
+  firstline=196,
+  lastline=197,
+]{../theories/BWT/BurrowsWheeler.v}
+
+Assuming that we can prove \var{recreate\_correct\_ind}, the
+correctness of of \text{unbwt} follows:
+
+\begin{theorem}[cols\_id]
+  For all $j$ and matrices $m : \var{list}~(\var{list}~A)$,
+  \begin{gather*}
+    (\forall r \in m,~\var{length}~r \le n) \implies \\
+    \var{cols}~n~m = m
+  \end{gather*}
+\end{theorem}
+\begin{proof}
+  $\var{cols}~j~m = \var{map}~(\var{firstn}~j)~m$, and we
+  know that $m = \var{map}~(\lambda x.x)~m$. So, it suffices to show that
+  for every $r \in m$, the mapping functions agree, that is
+  $\var{firstn}~j~r = r$. This follows from \var{firstn\_all2}
+  in the Coq standard library.
+\end{proof}
+
+\begin{theorem}[iter\_preserves]
+  We are given $f$, $n$, $z$ and a predicate $P : A \to \mathbb{P}$.
+  Suppose $f$ preserves $P$, which is to say that
   \begin{equation*}
-    (\mathtt{lexsort}~(\mathtt{rots}~l))_{(\mathtt{bwn}~l)} = l
+    \forall x,~P~x \implies P~(f~x)
+  \end{equation*}
+  then,
+  \begin{gather*}
+     P~z \implies \forall x \in \var{iter}~f~n~z,~P~x
+  \end{gather*}
+\end{theorem}
+\begin{proof}
+  This follows from induction on $n$.
+\end{proof}
+
+\begin{lemma}[sort\_rots\_all\_len]
+  For all lists $l$,
+  \begin{equation*}
+    \forall x \in \var{lexsort}~(\var{rots}~l),~\var{length}~x = \var{length}~l
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  Because \var{lexsort} permutes its argument, it suffices to show
+  that all the rows of $\var{rots}~l$ have the same length as $l$.
+  This follows from \var{iter\_preserves} and the fact that
+  \var{lrot} preserves length.
+\end{proof}
+
+\begin{corollary}[recreate\_correct]
+  For all lists $l$,
+  \begin{equation*}
+    \var{recreate}~(\var{length}~l)~(\var{bwp}~l) =
+    \var{lexsort}~(\var{rots}~l)
+  \end{equation*}
+\end{corollary}
+\begin{proof}
+  This follows from \var{recreate\_correct\_ind},
+  \var{cols\_id}, and \var{sort\_rots\_all\_len}.
+\end{proof}
+
+\begin{theorem}[unbwt\_correct]
+  For all lists $l$,
+  \begin{equation*}
+    \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l) = l
   \end{equation*}
 \end{theorem}
+\begin{proof}
+  The proof is almost identical to that given in
+  \Crefrange{eq:unbwt_correct_start}{eq:unbwt_correct_end}.
+\end{proof}
 
-How, then, do we implement \texttt{recreate}? We will recreate the
-sorted rotation matrix column by column, so \verb|recreate j| will
-recreate the first \verb|j| columns of the sorted rotation matrix.
-\begin{verbatim}
-Definition cols (j : nat) (l : list A) : list A := map (firstn j) l.
+To implement \var{recreate}, we need expressions for
+$\var{recreate}~0$ and $\var{recreate}~(j+1)$ that satisfy
+\Cref{eq:recreate_correct_ind} in \var{recrate\_correct\_ind}:
+\begin{equation}
+  \var{recreate}~j~(\var{bwp}~l) =
+  \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
+  \tag{\ref{eq:recreate_correct_ind}}
+\end{equation}
 
-Theorem recreate_correct_ind : forall j l,
-    j <= length l ->
-    recreate j (bwp l) = cols j (lexsort (rots l)).
-\end{verbatim}
+To derive an implementation, we can substitute for $j$ on the right-hand
+side of \Cref{eq:recreate_correct_ind} and manipulate that expression
+algebraically until it is in the form \var{\_ (bwp l)}. Then we can
+define $\var{recreate}~0$ and $\var{recreate}~(j+1)$ to be
+precisely those \var{\_}s, and by running our derivation in
+reverse, we obtain a proof.
 
-To derive an implementation of \verb|recreate| from this correctness
-specification, we need an expression for \verb|recreate 0| and
-\verb|recreate (S j)|. We can substitute for \verb|j| on the
-right-hand side and manipulate that expression algebraically until it
-is in the form \verb|_ (bwp l)|. Then we can define \verb|recreate 0|
-and \verb|recreate (S j)| to be precisely the \verb|_|, and by
-running our derivation in reverse, we obtain a proof for the base and
-inductive cases that uses only rewrites.
+We start with the base case, $\var{recreate}~0$:
+\begin{align*}
+     \var{cols}~0~(\var{lexsort}~(\var{rots}~l))
+  &= \var{map}~(\lambda x.[])~(\var{lexsort}~(\var{rots}~l))) \\
+  &= \var{repeat}~[]~(\var{length}~(\var{lexsort}~(\var{rots}~l))) \\
+  &= \var{repeat}~[]~(\var{length}~(\var{bwp}~l)) \\
+     \var{cols}~0~(\var{lexsort}~(\var{rots}~l))
+  &= \var{map}~(\lambda x.[])~(\var{bwp}~l)
+\end{align*}
 
-We start with the base case, \verb|recreate 0|:
-\begin{verbatim}
-  cols 0 (lexsort (rots l))
-= { by unfold cols }
-  map (firstn 0) (lexsort (rots l))
-= { by firstn_O}
-  map (const []) (lexsort (rots l))
-= { by map_const }
-  repeat [] (length (lexsort (rots l)))
-= { by bwp_length, rots_length, lexsort_length }
-  repeat [] (length (bwp l))
-= { by map_const }
-  map (const []) (bwp l)
-\end{verbatim}
-
-From this, we can define \verb|recreate 0 l = map (const []) l| and
-we already have the proof for the base case of the correctness
-specification.
+From this, we can define $\var{recreate}~0~l = \var{map}~(\lambda
+x.[])~l$ and we already have the proof for the base case of the
+correctness specification.
 
 For the inductive case, we have the expression
-\begin{verbatim}
-  cols (S j) (lexsort (rots l))
-\end{verbatim}
-on the right-hand side. Our plan is to express \verb|cols (S j)|
-in terms of \verb|cols j|, and then to use the inductive hypothesis
-\begin{verbatim}
-  cols j (lexsort (rots l)) = recreate j
-\end{verbatim}
+\begin{equation*}
+  \var{cols}~(j+1)~(\var{lexsort}~(\var{rots}~l))
+\end{equation*}
+on the right-hand side. Our plan is to express $\var{cols}~(j+1)$
+in terms of $\var{cols}~j$, and then to use the induction
+hypothesis
+\begin{equation*}
+  \var{cols}~j~(\var{lexsort}~(\var{rots}~l)) = \var{recreate}~j
+\end{equation*}
 to produce a well-formed recursive definition.
 
-One way to do this is to say that taking the first \verb|S j| columns
-of a matrix is the same as moving the first column to the back, taking
-the first \verb|j| columns, and then sticking the last column back on
-the front:
-\begin{verbatim}
-  cols (S j) l = prepend_col (map last (map lrot l)) (cols j (map lrot l))
-\end{verbatim}
-If we substitute \verb|map rrot l| for \verb|l| we find a simpler
-expression
-\begin{verbatim}
-  cols (S j) (map rrot l) = prepend_col (map last l) (cols j l)
-\end{verbatim}
+One way to do this is to note that taking the first $j+1$ columns of a
+matrix is the same as moving the first column to the back, taking the
+first $j$ columns, and then sticking the last column back on the
+front:
+\begin{equation*}
+  \var{cols}~(j+1)~l =
+  \var{prepend\_col}~(\var{map}~\var{hd}~l)~(\var{cols}~j~(\var{map}~\var{lrot}~l))
+\end{equation*}
+If we substitute $\var{map}~\var{rrot}~l$ for $l$ we find a simpler expression
+\begin{equation}
+  \label{eq:cols_rrot}
+  \var{cols}~(j+1)~(\var{map}~\var{rrot}~l) =
+  \var{prepend\_col}~(\var{map}~\var{last}~l)~(\var{cols}~j~l)
+\end{equation}
 
-Now we would like to find a way of introducing a \verb|map rrot|
-before the \verb|lexsort (rots l)|. The insight here is that we can
-assume that \verb|lexsort| is actually a radix-sort. That is, that
-\begin{verbatim}
-lexsort xs = rep (length xs) (hdsort ∘ map rrot) xs
-\end{verbatim}
-where \verb|hdsort| sorts the matrix stably on the the first column
-and \verb|rep| is defined as
-\begin{verbatim}
-Fixpoint rep (f : A -> A) (n : nat) (z : A) : A :=
-  match n with
-  | O => z
-  | S m => f (rep f m z)
-  end.
-\end{verbatim}
+Now we would like to find a way of introducing a $\var{map}~\var{rrot}$
+before the $\var{lexsort}~(\var{rots}~l)$. The insight here is that we can
+assume that \var{lexsort} is actually a radix-sort. That is, that
+\begin{equation*}
+\var{lexsort}~l = (\var{hdsort} \comp \var{map}~\var{rrot})^{\var{length}~l}~l
+\end{equation*}
+where \var{hdsort} sorts the matrix stably on the the first column.
 
 We can do this because there is only one way to sort the matrix, so
-assuming \verb|lexsort| and the radix-sort are correct, they must be
-doing the same thing. We will specify precisely when and why this is
-valid later.
+assuming \var{lexsort} and the radix-sort are correct, they must be
+doing the same thing. We will prove this in
+\cref{sec:sorted_perm_stable,sec:sort}.
 
-Stepping back, it is important to say that this insight, whether
-explicitly stated or not, is always the crucial step in inverting the
-Burrows-Wheeler transform. Because radix-sort only needs one column at
-a time, this transformation will allow us to reconstruct the sorted
-rotation matrix one column at a time.
+%% In seeking to introduce an extra \verb|map rrot| term, we can note
+%% that rotating and sorting an extra time doesn't change anything when
+%% we are sorting the rotation matrix. This works because rotating the
+%% last column to the front produces a matrix that is sorted on its last
+%% \verb|n - 1| characters, so sorting stably on the first column will
+%% re-sort the matrix. Crucially though, rotating the last column
+%% forwards also permutes the entire matrix, so in addition to having a
+%% sorted list, we have a sorted list of the same elements.
 
-In seeking to introduce an extra \verb|map rrot| term, we can note
-that rotating and sorting an extra time doesn't change anything when
-we are sorting the rotation matrix. This works because rotating the
-last column to the front produces a matrix that is sorted on its last
-\verb|n - 1| characters, so sorting stably on the first column will
-re-sort the matrix. Crucially though, rotating the last column
-forwards also permutes the entire matrix, so in addition to having a
-sorted list, we have a sorted list of the same elements.
+%% Using that we can show
+%% \begin{equation*}
+%%   lexsort (rots x)
+%% = { because rrot permutes its elements }
+%%   lexsort (rrot (rots x))
+%% = { because hdsort permutes its elements }
+%%   lexsort (hdsort (rrot (rots x)))
+%% = { radix-sort }
+%%   rep n (hdsort ∘ rrot) (hdsort (rrot (rots x)))
+%% = { rep_absorb_r }
+%%   rep (S n) (hdsort ∘ rrot) (rots x)
+%% = { rep_split_l }
+%%   hdsort (rrot (rep n (hdsort ∘ rrot) (rots x)))
+%% = { radix-sort }
+%%   hdsort (rrot (lexsort (rots x)))
+%% \end{equation*}
 
-More precisely, we can characterize how \verb|map rrot| permutes
-\verb|rots|
-\begin{verbatim}
-map rrot (rots x) = rrot (rots x)
-\end{verbatim}
+%% From here we can calculate
+%% \begin{equation*}
+%%   cols (S j) (lexsort (rots l))
+%% = cols (S j) (hdsort (map rrot (lexsort (rots l))))
+%% = { hdsort commutes with cols (S j) }
+%%   hdsort (cols (S j) (map rrot (lexsort (rots l))))
+%% = { reduce cols (S j) }
+%%   hdsort (prepend_col (map last (lexsort (rots l))) (cols j (lexsort (rots l)))
+%% = { by IH and def of bwp }
+%%   hdsort (prepend_col (bwp l) (recreate j (bwp l)))
+%% = (fun x => hdsort (prepend_col x (recreate j x))) bwp
+%% \end{equation*}
 
-Using that we can show
-\begin{verbatim}
-  lexsort (rots x)
-= { because rrot permutes its elements }
-  lexsort (rrot (rots x))
-= { because hdsort permutes its elements }
-  lexsort (hdsort (rrot (rots x)))
-= { radix-sort }
-  rep n (hdsort ∘ rrot) (hdsort (rrot (rots x)))
-= { rep_absorb_r }
-  rep (S n) (hdsort ∘ rrot) (rots x)
-= { rep_split_l }
-  hdsort (rrot (rep n (hdsort ∘ rrot) (rots x)))
-= { radix-sort }
-  hdsort (rrot (lexsort (rots x)))
-\end{verbatim}
+%% More precisely, we can characterize how \verb|map rrot| permutes
+%% \verb|rots|
+%% \begin{equation*}
+%% map rrot (rots x) = rrot (rots x)
+%% \end{equation*}
 
-From here we can calculate
-\begin{verbatim}
-  cols (S j) (lexsort (rots l))
-= cols (S j) (hdsort (map rrot (lexsort (rots l))))
-= { hdsort commutes with cols (S j) }
-  hdsort (cols (S j) (map rrot (lexsort (rots l))))
-= { reduce cols (S j) }
-  hdsort (prepend_col (map last (lexsort (rots l))) (cols j (lexsort (rots l)))
-= { by IH and def of bwp }
-  hdsort (prepend_col (bwp l) (recreate j (bwp l)))
-= (fun x => hdsort (prepend_col x (recreate j x))) bwp
-\end{verbatim}
+%% From here we can calculate
+%% \begin{equation*}
+%%   cols (S j) (lexsort (rots l))
+%% = cols (S j) (hdsort (map rrot (lexsort (rots l))))
+%% = { hdsort commutes with cols (S j) }
+%%   hdsort (cols (S j) (map rrot (lexsort (rots l))))
+%% = { reduce cols (S j) }
+%%   hdsort (prepend_col (map last (lexsort (rots l))) (cols j (lexsort (rots l)))
+%% = { by IH and def of bwp }
+%%   hdsort (prepend_col (bwp l) (recreate j (bwp l)))
+%% = (fun x => hdsort (prepend_col x (recreate j x))) bwp
+%% \end{equation*}
 
-We can use this to define \verb|recreate|:
-\begin{verbatim}
-Fixpoint recreate (j : nat) (l : list A) : list (list A) :=
-  match j with
-  | O    => map (const []) l
-  | S j' => hdsort (prepend_col l (recreate j' l))
-  end.
-\end{verbatim}
+%% We can use this to define \verb|recreate|:
+%% \begin{equation*}
+%% Fixpoint recreate (j : nat) (l : list A) : list (list A) :=
+%%   match j with
+%%   | O    => map (const []) l
+%%   | S j' => hdsort (prepend_col l (recreate j' l))
+%%   end.
+%% \end{equation*}
 
-\section{\texttt{Sorted}, \texttt{Permutation}, and \texttt{StablePermutation}}
+\section{\var{Sorted}, \var{Permutation}, and \var{StablePermutation}}
 \label{sec:sorted_perm_stable}
 
-In this section we prove the lemmas from the previous section:
-\begin{enumerate}
-\item
-\begin{verbatim}
-map rrot (rots x) = rrot (rots x)
-\end{verbatim}
+The previous section's derivation relied on the correctness and stability of
+various sorts. To make those parts of the derivation precise, we
+define here what it means for a sorting function to be correct, and
+what it means for a sort to be stable.
 
-\item \verb|hdsort| commutes with \verb|cols (S j)|
-  This proof is cool because in theory it can be proven via
-  parametricity, but that's a metatheorem in Coq.
+Intuitively, a sorting function must produce a sorted list that
+contains all of the elements that its input contained. So, given a function $f : \var{list}~A \to \var{list}~A$, we say that $\var{Sort}~f$ precisely when for all $l$,
+\begin{equation*}
+  \var{Sorted}~(f~l) \land \var{Permutation}~(f~l)~l
+\end{equation*}
+Now we need to define \var{Sorted} and \var{Permutation} to capture
+the idea that a list is in sorted order, and that two lists contain
+all of the same elements.
 
-\item
-\begin{verbatim}
-cols (S j) (map rrot l) = prepend_col (map last l) (cols j l)
-\end{verbatim}
-\end{enumerate}
+\subsection{\var{Sorted}}
+\label{subsec:sorted}
+
+For a sorting function to be correct, its output must be sorted with
+respect to the ordering defined by the \var{Preord} instance. There a
+a few ways we might define what it means for a list to be sorted. In
+mathematics, we might say that a list $l$ is sorted if for all indices
+$i$ and $j$,
+\begin{equation*}
+  i \le j \implies l_i \le l_j
+\end{equation*}
+We will call this property \var{SortedIx}.
+
+However, it is often easier to reason about inductively defined
+properties, so we would like to define \var{Sorted} inductively. So,
+we can say that as a base case, the empty list is sorted, and given a
+sorted list, we can add an element that is smaller than all the
+existing elements and the result will also be sorted:
+\begin{align*}
+  \infer[\var{Sorted\_nil}]{\var{Sorted}~[]}{}
+  &&
+  \infer[\var{Sorted\_cons}]{\var{Sorted}~(a :: l)}{%
+    (\forall~x \in l,~a < x) & \var{Sorted}~l
+  }
+\end{align*}
+
+By transitivity it suffices to check that the new element is
+smaller than just the first element of the sorted list. That leads to
+a third definition of sorted:
+\begin{gather*}
+  \begin{align*}
+    \infer[\var{SortedLocal\_nil}]{\var{SortedLocal}~[]}{}
+    &&
+    \infer[\var{SortedLocal\_1}]{\var{SortedLocal}~[a]}{}
+  \end{align*}
+  \\
+  \\
+  \infer[\var{SortedLocal\_cons}]{\var{SortedLocal}~(a :: b :: l)}{%
+    a < b & \var{SortedLocal}~(b :: l)
+  }
+\end{gather*}
+
+Although these two inductive definitions are intuitive, they are not
+obviously equivalent to the first definition, which is more clearly
+correct. We prove them equivalent all:
+\begin{theorem}[SortedLocal\_iff]
+  For all lists $l$,
+  \begin{equation*}
+    \var{Sorted}~l \iff \var{SortedLocal}~l
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  The $\implies$ direction follows by induction. The $\impliedby$
+  direction follows by induction on $l$, reasoning about the empty
+  list, the singleton list, and lists with at least two elements. In
+  the inductive case $l = a :: b :: l'$, we have that $a \le b$, and
+  that $\forall x \in l',~b \le x$, which gives us $\forall x \in b :: l',~a \le x$ by
+  transitivity.
+\end{proof}
+
+\begin{theorem}[SortedIx\_iff]
+  A list $l$ is sorted under the inductive definition (\var{Sorted}) if
+  and only if it is sorted under the index definition (\var{SortedIx}).
+\end{theorem}
+\begin{proof}
+  We reason by induction on $l$. The base cases are easy, so we
+  address the inductive case, $l = a :: l'$
+
+  ($\implies$) We have to prove that $(a :: l')_i \le (a :: l')_j$ given
+  $i \le j$ and $\var{Sorted}~l$. We reason by cases on $i$ and $j$:
+  \begin{itemize}[noitemsep]
+  \item $i = j = 0$. By reflexivity, $a \le a$.
+  \item $i = 0,~j=j' + 1$. We need to show $a \le l'_{j'}$. We know that
+    $a \le x$ for all $x \in l'$ by $\var{SortedIx}~(a::l')$, so we are
+    done.
+  \item $i = i' +1,~j = j' + 1$. We need to show $l'_{i'} \le l'_{j'}$,
+    which follows from the induction hypothesis.
+  \end{itemize}
+
+  ($\impliedby$) We have $\var{SortedIx}~(a :: l')$, and we need to
+  show $\var{Sorted}~a :: l'$ given the induction hypothesis
+  $\var{SortedIx}~l' \implies \var{Sorted}~l'$. By
+  \var{SortedLocal\_iff}, we can just prove $\var{SortedLocal}~(a ::
+  l')$. If $l' = []$, we can apply $\var{SortedLocal\_1}$, so assume
+  $l' = b :: l''$. To apply \var{SortedLocal\_cons} we need to show
+  $\var{SortedLocal}~(b :: l'')$, which we have by the induction
+  hypothesis and $\var{SortedIx}~(a :: b :: l'')$, and that $a \le b$.
+  That follows because $a = (a :: b :: l'')_0$ and $b = (a :: b ::
+  l'')_1$, so by $\var{SortedIx}~(a :: b :: l'')$, we are done.
+\end{proof}
+
+\subsection{\var{Permutation}}
+
+There are multiple ways of defining \var{Permutation} as well: some of
+them are more obviously correct, and some are more useful in certain
+situations. The ``mathematical'' definition of when two lists $l$
+and $l'$ are permutations of each other is:
+\begin{itemize}[nosep]
+  \item $\var{length}~l = \var{length}~l' = n$
+  \item there exists a permutation function $p : \mathbb{N}_{<n} \to
+    \mathbb{N}_{<n}$, where $p$ is a bijection, such that for all indices $i$,
+    \begin{equation*}
+      l_i = l'_{p(i)}
+    \end{equation*}
+\end{itemize}
+
+In Coq, we can represent a finite function as a list of its outputs:
+$f : \mathbb{N}_{<n} \to A$ is represented by the $\var{list}~A$
+$[f(0),~f(1),\ldots,f(n-1)]$. This is consistent with the mathematical
+notation where we denote by $2~1~0$ the permutation $p(0) = 2$, $p(1)
+= 1$, $p(2) = 0$. So, $p$ here would be represented by $[2,~1,~0]$.
+
+To restrict the range to $\mathbb{N}_{<n}$ we say that $\forall i,~i \in p
+\implies ~i<n$, to make the function surjective, we strengthen that to
+$\forall i,~i \in p \iff ~i<n$, and to make it injective we require that $p$ not
+contain any duplicate entries. The inductive property \var{NoDup} from
+the Coq standard library captures that requirement:
+\begin{align*}
+  \infer[\var{NoDup\_nil}]{\var{NoDup}~[]}{}
+  &&
+  \infer[\var{NoDup\_cons}]{\var{NoDup}~(x :: l)}{%
+    x \notin l & \var{NoDup}~l
+  }
+\end{align*}
+This leads to the following definition of the property
+$\var{PermFun}~n~p$, which holds when $p : \var{list}~\mathbb{N}$
+represents a permutation function on lists of length $n$:
+\lstinputlisting[
+  firstline=14, lastline=15
+]{../theories/BWT/Sorting/PermFun.v}
+
+%% In this section we prove the lemmas from the previous section:
+%% \begin{enumerate}
+%% \item
+%% \begin{equation*}
+%% map rrot (rots x) = rrot (rots x)
+%% \end{equation*}
+
+%% \item \verb|hdsort| commutes with \verb|cols (S j)|
+%%   This proof is cool because in theory it can be proven via
+%%   parametricity, but that's a metatheorem in Coq.
+
+%% \item
+%% \begin{equation*}
+%% cols (S j) (map rrot l) = prepend_col (map last l) (cols j l)
+%% \end{equation*}
+%% \end{enumerate}
 
@@
 
-We prove this using a few lemmas about \texttt{findIndex},
-\texttt{lexsort}, and \texttt{rots}.
+%% We prove this using a few lemmas about \var{findIndex},
+%% \var{lexsort}, and \var{rots}.
 
-\begin{lemma}[findIndex\_correct]
-  For all elements $x$, lists $l$,
-  \begin{equation*}
-    (\exists y \in l,~x \equiv y) \implies l_{(\mathtt{findIndex}~x~xs)} \equiv x
-  \end{equation*}
-  \begin{proof}
-    This follows by induction on $xs$: At every step, either the
-    current element is equivalent to $x$ and we return $0$, or there
-    must be such a $y$ in the rest of the list.
-  \end{proof}
-\end{lemma}
+%% \begin{lemma}[findIndex\_correct]
+%%   For all elements $x$, lists $l$,
+%%   \begin{equation*}
+%%     (\exists y \in l,~x \equiv y) \implies l_{(\var{findIndex}~x~xs)} \equiv x
+%%   \end{equation*}
+%%   \begin{proof}
+%%     This follows by induction on $xs$: At every step, either the
+%%     current element is equivalent to $x$ and we return $0$, or there
+%%     must be such a $y$ in the rest of the list.
+%%   \end{proof}
+%% \end{lemma}
 
-\begin{lemma}[orig\_in\_sorted\_rots]
-  For every non-empty list $l$,
-  \begin{equation*}
-    \exists y \in \mathtt{lexsort}~(\mathtt{rots}~l),~x = y
-  \end{equation*}
-  \begin{proof}
-    Clearly $l$ is in $\mathtt{rots}~l$ at index 0, and because
-    \texttt{lexsort} produces a permutation of its input, $l$ must be
-    in $\mathtt{lexsort}~(\mathtt{rots}~l)$ as well.
-  \end{proof}
-\end{lemma}
+%% \begin{lemma}[orig\_in\_sorted\_rots]
+%%   For every non-empty list $l$,
+%%   \begin{equation*}
+%%     \exists y \in \var{lexsort}~(\var{rots}~l),~x = y
+%%   \end{equation*}
+%%   \begin{proof}
+%%     Clearly $l$ is in $\var{rots}~l$ at index 0, and because
+%%     \var{lexsort} produces a permutation of its input, $l$ must be
+%%     in $\var{lexsort}~(\var{rots}~l)$ as well.
+%%   \end{proof}
+%% \end{lemma}
 
-\begin{theorem}[bwn\_correct]
-  For every non-empty list $xs$,
-  \begin{equation*}
-    (\mathtt{lexsort}~(\mathtt{rots}~xs))_{(\mathtt{bwn}~xs)} = xs
-  \end{equation*}
-  \begin{proof}
-    This follows from \textsf{findIndex\_correct} if we can show that
-    $\exists y \in xs,~x = y$, which we have by \textsf{orig\_in\_sorted\_rots}.
-  \end{proof}
-\end{theorem}
+%% \begin{theorem}[bwn\_correct]
+%%   For every non-empty list $xs$,
+%%   \begin{equation*}
+%%     (\var{lexsort}~(\var{rots}~xs))_{(\var{bwn}~xs)} = xs
+%%   \end{equation*}
+%%   \begin{proof}
+%%     This follows from \textsf{findIndex\_correct} if we can show that
+%%     $\exists y \in xs,~x = y$, which we have by \textsf{orig\_in\_sorted\_rots}.
+%%   \end{proof}
+%% \end{theorem}

### 2019-05-03 draft_paper/draft_paper.tex
Finish permutation

@@
 
-There are multiple ways of defining \var{Permutation} as well: some of
-them are more obviously correct, and some are more useful in certain
-situations. The ``mathematical'' definition of when two lists $l$
-and $l'$ are permutations of each other is:
-\begin{itemize}[nosep]
+The word \textit{permutation} has two related meanings that are often
+used interchangably. We sometimes say that two lists are permutations of each
+other if they contain all of the same elements, just appearing in a
+different order. However, in mathematics we often say that a
+permutation is actually the bijection on indices that defines the
+correspondence between the two lists. For example, we say that $p$ is
+the permutation that relates $l$ to $l'$ if for all indices $i$ and
+$j$,
+\begin{equation*}
+  l_i = l'_{p(i)}
+\end{equation*}
+For consistency with Coq's built-in \var{Permuation} relation, in a
+situation like this we will say that $l$ and $l'$ are permutations of
+each other, and that $p$ is the \textit{permutation function} that
+relates $l$ to $l'$. We say that two lists $l$ and $l'$ are
+permutations of each other is precisely when
+\vspace{-\parskip}
+\begin{itemize}[nolistsep]
   \item $\var{length}~l = \var{length}~l' = n$
@@
 $f : \mathbb{N}_{<n} \to A$ is represented by the $\var{list}~A$
-$[f(0),~f(1),\ldots,f(n-1)]$. This is consistent with the mathematical
+$[f(0),~f(1),\ldots,f(n-1)]$. For permutations, this is consistent with the
 notation where we denote by $2~1~0$ the permutation $p(0) = 2$, $p(1)
@@
 \end{align*}
+
 This leads to the following definition of the property
@@
 \lstinputlisting[
-  firstline=14, lastline=15
+  firstline=14, lastline=15,
 ]{../theories/BWT/Sorting/PermFun.v}
+Then, we can define \var{apply}, which permutes a list according to a
+given permutation:
+\lstinputlisting[
+  widthgobble=2, firstline=161, lastline=165,
+]{../theories/BWT/Sorting/PermFun.v}
+And finally, we can define \var{PermutationEx}, which holds between
+two lists when they are permutations of each other.
+\lstinputlisting[
+  widthgobble=2, firstline=510, lastline=511,
+]{../theories/BWT/Sorting/PermFun.v}
+
+As with \var{Sorted}, we would also like an inductive definition of
+\var{Permutation}. The Coq standard library includes one:
+\begin{align*}
+  \infer[\var{perm\_nil}]{\var{Permutation}~[]~[]}{}
+  &&
+  \infer[\var{perm\_skip}]{\var{Permutation}~(x :: l)~(x :: l')}{%
+    \var{Permutation}~l~l'
+  }
+  \\
+  \\
+  \infer[\var{perm\_swap}]{\var{Permutation}~(x :: y :: l)~(y :: x :: l)}{}
+  &&
+  \infer[\var{perm\_trans}]{\var{Permutation}~l~l''}{
+    \var{Permutation}~l~l' & \var{Permutation}~l'~l''
+  }
+\end{align*}
+
+We will prove these two definitions equivalent, but first we confirm that our definition of \var{PermFun} ensures that .
+
+We know we will
+eventually need to reason inductively about permutation functions, so
+we would like to define the relationship between applying a
+permutation function $i :: p$ and applying $p$. Conceptually, the
+first element of $\var{apply}~(i::p)~l$ will be $l_i$. The rest will
+be the result of applying some new permutation $p'$, where everything
+greater than $i$ has been shifted down by one, to some new list $l'$,
+which has its $i$th element removed. So, we define \var{rem\_PermFun}
+and \var{rem\_nth} respectively, hoping to prove that
+\begin{equation*}
+  \var{apply}~(i::p)~l =
+  l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l)
+\end{equation*}
+\lstinputlisting[
+  widthgobble=4, firstline=229, lastline=229,
+]{../theories/BWT/Sorting/PermFun.v}
+\vspace{\parskip}
+\lstinputlisting[
+  widthgobble=2, firstline=599, lastline=607,
+]{../theories/BWT/Lib/List.v}
+
+These two theorems define how \var{rem\_nth} affects indexing.
+\begin{theorem}[nth\_lt\_rem\_nth, nth\_ge\_rem\_nth]
+  For all lists $l$ and indices $i$ and $j$,
+  \begin{equation*}
+    j < i \implies (\var{rem\_nth}~i~l)_j = l_j
+  \end{equation*}
+  and
+  \begin{equation*}
+    j \ge i \implies (\var{rem\_nth}~i~l)_j = l_{j+1}
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  Both theorems follow by induction on $i$.
+\end{proof}
+
+As we expect, removing the $n$th element and sticking it on the front
+preserves all elements.
+\begin{theorem}[rem\_nth\_Perm]
+  For all list $l$ and indices $i$
+  \begin{equation*}
+    \var{Permutation}~l~(l_i :: \var{rem\_nth}~i~l)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  This follows from induction on $l$.
+\end{proof}
+
+\begin{theorem}[rem\_PermFun\_correct]
+  For all permutation functions $p$, lists $l$, and indices $i$.
+  \begin{equation*}
+    \var{apply}~(i::p)~l =
+    l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  Working backwards from our goal,
+  \begin{align*}
+    \var{apply}~(i::p)~l &=
+    l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l) \\
+    l_i :: \var{map}~(\lambda j.~l_j)~p &=
+    l_i :: \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_j)~(\var{rem\_PermFun}~i~p) \\
+    \var{map}~(\lambda j.~l_j)~p &=
+    \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_j)~(\var{map}~(\lambda j. \bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j)~p) \\
+    \var{map}~(\lambda j.~l_j)~p &=
+    \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_{\bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j})~p \\
+  \end{align*}
+  From here, we need to prove that the mapping functions agree on all
+  the elements of $p$, namely that for all $j \in p$
+  \begin{equation*}
+    l_j = (\var{rem\_nth}~i~l)_{\bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j}
+  \end{equation*}
+  We know that $i \neq j$, because $j \in p$ and $\var{NoDup}~(i :: p)$.
+  Therefore there are two cases, $i < j$ and $i > j$. When $i < j$, we have
+  \begin{equation*}
+    l_j = (\var{rem\_nth}~i~l)_{j-1}
+  \end{equation*}
+  by \var{nth\_ge\_rem\_nth}, and when $i > j$, we have
+  \begin{equation*}
+    l_j = (\var{rem\_nth}~i~l)_j
+  \end{equation*}
+  by \var{nth\_lt\_rem\_nth}.
+\end{proof}
+
+\begin{theorem}[apply\_correct]
+  For all permutation functions $p$ and lists $l$,
+  \begin{equation*}
+    \var{Permutation}~(\var{apply}~p~l)~l
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  First we let $n = \var{length}~l$, so that we can reason by
+  induction on $n$. The base case is uninteresting. In the inductive
+  case, we have the induction hypothesis that for all lists of length
+  $n$ and permutation functions on such lists,
+  \begin{equation*}
+    \var{Permutation}~(\var{apply}~p~l)~l
+  \end{equation*}
+  and we must show that
+  \begin{equation*}
+    \var{Permutation}~(\var{apply}~(i :: p)~(a :: l))~(a :: l)
+  \end{equation*}
+
+  We reason as follows, using the notation $\eqperm$ to
+  chain together \var{Permutation}s:
+  \begin{align*}
+    \var{apply}~(i :: p')~(a :: l) &\eqperm
+    (a :: l)_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a::l)) \\
+    &\eqperm (a :: l)_i :: \var{rem\_nth}~i~(a::l) \\
+    &\eqperm a :: l
+  \end{align*}
+  where the first step follows from \var{rem\_PermFun\_correct}, the
+  second from the induction hypothesis, and the last from
+  \var{rem\_nth\_Perm}.
+\end{proof}
+
+\begin{theorem}[PermutationEx\_iff]
+  For all lists $l$ and $l'$,
+  \begin{equation*}
+    \var{Permutation}~l~l' \iff \var{PermutationEx}~l~l'
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  The $\impliedby$ direction follows from \var{apply\_correct}.
+
+  ($\implies$) We reason by induction on the evidence of
+  $\var{Permutation}~l~l'$. The empty case is simple.
+  \begin{itemize}
+  \item We have a permutation function $p$ from the induction
+    hypothesis such that $\var{apply}~p~l = l'$, and we need to
+    construct a permutation function $p'$ such that
+    $\var{apply}~p'~(x::l) = x :: l'$. We let $p' = 0 ::
+    \var{map}~(+1)~p$, shifting every index up by one and mapping
+    the first element of the list to itself.
+  \item We have to construct a permutation function $p$ such that
+    $\var{apply}~p~(y :: x :: l) = x :: y :: l$. We let $p = 1 :: 0 ::
+    \var{map}~(+2)~(\var{seq}~0~(\var{length}~l))$, leaving every
+    element in place except for the first 2.
+  \item We have permutation functions $p_1$ and $p_2$ from our two
+    induction hypotheses, such that
+    \begin{align*}
+      \var{apply}~p_1~l = l' && \var{apply}~p_2~l' = l''
+    \end{align*}
+    We need to construct a $p'$ such that $\var{apply}~p'~l=l''$, so
+    we let $p' = p_2 \comp p_1$. For the definition of composition of
+    permutations, see \cref{appendix:perm_comp}. Then,
+    \begin{equation*}
+      \var{apply}~(p_2 \comp p_1)~l =
+      \var{apply}~p_2~(\var{apply}~p_1~l) = l'' \qedhere
+    \end{equation*}
+  \end{itemize}
+\end{proof}
+
+For completeness, we will briefly mention a third definition of a
+permutation relation that says that two lists are permutations of one
+another when any element occurs an equal number of times in each:
+\lstinputlisting[
+  widthgobble=2, firstline=14, lastline=15,
+]{../theories/BWT/Sorting/PermutationCount.v}
+\var{count\_occ} counts the number of occurences of a given variable
+using a decision procedure for Leibniz equality. Here that decision
+procedure is \texttt{equiv\_dec} from an \var{EqDec} instance where we
+require the equivalence relation itself by Leibniz equality.
+
+\begin{theorem}[PermutationCount\_iff]
+  For all lists $l$ and $l'$,
+  \begin{equation*}
+    \var{Permutation}~l~l' \iff \var{PermutationCount}~l~l'
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  The ($\implies$) direction follows from induction in the evidence of
+  $\var{Permutation}~l~l'$.
+
+  In the ($\impliedby$) direction, we reason by induction on $l$. In
+  the inductive case, we have to show $\var{Permutation}~(a::l)~l'$.
+  We know that $a \in l'$ because $\var{count\_occ}~a~l' =
+  \var{count\_occ}~a~(a::l) > 0$. That means we can deconstruct $l'$
+  as $L \dplus a :: R$, and
+  \begin{equation*}
+    L \dplus a :: R \eqperm a :: L \dplus R \eqperm a :: l
+  \end{equation*}
+  where the last step follows from the induction hypothesis.
+\end{proof}
 
@@
 \label{sec:BWT_proof}
+
+\appendix
+
+\section{Permutation Composition}
+\label{appendix:perm_comp}
 

### 2019-05-03 draft_paper/draft_paper.tex
stash

@@
 \Cref{sec:sort} implements insertion sort and radixsort, and proves
-their correctness. \Cref{sec:BWT_proof} more formally proves the
+their correctness. \Cref{appendix:BWT_proof} more formally proves the
 theorems about the BWT mentioned in previous sections, including
@@
 doing the same thing. We will prove this in
-\cref{sec:sorted_perm_stable,sec:sort}.
-
-%% In seeking to introduce an extra \verb|map rrot| term, we can note
-%% that rotating and sorting an extra time doesn't change anything when
-%% we are sorting the rotation matrix. This works because rotating the
-%% last column to the front produces a matrix that is sorted on its last
-%% \verb|n - 1| characters, so sorting stably on the first column will
-%% re-sort the matrix. Crucially though, rotating the last column
-%% forwards also permutes the entire matrix, so in addition to having a
-%% sorted list, we have a sorted list of the same elements.
-
-%% Using that we can show
-%% \begin{equation*}
-%%   lexsort (rots x)
-%% = { because rrot permutes its elements }
-%%   lexsort (rrot (rots x))
-%% = { because hdsort permutes its elements }
-%%   lexsort (hdsort (rrot (rots x)))
-%% = { radix-sort }
-%%   rep n (hdsort ∘ rrot) (hdsort (rrot (rots x)))
-%% = { rep_absorb_r }
-%%   rep (S n) (hdsort ∘ rrot) (rots x)
-%% = { rep_split_l }
-%%   hdsort (rrot (rep n (hdsort ∘ rrot) (rots x)))
-%% = { radix-sort }
-%%   hdsort (rrot (lexsort (rots x)))
-%% \end{equation*}
-
-%% From here we can calculate
-%% \begin{equation*}
-%%   cols (S j) (lexsort (rots l))
-%% = cols (S j) (hdsort (map rrot (lexsort (rots l))))
-%% = { hdsort commutes with cols (S j) }
-%%   hdsort (cols (S j) (map rrot (lexsort (rots l))))
-%% = { reduce cols (S j) }
-%%   hdsort (prepend_col (map last (lexsort (rots l))) (cols j (lexsort (rots l)))
-%% = { by IH and def of bwp }
-%%   hdsort (prepend_col (bwp l) (recreate j (bwp l)))
-%% = (fun x => hdsort (prepend_col x (recreate j x))) bwp
-%% \end{equation*}
-
-%% More precisely, we can characterize how \verb|map rrot| permutes
-%% \verb|rots|
-%% \begin{equation*}
-%% map rrot (rots x) = rrot (rots x)
-%% \end{equation*}
-
-%% From here we can calculate
-%% \begin{equation*}
-%%   cols (S j) (lexsort (rots l))
-%% = cols (S j) (hdsort (map rrot (lexsort (rots l))))
-%% = { hdsort commutes with cols (S j) }
-%%   hdsort (cols (S j) (map rrot (lexsort (rots l))))
-%% = { reduce cols (S j) }
-%%   hdsort (prepend_col (map last (lexsort (rots l))) (cols j (lexsort (rots l)))
-%% = { by IH and def of bwp }
-%%   hdsort (prepend_col (bwp l) (recreate j (bwp l)))
-%% = (fun x => hdsort (prepend_col x (recreate j x))) bwp
-%% \end{equation*}
-
-%% We can use this to define \verb|recreate|:
-%% \begin{equation*}
-%% Fixpoint recreate (j : nat) (l : list A) : list (list A) :=
-%%   match j with
-%%   | O    => map (const []) l
-%%   | S j' => hdsort (prepend_col l (recreate j' l))
-%%   end.
-%% \end{equation*}
+\cref{sec:unique,sec:sort}.
+
+Using that we can show
+\begin{align*}
+     \var{lexsort}~(\var{rots}~x)
+  &= \var{lexsort}~(\var{rrot}~(\var{rots}~x))
+     && \text{because rrot permutes its elements} \\
+  &= \var{lexsort}~(\var{hdsort}~(\var{rrot}~(\var{rots}~x)))
+     && \text{because hdsort permutes its elements} \\
+  &= (\var{hdsort} \comp \var{rrot})^n~(\var{hdsort}~(\var{rrot}~(\var{rots}~x)))
+     && \text{radix-sort} \\
+  &= (\var{hdsort} \comp \var{rrot})^{n+1}~(\var{rots}~x)
+  && \text{rep\_absorb\_r} \\
+  &= \var{hdsort}~(\var{rrot}~((\var{hdsort} \comp \var{rrot})^n~(\var{rots}~x)))
+     && \text{ rep\_split\_l } \\
+  &= \var{hdsort}~(\var{rrot}~(\var{lexsort}~(\var{rots}~x)))
+     && \text{ radix-sort }
+\end{align*}
+
+In seeking to introduce an extra \var{map rrot} term, we can note that
+rotating and sorting an extra time doesn't change anything when we are
+sorting the rotation matrix. This works because rotating the last
+column to the front produces a matrix that is sorted on its last $n -
+1$ characters, so sorting stably on the first column will re-sort the
+matrix. Crucially though, rotating the last column forwards also
+permutes the entire matrix, so in addition to having a sorted list, we
+have a sorted list of the same elements.
+
+More precisely, we can characterize how $\var{map}~\var{rrot}$
+permutes \var{rots}.
+\begin{equation*}
+  \var{map}~\var{rrot}~(\var{rots}~x) = \var{rrot}~(\var{rots}~x)
+\end{equation*}
+
+From here we can calculate
+\begin{align*}
+  &\var{cols}~(j+1)~(\var{lexsort}~(\var{rots}~l)) \\
+&= \var{cols}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
+&= \var{hdsort}~(\var{cols}~(j+1)~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l))))
+&&\text{ \var{hdsort} commutes with $\var{cols}~(j+1)$ } \\
+&= \var{hdsort}~(\var{prepend\_col}~(\var{map}~\var{last}~(\var{lexsort}~(\var{rots}~l)))~(\var{cols}~j~(\var{lexsort}~(\var{rots}~l)))
+  &&\text{ reduce $\var{cols}~(j+1)$ } \\
+&= \var{hdsort}~(\var{prepend\_col}~(\var{bwp}~l)~(\var{recreate}~j~(\var{bwp}~l)))
+  &&\text{ by IH and def of bwp } \\
+&= (\lambda x.~\var{hdsort}~(\var{prepend\_col}~x~(\var{recreate}~j~x)))~\var{bwp}
+\end{align*}
+
+We can use this to define \var{recreate}:
+\lstinputlisting[
+  widthgobble=2, firstline=156, lastline=160,
+]{../theories/BWT/BurrowsWheeler.v}
 
@@
 By transitivity it suffices to check that the new element is
-smaller than just the first element of the sorted list. That leads to
-a third definition of sorted:
+smaller than just the first element of the sorted list. That leads to a third definition of sorted:
 \begin{gather*}
@@
 \subsection{\var{Permutation}}
+\label{subsec:perm}
 
@@
 \begin{equation*}
-  l_i = l'_{p(i)}
+  l_{p(i)} = l'_i
 \end{equation*}
@@
     \begin{equation*}
-      l_i = l'_{p(i)}
+      l_{p(i)} = l'_i
     \end{equation*}
@@
 
-%% In this section we prove the lemmas from the previous section:
-%% \begin{enumerate}
-%% \item
-%% \begin{equation*}
-%% map rrot (rots x) = rrot (rots x)
-%% \end{equation*}
-
-%% \item \verb|hdsort| commutes with \verb|cols (S j)|
-%%   This proof is cool because in theory it can be proven via
-%%   parametricity, but that's a metatheorem in Coq.
-
-%% \item
-%% \begin{equation*}
-%% cols (S j) (map rrot l) = prepend_col (map last l) (cols j l)
-%% \end{equation*}
-%% \end{enumerate}
+\subsection{\var{StablePerm}}
+\label{subsec:stable}
+
+A stable sort is a sort that doesn't swap any two elements considered
+equivalent by the ordering. This is relevant when the induced
+equivalence relation does not imply equality, so that we can
+distinguish between elements that the sort considers identical. A good
+example of this is in radixsort: when we sort on a single column, the
+sort can't tell the difference between strings that contain equal
+characters in that column, but it is crucial that it respect the
+initial ordering in cases of a tie.
+
+For stable sorts, we need a stronger property than \var{Permutation}.
+We can extend our mathematical definition. We say that two lists $l$
+and $l'$ are stable permutations of each other when we have a
+permutation $p$ that relates them, and that satisfies an additional
+stability property: for all indices $i$ and $j$ where
+$(\var{apply}~p~l)_i \equiv (\var{apply}~p~l)_j$
+\begin{gather*}
+  i < j \implies p(i) < p(j)
+\end{gather*}
+Equivalently, we can say that for all indices $i$ and $j$ where $l_j \equiv
+l_j$ we must have
+\begin{gather*}
+  i < j \implies p^{-1}(i) < p^{-1}(j)
+\end{gather*}
+
+Based on this, we define a stronger version of the \var{PermFun}
+property, \var{StablePermFun}. However, while $\var{PermFun}~n~p$
+means that $p$ represents a permutation function on lists of length
+$n$, we need to specify the list with resepct to which $p$ is stable,
+so we say $\var{StablePermFun}~l~p$ for some list $l$.
+\lstinputlisting[
+  widthgobble=2, firstline=557, lastline=561,
+]{../theories/BWT/Sorting/PermFun.v}
+As above, we can also define this property using the inverse
+permutation:
+\lstinputlisting[
+  widthgobble=2, firstline=563, lastline=567,
+]{../theories/BWT/Sorting/PermFun.v}
+As usual, we will prove these definitions equivalent. First, though,
+we need to define \var{image} and \var{preimage}.
+\lstinputlisting[
+  widthgobble=2, firstline=85, lastline=86,
+]{../theories/BWT/Sorting/PermFun.v}
+We will not prove it here, but $\var{preimage}~p$ and $\var{image}~p$
+are both of bijections of type $\mathbb{N}_{<n} \to \mathbb{N}_{<n}$
+and are invereses of each other. We also have the following two
+theorems:
+\begin{theorem}[nth\_preimage\_apply, nth\_image\_apply]
+  For all lists $l$, permutations $p$, and indices $i$,
+  \begin{align*}
+    (\var{apply}~p~l)_{p^{-1}(i)} = l_i && l_{p(i)} = (\var{apply}~p~l)_i
+  \end{align*}
+\end{theorem}
+\begin{proof}
+  The first equation (\var{nth\_preimage\_apply}) holds because:
+  \begin{align*}
+       (\var{apply}~p~l)_{p^{-1}(i)}
+    &= (\var{map}~(\lambda j.~l_j)~p)_{\var{findIndex}~i~p} \\
+    &= l_{p_{(\var{findIndex}~i~p)}} \\
+    &= l_i
+  \end{align*}
+  and the second follows from this:
+  \begin{equation*}
+      (\var{apply}~p~l)_i
+    = (\var{apply}~p~l)_{p^{-1}(p(i))}
+    = l_{p(i)}
+    \qedhere
+  \end{equation*}
+\end{proof}
+
+\begin{theorem}[StablePermFun\_iff]
+  For all lists $l$ and permutation $p$,
+  \begin{equation*}
+    \var{StablePermFun}~l~p \iff \var{StablePermFun\_preimage}~l~p
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We will only prove the $\implies$ direction; the other direction is
+  very similar. We have that $l_i \equiv l_j$, and that $i < j$, and we
+  have to show that $p^{-1}(i) < p^{-1}(j)$. We will show this by
+  proving that $p^{-1}(i) \ge p^{-1}(j)$ leads to a contradiction.
+
+  Suppose $p^{-1}(i) \ge p^{-1}(j)$. We actually know that $p^{-1}(i) >
+  p^{-1}(j)$ because $i \neq j$. From $l_i \equiv l_j$ and
+  $\var{StablePermFun}~l~p$, along with $p^{-1}(j) < p^{-1}(i)$, we
+  know that $p(p^{-1}(j)) < p(p^{-1}(i))$. However, this implies $j <
+  i$, which is a contradiction. Therefore, $p^{-1}(i) < p^{-1}(j)$.
+\end{proof}
+
+Now, we can define \var{StablePermEx} to describe when two lists are
+stable permutations of each other:
+\lstinputlisting[
+  widthgobble=2, firstline=607, lastline=608,
+]{../theories/BWT/Sorting/PermFun.v}
+
+While \var{StablePermEx} corresponds to \var{PermutationEx}, how can
+we modify \var{Permutation} to create an inductive relation for stable
+lists? Intuitively, the problematic inference rule is
+\var{Permutation\_swap}: we can only swap two elements and preserve
+stability if they are not equivalent. With this fix, we have \var{StablePermInd}:
+\begin{gather*}
+  \infer[\var{StablePermInd\_nil}]{\var{StablePermInd}~[]~[]}{}
+  \\ \\
+  \infer[\var{StablePermInd\_skip}]{\var{StablePermInd}~(x :: l)~(x :: l')}{%
+    \var{StablePermInd}~l~l'
+  }
+  \\ \\
+  \infer[\var{StablePermInd\_swap}]
+        {\var{StablePermInd}~(x :: y :: l)~(y :: x :: l)}
+        {x \not\equiv y}
+  \\ \\
+  \infer[\var{StablePermInd\_trans}]{\var{StablePermInd}~l~l''}{
+    \var{StablePermInd}~l~l' & \var{StablePermInd}~l'~l''
+  }
+\end{gather*}
+
+Unforunately, neither of these two definitions will be particularly
+useful for proving the theorems we would like to prove with them. The
+best definition for stability does not come by analogy with
+\var{Permutation}. The intuition is that for any two lists that are
+stable permutations of each other, if we filter each list by
+equivalence with any element, the resulting lists should be identical.
+This is because they must have the same elements, since the original
+lists are permutations of one another, and those elements must be in
+the same order, since the original lists are specifically stable
+permutations of one another.
+\lstinputlisting[
+  widthgobble=2, firstline=21, lastline=22,
+]{../theories/BWT/Sorting/Stable.v}
+
+We will prove equivalence between these three definitions, but first
+we will mention some properties of \var{Stable}. \var{Stable} is an
+equivalence relation, meaning it is reflexive, transitive, and symmetric.
+
+\begin{lemma}[StablePerm\_app]
+  For all lists $l$, $l'$, $m$ and $m'$,
+  \begin{equation*}
+    \var{StablePerm}~l~l' \implies \var{StablePerm}~m~m' \implies
+    \var{StablePerm}~(l \dplus m)~(l' \dplus m')
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  This follows from the distributivity of \var{filter} over \var{app}.
+\end{proof}
+
+\begin{lemma}[StablePerm\_skip]
+  For all lists $l$ and $l'$, and elements $a$,
+  \begin{equation*}
+    \var{StablePerm}~l~l' \implies \var{StablePerm}~(a::l)~(a::l')
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  For any $x$, we have to show $\var{filter}~(\equiv x)~(a::l) =
+  \var{filter}~(\equiv x)~(a::l')$. If $x \equiv a$, $a$ is added to both lists,
+  and if $x \not\equiv a$ it is added to neither.
+\end{proof}
+
+\begin{lemma}[StablePerm\_swap]
+  For all lists $l$ and $l'$, and elements $a$ and $b$ where $a \not\equiv b$,
+  \begin{equation*}
+     \var{StablePerm}~(a::b::l)~(b::a::l)
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  For any $x$, we have to show $\var{filter}~(\equiv x)~(a::b::l) =
+  \var{filter}~(\equiv x)~(b::a::l)$. The proof follows from case analysis
+  on $x \equiv a$ and $x \equiv b$.
+\end{proof}
+
+\begin{lemma}[StablePerm\_cons\_app]
+  For all lists $l$, $l_1$ and $l_2$, and elements $a$ such that $\forall b
+  \in l_1,~a \not\equiv b$,
+  \begin{equation*}
+     \var{StablePerm}~l~(l_1 \dplus l_2) \implies \var{StablePerm}~(a
+     :: l)~(l_1 \dplus a :: l_2)
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  For any $x$, we have to show
+  \begin{equation*}
+    \var{filter}~(\equiv x)~(a::l) = \var{filter}~(\equiv x)~(l_1 \dplus a :: l_2)
+  \end{equation*}
+  If $x \not\equiv a$ the proof is easy, so assume $x \equiv a$. Then we have
+  \begin{align*}
+      \var{filter}~(\equiv x)~(l_1 \dplus a :: l_2)
+    &= \var{filter}~(\equiv x)~l_1 \dplus \var{filter}~(a :: l_2) \\
+    &= [] \dplus a::\var{filter}~l_2 \\
+    &= a::\var{filter}~l
+  \end{align*}
+  where we know that $\var{filter}~(\equiv x)~l_1 = []$ because $\forall b \in
+  l_1,~a \not\equiv b$, and the last step follows from
+  $\var{StablePerm}~l~(l_1 \dplus l_2)$.
+\end{proof}
+
+\begin{theorem}[StablePerm\_destr]
+  For all lists $l$ and $l'$, and elements $h$ and $h'$ such that $h
+  \not\equiv h'$,
+  \begin{gather*}
+    \var{StablePerm}~(h :: l)~(h' :: l') \implies \\
+    \exists~l_1~l_2, ~l = l_1 \dplus h' :: l_2 \land (\forall x \in l_1, h' \not\equiv x)
+  \end{gather*}
+\end{theorem}
+\begin{proof}
+  Let $l_1 = \var{take\_while}~(\not\equiv h')~l$ and $l_2 =
+  \var{drop\_while}~(\not\equiv h')~l$. Clearly $l = l_1 \dplus l_2$, and
+  because $\var{StablePerm}~(h :: l)~(h' :: l')$, we know the first
+  element of $l_2$ must be $h'$. Let $l_2'$ be the rest of $l_2$, and
+  we have two lists $l_1$ and $l_2'$ that satisfy the theorem.
+\end{proof}
+
+\begin{theorem}[StablePerm\_length]
+  For all lists $l$ and $l'$
+  \begin{equation*}
+    \var{StablePerm}~l~l' \implies \var{length}~l = \var{length}~l'
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We reason by induction on $n = \var{length}~l$. In the inductive
+  case, we need to show that $\var{length}~(h::t) =
+  \var{length}~(h'::t')$ given that $\var{StablePerm}~(h::t)~(h'::t')$
+  and the induction hypothesis that for all lists of length $n$,
+  \var{StablePerm} implies their lengths are equal.
+
+  If $h \equiv h'$, then from $\var{StablePerm}~(h::t)~(h'::t')$ we know
+  that $h = h'$. We can then remove $h$ from the front of $h::t$ and
+  $h::t'$ and apply the induction hypothesis.
+
+  If $h \not\equiv h'$, we can destructure $t$ and $t'$ as $l_1 \dplus h'
+  :: l_2$ and $l_1' \dplus h :: l_2'$ respectively, using
+  \var{StablePerm\_destr}. Then,
+  \begin{equation*}
+      \var{length}~t
+    = \var{length}~(h' :: l_1  \dplus l_2)
+    = \var{length}~(h  :: l_1' \dplus l_2')
+    =\var{length}~t'
+  \end{equation*}
+  where the first and last steps follow from the induction hypothesis,
+  using the fact that $\var{StablePerm}~t~(h'::l_1 \dplus l_2)$ and
+  $\var{StablePerm}~t'~(h::l_1' \dplus l_2')$.
+\end{proof}
+
+\begin{theorem}[StablePerm\_Perm]
+  For all lists $l$ and $l'$,
+  \begin{equation*}
+    \var{StablePerm}~l~l' \implies \var{Permutation}~l~l'
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We reason by strong induction on the length of both lists (we know
+  the lengths are equal by \var{StablePerm\_length}). In the inductive
+  case, we need to show $\var{Permutation}~(h :: t)~(h' :: t')$. If $h
+  \equiv h$, we know that $h = h'$, and the proof follows by
+  \var{perm\_skip}.
+
+  If $h \not\equiv$, we again destructure $t$ and $t'$ as $l_1 \dplus h' ::
+  l_2$ and $l_1' \dplus h :: l_2'$ respectively, using
+  \var{StablePerm\_destr}.
+  \begin{align*}
+    h::t &= h :: l_1 \dplus h' :: l_2 \\
+    &\eqperm h :: h' :: l_1 \dplus l_2 \\
+    &\eqperm h' :: h :: l_1 \dplus l_2 \\
+    &\eqperm h' :: h :: l_1' \dplus l_2' \\
+    &\eqperm h' :: l_1' \dplus h :: l_2' = h'::t
+  \end{align*}
+  The crucial step is $l_1 \dplus l_2 \eqperm l_1' \dplus l_2'$, which
+  follows from the induction hypothesis.
+\end{proof}
+
+\begin{theorem}[StablePermInd\_iff]
+  For all lists $l$ and $l'$
+  \begin{equation*}
+    \var{StablePerm}~l~l' \iff \var{StablePermInd}~l~l'
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  The ($\impliedby$) direction follows easily by induction on
+  $\var{StablePermInd}~l~l'$, using \var{StablePerm\_skip},
+  \var{StablePerm\_swap}, and \var{StablePerm\_trans}.
+
+  The ($\implies$) direction is proved very similarly to
+  \var{StablePerm\_Perm}, using \var{StablePerm\_destr} to split up
+  the lists.
+\end{proof}
+
+Now we prove equivalence with \var{StablePermEx}.
+
+\begin{theorem}[apply\_correct\_stable]
+  For all lists $l$ and permutation functions $p$,
+  \begin{equation*}
+    \var{StablePermFun}~l~p \implies \var{StablePerm}~(\var{apply}~p~l)~l
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We reason by induction on the length of $l$. In the inductive case,
+  we need to show
+  \begin{equation*}
+    \var{StablePerm}~(\var{apply}~(i :: p)~(a :: l))~(a :: l)
+  \end{equation*}
+
+  Using $\eqstable$ to chain \var{StablePerm}s transitively,
+  \begin{align*}
+    \var{apply}~(i :: p)~(a :: l)
+    &= (a :: l)_i  ::
+    \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a :: l)) \\
+    &\eqstable (a :: l)_i :: \var{rem\_nth}~i~(a :: l) \\
+    &\eqstable (a :: l)_i
+  \end{align*}
+  where we rely on the induction hypothesis between the first and
+  second lines.
+\end{proof}
+
+\begin{theorem}[StablePermEx\_iff]
+  For all lists $l$ and $l'$
+  \begin{equation*}
+    \var{StablePerm}~l~l' \iff \var{StablePermEx}~l~l'
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We use \var{StablePermInd\_iff} and prove $\var{StablePermInd}~l~l'
+  \iff \var{StablePermEx}~l~l'$.
+
+  The $\implies$ direction is very similar to the $\implies$ direction
+  of the \var{PermutationEx\_iff} proof: we reason by induction on
+  $\var{StablePermInd}~l~l'$, constructing the same permutation
+  functions in each case, but this time we have to prove that it is
+  stable as well. The $\impliedby$ direction follows from
+  \var{apply\_correct\_stable}.
+\end{proof}
+
+\section{Uniqueness of Stable Sorts}
+\label{sec:unique}
+
+Using our definitions of \var{StablePerm} and \var{Sorted}, we can
+prove the theorem that allowed us to replace \var{lexsort} with
+radixsort.
+
+\begin{theorem}[stable\_sort\_unique]
+  \begin{equation*}
+    \var{Sorted}~l \implies \var{Sorted}~l' \implies
+    \var{StablePerm}~l~l' \implies l = l'.
+  \end{equation*}
+\end{theorem}
+
+With \var{lexsort}, we are guaranteed stability because we are using
+an order, not just a preorder, and for orders all sorts are stable.
+
+\begin{theorem}[all\_perm\_stable]
+  If we have that equivalence implies Leibniz equality, then for all
+  lists $l$ and $l'$
+  \begin{equation*}
+      \var{Permutation}~l~l' \implies \var{StablePerm}~l~l'
+  \end{equation*}
+\end{theorem}
 
@@
 
+\lstinputlisting[
+  widthgobble=2, firstline=11, lastline=21,
+]{../theories/BWT/Sorting/InsertionSort.v}
+
+\begin{theorem}[sort\_sorted]
+  \begin{equation*}
+    \var{Sorted}~(\var{sort}~l)
+  \end{equation*}
+\end{theorem}
+
+\begin{theorem}[sort\_stable]
+  \begin{equation*}
+    \var{Stable}~(\var{sort}~l)
+  \end{equation*}
+\end{theorem}
+
+\lstinputlisting[
+  widthgobble=2, firstline=22, lastline=23,
+]{../theories/BWT/Sorting/RadixSort.v}
+
+\begin{theorem}[radixsort\_sorted]
+  \begin{equation*}
+    (\forall x \in l,~\var{length}~x = n) \implies \var{Sorted}~(\var{radixsort}~l~n)
+  \end{equation*}
+\end{theorem}
+
+\appendix
+
 \section{Proving BWT Correct}
-\label{sec:BWT_proof}
-
-\appendix
+\label{appendix:BWT_proof}
+
+In this section we prove the lemmas from the derivation that we left out:
+\begin{enumerate}
+\item
+\begin{equation*}
+  \var{map}~\var{rrot}~(\var{rots}~x) = \var{rrot}~(\var{rots}~x)
+\end{equation*}
+
+\item \var{hdsort} commutes with $\var{cols} (j+1)$.
+  This proof is cool because in theory it can be proven via
+  parametricity, but that's a metatheorem in Coq.
+
+\item
+  \begin{equation*}
+    \var{cols}~(j+1)~(\var{map}~\var{rrot}~l) = \var{prepend\_col}~(\var{map}~\var{last}~l)~(\var{cols}~j~l)
+  \end{equation*}
+\end{enumerate}
 
@@
 
+\begin{theorem}[compose\_apply]
+  For all lists $l$ and permutation functions $p_1$ and $p_2$
+  \begin{equation*}
+      \var{apply}~(p_2 \comp p_1)~l = \var{apply}~p2~(\var{apply}~p1~l)
+  \end{equation*}
+\end{theorem}
+
 \printbibliography{}
@@
 \end{document}
-
-%% We prove this using a few lemmas about \var{findIndex},
-%% \var{lexsort}, and \var{rots}.
-
-%% \begin{lemma}[findIndex\_correct]
-%%   For all elements $x$, lists $l$,
-%%   \begin{equation*}
-%%     (\exists y \in l,~x \equiv y) \implies l_{(\var{findIndex}~x~xs)} \equiv x
-%%   \end{equation*}
-%%   \begin{proof}
-%%     This follows by induction on $xs$: At every step, either the
-%%     current element is equivalent to $x$ and we return $0$, or there
-%%     must be such a $y$ in the rest of the list.
-%%   \end{proof}
-%% \end{lemma}
-
-%% \begin{lemma}[orig\_in\_sorted\_rots]
-%%   For every non-empty list $l$,
-%%   \begin{equation*}
-%%     \exists y \in \var{lexsort}~(\var{rots}~l),~x = y
-%%   \end{equation*}
-%%   \begin{proof}
-%%     Clearly $l$ is in $\var{rots}~l$ at index 0, and because
-%%     \var{lexsort} produces a permutation of its input, $l$ must be
-%%     in $\var{lexsort}~(\var{rots}~l)$ as well.
-%%   \end{proof}
-%% \end{lemma}
-
-%% \begin{theorem}[bwn\_correct]
-%%   For every non-empty list $xs$,
-%%   \begin{equation*}
-%%     (\var{lexsort}~(\var{rots}~xs))_{(\var{bwn}~xs)} = xs
-%%   \end{equation*}
-%%   \begin{proof}
-%%     This follows from \textsf{findIndex\_correct} if we can show that
-%%     $\exists y \in xs,~x = y$, which we have by \textsf{orig\_in\_sorted\_rots}.
-%%   \end{proof}
-%% \end{theorem}

### 2019-05-04 draft_paper/draft_paper.tex
stash

@@
 \maketitle
+\vfill
 
@@
 that makes a string more amenable to compression by other methods
-\cite{bw}. The forwards transform is straightforward to describe: for
+\cite{bw}. It accomplishes this by creating runs of repeated letters
+in the produced string. Of course, a simple way to create runs is to
+sort the string, but that would be useless for compression because the
+original string would be unrecoverable.
+
+The forwards transform is straightforward to describe: for
 a string of length \(n\), form the \(n \times n\) rotaion matrix of all
@@
 
-\begin{figure}
+\begin{figure}[!hb]
   \centering
@@
   widthgobble=2, firstline=21, lastline=22,
-]{../theories/BWT/Sorting/Stable.v}
+]{../theories/BWT/Sorting/StablePerm.v}
 
@@
 
-\section{Insertion Sort, Radixsort, and \texttt{lexsort}}
+\section{Radixsort}
 \label{sec:sort}
+
+As
 

### 2019-05-05 draft_paper/draft_paper.tex
Stash

@@
-\documentclass[11pt]{thesis}
-
-\title{Towards a Verified Bzip2: A Functional Specification for Burrows-Wheeler}
+\documentclass[11pt, titlepage]{thesis}
+
+\title{Towards a Verified Bzip2: A Functional Specification for Burrows--Wheeler}
 \author{Jake Waksbaum}
@@
 \maketitle
-\vfill
+
+\tableofcontents
+\clearpage
 
@@
 
-The Burrows-Wheeler transform (BWT) is an invertible transformation
+The Burrows--Wheeler transform (BWT) is an invertible transformation
 that makes a string more amenable to compression by other methods
@@
 
-Bird and Mu \cite{birdmu,pearls} analyze the Burrows-Wheeler algorithm
+Bird and Mu \cite{birdmu,pearls} analyze the Burrows--Wheeler algorithm
 in a functional setting. They express the algorithm as the composition
@@
 inverse transform, this provides the outline of a proof of the
-correctness of the Burrows-Wheeler algorithm. We give our version of
+correctness of the Burrows--Wheeler algorithm. We give our version of
 their derivation in \cref{sec:forwards_BWT,sec:inverse_BWT}.
@@
 specification for a potential verified implementation of a program
-that relies on Burrows-Wheeler \cite{appel-func-spec}. But more
+that relies on Burrows--Wheeler \cite{appel-func-spec}. But more
 generally, this provides evidence that Bird's approach of program
@@
 
-\section{Burrows-Wheeler}
+\section{Burrows--Wheeler}
 \label{sec:bwt}
 
-The Burrows-Wheeler transform was originally introduced as a pass in a
+The Burrows--Wheeler transform was originally introduced as a pass in a
 compression algorithm. First the input is transformed with BWT, which
@@
 and \Cref{eq:unbwt} would follow directly from the \var{bwn\_correct}:
-\begin{align}
-     \label{eq:unbwt_correct_start}
+\begin{align*}
      \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l)
-  &= (\var{recreate}~(\var{bwp}~l))_{(\var{bwn})~l} &&\\
-  &= (\var{lexsort}~(\var{rots}~l))_{(\var{bwn})~l}
-     &&\text{\{by \Cref{eq:recreate}\}}\\
-  &= l &&\text{\{\var{recreate\_correct}\}}
-     \label{eq:unbwt_correct_end}
-\end{align}
-
-How, then, do we implement \var{recreate} such that
-\Cref{eq:recreate} holds? One idea is to recreate the sorted rotation
-matrix column by column. We define $\var{cols}~j$ to take the first
-$j$ columns of a matrix:
+  &= (\var{recreate}~(\var{bwp}~l))_{(\var{bwn})~l}\\
+  &= (\var{lexsort}~(\var{rots}~l))_{(\var{bwn})~l} \\
+  &= l
+\end{align*}
+
+How, then, do we implement \var{recreate} such that \Cref{eq:recreate}
+holds? A key insight is that because any two methods of sorting the
+rotation matrix must produce the same output, we can pretend that
+\var{lexsort} is implemented using an sorting algorithm we find
+convenient. Specifically, if we examine how an LSD radixsort sorts the
+rotation matrix (\Cref{fig:radixsort_rots}) we notice an interesting
+pattern. Each column of the rotation matrix is a permutation of the
+original string, so when we sort a on particular column, we always
+produce the same result in that column. As we move right-to-left, the
+column containing the sorted input string moves with us until it
+reaches the left edge of the matrix.
+
+\begin{figure}[!hb]
+  \centering
+  \begin{tt}
+  \setlength{\tabcolsep}{0pt}
+  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    abracadabra&!& \\
+    bracadabra!&a& \\
+    cadabra!abr&a& \\
+    dabra!abrac&a& \\
+    bra!abracad&a& \\
+    !abracadabr&a& \\
+    racadabra!a&b& \\
+    ra!abracada&b& \\
+    adabra!abra&c& \\
+    abra!abraca&d& \\
+    acadabra!ab&r& \\
+    a!abracadab&r&
+  \end{tabular}
+  $\rightarrow \cdots \rightarrow$
+  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    racadabra&!&ab \\
+    bracadabr&a&!a \\
+    acadabra!&a&br \\
+    a!abracad&a&br \\
+    dabra!abr&a&ca \\
+    bra!abrac&a&da \\
+    cadabra!a&b&ra \\
+    !abracada&b&ra \\
+    abra!abra&c&ad \\
+    ra!abraca&d&ab \\
+    abracadab&r&a! \\
+    adabra!ab&r&ac
+  \end{tabular}
+  $\rightarrow \cdots \rightarrow$
+  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    a&!&abracadabr \\
+    r&a&!abracadab \\
+    d&a&bra!abraca \\
+    !&a&bracadabra \\
+    r&a&cadabra!ab \\
+    c&a&dabra!abra \\
+    a&b&ra!abracad \\
+    a&b&racadabra! \\
+    a&c&adabra!abr \\
+    a&d&abra!abrac \\
+    b&r&a!abracada \\
+    b&r&acadabra!a
+  \end{tabular}
+  $\rightarrow$
+  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    &!&abracadabra \\
+    &a&!abracadabr \\
+    &a&bra!abracad \\
+    &a&bracadabra! \\
+    &a&cadabra!abr \\
+    &a&dabra!abrac \\
+    &b&ra!abracada \\
+    &b&racadabra!a \\
+    &c&adabra!abra \\
+    &d&abra!abraca \\
+    &r&a!abracadab \\
+    &r&acadabra!ab
+  \end{tabular}
+  \end{tt}
+  \caption{Sorting the rotation matrix with LSD radixsort. The
+    highlighted column contains the sorted original string.}
+  \label{fig:radixsort_rots}
+\end{figure}
+
+We also can notice that the columns to the right of the sorted column
+move with it for similar reasons. The column directly to its right is
+the result of sorting the input string, and then stably sorting it
+again by the character that precedes it in the original string. The
+next column is the result of sorting by each character, sorting by the
+preceding chracter, and then sorting by the character 2 spaces in
+front.
+
+All this means that the columns to the left of the sorted column seem
+like they don't matter, which suggests that we can recreate the entire
+rotation matrix from just the last column (\Cref{fig:recreate}).
+However, the column direclty to the left of the sorted column is very
+important, as it is the next column to be sorted. And we can see that
+this column changes, only becoming the eventual last column of the
+matrix in the second to last iteration. Why then, does it still work
+if we use the last column everywhere?
+
+We can think of the process of sorting the matrix as computing the
+last column subject to the constraints of the lexicographic ordering.
+For example, we know that the collection of all characters preceding
+an \texttt{a} in the sorted column must be some permutation of
+$[\texttt{c},~\texttt{d},~\texttt{r},~\texttt{r},~\texttt{!}]$,
+because those are the characters that preced \texttt{a}s in the
+original string. However, the order of these characters in the final
+string
+($[\texttt{r},~\texttt{d},~\texttt{!},~\texttt{r},~\texttt{c}]$) is
+important. An \texttt{r} must be first, because in the original string
+there is an \texttt{a} following an \texttt{r} which is itself
+followed by a \texttt{!}, the smallest character. Similarly, although
+the \texttt{d} and \texttt{!} precede \texttt{a}s that are both
+followed by \texttt{bra}s, the next character for the \texttt{d} is a
+\texttt{!}, and the next character for the \texttt{!} is a \texttt{c}.
+This requires that the \texttt{d} preced the \texttt{!} in the last
+column.
+
+Generally, there are constraints on the relative order of the letters
+in the column before the sorted column, and in each step we are
+ensuring that these constraints are not violated one more character
+deep in the string. By the time we have reached the last column, we
+have satisfied all the constraints.
+
+We can also think of this as computing the fixpoint of $\var{hdsort} \circ
+\var{map}~\var{rrot}$: we continually move the rightmost column to the
+front and sort on it, until we have sorted the whole matrix and this
+becomes a no-op. The fact that the last column corresponds to the
+fixpoint justifies why in \cref{fig:recreate} we can replace every
+column left of the sorted column with the output of the BWT.
+
+\begin{figure}[!hb]
+  \centering
+  \begin{tt}
+  \setlength{\tabcolsep}{0pt}
+  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    \textcolor{lightgray}{aaaaaaaaaaa}&!& \\
+    \textcolor{lightgray}{rrrrrrrrrrr}&a& \\
+    \textcolor{lightgray}{ddddddddddd}&a& \\
+    \textcolor{lightgray}{!!!!!!!!!!!}&a& \\
+    \textcolor{lightgray}{rrrrrrrrrrr}&a& \\
+    \textcolor{lightgray}{ccccccccccc}&a& \\
+    \textcolor{lightgray}{aaaaaaaaaaa}&b& \\
+    \textcolor{lightgray}{aaaaaaaaaaa}&b& \\
+    \textcolor{lightgray}{aaaaaaaaaaa}&c& \\
+    \textcolor{lightgray}{aaaaaaaaaaa}&d& \\
+    \textcolor{lightgray}{bbbbbbbbbbb}&r& \\
+    \textcolor{lightgray}{bbbbbbbbbbb}&r&
+  \end{tabular}
+  $\rightarrow \cdots \rightarrow$
+  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    \textcolor{lightgray}{aaaaaaaaa}&!&ab \\
+    \textcolor{lightgray}{rrrrrrrrr}&a&!a \\
+    \textcolor{lightgray}{ddddddddd}&a&br \\
+    \textcolor{lightgray}{!!!!!!!!!}&a&br \\
+    \textcolor{lightgray}{rrrrrrrrr}&a&ca \\
+    \textcolor{lightgray}{ccccccccc}&a&da \\
+    \textcolor{lightgray}{aaaaaaaaa}&b&ra \\
+    \textcolor{lightgray}{aaaaaaaaa}&b&ra \\
+    \textcolor{lightgray}{aaaaaaaaa}&c&ad \\
+    \textcolor{lightgray}{aaaaaaaaa}&d&ab \\
+    \textcolor{lightgray}{bbbbbbbbb}&r&a! \\
+    \textcolor{lightgray}{bbbbbbbbb}&r&ac
+  \end{tabular}
+  $\rightarrow \cdots \rightarrow$
+  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    \textcolor{lightgray}{a}&!&abracadabr \\
+    \textcolor{lightgray}{r}&a&!abracadab \\
+    \textcolor{lightgray}{d}&a&bra!abraca \\
+    \textcolor{lightgray}{!}&a&bracadabra \\
+    \textcolor{lightgray}{r}&a&cadabra!ab \\
+    \textcolor{lightgray}{c}&a&dabra!abra \\
+    \textcolor{lightgray}{a}&b&ra!abracad \\
+    \textcolor{lightgray}{a}&b&racadabra! \\
+    \textcolor{lightgray}{a}&c&adabra!abr \\
+    \textcolor{lightgray}{a}&d&abra!abrac \\
+    \textcolor{lightgray}{b}&r&a!abracada \\
+    \textcolor{lightgray}{b}&r&acadabra!a
+  \end{tabular}
+  $\rightarrow$
+  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    &!&abracadabra \\
+    &a&!abracadabr \\
+    &a&bra!abracad \\
+    &a&bracadabra! \\
+    &a&cadabra!abr \\
+    &a&dabra!abrac \\
+    &b&ra!abracada \\
+    &b&racadabra!a \\
+    &c&adabra!abra \\
+    &d&abra!abraca \\
+    &r&a!abracadab \\
+    &r&acadabra!ab
+  \end{tabular}
+  \end{tt}
+  \caption{Recreating the rotation matrix from the last column.}
+  \label{fig:recreate}
+\end{figure}
+
+We implement \var{recreate} as follows, introducing a parameter $j$ so
+that we can recursively recreate the first $j$ columns of the matrix,
+one column at a time. Here $l$ is meant to be the output of the BWT,
+the last column of the sorted rotation matrix.
+\lstinputlisting[
+  widthgobble=2, firstline=156, lastline=160,
+]{../theories/BWT/BurrowsWheeler.v}
+This definition relies on \var{prepend\_col} and \var{zipWith}, defined below.
+\lstinputlisting[
+  widthgobble=2, firstline=47, lastline=48,
+]{../theories/BWT/Columns.v}
+\lstinputlisting[
+  widthgobble=2, firstline=16, lastline=20,
+]{../theories/BWT/Lib/ZipWith.v}
+
+Now we can redefine \var{unbwt} as
+\lstinputlisting[
+  widthgobble=2,
+  firstline=196,
+  lastline=197,
+]{../theories/BWT/BurrowsWheeler.v}
+
+Since we recreate the matrix column by column, we can define an
+invariant using $\var{cols}~j$, which takes the first $j$ columns of a matrix.
 \lstinputlisting[
@@
 ]{../theories/BWT/Columns.v}
-Then we introduce a parameter $j$ to \var{recreate}, and our new
-specification is that $\var{recreate}~j$ recreates the first $j$
-columns of the sorted rotation matrix:
-\begin{theorem}[recreate\_correct\_ind]
+
+\begin{theorem}[recreate\_correct\_inv]
   For all lists $l$ and $j \le \var{length}~l$,
-  \begin{equation}
+  \begin{equation*}
     \var{recreate}~j~(\var{bwp}~l) =
     \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
-    \label{eq:recreate_correct_ind}
-  \end{equation}
-\end{theorem}
-
-Now we can redefine \var{unbwt} as
-\lstinputlisting[
-  widthgobble=2,
-  firstline=196,
-  lastline=197,
-]{../theories/BWT/BurrowsWheeler.v}
-
-Assuming that we can prove \var{recreate\_correct\_ind}, the
-correctness of of \text{unbwt} follows:
-
-\begin{theorem}[cols\_id]
-  For all $j$ and matrices $m : \var{list}~(\var{list}~A)$,
-  \begin{gather*}
-    (\forall r \in m,~\var{length}~r \le n) \implies \\
-    \var{cols}~n~m = m
-  \end{gather*}
-\end{theorem}
-\begin{proof}
-  $\var{cols}~j~m = \var{map}~(\var{firstn}~j)~m$, and we
-  know that $m = \var{map}~(\lambda x.x)~m$. So, it suffices to show that
-  for every $r \in m$, the mapping functions agree, that is
-  $\var{firstn}~j~r = r$. This follows from \var{firstn\_all2}
-  in the Coq standard library.
-\end{proof}
-
-\begin{theorem}[iter\_preserves]
-  We are given $f$, $n$, $z$ and a predicate $P : A \to \mathbb{P}$.
-  Suppose $f$ preserves $P$, which is to say that
-  \begin{equation*}
-    \forall x,~P~x \implies P~(f~x)
-  \end{equation*}
-  then,
-  \begin{gather*}
-     P~z \implies \forall x \in \var{iter}~f~n~z,~P~x
-  \end{gather*}
-\end{theorem}
-\begin{proof}
-  This follows from induction on $n$.
-\end{proof}
-
-\begin{lemma}[sort\_rots\_all\_len]
+  \end{equation*}
+\end{theorem}
+
+Assuming we can prove this theorem, we will be able to prove that our
+implementation of \var{unbwt} is correct along the same lines as before.
+
+To prove this invariant, we need to prove a few other theorems that
+have already come up when explaining the algorithm conceptually.
+First, we note that moving the last column of the rotation matrix to the front
+is the same as moving the last row to the top, because the rotation
+matrix is symmetric.
+
+\begin{theorem}[map\_rrot\_rots]
   For all lists $l$,
   \begin{equation*}
-    \forall x \in \var{lexsort}~(\var{rots}~l),~\var{length}~x = \var{length}~l
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  Because \var{lexsort} permutes its argument, it suffices to show
-  that all the rows of $\var{rots}~l$ have the same length as $l$.
-  This follows from \var{iter\_preserves} and the fact that
-  \var{lrot} preserves length.
-\end{proof}
-
-\begin{corollary}[recreate\_correct]
+    \var{map}~\var{rrot}~(\var{rots}~l) = \var{rrot}~(\var{rots}~l)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  Letting $n = \var{length}~l$
+  \begin{align*}
+       &\var{map}~\var{rrot}~(\var{rots}~l) \\
+    ={}&\var{map}~\var{rrot}~[l,~\var{lrot}~l,~\var{lrot}^2~l,~\ldots,~\var{lrot}^{n-1}~l] \\
+    ={}&[\var{rrot}~l,~l,~\var{lrot}~l,~\ldots,~\var{lrot}^{n-2}~l] \\
+    ={}&[\var{lrot}^{n-1}~l,~l,~\var{lrot}~l,~\ldots,~\var{lrot}^{n-2}~l] \\
+    ={}&\var{rrot}~[l,~\var{lrot}~l,~\ldots,~\var{lrot}^{n-1}~l] \\
+    ={}&\var{rrot}~(\var{rots}~l) \qedhere
+  \end{align*}
+\end{proof}
+
+Now we can prove that the sorted rotation matrix is indeed a fixpoint.
+\begin{theorem}[lexsort\_rots\_hdsort]
   For all lists $l$,
   \begin{equation*}
-    \var{recreate}~(\var{length}~l)~(\var{bwp}~l) =
+    (\var{hdsort} \circ
+    \var{map}~\var{rrot})~(\var{lexsort}~(\var{rots}~l)) =
     \var{lexsort}~(\var{rots}~l)
   \end{equation*}
-\end{corollary}
-\begin{proof}
-  This follows from \var{recreate\_correct\_ind},
-  \var{cols\_id}, and \var{sort\_rots\_all\_len}.
-\end{proof}
-
-\begin{theorem}[unbwt\_correct]
-  For all lists $l$,
-  \begin{equation*}
-    \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l) = l
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  The proof is almost identical to that given in
-  \Crefrange{eq:unbwt_correct_start}{eq:unbwt_correct_end}.
-\end{proof}
-
-To implement \var{recreate}, we need expressions for
-$\var{recreate}~0$ and $\var{recreate}~(j+1)$ that satisfy
-\Cref{eq:recreate_correct_ind} in \var{recrate\_correct\_ind}:
-\begin{equation}
+\end{theorem}
+\begin{proof}
+  By \var{StablePerm\_Sorted\_eq}, proved in \cref{sec:unique}, any
+  two sorts are equivalent, and we can replace \var{lexsort} with an
+  LSD radixsort implemented as
+  \begin{equation*}
+    \var{radixsort}~m \coloneqq (\var{hdsort} \circ \var{map}~\var{rrot})^n~m
+  \end{equation*}
+  where $m$ is a matrix with $n$ columns. In this case, we are sorting
+  the rotation matrix and we let $n = \var{length}~l$. Then we have
+  \begin{align*}
+       &(\var{hdsort} \circ \var{map}~\var{rrot})~(\var{lexsort}~(\var{rots}~l)) \\
+    ={}&(\var{hdsort} \circ \var{map}~\var{rrot})~((\var{hdsort} \circ \var{map}~\var{rrot})^n~(\var{rots}~l)) \\
+    ={}&(\var{hdsort} \circ \var{map}~\var{rrot})^{n+1}~(\var{rots}~l) \\
+    ={}&(\var{hdsort} \circ \var{map}~\var{rrot})^n~((\var{hdsort} \circ \var{map}~\var{rrot})~(\var{rots}~l)) \\
+    ={}&\var{lexsort}~((\var{hdsort} \circ \var{map}~\var{rrot})~(\var{rots}~l)) \\
+    ={}&\var{lexsort}~((\var{hdsort} \circ \var{rrot})~(\var{rots}~l)) \\
+    ={}&\var{lexsort}~(\var{rots}~l)
+  \end{align*}
+  The last step follows because the function $\var{hdsort} \circ
+  \var{rrot}$ permutes the rotation matrix
+\end{proof}
+
+We need two more lemmas that describe the interaction between
+\var{cols} and $\var{map}~\var{rrot}$.
+
+\begin{theorem}[cols\_hdsort\_comm]
+  For all matrices $m$ and naturals $j$,
+  \begin{equation*}
+    \var{cols}~(j+1)~(\var{hdsort}~m) = \var{hdsort}~(\var{cols}~(j+1)~m)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  \var{hdsort} only examines the first elements of the rows of $m$,
+  and $\var{cols}~(j+1)$ doesn't alter the first elements of the rows
+  of $m$.\footnote{This argument is made precise by
+    \var{key\_sort\_inv} in \cref{sec:lex_ord}.}
+\end{proof}
+
+\begin{theorem}[cols\_map\_rrot]
+  For all $j$ and matrices $m$ that has rows all of length less than $j$,
+  \begin{equation*}
+    \var{cols}~(j+1)~(\var{map}~\var{rrot}~m) =
+    \var{prepend\_col}~(\var{map}~\var{last}~m)~(\var{cols}~j~m)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We reason by induction on $m$. In the inductive case we need to show that
+  \begin{equation*}
+    \var{cols}~(j+1)~(\var{map}~\var{rrot}~(r::m)) =
+    \var{prepend\_col}~(\var{map}~\var{last}~(r::m))~(\var{cols}~j~(r::m))
+  \end{equation*}
+  Simplifying, we have that the tails are equal by the induction
+  hypothesis, and we must show that the heads are equal.
+  \begin{align*}
+    \var{firstn}~(j+1)~(\var{rrot}~r) &= \var{last}~r :: \var{firstn}~j~r \\
+    \var{firstn}~(j+1)~(\var{last}~r :: \var{init}~r) &= \var{last}~r
+    :: \var{firstn}~j~r \\
+    \var{last}~r :: \var{firstn}~j~(\var{init}~r) &= \var{last}~r :: \var{firstn}~j~r \\
+    \var{firstn}~j~(\var{init}~r) &= \var{firstn}~j~r
+  \end{align*}
+  The bottom equation is true by another theorem we will not prove here.
+\end{proof}
+
+Now we are ready to prove the invariant:
+\begin{theorem}[recreate\_correct\_inv]
+  For all lists $l$ and $j \le \var{length}~l$,
+  \begin{equation*}
     \var{recreate}~j~(\var{bwp}~l) =
     \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
-  \tag{\ref{eq:recreate_correct_ind}}
-\end{equation}
-
-To derive an implementation, we can substitute for $j$ on the right-hand
-side of \Cref{eq:recreate_correct_ind} and manipulate that expression
-algebraically until it is in the form \var{\_ (bwp l)}. Then we can
-define $\var{recreate}~0$ and $\var{recreate}~(j+1)$ to be
-precisely those \var{\_}s, and by running our derivation in
-reverse, we obtain a proof.
-
-We start with the base case, $\var{recreate}~0$:
-\begin{align*}
-     \var{cols}~0~(\var{lexsort}~(\var{rots}~l))
-  &= \var{map}~(\lambda x.[])~(\var{lexsort}~(\var{rots}~l))) \\
-  &= \var{repeat}~[]~(\var{length}~(\var{lexsort}~(\var{rots}~l))) \\
-  &= \var{repeat}~[]~(\var{length}~(\var{bwp}~l)) \\
-     \var{cols}~0~(\var{lexsort}~(\var{rots}~l))
-  &= \var{map}~(\lambda x.[])~(\var{bwp}~l)
-\end{align*}
-
-From this, we can define $\var{recreate}~0~l = \var{map}~(\lambda
-x.[])~l$ and we already have the proof for the base case of the
-correctness specification.
-
-For the inductive case, we have the expression
-\begin{equation*}
-  \var{cols}~(j+1)~(\var{lexsort}~(\var{rots}~l))
-\end{equation*}
-on the right-hand side. Our plan is to express $\var{cols}~(j+1)$
-in terms of $\var{cols}~j$, and then to use the induction
-hypothesis
-\begin{equation*}
-  \var{cols}~j~(\var{lexsort}~(\var{rots}~l)) = \var{recreate}~j
-\end{equation*}
-to produce a well-formed recursive definition.
-
-One way to do this is to note that taking the first $j+1$ columns of a
-matrix is the same as moving the first column to the back, taking the
-first $j$ columns, and then sticking the last column back on the
-front:
-\begin{equation*}
-  \var{cols}~(j+1)~l =
-  \var{prepend\_col}~(\var{map}~\var{hd}~l)~(\var{cols}~j~(\var{map}~\var{lrot}~l))
-\end{equation*}
-If we substitute $\var{map}~\var{rrot}~l$ for $l$ we find a simpler expression
-\begin{equation}
-  \label{eq:cols_rrot}
-  \var{cols}~(j+1)~(\var{map}~\var{rrot}~l) =
-  \var{prepend\_col}~(\var{map}~\var{last}~l)~(\var{cols}~j~l)
-\end{equation}
-
-Now we would like to find a way of introducing a $\var{map}~\var{rrot}$
-before the $\var{lexsort}~(\var{rots}~l)$. The insight here is that we can
-assume that \var{lexsort} is actually a radix-sort. That is, that
-\begin{equation*}
-\var{lexsort}~l = (\var{hdsort} \comp \var{map}~\var{rrot})^{\var{length}~l}~l
-\end{equation*}
-where \var{hdsort} sorts the matrix stably on the the first column.
-
-We can do this because there is only one way to sort the matrix, so
-assuming \var{lexsort} and the radix-sort are correct, they must be
-doing the same thing. We will prove this in
-\cref{sec:unique,sec:sort}.
-
-Using that we can show
-\begin{align*}
-     \var{lexsort}~(\var{rots}~x)
-  &= \var{lexsort}~(\var{rrot}~(\var{rots}~x))
-     && \text{because rrot permutes its elements} \\
-  &= \var{lexsort}~(\var{hdsort}~(\var{rrot}~(\var{rots}~x)))
-     && \text{because hdsort permutes its elements} \\
-  &= (\var{hdsort} \comp \var{rrot})^n~(\var{hdsort}~(\var{rrot}~(\var{rots}~x)))
-     && \text{radix-sort} \\
-  &= (\var{hdsort} \comp \var{rrot})^{n+1}~(\var{rots}~x)
-  && \text{rep\_absorb\_r} \\
-  &= \var{hdsort}~(\var{rrot}~((\var{hdsort} \comp \var{rrot})^n~(\var{rots}~x)))
-     && \text{ rep\_split\_l } \\
-  &= \var{hdsort}~(\var{rrot}~(\var{lexsort}~(\var{rots}~x)))
-     && \text{ radix-sort }
-\end{align*}
-
-In seeking to introduce an extra \var{map rrot} term, we can note that
-rotating and sorting an extra time doesn't change anything when we are
-sorting the rotation matrix. This works because rotating the last
-column to the front produces a matrix that is sorted on its last $n -
-1$ characters, so sorting stably on the first column will re-sort the
-matrix. Crucially though, rotating the last column forwards also
-permutes the entire matrix, so in addition to having a sorted list, we
-have a sorted list of the same elements.
-
-More precisely, we can characterize how $\var{map}~\var{rrot}$
-permutes \var{rots}.
-\begin{equation*}
-  \var{map}~\var{rrot}~(\var{rots}~x) = \var{rrot}~(\var{rots}~x)
-\end{equation*}
-
-From here we can calculate
+  \end{equation*}
+\end{theorem}
+\begin{proof}
 \begin{align*}
   &\var{cols}~(j+1)~(\var{lexsort}~(\var{rots}~l)) \\
-&= \var{cols}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
-&= \var{hdsort}~(\var{cols}~(j+1)~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l))))
-&&\text{ \var{hdsort} commutes with $\var{cols}~(j+1)$ } \\
-&= \var{hdsort}~(\var{prepend\_col}~(\var{map}~\var{last}~(\var{lexsort}~(\var{rots}~l)))~(\var{cols}~j~(\var{lexsort}~(\var{rots}~l)))
-  &&\text{ reduce $\var{cols}~(j+1)$ } \\
-&= \var{hdsort}~(\var{prepend\_col}~(\var{bwp}~l)~(\var{recreate}~j~(\var{bwp}~l)))
-  &&\text{ by IH and def of bwp } \\
-&= (\lambda x.~\var{hdsort}~(\var{prepend\_col}~x~(\var{recreate}~j~x)))~\var{bwp}
-\end{align*}
-
-We can use this to define \var{recreate}:
-\lstinputlisting[
-  widthgobble=2, firstline=156, lastline=160,
-]{../theories/BWT/BurrowsWheeler.v}
+={}& \var{cols}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
+={}&
+\var{hdsort}~(\var{cols}~(j+1)~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
+={}&
+\var{hdsort}~(\var{prepend\_col}~(\var{map}~\var{last}~(\var{lexsort}~(\var{rots}~l)))~(\var{cols}~j~(\var{lexsort}~(\var{rots}~l))) \\
+={}&
+\var{hdsort}~(\var{prepend\_col}~(\var{bwp}~l)~(\var{recreate}~j~(\var{bwp}~l))) \\
+={}& \var{recreate}~j~(\var{bwp}~l) \qedhere
+\end{align*}
+\end{proof}
 
@@
   \begin{align*}
-    h::t &= h :: l_1 \dplus h' :: l_2 \\
-    &\eqperm h :: h' :: l_1 \dplus l_2 \\
-    &\eqperm h' :: h :: l_1 \dplus l_2 \\
-    &\eqperm h' :: h :: l_1' \dplus l_2' \\
-    &\eqperm h' :: l_1' \dplus h :: l_2' = h'::t
+    h::t ={}& h :: l_1 \dplus h' :: l_2 \\
+    \eqperm{}& h :: h' :: l_1 \dplus l_2 \\
+    \eqperm{}& h' :: h :: l_1 \dplus l_2 \\
+    \eqperm{}& h' :: h :: l_1' \dplus l_2' \\
+    \eqperm{}& h' :: l_1' \dplus h :: l_2' \\
+    ={}& h'::t
   \end{align*}
@@
     \var{apply}~(i :: p)~(a :: l)
-    &= (a :: l)_i  ::
+    ={}& (a :: l)_i  ::
     \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a :: l)) \\
-    &\eqstable (a :: l)_i :: \var{rem\_nth}~i~(a :: l) \\
-    &\eqstable (a :: l)_i
+    \eqstable{}& (a :: l)_i :: \var{rem\_nth}~i~(a :: l) \\
+    \eqstable{}& (a :: l)_i
   \end{align*}
@@
 
-\begin{theorem}[stable\_sort\_unique]
+\begin{theorem}[StablePerm\_Sorted\_eq]
+  For all lists $l$,
   \begin{equation*}
@@
 \end{theorem}
-
-With \var{lexsort}, we are guaranteed stability because we are using
-an order, not just a preorder, and for orders all sorts are stable.
+\begin{proof}
+  We reason by induction on $l$. In the inductive case, we have to
+  show that
+  \begin{equation*}
+    h :: t = h' :: t'
+  \end{equation*}
+  given $\var{Sorted}~(h :: t)$, $\var{Sorted}~(h' :: t')$, and
+  $\var{StablePerm}~(h :: t)~(h' :: t')$.
+
+  First, we can say that $h \equiv h'$: we know that $h \in h' :: t'$ and $h'
+  \in h :: t$. If $h = h'$, then we are done, so assuming $h \neq h'$ we
+  know $h \in t'$ and $h' \in t$. Because both $h::t$ and $h'::t'$ are
+  sorted, we knnow that $h \le h'$ and $h' \le h$, implying $h \equiv h'$.
+
+  This means that $h = h'$: from $\var{StablePerm}~(h :: t)~(h' ::
+  t')$ we have
+  \begin{align*}
+    \var{filter}~(\equiv h)~(h :: t) &= \var{filter}~(\equiv h)~(h' :: t') \\
+    h :: \var{filter}~(\equiv h)~t &= h' :: \var{filter}~(\equiv h)~t'
+  \end{align*}
+  We also have $t = t'$ by the induction hypothesis, so we are done.
+\end{proof}
+
+If we define \var{Sort} and \var{StableSort} as we did
+\cref{sec:sorted_perm_stable}, we can restate the theorem slightly
+more elegantly.
+\lstinputlisting[
+  widthgobble=2, firstline=15, lastline=19,
+]{../theories/BWT/Sorting/Sort.v}
+
+\begin{corollary}[StableSort\_unique]
+  For all functions $f, g: \var{list}~A \to \var{list}~A$,
+  \begin{equation*}
+    \var{StableSort}~f \implies \var{StableSort}~g \implies
+    (\forall l,~f~l = g~l)
+  \end{equation*}
+\end{corollary}
+\begin{proof}
+  This follows from \var{StablePerm\_Sorted\_eq}.
+\end{proof}
+
+However, we do not want to require that \var{lexsort} be implemented
+using a stable sorting algorithm. Luckily, because we require that $A$
+be equipped with an order and not just a preorder, we know that
+equivalent elements are actually indistinguishable. This means that
+all sorting algorithms can be considered to be stable: it is
+impossible to tell if any two equivalent elements were swapped because
+you cannot distinguish between equivalent elements.
 
@@
 \end{theorem}
+\begin{proof}
+  We reason by induction on the evidence of $\var{Permutation}~l~l'$.
+  Every case other than the \var{perm\_swap} case follows easily.
+  In the \var{perm\_swap} case, we must show that
+  \begin{equation*}
+    \var{StablePerm}~(y :: x :: l)~(x :: y :: l)
+  \end{equation*}
+  for all $x$ and $y$. If $x \not\equiv y$, we can just apply
+  \var{StablePerm\_swap}. If $x \equiv y$, then we know that $x = y$, and
+  we can show
+  \begin{equation*}
+    \var{StablePerm}~(x :: x :: l)~(x :: x :: l)
+  \end{equation*}
+  by applying \var{perm\_skip} twice.
+\end{proof}
+
+\begin{corollary}[Sort\_StableSort\_Ord]
+  When sorting using an order,
+  \begin{equation*}
+    \var{Sort}~f \iff \var{StableSort}~f
+  \end{equation*}
+  for any $f$.
+\end{corollary}
+\begin{proof}
+  This follows from \var{all\_perm\_stable}.
+\end{proof}
+
+\begin{corollary}[Sort\_Ord\_unique]
+  When sorting using an order,
+  \begin{equation*}
+    \var{Sort}~f \implies \var{Sort}~g \implies
+    (\forall l,~f~l = g~l)
+  \end{equation*}
+  for any $f$ and $g$.
+\end{corollary}
+\begin{proof}
+  This follows from \var{Sort\_StableSort\_Ord} and \var{StableSort\_unique}.
+\end{proof}
+
+\section{Lexicographic Orders and Keys}
+\label{sec:lex_ord}
+
+Given a preorder on a type $A$, we can define a lexicographic preorder
+on $\var{list}~A$. Intuitively, we want to say that to compare two
+lists, we compare their elements from left to right. As soon as we
+encounter two elements that are not equivalent, their relationship
+defines the relationship between the two lists. If we make it all the
+way through one of the lists and all the elements are equivalent, we
+say that the shorter list is smaller. If they are the same length and
+all the elements are equivalent, then the lists are equivalent. To
+capture this, we define the inductive relation \var{lex\_le}, which we
+will denote by $\lexle$ in proofs.
+\begin{gather*}
+  \infer[\var{lex\_le\_nil}]{[] \lexle ys}{}
+  \\
+  \\
+  \infer[\var{lex\_le\_cons\_lt}]{(x :: xs) \lexle (y :: ys)}{%
+    x < y
+  }
+  \\
+  \\
+  \infer[\var{lex\_le\_cons\_eq}]{(x :: xs) \lexle (y :: ys)}{%
+    x \equiv y & xs \lexle ys
+  }
+\end{gather*}
+
+Given a preorder on a type $K$ and a key function $\var{key} : A \to K$,
+we can define a preorder on $A$ that compares all elements using the
+key function. That is, we can define the relation $\le_{\var{key}}$ on
+$A$ such that for all $x, y : A$
+\begin{equation*}
+  x \le_{\var{key}} y \coloneqq (\var{key}~x) \le (\var{key}~y)
+\end{equation*}
+
+\begin{theorem}[key\_sort\_inv]
+  Given a key function $\var{key} : A \to K$ and a function $f : A \to A$
+  such that for all $x : A$,
+  \begin{equation*}
+    \var{key}~(f~x) = \var{key}~x
+  \end{equation*}
+  we have that for all lists $l$
+  \begin{equation*}
+    \var{sort}~(\var{map}~\var{f}~l) = \var{map}~\var{f}~(\var{sort}~l).
+  \end{equation*}
+  where \var{sort} is insertion sort.\footnote{This theorem actually
+    holds for any function with type $\forall X, (X \to X \to \mathbb{B}) \to
+    \var{list}~X \to \var{list}~X$ due to parametricity \cite{free}.}
+\end{theorem}
+\begin{proof}
+  The proof follows by induction, unfolding the definition of
+  \var{sort}. All decisions are made by calling \var{key}, so both
+  sides are equal.
+\end{proof}
+
+Taken together, we can define prefix lexicographic orderings on lists,
+where the lexicographic ordering is only allowed to compare up to the
+first $n$ elements of two lists. We define the relation $\le_n$ on
+$\var{list}~A$ such that for all $xs, ys : \var{list}~A$,
+\begin{equation*}
+  xs \le_n ys \coloneqq (\var{firstn}~n~xs) \lexle (\var{firstn}~n~ys)
+\end{equation*}
+For convenience, we define $\var{PrefixSorted}~n$ to hold when a list
+of lists is sorted if you only look at the first $n$ elements of each:
+\lstinputlisting[
+  widthgobble=2, firstline=323, lastline=324,
+]{../theories/BWT/Sorting/Key.v}
+
+Although it is difficult to prove, it is intuitive that prefix orders
+that can look at more elements of the lists are ``stronger'' than
+those that can look at fewer.
+
+\begin{theorem}[key\_le\_firstn\_ge]
+  For all $j$ and $k$ such that $k \ge j$,
+  \begin{equation*}
+    x \le_k y \implies x \le_j y
+  \end{equation*}
+\end{theorem}
+
+If we consider the implications this has for stability, a weaker order
+can distinguish between fewer elements, which means that it considers
+more elements equivalent. This means that stability with respect to a
+weaker order is actually a stronger property, because the weaker
+order, considering more elements equivalent, is more restrictive about
+which swaps it allows.
+
+\begin{theorem}[StablePerm\_weaken]
+  Suppose we have a two preorders $O$ and $O'$ on $A$, and two lists
+  $l_1$ and $l_2$, with elements of type $A$.
+
+  We also know that on the elements of $l_1$, $O'$ is stronger than
+  $O$. If we denote by $\le$ and $\le'$ the relations for $O$ and $O'$
+  respectively, this means that for all $x$ and $y$ in $l_1$,
+  \begin{equation*}
+    x \le' y \implies x \le y.
+  \end{equation*}
+  Then, denoting by \var{StablePerm} stability relative to the
+  equivalence relation of $O$, and by $\var{StablePerm}'$ stability
+  relative to the equivalence relation of $O'$, we have
+  \begin{equation*}
+    \var{StablePerm}~l_1~l_2 \implies \var{StablePerm}'~l_1~l_2
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We reason inductively on the proof of $\var{StablePerm}~l_1~l_2$
+  using \var{StablePermInd\_iff}. The important case is the
+  \var{StablePermInd\_swap} case, where we have to show
+  \begin{equation*}
+    \var{StablePerm}'~(y :: x :: l)~(x :: y :: l)
+  \end{equation*}
+  given that $x \not\equiv y$. To show this, we need to know that $x \not\equiv'
+  y$, which follows from the fact that $O'$ is stronger than $O$.
+\end{proof}
 
@@
 
-As
-
-\lstinputlisting[
-  widthgobble=2, firstline=11, lastline=21,
+In a least-significant-digit radix sort, we lexicographically sort a
+list of strings of equal length by stably sorting on each column, from
+the end of the string to the front (\Cref{fig:lsd_radixsort}). This
+works because the lexicographic order says that if two strings are
+equal up to a column then the tie is broken in according to the rest
+of the string. This is reflected in the fact that when we stably sort
+on a column, ties are broken in favor or the existing order, which was
+established by sorted on the following columns.
+
+\begin{figure}[!hb]
+  \centering
+  \begin{tt}
+    \begin{tabular}{c}
+      d a \textbf{b} \\
+      a d \textbf{d} \\
+      c a \textbf{b} \\
+      f a \textbf{d} \\
+      f e \textbf{e} \\
+      b a \textbf{d} \\
+      d a \textbf{d} \\
+      b e \textbf{e} \\
+      f e \textbf{d} \\
+      b e \textbf{d} \\
+      e b \textbf{b} \\
+      a c \textbf{e}
+    \end{tabular}
+    $\rightarrow$
+    \begin{tabular}{c}
+      d \textbf{a} b \\
+      c \textbf{a} b \\
+      e \textbf{b} b \\
+      a \textbf{d} d \\
+      f \textbf{a} d \\
+      b \textbf{a} d \\
+      d \textbf{a} d \\
+      f \textbf{e} d \\
+      b \textbf{e} d \\
+      f \textbf{e} e \\
+      b \textbf{e} e \\
+      a \textbf{c} e
+    \end{tabular}
+    $\rightarrow$
+    \begin{tabular}{c}
+      \textbf{d} a b \\
+      \textbf{c} a b \\
+      \textbf{f} a d \\
+      \textbf{b} a d \\
+      \textbf{d} a d \\
+      \textbf{e} b b \\
+      \textbf{a} c e \\
+      \textbf{a} d d \\
+      \textbf{f} e d \\
+      \textbf{b} e d \\
+      \textbf{f} e e \\
+      \textbf{b} e e
+    \end{tabular}
+    $\rightarrow$
+    \begin{tabular}{c}
+      a c e \\
+      a d d \\
+      b a d \\
+      b e d \\
+      b e e \\
+      c a b \\
+      d a b \\
+      d a d \\
+      e b b \\
+      f a d \\
+      f e d \\
+      f e e
+    \end{tabular}
+  \end{tt}
+  \caption{LSD Radixsort}
+  \label{fig:lsd_radixsort}
+\end{figure}
+
+In a functional setting, we can define an LSD radixsort elegantly by
+repeatedly moving the last column to the front of the matrix and
+sorting on the first column. For a matrix with $n$ columns,
+\begin{equation*}
+  \var{radixsort} = (\var{hdsort} \circ \var{map}~\var{rrot})^n
+\end{equation*}
+Or, in Coq:
+\lstinputlisting[
+  widthgobble=2, firstline=26, lastline=27,
+]{../theories/BWT/Sorting/RadixSort.v}
+Here \var{hdsort} is just an insertion sort that uses a lexicographic
+ordering restricted to the first element of a list.
+
+First we want to prove that \var{radixsort} permutes the rows of the
+matrix. We will start by proving the invariant that if we rotate the
+last column to the front and sort $j$ times, the resulting matrix is a
+permtuation of simply rotating the last column to the front $j$
+times.
+\begin{lemma}[radixsort\_perm\_inv]
+  For all matrices $m$ and $j : \mathbb{N}$,
+  \begin{equation*}
+    \var{Permutation}~((\var{map}~\var{rrot})^j~m)~
+                      ((\var{hdsort} \circ \var{map}~\var{rrot})^j~m)
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  We reason by induction on $j$. In the inductive case,
+  \begin{align*}
+    (\var{hdsort} \circ \var{map}~\var{rrot})^{j+1}~m ={}&
+    \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \circ
+    \var{map}~\var{rrot})^j~m) \\
+    \eqperm{}& \var{map}~\var{rrot}~((\var{hdsort} \circ
+    \var{map}~\var{rrot})^j~m) \\
+    \eqperm{}& \var{map}~\var{rrot}~((\var{map}~\var{rrot})^j~m) \\
+    ={}& (\var{map}~\var{rrot})^{j+1}~m \qedhere
+  \end{align*}
+\end{proof}
+
+\begin{theorem}[radixsort\_perm]
+  For all matrices $m$ with rows of length $n$,
+  \begin{equation*}
+    \var{Permutation}~m~(\var{radixsort}~m~n)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  \begin{align*}
+    m ={}&
+    \var{map}~\var{rrot}^n~m \\
+    ={}& (\var{map}~\var{rrot})^n~m \\
+    \eqperm{}& (\var{hdsort} \circ \var{map}~\var{rrot})^n~m \\
+    ={}& \var{radixsort}~m~n \qedhere
+  \end{align*}
+\end{proof}
+
+We will prove the stability of \var{radixsort} in a similar way, but
+first we need a lemma.
+\begin{lemma}[StablePerm\_map\_rrot]
+  For all matrices $m$ and $m'$,
+  \begin{equation*}
+    \var{StablePerm}~m~m' \implies
+    \var{StablePerm}~(\var{map}~\var{rrot}~m)~(\var{map}~\var{rrot}~m').
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  We reason by induction on the evidence that $\var{StablePerm}~m~m'$
+  using \var{StablePermInd\_iff}. The important case is the swap case,
+  where we have to show
+  \begin{equation*}
+    \var{StablePerm}~(\var{map}~(y :: x :: l))~(\var{map}~(x :: y :: l))
+  \end{equation*}
+  given that $x \not\equivle y$. For this it suffices to show that
+  $\var{rrot}~x \not\equivle \var{rrot}~y$. Since lexicographic
+  equivalence means that all the pairs of elements at corresponding
+  indices are equivalent, when \var{rrot} moves the first elements of
+  both lists to the end of both lists, it preserves lexicographic
+  equivalence. Therefore
+  \begin{equation*}
+    x \equivle y \iff \var{rrot}~x \equivle \var{rrot}~y \qedhere
+  \end{equation*}
+\end{proof}
+
+\begin{lemma}[radixsort\_stable\_inv]
+  For all matrices $m$ and $j : \mathbb{N}$,
+  \begin{equation*}
+    \var{StablePerm}~((\var{map}~\var{rrot})^j~m)~
+                     ((\var{hdsort} \circ \var{map}~\var{rrot})^j~m)
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  We reason by induction on $j$. In the inductive case,
+  \begin{align*}
+    (\var{hdsort} \circ \var{map}~\var{rrot})^{j+1}~m ={}&
+    \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \circ
+    \var{map}~\var{rrot})^j~m) \\
+    \eqstable{}& \var{map}~\var{rrot}~((\var{hdsort} \circ
+    \var{map}~\var{rrot})^j~m) \\
+    \eqstable{}& \var{map}~\var{rrot}~((\var{map}~\var{rrot})^j~m) \\
+    ={}& (\var{map}~\var{rrot})^{j+1}~m
+  \end{align*}
+  Going from the first line to the second line requires using
+  \var{StablePerm\_weaken} to show that the stability we are
+  guaranteed by \var{hdsort}, namely stability with respect to an
+  ordering that examines at most one element, implies stability with
+  respect to a regular lexicographic ordering.
+\end{proof}
+
+\begin{theorem}[radixsort\_stable]
+  For all matrices $m$ with rows of length $n$,
+  \begin{equation*}
+    \var{StablePerm}~m~(\var{radixsort}~m~n)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  The proof is almost identical to the proof of \var{radixsort\_perm}.
+\end{proof}
+
+We would also like to establish an invariant to prove that
+\var{radixsort} produces a sorted list. To do so, we need to be able
+to reason about the process of adding a column to the front of a
+sorted matrix and sorting on that column. Intuitively, we need to know
+that if we have a matrix that is sorted on the first $j$ columns, and
+we add a new column to the front and sort on it, the result list will
+be sorted on the first $j+1$ columns. For a proof, see
+\cref{appendix:hdsort_sorted_S}.
+\begin{theorem}[hdsort\_sorted\_S]
+  For all matrices $m$ and naturals $j$,
+  \begin{equation*}
+    \var{PrefixSorted}~j~(\var{map}~\var{tl}~m) \implies
+    \var{PrefixSorted}~(j+1)~(\var{hdsort}~m)
+  \end{equation*}
+\end{theorem}
+
+We also need the following property about \var{prepend\_col}, that
+states that prepending a column and then removing the first column is
+a no-op.
+\begin{theorem}[map\_tl\_prepend]
+  For all matrices $m : \var{list}~(\var{list}~A)$ and columns $c :
+  \var{list}~A$, where $c$ has at least as many elements as $m$ has columns,
+  \begin{equation*}
+      \var{map}~\var{tl}~(\var{prepend\_col}~c~m) = m
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We reason by induction on $m$. In the inductive case, we have to
+  prove that
+  \begin{align*}
+    &\var{map}~\var{tl}~(\var{prepend\_col}~(h_c :: t_c)~(r :: m)) \\
+    ={}&\var{map}~\var{tl}~((h_c :: r) :: \var{prepend\_col}~t_c~m) \\
+    ={}&r :: \var{map}~\var{tl}~(\var{prepend\_col}~t_c~m) \\
+    ={}&r :: m
+  \end{align*}
+  where the last line follows from the induction hypothesis.
+\end{proof}
+
+\begin{corollary}
+  For all matrices $m$ and rows $r$,
+  \begin{equation*}
+    \var{map}~\var{tl}~(\var{map}~\var{rrot}~m) = \var{map}~\var{init}~m
+  \end{equation*}
+\end{corollary}
+\begin{proof}
+  \begin{align*}
+    \var{map}~\var{tl}~(\var{map}~\var{rrot}~m) &= \var{map}~\var{tl}~(\var{prepend\_col}~(\var{map}~\var{last}~m)~(\var{map}~\var{init}~m) \\
+    &= \var{map}~\var{init}~m \qedhere
+  \end{align*}
+\end{proof}
+
+Now we can establish our invariant:
+
+\begin{lemma}[radixsort\_sorted\_inv]
+  For all matrices $m$ with rows of length $n$, and $j \le n$,
+  \begin{equation*}
+    \var{PrefixSorted}~j~(\var{radixsort}~m~j)
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  We reason by induction on $j$. The base case is true because every
+  list is sorted under ordering where you examine no elements of the
+  list. In the inductive case, we reason backwards from our goal:
+  \begin{align*}
+       &\var{PrefixSorted}~(j+1)~(\var{radixsort}~(r :: m)~(j+1)) \\
+    \iff{}&\var{PrefixSorted}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{radixsort}~(r :: m)~j))) \\
+    \iff{}&\var{PrefixSorted}~j~(\var{map}~\var{tl}~(\var{map}~\var{rrot}~(\var{radixsort}~(r :: m)~j))) \\
+    \iff{}&\var{PrefixSorted}~j~(\var{map}~\var{init}~(\var{radixsort}~(r :: m)~j)) \\
+    \iff{}&\var{Sorted}~(\var{map}~(\var{firstn}~j)~(\var{map}~\var{init}~(\var{radixsort}~(r :: m)~j))) \\
+    \iff{}&\var{Sorted}~(\var{map}~(\var{firstn}~j)~(\var{radixsort}~(r :: m)~j)) \\
+    \iff{}&\var{PrefixSorted}~j~(\var{radixsort}~(r :: m)~j)
+  \end{align*}
+  The last line follows from the induction hypothesis.
+\end{proof}
+
+\begin{theorem}[radixsort\_sorted]
+  For all matrices $m$ with rows of length $n$,
+  \begin{equation*}
+      \var{Sorted}~(\var{radixsort}~m~n)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  \begin{align*}
+      &\var{Sorted}~(\var{radixsort}~m~n) \\
+    \iff{}&\var{Sorted}~(\var{map}~(\var{firstn}~n)~(\var{radixsort}~m~n)) \\
+    \iff{}&\var{PrefixSorted}~n~(\var{radixsort}~m~n))
+  \end{align*}
+  The last line follows from \var{radixsort\_sorted\_inv}.
+\end{proof}
+
+\appendix
+
+\section{Insertion Sort}
+\label{appendix:insertion_sort}
+
+Insertion sort is a simple and elegant stable sort that is easy to
+reason about. The theorems proved in \cref{sec:unique} allow us to
+prove things using insertion sort that apply to any stable sort. For
+example, our proof of correctness for radixsort relies on the fact
+that it uses insertion sort as the innner sort, but
+\var{StableSort\_unique} allows us to extend that proof to a radixsort
+that uses any stable sort. We also use it as the implementation of
+\var{lexsort} because it is convenient, but similarly
+\var{Sort\_Ord\_unique} ensures that it can be replaced with any other
+sort.
+
+\lstinputlisting[
+  widthgobble=2, firstline=15, lastline=26,
 ]{../theories/BWT/Sorting/InsertionSort.v}
 
+\begin{lemma}[insert\_perm]
+  For all lists $l$ and elements $x$,
+  \begin{equation*}
+    \var{Permutation}~(x::l)~(\var{insert}~x~l)
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  We reason by induction on $l$. In the inductive case, we need to show
+  \begin{equation*}
+    \var{Permutation}~(x::h::t)~(\var{insert}~x~(h::t))
+  \end{equation*}
+  given the induction hypothesis that
+  \begin{equation*}
+    \var{Permutation}~(x::t)~(\var{insert}~x~t)
+  \end{equation*}
+  If $x \le h$, then $\var{insert}~x~(h::t) = x :: h :: t$ and we are
+  done. Otherwise, $\var{insert}~x~(h::t) = h :: \var{insert}~x~t$ and
+  we can apply the induction hypothesis.
+\end{proof}
+
+\begin{theorem}[sort\_perm]
+  For all lists $l$,
+  \begin{equation*}
+    \var{Permutation~l}~(\var{sort}~l)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  The proof follows by induction on $l$. The inductive case is
+  \begin{equation*}
+    h :: t \eqperm h :: \var{sort}~t \eqperm
+    \var{insert}~h~(\var{sort}~t) = \var{sort}~(h :: t)
+  \end{equation*}
+  where the first step follows from the induction hypothesis, and the
+  second from \var{insert\_perm}.
+\end{proof}
+
+\begin{lemma}[insert\_sorted]
+  For all lists $l$ and elements $x$,
+  \begin{equation*}
+    \var{Sorted}~l \implies \var{Sorted}~(\var{insert}~x~l)
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  We reason by induction on the proof that $\var{Sorted}~l$. In the
+  inductive case we need prove that
+  \begin{equation*}
+    \var{Sorted}~(\var{insert}~x~(h::t))
+  \end{equation*}
+  If $x \le h$, then $\var{insert}~x~(h::t) = x :: h :: t$, and by
+  \var{SortedLocal\_cons} we are done.
+
+  If $x > h$, $\var{insert}~x~(h::t) = h :: \var{insert}~x~t$, and it
+  suffices to show that $\forall y \in \var{insert}~x~t,~h \le y$. By
+  \var{insert\_perm}, this is equivalent to $\forall y \in x::t,~h \le y$ From
+  the the fact that $\var{Sorted}~(h::t)$, we know that $\forall y \in t,~h \le
+  y$, and we have that $h < x$, so we are done.
+\end{proof}
+
 \begin{theorem}[sort\_sorted]
+  For all lists $l$,
   \begin{equation*}
@@
 \end{theorem}
+\begin{proof}
+  The proof follows by induction on $l$ and \var{insert\_sorted}.
+\end{proof}
+
+\begin{lemma}[insert\_stable]
+  For all lists $l$ and elements $x$,
+  \begin{equation*}
+    \var{StablePerm}~(x::l)~(\var{insert}~x~l)
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  We reason by induction on $l$. In the inductive case, we need to show
+  \begin{equation*}
+    \var{StablePerm}~(x::h::t)~(\var{insert}~x~(h::t))
+  \end{equation*}
+  given the induction hypothesis that
+  \begin{equation*}
+    \var{StablePerm}~(x::t)~(\var{insert}~x~t)
+  \end{equation*}
+  If $x \le h$, then $\var{insert}~x~(h::t) = x :: h :: t$ and we are
+  done. Otherwise, $x > h$ and $\var{insert}~x~(h::t) = h ::
+  \var{insert}~x~t$.
+  \begin{equation*}
+    x :: h :: t \eqstable h :: x :: t \eqstable h :: \var{insert}~x~t
+  \end{equation*}
+  where the first step follows from the fact that $x \not\equiv h$ and the
+  second from the induction hypothesis.
+\end{proof}
 
 \begin{theorem}[sort\_stable]
-  \begin{equation*}
-    \var{Stable}~(\var{sort}~l)
-  \end{equation*}
-\end{theorem}
-
-\lstinputlisting[
-  widthgobble=2, firstline=22, lastline=23,
-]{../theories/BWT/Sorting/RadixSort.v}
-
-\begin{theorem}[radixsort\_sorted]
-  \begin{equation*}
-    (\forall x \in l,~\var{length}~x = n) \implies \var{Sorted}~(\var{radixsort}~l~n)
-  \end{equation*}
-\end{theorem}
-
-\appendix
-
-\section{Proving BWT Correct}
-\label{appendix:BWT_proof}
-
-In this section we prove the lemmas from the derivation that we left out:
-\begin{enumerate}
-\item
-\begin{equation*}
-  \var{map}~\var{rrot}~(\var{rots}~x) = \var{rrot}~(\var{rots}~x)
-\end{equation*}
-
-\item \var{hdsort} commutes with $\var{cols} (j+1)$.
-  This proof is cool because in theory it can be proven via
-  parametricity, but that's a metatheorem in Coq.
-
-\item
-  \begin{equation*}
-    \var{cols}~(j+1)~(\var{map}~\var{rrot}~l) = \var{prepend\_col}~(\var{map}~\var{last}~l)~(\var{cols}~j~l)
-  \end{equation*}
-\end{enumerate}
+  For all lists $l$,
+  \begin{equation*}
+    \var{StablePerm}~l~(\var{sort}~l)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  The proof is almost identical to the proof for \var{sort\_perm}.
+\end{proof}
+
+\begin{theorem}[sort\_StableSort]
+  \begin{equation*}
+    \var{StableSort}~\var{sort}
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  This follows from \var{sort\_sorted} and \var{sort\_stable}.
+\end{proof}
+
+\section{\var{hdsorted\_sorted\_S}}
+\label{appendix:hdsort_sorted_S}
 
@@
 
+\nocite{*}
 \printbibliography{}

### 2019-05-05 BW.hs

@@ -1,0 +1,29 @@
+import Data.List (sortOn, sort)
+
+lrot :: [a] -> [a]
+lrot [] = []
+lrot xs = tail xs ++ [head xs]
+
+rrot :: [a] -> [a]
+rrot [] = []
+rrot xs = last xs : init xs
+
+rots :: [a] -> [[a]]
+rots xs = take (length xs) $ iterate lrot xs
+
+printMat :: [String] -> IO ()
+printMat = mapM_ putStrLn
+
+sortFrom :: Int -> [String] -> [String]
+sortFrom 0 m = sortOn (!! i') $ m
+    where i' = length m - 1
+sortFrom i m = sortOn (!! i') $ sortFrom (i-1) m
+    where i' = length m - 1 - i
+
+bwp :: Ord a => [a] -> [a]
+bwp = map last . sort . rots
+
+recreate :: Ord a => Int -> [a] -> [[a]]
+recreate 0 l = map (const []) l
+recreate j l = sortOn head . zipWith (:) l $ recreate (j-1) l
+

### 2019-05-05 draft_paper/draft_paper.tex
Finish permutation comp appendix

@@
 
-We can also think of this as computing the fixpoint of $\var{hdsort} \circ
+We can also think of this as computing the fixpoint of $\var{hdsort} \comp
 \var{map}~\var{rrot}$: we continually move the rightmost column to the
@@
   \begin{equation*}
-    (\var{hdsort} \circ
+    (\var{hdsort} \comp
     \var{map}~\var{rrot})~(\var{lexsort}~(\var{rots}~l)) =
@@
   \begin{equation*}
-    \var{radixsort}~m \coloneqq (\var{hdsort} \circ \var{map}~\var{rrot})^n~m
+    \var{radixsort}~m \coloneqq (\var{hdsort} \comp \var{map}~\var{rrot})^n~m
   \end{equation*}
@@
   \begin{align*}
-       &(\var{hdsort} \circ \var{map}~\var{rrot})~(\var{lexsort}~(\var{rots}~l)) \\
-    ={}&(\var{hdsort} \circ \var{map}~\var{rrot})~((\var{hdsort} \circ \var{map}~\var{rrot})^n~(\var{rots}~l)) \\
-    ={}&(\var{hdsort} \circ \var{map}~\var{rrot})^{n+1}~(\var{rots}~l) \\
-    ={}&(\var{hdsort} \circ \var{map}~\var{rrot})^n~((\var{hdsort} \circ \var{map}~\var{rrot})~(\var{rots}~l)) \\
-    ={}&\var{lexsort}~((\var{hdsort} \circ \var{map}~\var{rrot})~(\var{rots}~l)) \\
-    ={}&\var{lexsort}~((\var{hdsort} \circ \var{rrot})~(\var{rots}~l)) \\
+       &(\var{hdsort} \comp \var{map}~\var{rrot})~(\var{lexsort}~(\var{rots}~l)) \\
+    ={}&(\var{hdsort} \comp \var{map}~\var{rrot})~((\var{hdsort} \comp \var{map}~\var{rrot})^n~(\var{rots}~l)) \\
+    ={}&(\var{hdsort} \comp \var{map}~\var{rrot})^{n+1}~(\var{rots}~l) \\
+    ={}&(\var{hdsort} \comp \var{map}~\var{rrot})^n~((\var{hdsort} \comp \var{map}~\var{rrot})~(\var{rots}~l)) \\
+    ={}&\var{lexsort}~((\var{hdsort} \comp \var{map}~\var{rrot})~(\var{rots}~l)) \\
+    ={}&\var{lexsort}~((\var{hdsort} \comp \var{rrot})~(\var{rots}~l)) \\
     ={}&\var{lexsort}~(\var{rots}~l)
   \end{align*}
-  The last step follows because the function $\var{hdsort} \circ
+  The last step follows because the function $\var{hdsort} \comp
   \var{rrot}$ permutes the rotation matrix
@@
 \begin{equation*}
-  \var{radixsort} = (\var{hdsort} \circ \var{map}~\var{rrot})^n
+  \var{radixsort} = (\var{hdsort} \comp \var{map}~\var{rrot})^n
 \end{equation*}
@@
     \var{Permutation}~((\var{map}~\var{rrot})^j~m)~
-                      ((\var{hdsort} \circ \var{map}~\var{rrot})^j~m)
+                      ((\var{hdsort} \comp \var{map}~\var{rrot})^j~m)
   \end{equation*}
@@
   \begin{align*}
-    (\var{hdsort} \circ \var{map}~\var{rrot})^{j+1}~m ={}&
-    \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \circ
+    (\var{hdsort} \comp \var{map}~\var{rrot})^{j+1}~m ={}&
+    \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \comp
     \var{map}~\var{rrot})^j~m) \\
-    \eqperm{}& \var{map}~\var{rrot}~((\var{hdsort} \circ
+    \eqperm{}& \var{map}~\var{rrot}~((\var{hdsort} \comp
     \var{map}~\var{rrot})^j~m) \\
@@
     ={}& (\var{map}~\var{rrot})^n~m \\
-    \eqperm{}& (\var{hdsort} \circ \var{map}~\var{rrot})^n~m \\
+    \eqperm{}& (\var{hdsort} \comp \var{map}~\var{rrot})^n~m \\
     ={}& \var{radixsort}~m~n \qedhere
@@
     \var{StablePerm}~((\var{map}~\var{rrot})^j~m)~
-                     ((\var{hdsort} \circ \var{map}~\var{rrot})^j~m)
+                     ((\var{hdsort} \comp \var{map}~\var{rrot})^j~m)
   \end{equation*}
@@
   \begin{align*}
-    (\var{hdsort} \circ \var{map}~\var{rrot})^{j+1}~m ={}&
-    \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \circ
+    (\var{hdsort} \comp \var{map}~\var{rrot})^{j+1}~m ={}&
+    \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \comp
     \var{map}~\var{rrot})^j~m) \\
-    \eqstable{}& \var{map}~\var{rrot}~((\var{hdsort} \circ
+    \eqstable{}& \var{map}~\var{rrot}~((\var{hdsort} \comp
     \var{map}~\var{rrot})^j~m) \\
@@
 
+In this section, we define \var{compose} to compose permutations,
+which we will denote using $\comp$ in proofs. We prove the following
+property:
+
 \begin{theorem}[compose\_apply]
@@
   \begin{equation*}
-      \var{apply}~(p_2 \comp p_1)~l = \var{apply}~p2~(\var{apply}~p1~l)
-  \end{equation*}
-\end{theorem}
+      \var{apply}~(p_2 \comp p_1)~l = \var{apply}~p_2~(\var{apply}~p_1~l)
+  \end{equation*}
+\end{theorem}
+
+Because we represent the permutations as lists of outputs, we can
+simply apply one permutation to the other permutation to compose them.
+\lstinputlisting[
+  widthgobble=2, firstline=409, lastline=409
+]{../theories/BWT/Sorting/PermFun.v}
+This definition, while simple, does not make it obvious that
+\var{compose\_apply} holds.
+\begin{equation}
+  \label{eq:compose_apply}
+  \var{apply}~(\var{apply}~p_2~p_1)~l = \var{apply}~p_2~(\var{apply}~p_1~l)
+\end{equation}
+
+Conceptually, the theorem works because when we permute the list $l$,
+we can think of it as permuting the original indices $\iota =
+[0,~1,~\ldots,~n-1]$. If we rephrase \Cref{eq:compose_apply} the effect on
+on $I$, we get
+\begin{equation*}
+  \var{apply}~(\var{apply}~p_2~p_1)~\iota = \var{apply}~p_2~(\var{apply}~p_1~\iota)
+\end{equation*}
+This equation clearly holds if we can prove that $\iota$ is the identity
+of \var{apply}, giving as an outline of a proof.
+
+We will use the function \var{combine} defined in the Coq standared
+library to annotate every element of $l$ with its index.
+\begin{lstlisting}
+  Fixpoint combine (l : list A) (l' : list B) : list (A*B) :=
+    match l,l' with
+    | x :: tl, y :: tl' => (x,y) :: (combine tl tl')
+    | _, _ => nil
+    end.
+\end{lstlisting}
+
+To formalize the idea that we can reason about the indices of the
+original list $l$ in place of $l$, we need the following theorem.
+Intuitively, it says that if we have two lists of pairs, if we know
+that the $x$-coordinates are equal and we know that the list of pairs
+are permutations of one another, then we know the $y$-coordinates must
+be equal as well, because the permutation condition guarantees that
+each $x$-coordinate is paired with the same $y$-coordinate in both
+lists.
+\begin{theorem}[Permutation\_combine\_eq]
+  For all lists $xs$, $ys_1$, and $ys_2$ with all the same length,
+  if $xs$ has no duplicates then
+  \begin{equation*}
+    \var{Permutation}~(\var{combine}~xs~ys_1)~(\var{combine}~xs~ys_2) \implies ys_1 = ys_2
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We reason by induction on the list $\var{combine}~xs~ys_1$. In the
+  inductive case, we have to show that
+  \begin{equation*}
+    y_1 :: ys_1 = y_2 :: ys_2
+  \end{equation*}
+  given that
+  \begin{equation*}
+    (x,~y_1) :: \var{combine}~xs~ys_1 \eqperm (x, y_2) :: \var{combine}~xs~ys_2
+  \end{equation*}
+  This means that $(x, y_1) \in (x, y_2) :: \var{combine}~xs~ys_2$, so
+  either $(x, y_1) = (x, y_2)$, in which case we are done, or $(x,
+  y_1) \in \var{combine}~xs~ys_2$. In the latter case, we have a
+  contradiction, because we know that $x :: xs$ contains no
+  duplicates, so $x$ cannot be in $xs$.
+\end{proof}
+
+We also need to show that \var{apply} distributes over \var{combine}:
+\begin{theorem}[apply\_combine]
+  For all lists $xs$ and $ys$ with equal length,
+  \begin{equation*}
+    \var{apply}~p~(\var{combine}~xs~ys) =
+    \var{combine}~(\var{apply}~p~xs)~(\var{apply}~p~ys)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  Let $n = \var{length}~xs = \var{length}~ys$. We reason by induction
+  on n. In the inductive case, we have to show
+  \begin{align*}
+       & \var{apply}~(i::p)~(\var{combine}~(x::xs)~(y::ys)) \\
+    ={}& \var{combine}~(\var{apply}~(i::p)~(x::xs))~(\var{apply}~(i::p)~(y::ys))
+  \end{align*}
+  We can use \var{rem\_PermFun\_correct} to simplify both sides of the
+  equation. We have that the heads of both lists must are equal:
+  \begin{equation*}
+    (\var{combine}~(x::xs)~(y::ys))_i = ((x::xs)_i, (y::ys)_i)
+  \end{equation*}
+  That the tails are equal follows from the induction hypothesis, so
+  we are done.
+\end{proof}
+
+Now we show that $\iota$ is the identity of $\var{compose}$/$\var{apply}$.
+\begin{theorem}[compose\_id\_l]
+  For all permutation functions $p$ on lists of length $n$,
+  \begin{equation*}
+    \var{compose}~(\var{seq}~0~n)~p = p
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  The proof follows from induction on $p$. In the inductive case,
+  \begin{align*}
+       \var{compose}~(\var{seq}~0~(n+1))~(i::p)
+    &= \var{map}~(\lambda j.~(i::p)_j)~(\var{seq}~0~(n+1)) \\
+    &= \var{map}~(\lambda j.~(i::p)_j)~(0 :: \var{seq}~1~(n+1)) \\
+    &= i :: \var{map}~(\lambda j.~(i::p)_j)~(\var{seq}~1~(n+1)) \\
+    &= i :: \var{map}~(\lambda j.~(i::p)_j)~(\var{map}~(\lambda j.~j+1)~(\var{seq}~0~n)) \\
+    &= i :: \var{map}~(\lambda j.~(i::p)_{j+1})~(\var{seq}~0~n) \\
+    &= i :: \var{map}~(\lambda j.~p_j)~(\var{seq}~0~n) \\
+    &= i :: \var{compose}~(\var{seq}~0~n)~p \\
+    &= i :: p \qedhere
+  \end{align*}
+\end{proof}
+
+\begin{theorem}[compose\_id\_r]
+  For all permutation functions $p$ on lists of length $n$,
+  \begin{equation*}
+    \var{compose}~p~(\var{seq}~0~n) = p
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  \begin{equation*}
+       \var{compose}~p~(\var{seq}~0~n)
+    = \var{map}~(\lambda j.~(\var{seq}~0~n)_j)~p
+    = \var{map}~(\lambda j.~j)~p
+    = p \qedhere
+  \end{equation*}
+\end{proof}
+
+Now we can prove the main theorem:
+\begin{theorem}[compose\_apply]
+  For all lists $l$ and permutation functions $p_1$ and $p_2$
+  \begin{equation*}
+      \var{apply}~(p_2 \comp p_1)~l = \var{apply}~p_2~(\var{apply}~p_1~l)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  Letting $\iota = \var{seq}~0~(\var{length}~l)$, we annotate $l$ with
+  indices. By \var{apply\_correct}, we have
+  \begin{equation*}
+     \var{combine}~\iota~l \eqperm \var{apply}~(p_2 \comp p_1)~(\var{combine}~\iota~l)
+  \end{equation*}
+  and
+  \begin{equation*}
+    \var{combine}~\iota~l
+    \eqperm
+    \var{apply}~p_1~(\var{combine}~\iota~l)
+    \eqperm
+    \var{apply}~p_2~(\var{apply}~p_1~(\var{combine}~\iota~l))
+  \end{equation*}
+  We combine these by transtivity, and then simplify, using \var{app}
+  as an abbreviation for \var{apply}.
+  \begin{align*}
+    \var{pap}~(p_2 \comp p_1)~(\var{combine}~\iota~l)
+    &\eqperm
+    \var{app}~p_2~(\var{app}~p_1~(\var{combine}~\iota~l)) \\
+    \var{combine}~(\var{app}~(p_2 \comp p_1)~\iota)~(\var{app}~(p_2 \comp p_1)~l)
+    &\eqperm
+    \var{app}~p_2~(\var{combine}~(\var{app}~p_1~\iota)~(\var{app}~p_1~l)) \\
+    \var{combine}~(p_2 \comp p_1)~(\var{app}~(p_2 \comp p_1)~l)
+    &\eqperm
+    \var{app}~p_2~(\var{combine}~p_1~(\var{app}~p_1~l)) \\
+    \var{combine}~(p_2 \comp p_1)~(\var{app}~(p_2 \comp p_1)~l)
+    &\eqperm
+    \var{combine}~(\var{app}~p_2~p_1)~(\var{app}~p_2~(\var{app}~p_1~l))
+  \end{align*}
+  Because $p_2 \comp p_1 = \var{app}~p_2~p_1$, we can apply
+  \var{Permutation\_combine\_eq}.
+\end{proof}
 

### 2019-05-06 draft_paper/draft_paper.tex
Finish appendices

@@
 
-Although it is difficult to prove, it is intuitive that prefix orders
-that can look at more elements of the lists are ``stronger'' than
-those that can look at fewer.
+We can describe the relationship between orders that can look at fewer
+or more characters with the following two theorems. We will briefly
+offer an explanation of why these theorems makes sense, but these
+explanations do not mirror the Coq proofs we have for these theorems.
+
+\begin{theorem}[key\_lt\_firstn\_ge]
+  For all $j$ and $k$ such that $j \le k$,
+  \begin{equation*}
+    x <_j y \implies x <_k y
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  If $x <_j y$, then there must be some index $i < j$ such that $x_i <
+  y_i$. Then $i < k$ as well, so $x <_k y$.
+\end{proof}
 
@@
 \end{theorem}
+\begin{proof}
+  If $x \le_k y$, then either the first $k$ characters are all
+  equivalent, in which case the first $j$ characters are all
+  equivalent as well, or there was some $i < k$ such that $x_i < y_i$,
+  but at every index before $i$ $x$ and $y$ are equivalent. If $j <
+  i$, then the first $j$ characters are still equivalent, and we have
+  $x \equiv y \implies x \le_j y$. If $j \ge i$, then the deciding character is
+  among the first $j$ and $x \le_j k$.
+\end{proof}
 
@@
 
+In this section, we prove the following theorem:
+\begin{theorem}[hdsort\_sorted\_S]
+  For all matrices $m$ and naturals $j$,
+  \begin{equation*}
+    \var{PrefixSorted}~j~(\var{map}~\var{tl}~m) \implies
+    \var{PrefixSorted}~(j+1)~(\var{hdsort}~m)
+  \end{equation*}
+\end{theorem}
+
+First we prove this lemma:
+\begin{lemma}[insert\_Sorted\_S]
+  For all matrices $m$, rows $r$, and naturals $j$,
+  \begin{gather*}
+    \var{PrefixSorted}~(j+1)~(\var{hdsort}~m) \implies \\
+    \var{PrefixSorted}~j~(\var{tl}~r :: \var{map}~{tl}~m) \implies \\
+    \var{PrefixSorted}~(j+1)~(\var{insert}_1~r~(\var{hdsort}~m))
+  \end{gather*}
+  where $\var{insert}_1$ is \var{insert} that uses the 1-prefix
+  lexicographic ordering used by \var{hdsort}.
+\end{lemma}
+\begin{proof}
+  We can destruct $\var{insert}_1~r~(\var{hdsort}~m)$ as $m_1 \dplus r
+  :: m_2$, where $m_1 \dplus m_2 = \var{hdsort}~m$. Then by
+  \var{Sorted\_app\_cons}, it suffices to show
+  \begin{gather*}
+    \var{PrefixSorted}~(j+1)~m_1 \land \\
+    \var{PrefixSorted}~(j+1)~(r :: m_2) \land \\
+    (\forall x \in m_1,~ r >_{j+1} x)
+  \end{gather*}
+  \begin{itemize}
+  \item $\var{PrefixSorted}~(j+1)~m_1$ follows from the fact that
+    $\var{hdsort}~m = m_1 \dplus m_2$ and
+    $\var{PrefixSorted}~(j+1)~(\var{hdsort}~m)$.
+  \item By similar reasoning, we know that
+    $\var{PrefixSorted}~(j+1)~m_2$, so in order to show
+    $\var{PrefixSorted}~(j+1)~(r :: m_2)$ all we need to prove is that
+    for every row $x$ in $m_2$,
+    \begin{equation*}
+      r \le_{j+1} x
+    \end{equation*}
+    We know from the fact that $\var{insert}_1~r~(\var{hdsort}~m) =
+    m_1 \dplus r :: m_2$ that $r \le_1 x$. Then by \var{key\_firstn\_S},
+    we only need to show that
+    \begin{equation*}
+      \var{tl}~r \le_j \var{tl}~x
+    \end{equation*}
+    This follows from $\var{PrefixSorted}~j~(\var{tl}~r :: \var{map}~{tl}~m)$.
+  \item We have that for any $x \in m_1$, $r >_1 x$ by the fact that
+    $\var{insert}_1~r~(\var{hdsort}~m) = m_1 \dplus r :: m_2$. This
+    gives us $r >_{j+1} x$ by \var{key\_lt\_firstn\_ge}. \qedhere
+  \end{itemize}
+\end{proof}
+
+%% \begin{proof}
+%%   We reason by induction on $m$. In the inductive case, we need to show
+%%   \begin{equation*}
+%%     \var{PrefixSorted}~(j+1)~(\var{hdsort}~(r::m))
+%%   \end{equation*}
+%%   If we simplify \var{hdsort}, this becomes
+%%   \begin{equation*}
+%%     \var{PrefixSorted}~(j+1)~(\var{insert}_1~r~(\var{hdsort}~m))
+%%   \end{equation*}
+%%   where $\var{insert}_1$ is \var{insert} using the $1$-prefix
+%%   lexicographic ordering used in \var{hdsort}.
+%%   We know from our assumption that
+%%   \begin{equation*}
+%%     \var{PrefixSorted}~j~(\var{tl}~x :: \var{map}~\var{tl}~m)
+%%   \end{equation*}
+%%   and from our induction hypothesis we have
+%%   \begin{equation*}
+%%     \var{PrefixSorted}~(j+1)~(\var{hdsort}~m)
+%%   \end{equation*}
+%% \end{proof}
+
+  %% We can deconstruct $\var{insert}_1~r~(\var{hdsort}~m)$ as
+  %% $m_1 \dplus r :: m_2$, where we also know that $m = m_1 \dplus m_2$.
+  %% Now we have to show
+  %% \begin{equation*}
+  %%   \var{PrefixSorted}~(j+1)~(m_1 \dplus r :: m_2)
+  %% \end{equation*}
+
 \section{Permutation Composition}

### 2019-05-06 draft_paper/draft_paper.tex
stash

@@
 
-The forwards transform is straightforward to describe: for
-a string of length \(n\), form the \(n \times n\) rotaion matrix of all
-cyclic shifts of the original string, sort it lexicographically, and
-take the last column of the matrix. In addition, we note the index
-where the original string appears in the sorted rotation matrix. As
-shown in \Cref{fig:bw_ex}, applying the algorithm to the string
-\var{abracadabra!} produces the new string \var{ard!rcaaabb} and
-the index \var{3}. The transformation tends to produce runs like
-the three \var{a}s or the two \var{b}s in \var{ard!rcaaabb},
-and it is this feature that makes the resulting string easier to
-compress. Although it is not obvious, the transformed string, together
-with the index, provide enough information to recover the the original
-string. The BWT is discussed more in \cref{sec:bwt}.
+The forwards transform is straightforward to describe: for a string of
+length \(n\), form the \(n \times n\) matrix of all cyclic shifts of
+the original string (also referred to as the rotation matrix), sort
+the rows lexicographically, and take the last column of the matrix. In
+addition, we note the index where the original string appears in the
+sorted rotation matrix. As shown in \Cref{fig:bw_ex}, applying the
+algorithm to the string \var{abracadabra!} produces the new string
+\var{ard!rcaaabb} and the index \var{3}. The transformation tends to
+produce runs like the three \var{a}s or the two \var{b}s in
+\var{ard!rcaaabb}, and it is this feature that makes the resulting
+string easier to compress. Although it is not obvious, the transformed
+string, together with the index, provide enough information to recover
+the the original string. The BWT is discussed more in \cref{sec:bwt}.
 
@@
 the BWT, together with a machine-checked proof that our implementation
-is correct. Specifically, we implement the functions $\var{bwp} :
-\var{list}~A \to \var{list}~A$ and $\var{bwn} : \var{list}~A
-\to \mathbb{N}$ that produce the string and index part of the BWT
+is correct. Specifically, we implement the functions $\var{bwl} :
+\var{list}~A \to \var{list}~A$ and $\var{bwn} : \var{list}~A \to
+\mathbb{N}$ that produce the string and index part of the BWT
 respectively, and the function $\var{unbwt} : \mathbb{N} \to
-\var{list}~A \to \var{list}~A$ that reverses the transform. We
-prove the following theorem:
+\var{list}~A \to \var{list}~A$ that reverses the transform. We prove the
+following theorem:
 \begin{theorem}[unbwt\_correct]
@@
   \begin{equation*}
-    \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l) = l
+    \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l) = l
   \end{equation*}
@@
 
-As a part of the proof of \Cref{thm:unbwt_correct}, we implemented
-radix-sort and proved its correctness,
+As a part of the proof of \Cref{thm:unbwt_correct}, we proved two
+other noteworthy results that have not, to our knowledge, been proved
+before in Coq. We implemented a least-significant-digit (LSD)
+radixsort and proved its correctness,
 \begin{theorem}[radixsort\_correct]
   \label{thm:radixsort_correct}%
-  For all lists $l$,
-  \begin{equation*}
-    \var{Sorted}~(\var{radixsort}~l) \land \var{Permutation}~(\var{radixsort}~l)~l
-  \end{equation*}
-\end{theorem}
-and proved that any two stable sorts are interchangable,
-\begin{theorem}[stable\_sort\_unique]
-  \label{thm:radixsort_correct}%
-  For all functions $f, g : \var{list}~A \to \var{list}~A$,
-  \begin{gather*}
-    (\forall l,~\var{Sorted}~(f~l) \land \var{StablePermutation}~(f~l)~l) \implies \\
-    (\forall l,~\var{Sorted}~(g~l) \land \var{StablePermutation}~(g~l)~l) \implies \\
-    \forall l,~f~l = g~l
-  \end{gather*}
-\end{theorem}
+  For all matrices $m$ with $n$ columns,
+  \begin{equation*}
+    \var{Sorted}~(\var{radixsort}~m~n) \land \var{Permutation}~(\var{radixsort}~m~n)~m
+  \end{equation*}
+\end{theorem}
+and we proved that any two stable sorts are interchangable,
+\begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
+  For all functions $f, g: \var{list}~A \to \var{list}~A$,
+  \begin{equation*}
+    \var{StableSort}~f \implies \var{StableSort}~g \implies
+    (\forall l,~f~l = g~l)
+  \end{equation*}
+\end{restatable*}
 This in turn required developing suitable definitions for
-\var{Sorted}, \var{Permutation}, and \var{StablePermutation}.
-We explored multiple definitions for each of these properties, and
-proved equivalence between them.
-
-The rest of this paper is organize as follows. \Cref{sec:bwt} details
+\var{Sorted}, \var{Permutation}, and \var{StablePerm}. We explored
+multiple definitions for each of these properties, and proved
+equivalence between them.
+
+The rest of this paper is organized as follows. \Cref{sec:bwt} details
 the history of the BWT and its applications. \Cref{sec:ord} defines
@@
 \Cref{sec:sort} implements insertion sort and radixsort, and proves
-their correctness. \Cref{appendix:BWT_proof} more formally proves the
-theorems about the BWT mentioned in previous sections, including
-\Cref{thm:unbwt_correct}.
+their correctness.
 
@@
 
-\section{Burrows--Wheeler}
+\section{History and Applications of the BWT}
 \label{sec:bwt}
@@
 from the runs in the output of BWT. On English texts, often 50-60\% of
-the values produced by the MTF pass are zeroes
-\cite{fenwick2007,bw-analysis}. Often a Run-Length Encoding (RLE) pass
-is then used to encode only the runs of zeroes with their lengths,
-using a bijective base-2 encoding \cite{bw-analysis, tsai_2016}.
-Finally, some sort of zeroth-order coder like a Huffman coder is used.
+the values produced by the MTF pass are zeroes \cite{fenwick2007}.
+Often a Run-Length Encoding (RLE) pass is then used to encode only the
+runs of zeroes with their lengths, using a bijective base-2 encoding
+\cite{bw-analysis, tsai_2016}. Finally, some sort of zeroth-order
+coder like a Huffman coder is used.
 
@@
 in order to search for patterns in a compressed text
-\cite{ferragina_index}.
+\cite{ferragina_index}. These techniques have been applied to enable
+fast matching of DNA subsequences.
 
@@
 that this forms a preorder on $A$ when $\le$ is reflexive and transitive, that is
-$x \le x$ for all $xA$, and
+$x \le x$ for all $x:A$, and
 \begin{equation*}
@@
 \end{equation*}
-for all $x,y,z$.
+for all $x,y,z:A$.
 
@@
 
-From $\le$, we can define $x \ge y \coloneqq y \le x$, $x < y \coloneqq
-\lnot (y \le x)$, and $x > y \coloneqq \lnot (x \le y)$, and prove
-transitivity and reflexivity where appropriate.
-
-We can also define $x \equiv y \coloneqq x \le y \land y \le x$. $\equiv$ is a
+From $\le$, we can define
+\begin{align*}
+x \ge y \coloneqq y \le x && x < y \coloneqq \lnot (y \le x) && x > y \coloneqq \lnot (x \le y)
+\end{align*}
+
+We can also define
+\begin{equation*}
+  x \equiv y \coloneqq x \le y \land y \le x
+\end{equation*}
+$\equiv$ is a
 decidable equivalence relation, which means it is reflexive,
 transitive, and symmetric, and we have a decision procedure for it.
-\lstinputlisting[
-  firstline=116, lastline=120, widthgobble=2
-]{../theories/BWT/Sorting/Ord.v}
-
-The builtin class \var{EqDec} is used to describe types with
-decidable equivalence relations, so for any \var{Preord} instance
-we have the \var{EqDec} instance \var{Preord\_EqDec} using
-\var{eqv\_dec}. In Coq, the notation \var{===} is used for the
-equivalence relation of an \var{EqDec} instance and \var{==} for
-its decision procedure.
+
+The builtin class \var{EqDec} is used to describe types with decidable
+equivalence relations, so for any \var{Preord} instance we have the
+\var{EqDec} instance \var{Preord\_EqDec}. In Coq, the notation
+\var{===} is used for the equivalence relation of an \var{EqDec}
+instance and \var{==} for its decision procedure.
 
@@
 We can prove that this definition satisfies \Cref{eq:iter} by defining
-the function \var{rep} for repeated function application.
+the function \var{rep}\footnote{Alternatively, we can say that
+  $\lambda f.~\var{rep}~n~f$ is the $n$th Church numeral.} for
+repeated function application.
 \lstinputlisting[
@@
 
-\var{bwp} produces the transformed string by taking the last column
+\var{bwl} produces the transformed string by taking the last column
 of the sorted rotation matrix. \var{lexsort} sorts the rotation
@@
 \begin{equation}
-  \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l) = l
+  \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l) = l
   \label{eq:unbwt}
@@
 If we could recreate the entire sorted rotation matrix from
-$\var{bwp}~l$, it would be simple to recover $l$ by indexing into
+$\var{bwl}~l$, it would be simple to recover $l$ by indexing into
 the matrix at $\var{bwn}~l$.
@@
 \begin{equation}
-  \var{recreate}~(\var{bwp}~l) = \var{lexsort}~(\var{rots}~l)
+  \var{recreate}~(\var{bwl}~l) = \var{lexsort}~(\var{rots}~l)
   \label{eq:recreate}
@@
 \begin{align*}
-     \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l)
-  &= (\var{recreate}~(\var{bwp}~l))_{(\var{bwn})~l}\\
+     \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l)
+  &= (\var{recreate}~(\var{bwl}~l))_{(\var{bwn})~l}\\
   &= (\var{lexsort}~(\var{rots}~l))_{(\var{bwn})~l} \\
@@
   \begin{equation*}
-    \var{recreate}~j~(\var{bwp}~l) =
+    \var{recreate}~j~(\var{bwl}~l) =
     \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
@@
 ={}&
-\var{hdsort}~(\var{prepend\_col}~(\var{bwp}~l)~(\var{recreate}~j~(\var{bwp}~l))) \\
-={}& \var{recreate}~j~(\var{bwp}~l) \qedhere
+\var{hdsort}~(\var{prepend\_col}~(\var{bwl}~l)~(\var{recreate}~j~(\var{bwl}~l))) \\
+={}& \var{recreate}~j~(\var{bwl}~l) \qedhere
 \end{align*}
@@
 
-The previous section's derivation relied on the correctness and stability of
-various sorts. To make those parts of the derivation precise, we
-define here what it means for a sorting function to be correct, and
-what it means for a sort to be stable.
-
-Intuitively, a sorting function must produce a sorted list that
-contains all of the elements that its input contained. So, given a function $f : \var{list}~A \to \var{list}~A$, we say that $\var{Sort}~f$ precisely when for all $l$,
-\begin{equation*}
-  \var{Sorted}~(f~l) \land \var{Permutation}~(f~l)~l
-\end{equation*}
-Now we need to define \var{Sorted} and \var{Permutation} to capture
-the idea that a list is in sorted order, and that two lists contain
-all of the same elements.
+The previous section's proofs relied on the correctness and stability
+of various sorts. The Coq standard library includes definitions for
+\var{Sorted} and \var{Permutation}, but no definition for stability.
+The Coq standard library also includes an implementation of mergesort.
+However, it uses Coq's module system to enable parametricity over
+ordered types, which is much less convenient to use than typeclasses.
+In this section, we define our own version of \var{Sorted} and
+\var{StablePerm} that are based on our \var{Preord} typeclass.
+
+When defining these properties, it is very important that they mean
+what we would like them to mean. If we prove our implementations
+satisfy our specifications, but our specifications are incorrect, then
+the proof is worthless. In addition, it would be nice if the
+properties are defined in a way that is elegant to reason about. So,
+we define each property in multiple ways and prove equivalence between
+them: some definitions are more obviously correct, while others might
+be easier to reason about it different situations. Even if a
+definition is never used, proving equivalence gives us confidence
+about the correctness of the definitions.
 
@@
 However, it is often easier to reason about inductively defined
-properties, so we would like to define \var{Sorted} inductively. So,
-we can say that as a base case, the empty list is sorted, and given a
+properties, so we would like to define \var{Sorted} inductively. We
+can say that as a base case, the empty list is sorted, and given a
 sorted list, we can add an element that is smaller than all the
@@
 obviously equivalent to the first definition, which is more clearly
-correct. We prove them equivalent all:
+correct. We will prove all three equivalent.
+
 \begin{theorem}[SortedLocal\_iff]
@@
 \begin{proof}
-  The $\implies$ direction follows by induction. The $\impliedby$
-  direction follows by induction on $l$, reasoning about the empty
-  list, the singleton list, and lists with at least two elements. In
-  the inductive case $l = a :: b :: l'$, we have that $a \le b$, and
-  that $\forall x \in l',~b \le x$, which gives us $\forall x \in b :: l',~a \le x$ by
-  transitivity.
+  Both directions follow by induction on $l$, making use of the
+  transitivity of $\le$.
 \end{proof}
@@
 \begin{theorem}[SortedIx\_iff]
-  A list $l$ is sorted under the inductive definition (\var{Sorted}) if
-  and only if it is sorted under the index definition (\var{SortedIx}).
+  For all lists $l$,
+  \begin{equation*}
+    \var{Sorted}~l \iff \var{SortedIx}~l
+  \end{equation*}
 \end{theorem}
@@
   $i \le j$ and $\var{Sorted}~l$. We reason by cases on $i$ and $j$:
-  \begin{itemize}[noitemsep]
+  \begin{itemize}
   \item $i = j = 0$. By reflexivity, $a \le a$.
   \item $i = 0,~j=j' + 1$. We need to show $a \le l'_{j'}$. We know that
-    $a \le x$ for all $x \in l'$ by $\var{SortedIx}~(a::l')$, so we are
+    $a \le x$ for all $x \in l'$ by $\var{Sorted}~(a::l')$, so we are
     done.
@@
 permutations of each other is precisely when
-\vspace{-\parskip}
-\begin{itemize}[nolistsep]
+\begin{itemize}
   \item $\var{length}~l = \var{length}~l' = n$
@@
 
-We will prove these two definitions equivalent, but first we confirm that our definition of \var{PermFun} ensures that .
-
-We know we will
-eventually need to reason inductively about permutation functions, so
-we would like to define the relationship between applying a
-permutation function $i :: p$ and applying $p$. Conceptually, the
-first element of $\var{apply}~(i::p)~l$ will be $l_i$. The rest will
-be the result of applying some new permutation $p'$, where everything
-greater than $i$ has been shifted down by one, to some new list $l'$,
-which has its $i$th element removed. So, we define \var{rem\_PermFun}
-and \var{rem\_nth} respectively, hoping to prove that
+We know we will eventually need to reason inductively about
+permutation functions, so we would like to define the relationship
+between applying a permutation function $i :: p$ and applying $p$.
+Conceptually, the first element of $\var{apply}~(i::p)~l$ will be
+$l_i$. The rest will be the result of applying some new permutation
+$p'$, where everything greater than $i$ has been shifted down by one,
+to some new list $l'$, which has its $i$th element removed. So, we
+define \var{rem\_PermFun} and \var{rem\_nth} respectively, hoping to
+prove that
 \begin{equation*}
@@
 
-\begin{corollary}[StableSort\_unique]
-  For all functions $f, g: \var{list}~A \to \var{list}~A$,
-  \begin{equation*}
-    \var{StableSort}~f \implies \var{StableSort}~g \implies
-    (\forall l,~f~l = g~l)
-  \end{equation*}
-\end{corollary}
 \begin{proof}
@@
 
-\begin{corollary}[Sort\_Ord\_unique]
-  When sorting using an order,
-  \begin{equation*}
-    \var{Sort}~f \implies \var{Sort}~g \implies
-    (\forall l,~f~l = g~l)
-  \end{equation*}
-  for any $f$ and $g$.
-\end{corollary}
+\stablesortunique
 \begin{proof}
@@
 
-%% \begin{proof}
-%%   We reason by induction on $m$. In the inductive case, we need to show
-%%   \begin{equation*}
-%%     \var{PrefixSorted}~(j+1)~(\var{hdsort}~(r::m))
-%%   \end{equation*}
-%%   If we simplify \var{hdsort}, this becomes
-%%   \begin{equation*}
-%%     \var{PrefixSorted}~(j+1)~(\var{insert}_1~r~(\var{hdsort}~m))
-%%   \end{equation*}
-%%   where $\var{insert}_1$ is \var{insert} using the $1$-prefix
-%%   lexicographic ordering used in \var{hdsort}.
-%%   We know from our assumption that
-%%   \begin{equation*}
-%%     \var{PrefixSorted}~j~(\var{tl}~x :: \var{map}~\var{tl}~m)
-%%   \end{equation*}
-%%   and from our induction hypothesis we have
-%%   \begin{equation*}
-%%     \var{PrefixSorted}~(j+1)~(\var{hdsort}~m)
-%%   \end{equation*}
-%% \end{proof}
-
-  %% We can deconstruct $\var{insert}_1~r~(\var{hdsort}~m)$ as
-  %% $m_1 \dplus r :: m_2$, where we also know that $m = m_1 \dplus m_2$.
-  %% Now we have to show
-  %% \begin{equation*}
-  %%   \var{PrefixSorted}~(j+1)~(m_1 \dplus r :: m_2)
-  %% \end{equation*}
-
 \section{Permutation Composition}

### 2019-05-06 draft_paper/draft_paper.tex
Stash

@@
 approach to greedy algorithms, the Boyer-Moore algorithm, the
-Knuth-Morris-Pratt algoirthm, Sudoku solvers, arithmetic coding, and
-the Schorr-Waite algorithm, the Johnson-Trotter algorithm, and many
-more problems \cite{pearls}.
+Knuth-Morris-Pratt algorithm, Sudoku solvers, arithmetic coding, the
+Schorr-Waite algorithm, the Johnson-Trotter algorithm, and many more
+problems \cite{pearls}.
 
@@
 \end{theorem}
-and we proved that any two stable sorts are interchangable,
+and we proved that any two stable sorts are interchangeable,
 \begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
@@
 
-\section{History and Applications of the BWT}
+\section{History and applications of the BWT}
 \label{sec:bwt}
@@
 is represented by the number of different characters encountered since
-the last occurence of this character. This creates many runs of zeroes
+the last occurrence of this character. This creates many runs of zeroes
 from the runs in the output of BWT. On English texts, often 50-60\% of
@@
 
-\section{Orders and Preorders}
+\section{Orders and preorders}
 \label{sec:ord}
@@
 \var{EqDec} instance \var{Preord\_EqDec}. In Coq, the notation
-\var{===} is used for the equivalence relation of an \var{EqDec}
-instance and \var{==} for its decision procedure.
+\var{=\hspace{0pt}=\hspace{0pt}=} is used for the equivalence relation of an \var{EqDec}
+instance and \var{=\hspace{0pt}=} for its decision procedure.
 
@@
 
-\section{The Forwards BWT}
+\section{The forwards BWT}
 \label{sec:forwards_BWT}
@@
 
-Now we would likel to produce the rotation matrix by repeatedly
-applying \var{lrot}. We will use \var{iter}, which takes a function
-$f : A \to A$, a number $n$, and an initial value $z$, and returns the
-list of intermediate values obtained by applying $f$ to $z$ $n$ times.
-Or,
+Now we produce the rotation matrix by repeatedly applying \var{lrot}.
+We will use \var{iter}, which takes a function $f : A \to A$, a number
+$n$, and an initial value $z$, and returns the list of intermediate
+values obtained by applying $f$ to $z$ $n$ times. Or,
 \begin{equation}
-  \var{iter}~f~n~z = [z,~f~z,~f~(f~z),~\ldots,~f^n z]
+  \var{iter}~f~n~z = [z,~f~z,~f~(f~z),~\ldots,~f^{n-1} z]
   \label{eq:iter}
@@
 We can prove that this definition satisfies \Cref{eq:iter} by defining
-the function \var{rep}\footnote{Alternatively, we can say that
-  $\lambda f.~\var{rep}~n~f$ is the $n$th Church numeral.} for
-repeated function application.
+the function \var{rep} for repeated function application.
 \lstinputlisting[
@@
 matrix lexicographically using the \var{Ord} instance; we will
-discuss it's implementation in \cref{sec:sort}.
+discuss its implementation in \cref{sec:sort}.
 \lstinputlisting[
@@
 
-\section{The Inverse BWT}
+\section{The inverse BWT}
 \label{sec:inverse_BWT}
@@
 next column is the result of sorting by each character, sorting by the
-preceding chracter, and then sorting by the character 2 spaces in
+preceding character, and then sorting by the character 2 spaces in
 front.
@@
 rotation matrix from just the last column (\Cref{fig:recreate}).
-However, the column direclty to the left of the sorted column is very
+However, the column directly to the left of the sorted column is very
 important, as it is the next column to be sorted. And we can see that
@@
 $[\texttt{c},~\texttt{d},~\texttt{r},~\texttt{r},~\texttt{!}]$,
-because those are the characters that preced \texttt{a}s in the
+because those are the characters that precede \texttt{a}s in the
 original string. However, the order of these characters in the final
@@
 \texttt{!}, and the next character for the \texttt{!} is a \texttt{c}.
-This requires that the \texttt{d} preced the \texttt{!} in the last
+This requires that the \texttt{d} precede the \texttt{!} in the last
 column.
@@
 The word \textit{permutation} has two related meanings that are often
-used interchangably. We sometimes say that two lists are permutations of each
+used interchangeably. We sometimes say that two lists are permutations of each
 other if they contain all of the same elements, just appearing in a
@@
 To restrict the range to $\mathbb{N}_{<n}$ we say that $\forall i,~i \in p
-\implies ~i<n$, to make the function surjective, we strengthen that to
-$\forall i,~i \in p \iff ~i<n$, and to make it injective we require that $p$ not
+\implies ~i<n$; to make the function surjective, we strengthen that to
+$\forall i,~i \in p \iff ~i<n$; and to make it injective we require that $p$ not
 contain any duplicate entries. The inductive property \var{NoDup} from
@@
   induction on $n$. The base case is uninteresting. In the inductive
-  case, we have the induction hypothesis that for all lists of length
-  $n$ and permutation functions on such lists,
-  \begin{equation*}
+  case, we have the induction hypothesis that for $p :
+  \var{list}~\mathbb{N}$ and $l : \var{list}~A$
+  \begin{equation*}
+    \var{lenth}~l = n \implies \var{PermFun}~n~p \implies
     \var{Permutation}~(\var{apply}~p~l)~l
   \end{equation*}
-  and we must show that
-  \begin{equation*}
-    \var{Permutation}~(\var{apply}~(i :: p)~(a :: l))~(a :: l)
+  Given $\var{PermFun}~(n+1)~(i::p')$, we must show that
+  \begin{equation*}
+    \var{Permutation}~(\var{apply}~(i :: p')~(a :: l))~(a :: l)
   \end{equation*}
@@
   \begin{align*}
-    \var{apply}~(i :: p')~(a :: l) &\eqperm
-    (a :: l)_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a::l)) \\
-    &\eqperm (a :: l)_i :: \var{rem\_nth}~i~(a::l) \\
-    &\eqperm a :: l
-  \end{align*}
-  where the first step follows from \var{rem\_PermFun\_correct}, the
-  second from the induction hypothesis, and the last from
-  \var{rem\_nth\_Perm}.
+    &\var{apply}~(i :: p')~(a :: l) \\
+    \eqperm{}& \text{\{ by \var{rem\_PermFun\_correct} \}} \\
+    &(a :: l)_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a::l))
+    \\
+    \eqperm{}& \text{\{ by IH \}} \\
+    &(a :: l)_i :: \var{rem\_nth}~i~(a::l)
+    \\
+    \eqperm{}& \text{\{ \var{rem\_nth\_Perm} \}} \\
+    & a :: l \qedhere
+  \end{align*}
 \end{proof}
@@
 ]{../theories/BWT/Sorting/PermutationCount.v}
-\var{count\_occ} counts the number of occurences of a given variable
+\var{count\_occ} counts the number of occurrences of a given variable
 using a decision procedure for Leibniz equality. Here that decision
@@
 means that $p$ represents a permutation function on lists of length
-$n$, we need to specify the list with resepct to which $p$ is stable,
+$n$, we need to specify the list with respect to which $p$ is stable,
 so we say $\var{StablePermFun}~l~p$ for some list $l$.
-\lstinputlisting[
-  widthgobble=2, firstline=557, lastline=561,
-]{../theories/BWT/Sorting/PermFun.v}
-As above, we can also define this property using the inverse
-permutation:
-\lstinputlisting[
+\lstinputlisting[ widthgobble=2, firstline=557, lastline=561,
+]{../theories/BWT/Sorting/PermFun.v} As above, we can also define this
+property using the inverse permutation: \lstinputlisting[
   widthgobble=2, firstline=563, lastline=567,
-]{../theories/BWT/Sorting/PermFun.v}
-As usual, we will prove these definitions equivalent. First, though,
-we need to define \var{image} and \var{preimage}.
-\lstinputlisting[
-  widthgobble=2, firstline=85, lastline=86,
-]{../theories/BWT/Sorting/PermFun.v}
-We will not prove it here, but $\var{preimage}~p$ and $\var{image}~p$
-are both of bijections of type $\mathbb{N}_{<n} \to \mathbb{N}_{<n}$
-and are invereses of each other. We also have the following two
-theorems:
+]{../theories/BWT/Sorting/PermFun.v} As usual, we will prove these
+definitions equivalent. First, though, we need to define \var{image}
+and \var{preimage}. \lstinputlisting[ widthgobble=2, firstline=85,
+  lastline=86, ]{../theories/BWT/Sorting/PermFun.v} We will not prove
+it here, but $\var{preimage}~p$ and $\var{image}~p$ are both of
+bijections of type $\mathbb{N}_{<n} \to \mathbb{N}_{<n}$ and are
+inverses of each other. We also have the following two theorems:
 \begin{theorem}[nth\_preimage\_apply, nth\_image\_apply]
@@
 
-Unforunately, neither of these two definitions will be particularly
+Unfortunately, neither of these two definitions will be particularly
 useful for proving the theorems we would like to prove with them. The
@@
 
-\section{Uniqueness of Stable Sorts}
+\section{Uniqueness of stable sorts}
 \label{sec:unique}
@@
   know $h \in t'$ and $h' \in t$. Because both $h::t$ and $h'::t'$ are
-  sorted, we knnow that $h \le h'$ and $h' \le h$, implying $h \equiv h'$.
+  sorted, we know that $h \le h'$ and $h' \le h$, implying $h \equiv h'$.
 
@@
 
-\section{Lexicographic Orders and Keys}
+\section{Lexicographic orders and keys}
 \label{sec:lex_ord}
@@
 last column to the front and sort $j$ times, the resulting matrix is a
-permtuation of simply rotating the last column to the front $j$
+permutation of simply rotating the last column to the front $j$
 times.
@@
 
+\section{Relationship to the standard BWT}
+\label{sec:opt}
+
+It is important to note that standard implementations of the BWT do
+not actually reconstruct the entire rotation matrix, as this would be
+inefficient. Instead, they avoid this using the optimization described
+in this section. Although, we did not prove or implement this
+optimization in Coq, it would be straightforward to do so.
+
+In \var{recreate}, we repeatedly prepend the last column to the front
+of the matrix and sort on it. However, each time that column is the
+same, so the permutation that takes us from the unsorted column to the
+sorted column is the same. Therefore, we can calculate that
+permutation function once and store it in a variable $p$. The
+$i$th column of the recreated matrix is simply
+$(\var{apply}~p)^i~(\var{bwl}~l)$, so we can define
+\begin{lstlisting}
+Definition recreate' (p : list nat) (l : list A) : list (list A) :=
+  transpose (iter (apply p) (length l) (apply p l))
+\end{lstlisting}
+In the definition above, we are repeatedly applying \var{p} to the
+first column of the rotation matrix, $\var{apply}~p~l$. Then we
+take the transpose to create the matrix from the columns.
+
+Since in $\var{unbwt}~i~l$, we want the $i$th row of the matrix, we
+can eliminate the transpose by replacing the $\var{nth}~i$ with a
+$\var{map}~\var{nth}~i$ Assuming we already have calculated $p$,
+\begin{align*}
+   \var{unbwt}~i~l &= (\var{recreate}~p~l)_i \\
+  &= (\var{transpose}~(\var{iter}~(\var{apply}~p)~(\var{length}~l)~(\var{apply}~p~l)))_i \\
+  &= \var{map}~(\lambda
+  r.~r_i)~(\var{iter}~(\var{apply}~p)~(\var{length}~l)~(\var{apply}~p~l)) \\
+  &= \var{map}~(\lambda r.~r_i)~(\var{tl}~(\var{iter}~(\var{apply}~p)~(1 +
+  \var{length}~l)~l)) \\
+  &= \var{tl}~(\var{map}~(\lambda r.~r_i)~(\var{iter}~(\var{apply}~p)~(1 + \var{length}~l)~l))
+\end{align*}
+
+Now, instead of indexing into each column at $i$, where the columns
+are all just the result of successively applying $p$ to $l$, we will
+successively take the image of $i$ under $p$ to generate the indices
+into $l$ where those characters appear. This is justified via the
+following theorem:
+\begin{theorem}
+  For all $n$,
+  \begin{equation*}
+  \var{map}~(\lambda r.~r_i)~(\var{iter}~(\var{apply}~p)~n~l)
+  =
+  \var{map}~(\lambda j.~l_j)~(\var{iter}~(\var{image}~p)~n~i)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  \begin{align*}
+       & \var{map}~(\lambda r.~r_i)~(\var{iter}~(\var{apply}~p)~n~l) \\
+    ={}& \var{map}~(\lambda r.~r_i)~[l,~(\var{apply}~p)~l,~(\var{apply}~p)^2~l,~\ldots,~(\var{apply}~p)^{n-1}~l] \\
+    ={}& [l_i,~((\var{apply}~p)~l)_i,~((\var{apply}~p)^2~l)_i,~\ldots,~((\var{apply}~p)^{n-1}~l)_i] \\
+    ={}& [l_i,~l_{p(i)},~l_{p^2(i)},~\ldots,~l_{p^{n-1}~i}] \\
+    ={}& \var{map}~(\lambda j.~l_j)~[i,~p(i),~p^2(i),~\ldots,~p^{n-1}~i] \\
+    ={}& \var{map}~(\lambda j.~l_j)~(\var{iter}~(\var{image}~p)~n~i) \qedhere
+  \end{align*}
+\end{proof}
+
+So assuming we have some function $\var{calc\_sort\_perm} :
+\var{list}~A \to \var{list}~\mathbb{N}$ to calculate the permutation
+function that sorts $l$, we redefine \var{unbwt} as
+\begin{lstlisting}
+Definition unbwt' (i : nat) (l : list A) : list A :=
+  match l with
+  | [] => []
+  | d :: _ =>
+    let p := calc_sort_perm l in
+    let indices := tl (iter (image p) (S (length l)) i) in
+    map (fun j => nth l j d) indices
+\end{lstlisting}
+\var{calc\_sort\_perm} can be implemented in linear time using a
+counting sort, so if \var{nth} took constant time, then \var{unbwt}
+would take linear time as well. As a functional specification, this
+closely matches the standard imperative algorithm for the inverse BWT.
+
+\section{Conclusions}
+We have implemented and proved that the the inverse BWT actually
+inverts the forwards BWT, basing our work on Bird et al.'s program
+calculation derivation \cite{birdmu,pearls}. In the process, we
+defined the properties of stable sorts, and implemented and proved
+correct an LSD radixsort.
+
+This paper shows that Bird's style of program calculation is very
+useful for building machine-checked proofs. It leads to proofs that
+are composed of many smaller lemmas, each justifying a single rewrite
+step. Most of the time and effort went into proving the correctness of
+radixsort. If we exclude the parts related to sorting or to other
+general-purpose functions not specific to the BWT, the entire verified
+implementation of the BWT consists of 42 lines of Gallina definitions,
+7 lines of Gallina that are used only in proofs, and 823 lines of
+theorem statements and Ltac code.\footnote{This includes
+  \var{BurrowsWheeler.v}, \var{Columns.v}, \var{Rotations/Rotation.v},
+  \var{Rotations/Rots.v}, \var{Lib/Iterate.v}, \var{Lib/Repeat.v} and
+  \var{Lib/FindIndex.v}, which implement \var{bwl}, \var{bwn},
+  \var{unbwt}, \var{cols}, \var{prepend\_col}, \var{lrot}, \var{rrot},
+  \var{rots}, \var{iter}, and \var{rep}, \var{findIndex}.}
+
+This implementation, including the optimizations mentioned in
+\cref{sec:opt}, can be used as a functional specification to for a
+verified implementation of bzip2. Of course, it would also be
+necessary to have functional specifications for the other phases of
+bzip2, such as Run-Length Encoding, Move-to-Front Encoding, and
+arithmetic coding. As luck would have it, Bird has already analyzed
+arithmetic coding \cite{pearls}.
+\clearpage
+
 \appendix
 
-\section{Insertion Sort}
+\section{Insertion sort}
 \label{appendix:insertion_sort}
@@
 example, our proof of correctness for radixsort relies on the fact
-that it uses insertion sort as the innner sort, but
+that it uses insertion sort as the inner sort, but
 \var{StableSort\_unique} allows us to extend that proof to a radixsort
@@
 
-\section{Permutation Composition}
+\section{Permutation composition}
 \label{appendix:perm_comp}
@@
 
-We will use the function \var{combine} defined in the Coq standared
+We will use the function \var{combine} defined in the Coq standard
 library to annotate every element of $l$ with its index.
@@
   \end{equation*}
-  We combine these by transtivity, and then simplify, using \var{app}
+  We combine these by transitivity, and then simplify, using \var{app}
   as an abbreviation for \var{apply}.

### 2019-05-07 draft_paper/draft_paper.tex
Finish

@@
-\documentclass[11pt, titlepage]{thesis}
+\documentclass[11pt]{thesis}
 
@@
 
-\maketitle
+\pagenumbering{gobble}
+
+\begin{titlepage}
+  \centering \Large{\textbf{Towards a Verified Bzip2: \\ A Functional Specification for Burrows--Wheeler
+  }} \vspace{10mm}
+
+  \Large\textit{Jake Waksbaum} \\
+  \vspace{25mm} \large{Junior Two-semester Independent Work \\ Department of
+    Computer Science \\ Princeton University \\ Fall 2018--Spring 2019\\}
+  \vspace{8mm}
+  \begin{figure}[!h]
+    \centering
+    \includegraphics[scale=0.1]{seal.png}
+  \end{figure}
+  \vspace{1mm}
+  \normalsize{Advised by Andrew W. Appel (COS)}\\
+  \normalsize{Second Reader: David Walker (COS)}
+
+  \vfill
+  \pledge{Jake Waksbaum}
+\end{titlepage}
+
+\clearpage
+
+\section*{Acknowledgements}
+
+Thanks to Professor Andrew Appel for being extremely generous with his
+time and his guidance. He was always available for questions, and
+allowed me to focus my work on the areas I found interesting.
+
+I'd also like to thank Matthew Weaver, Lennart Berringer, and Qinshi
+Wang for answering my questions throughout the year. Zachary Kincaid
+provided useful comments on a draft of this paper.
+
+I'd like to thank Benjamin Huang, Yael Marans, and Eli Waksbaum for
+their help and support.
+
+\clearpage
 
@@
 \clearpage
+
+\pagenumbering{arabic}
 
@@
 
-The forwards transform is straightforward to describe: for a string of
-length \(n\), form the \(n \times n\) matrix of all cyclic shifts of
-the original string (also referred to as the rotation matrix), sort
-the rows lexicographically, and take the last column of the matrix. In
-addition, we note the index where the original string appears in the
-sorted rotation matrix. As shown in \Cref{fig:bw_ex}, applying the
-algorithm to the string \var{abracadabra!} produces the new string
-\var{ard!rcaaabb} and the index \var{3}. The transformation tends to
-produce runs like the three \var{a}s or the two \var{b}s in
-\var{ard!rcaaabb}, and it is this feature that makes the resulting
-string easier to compress. Although it is not obvious, the transformed
-string, together with the index, provide enough information to recover
-the the original string. The BWT is discussed more in \cref{sec:bwt}.
+To apply the transformation to a string of length $n$:
+\begin{enumerate}
+\item Form the \(n \times n\) matrix of all cyclic shifts of the
+  original string (also referred to as the rotation matrix). Starting
+  with the original string, each row is formed from the previous one
+  by removing the first character and sticking it on the end.
+\item Sort the rows lexicographically
+\item Take the last column of the matrix. That string is the result of the transformation.
+\item Note the index where the original string appears in the sorted
+  rotation matrix. This piece of information is necessary to reverse
+  the transformation.
+\end{enumerate}
+As shown in \Cref{fig:bw_ex}, applying the algorithm to the string
+\var{abracadabra!} produces the new string \var{ard!rcaaabb} and the
+index \var{3}. The transformation tends to produce runs like the three
+\var{a}s or the two \var{b}s in \var{ard!rcaaabb}, and it is this
+feature that makes the resulting string easier to compress. Although
+it is not obvious, the transformed string, together with the index,
+provide enough information to recover the the original string.
 
@@
 transform. In addition to providing an intuitive derivation of the
-inverse transform, this provides the outline of a proof of the
-correctness of the Burrows--Wheeler algorithm. We give our version of
-their derivation in \cref{sec:forwards_BWT,sec:inverse_BWT}.
+inverse transform, this provides the outline of a proof of for the
+inverse BWT.
 
@@
 following theorem:
-\begin{theorem}[unbwt\_correct]
+\begin{restatable*}[unbwt\_correct]{theorem}{unbwtcorrect}
   \label{thm:unbwt_correct}
@@
   \end{equation*}
-\end{theorem}
-The implementation of the functions and the outline of the proof are
-based on Bird et~al.'s derivations.
+\end{restatable*}
+Our implementation and proof are based on Bird et~al.'s derivations.
 
@@
 specification for a potential verified implementation of a program
-that relies on Burrows--Wheeler \cite{appel-func-spec}. But more
-generally, this provides evidence that Bird's approach of program
-calculation, in which programs are derived from their specifications
-via algebraic manipulations, is useful for proving the correctness of
-algorithms in a machine-checked setting. Bird has applied this
-approach to greedy algorithms, the Boyer-Moore algorithm, the
-Knuth-Morris-Pratt algorithm, Sudoku solvers, arithmetic coding, the
-Schorr-Waite algorithm, the Johnson-Trotter algorithm, and many more
-problems \cite{pearls}.
+that relies on Burrows--Wheeler, such as bzip2\cite{appel-func-spec}.
+But more generally, this provides evidence that Bird's approach of
+program calculation, in which programs are derived from their
+specifications via algebraic manipulations, is useful for proving the
+correctness of algorithms in a machine-checked setting. Bird has
+applied this approach to greedy algorithms, the Boyer-Moore algorithm,
+the Knuth-Morris-Pratt algorithm, Sudoku solvers, arithmetic coding,
+the Schorr-Waite algorithm, the Johnson-Trotter algorithm, and many
+more problems \cite{pearls}.
 
@@
 radixsort and proved its correctness,
-\begin{theorem}[radixsort\_correct]
+\begin{restatable*}[radixsort\_correct]{theorem}{radixsortcorrect}
   \label{thm:radixsort_correct}%
@@
   \begin{equation*}
-    \var{Sorted}~(\var{radixsort}~m~n) \land \var{Permutation}~(\var{radixsort}~m~n)~m
-  \end{equation*}
-\end{theorem}
+    \var{Sorted}~(\var{radixsort}~m~n) \land \var{StablePerm}~(\var{radixsort}~m~n)~m
+  \end{equation*}
+\end{restatable*}
 and we proved that any two stable sorts are interchangeable,
 \begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
+  \label{thm:stablesort_unique}
   For all functions $f, g: \var{list}~A \to \var{list}~A$,
@@
 This in turn required developing suitable definitions for
-\var{Sorted}, \var{Permutation}, and \var{StablePerm}. We explored
-multiple definitions for each of these properties, and proved
-equivalence between them.
+\var{Sorted}, \var{Permutation}, and \var{StablePerm}.
 
@@
 what we need to know about a type $A$ to sort a $\var{list}~A$ as a
-part of BWT. Following Bird et~al., \cref{sec:forwards_BWT} defines
-the implementation of the forwards BWT in Coq, and
-\cref{sec:inverse_BWT} derives an implementation of the inverse BWT
-via program calculation. \Cref{sec:sorted_perm_stable} provides
+part of BWT. \Cref{sec:forwards_BWT} implements the forwards BWT in
+Coq. \Cref{sec:inverse_BWT} implements the inverse BWT and proves
+\Cref{thm:unbwt_correct}. \Cref{sec:sorted_perm_stable} provides
 multiple definitions of \var{Sorted}, \var{Permutation}, and
 \var{StablePermutation}, and proves equivalence between them.
-\Cref{sec:sort} implements insertion sort and radixsort, and proves
-their correctness.
-
-The full source code can be found at
+\Cref{sec:unique} proves \Cref{thm:stablesort_unique}. \Cref{sec:ord}
+defines lexicographic orderings, need to sort the rotation matrix.
+\Cref{sec:sort} radixsort and proves \Cref{thm:radixsort_correct}.
+\var{sec:opt} explains the optimization that is most commonly applied
+to BWT that relates our implementation to the standard version of the
+algorithm.
+
+Every definition and theorem in this paper corresponds directly to a
+definition or theorem in our Coq sources, which can be found at
 \url{https://github.com/jbaum98/verifed_bzip2/tree/2019-05-06}.
+Definitions are copied verbatim. Theorems are rephrased for clarity,
+with the name of the corresponding Coq theorem given in parentheses.
+The only exceptions are \var{sec:opt} and places where we give
+hypothetical definitions to provide an explanation.
 
@@
 The Burrows--Wheeler transform was originally introduced as a pass in a
-compression algorithm. First the input is transformed with BWT, which
-tends to group together characters. The intuition for why this happens
-is that all adjacent characters in the output string are followed in
-the input string by characters that are adjacent in the sorted first
-column. If there is any correlation between a character and the
-character preceding it, bringing together the characters in the first
-column will tend to bring together similar characters in the last
-column.
+compression algorithm, and continues to be used as a part of the
+popular, open-source bzip2 compression software \cite{tsai_2016}.
+
+Here we describe a standard compression algorithm built around BWT.
+First the input is transformed with BWT, which tends to group together
+characters. Intuitively, the last column of the rotation matrix
+contains the characters that directly preced the characters in the
+first column of the rotation matrix. Sorting lexicographically, which
+completely sorts the first column. If there is a correlation between a
+character and the character preceding it, bringing together identical
+characters in the first column will tend to bring together identical
+characters in the last column.
 
@@
 \cite{ferragina_index}. These techniques have been applied to enable
-fast matching of DNA subsequences.
+fast matching of DNA subsequences \cite{dna}.
 
@@
 \end{equation}
+where $f^n$ denotes repeated function application.
 \lstinputlisting[
@@
 
-We can prove that this definition satisfies \Cref{eq:iter} by defining
-the function \var{rep} for repeated function application.
-\lstinputlisting[
-  widthgobble=2,
-  firstline=8,
-  lastline=12
+We can state \Cref{eq:iter} as a theorem by defining the
+function \var{rep} for repeated function application.
+\lstinputlisting[
+  widthgobble=2, firstline=8, lastline=12,
 ]{../theories/BWT/Lib/Repeat.v}
-In proofs, we will still use the notation $f^i z$ for $\var{rep}~f~i~z$.
-
-\begin{theorem}[rep\_r]
-  For all $f$, $n$, and $z$,
-  \begin{equation}
-     f^i~(f~z) = f^{i+1}~z
-  \end{equation}
-\end{theorem}
-\begin{proof}
-  The theorem follows by induction on $n$.
-\end{proof}
-
-\begin{theorem}[iter\_length]
-  For all $f$, $n$, and $z$,
-  \begin{equation}
-    \var{length}~(\var{iter}~f~n~z) = n
-  \end{equation}
-\end{theorem}
-\begin{proof}
-  The theorem follows by induction on $n$.
-\end{proof}
+In proofs, we will still use the notation $f^i z$ for
+$\var{rep}~f~i~z$. We will also use that notation $l_i$ for the $i$th
+element of a list. In Coq, that is given by $\var{nth}~i~l~d$, where
+$d:A$ is a default value in case $i \ge \var{length}~l$.
 
@@
   For all $f$, $n$, $z$, and $i < n$,
-  \begin{equation}
+  \begin{equation*}
     (\var{iter}~f~n~z)_i = f^i~z
-  \end{equation}
+  \end{equation*}
 \end{theorem}
@@
   \begin{align*}
-       (\var{iter}~f~(n+1)~z)_{i+1}
-    &= (z :: \var{iter}~f~n~(f~z))_{i+1} &&\{\text{\var{iter} def.}\} \\
-    &= (\var{iter}~f~n~(f~z))_i          &&\{\text{by \var{nth} def.}\} \\
-    &= f^i~(f~z)                            &&\{\text{by \var{rep\_r}}\} \\
-    &= f^{i+1}~z && \qedhere
+       (\var{iter}~f~(n+1)~z)_{i+1} \\
+    &= (z :: \var{iter}~f~n~(f~z))_{i+1} \\
+    &= (\var{iter}~f~n~(f~z))_i \\
+    &= f^i~(f~z) \\
+    &= f^{i+1}~z \qedhere
   \end{align*}
@@
 
-\var{bwl} produces the transformed string by taking the last column
-of the sorted rotation matrix. \var{lexsort} sorts the rotation
-matrix lexicographically using the \var{Ord} instance; we will
-discuss its implementation in \cref{sec:sort}.
+\var{bwl} produces the transformed string by taking the last column of
+the sorted rotation matrix. \var{lexsort} sorts the rotation matrix
+using insertion sort (\cref{appendix:insertion_sort}) and the induced
+lexicographic ordering(\cref{sec:lex_ord}). By
+\Cref{thm:stablesort_ord} though, \var{lexsort} can be implemented
+using any sorting algorithm.
 \lstinputlisting[
@@
   widthgobble=2,
-  firstline=73,
-  lastline=74,
+  firstline=75,
+  lastline=75,
 ]{../theories/BWT/BurrowsWheeler.v}
@@
 Given these definitions, we can quickly prove that \var{bwn}
-satisfies its specification, namely that for all non-empty $l$
-\begin{equation*}
-    (\var{lexsort}~(\var{rots}~l))_{(\var{bwn}~l)} = l
-\end{equation*}
-
+satisfies its specification:
+\begin{restatable*}[bwn\_correct]{theorem}{bwncorrect}
+  For every non-empty list $xs$,
+  \begin{equation*}
+    (\var{lexsort}~(\var{rots}~xs))_{(\var{bwn}~xs)} = xs
+  \end{equation*}
+\end{restatable*}
+
+First we prove a few lemmas:
 \begin{lemma}[findIndex\_correct]
@@
   \end{equation*}
+\end{lemma}
 \begin{proof}
@@
 \end{proof}
-\end{lemma}
 
@@
 \end{proof}
-\end{lemma}
-
-\begin{theorem}[bwn\_correct]
-  For every non-empty list $xs$,
-  \begin{equation*}
-    (\var{lexsort}~(\var{rots}~xs))_{(\var{bwn}~xs)} = xs
-  \end{equation*}
+
+\bwncorrect
 \begin{proof}
@@
 \end{proof}
-\end{theorem}
 
@@
 \begin{equation}
+  \label{eq:recreate}
   \var{recreate}~(\var{bwl}~l) = \var{lexsort}~(\var{rots}~l)
-  \label{eq:recreate}
 \end{equation}
@@
 \var{lexsort} is implemented using an sorting algorithm we find
-convenient. Specifically, if we examine how an LSD radixsort sorts the
-rotation matrix (\Cref{fig:radixsort_rots}) we notice an interesting
-pattern. Each column of the rotation matrix is a permutation of the
-original string, so when we sort a on particular column, we always
-produce the same result in that column. As we move right-to-left, the
-column containing the sorted input string moves with us until it
-reaches the left edge of the matrix.
+convenient (\Cref{thm:sort_ord_unique}). Specifically, if we examine
+how an LSD radixsort sorts the rotation matrix
+(\Cref{fig:radixsort_rots}) we notice an interesting pattern. Each
+column of the rotation matrix is a permutation of the original string,
+so when we sort a on particular column, we always produce the same
+result in that column. As we move right-to-left, the column containing
+the sorted input string moves with us until it reaches the left edge
+of the matrix.
 
@@
 
-All this means that the columns to the left of the sorted column seem
-like they don't matter, which suggests that we can recreate the entire
-rotation matrix from just the last column (\Cref{fig:recreate}).
-However, the column directly to the left of the sorted column is very
-important, as it is the next column to be sorted. And we can see that
-this column changes, only becoming the eventual last column of the
-matrix in the second to last iteration. Why then, does it still work
-if we use the last column everywhere?
-
-We can think of the process of sorting the matrix as computing the
-last column subject to the constraints of the lexicographic ordering.
-For example, we know that the collection of all characters preceding
-an \texttt{a} in the sorted column must be some permutation of
-$[\texttt{c},~\texttt{d},~\texttt{r},~\texttt{r},~\texttt{!}]$,
-because those are the characters that precede \texttt{a}s in the
-original string. However, the order of these characters in the final
-string
-($[\texttt{r},~\texttt{d},~\texttt{!},~\texttt{r},~\texttt{c}]$) is
-important. An \texttt{r} must be first, because in the original string
-there is an \texttt{a} following an \texttt{r} which is itself
-followed by a \texttt{!}, the smallest character. Similarly, although
-the \texttt{d} and \texttt{!} precede \texttt{a}s that are both
-followed by \texttt{bra}s, the next character for the \texttt{d} is a
-\texttt{!}, and the next character for the \texttt{!} is a \texttt{c}.
-This requires that the \texttt{d} precede the \texttt{!} in the last
-column.
-
-Generally, there are constraints on the relative order of the letters
-in the column before the sorted column, and in each step we are
-ensuring that these constraints are not violated one more character
-deep in the string. By the time we have reached the last column, we
-have satisfied all the constraints.
-
-We can also think of this as computing the fixpoint of $\var{hdsort} \comp
-\var{map}~\var{rrot}$: we continually move the rightmost column to the
-front and sort on it, until we have sorted the whole matrix and this
-becomes a no-op. The fact that the last column corresponds to the
-fixpoint justifies why in \cref{fig:recreate} we can replace every
-column left of the sorted column with the output of the BWT.
+This suggests that we can, as we had hoped, recreate the matrix from a
+single column (\Cref{fig:recreate}). But why specifically the last
+column?
 
@@
 
+The reason is that the last column is the solution to the fixpoint equation
+\begin{equation*}
+  \var{hdsort}~(\var{prefix\_shift}~c~m) = m
+\end{equation*}
+where \var{hdsort} sorts on the first column of a matrix,
+$\var{prefix\_shift}~c$ adds $c$ as the first column of a matrix,
+dropping the last column, and $m$ is the rotation matrix. This means
+as we repeatedly prefix $l$ and sort, we preserve the ``rest of the
+matrix'', even the parts that we have yet to reconstruct.
+
+We can think of the process of sorting the matrix as computing the
+solution to the fixpoint equation, subject to the constraints of the
+lexicographic ordering. For example, we know that the collection of
+all characters preceding an \texttt{a} in the sorted column must be
+some permutation of
+$[\texttt{c},~\texttt{d},~\texttt{r},~\texttt{r},~\texttt{!}]$,
+because those are the characters that precede \texttt{a}s in the
+original string. However, the order of these characters in the final
+string
+($[\texttt{r},~\texttt{d},~\texttt{!},~\texttt{r},~\texttt{c}]$) is
+important. An \texttt{r} must be first, because in the original string
+there is an \texttt{a} following an \texttt{r} which is itself
+followed by a \texttt{!}, the smallest character. Similarly, although
+the \texttt{d} and \texttt{!} both precede \texttt{a}s that are
+followed by \texttt{bra}s, the next character for the \texttt{d} is a
+\texttt{!}, and the next character for the \texttt{!} is a \texttt{c}.
+This requires that the \texttt{d} precede the \texttt{!} in the last
+column.
+
+Generally, there are constraints on the relative order of the letters
+in the column before the sorted column, and in each step we are
+ensuring that these constraints are not violated one more character
+deep in the string. By the time we have reached the last column, we
+have satisfied all the constraints.
+
 We implement \var{recreate} as follows, introducing a parameter $j$ so
@@
   widthgobble=2,
-  firstline=196,
-  lastline=197,
+  firstline=197,
+  lastline=198,
 ]{../theories/BWT/BurrowsWheeler.v}
@@
 
-\begin{theorem}[recreate\_correct\_inv]
+\begin{restatable*}[recreate\_correct\_inv]{theorem}{recreatecorrectinv}
   For all lists $l$ and $j \le \var{length}~l$,
@@
   \end{equation*}
-\end{theorem}
+\end{restatable*}
 
@@
 
-Now we can prove that the sorted rotation matrix is indeed a fixpoint.
+Now we can prove that the last column is indeed a fixpoint.
 \begin{theorem}[lexsort\_rots\_hdsort]
@@
 \begin{proof}
-  By \var{StablePerm\_Sorted\_eq}, proved in \cref{sec:unique}, any
-  two sorts are equivalent, and we can replace \var{lexsort} with an
-  LSD radixsort implemented as
+  By \Cref{thm:sort_ord_unique}, any two sorts are equivalent, and we
+  can replace \var{lexsort} with an LSD radixsort implemented as
   \begin{equation*}
@@
   The last step follows because the function $\var{hdsort} \comp
-  \var{rrot}$ permutes the rotation matrix
+  \var{rrot}$ permutes the rotation matrix.
 \end{proof}
@@
   \end{align*}
-  The bottom equation is true by another theorem we will not prove here.
+  The bottom equation is true because $j > \var{length}~r$, so if $j=
+  0$, $r = []$.
 \end{proof}
@@
 Now we are ready to prove the invariant:
-\begin{theorem}[recreate\_correct\_inv]
-  For all lists $l$ and $j \le \var{length}~l$,
-  \begin{equation*}
-    \var{recreate}~j~(\var{bwl}~l) =
-    \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
-  \end{equation*}
-\end{theorem}
+\recreatecorrectinv
 \begin{proof}
@@
 
+\begin{theorem}[cols\_id]
+  For all matrices $m$ such that all the rows of $m$ have length less
+  than or equal to $n$,
+  \begin{equation*}
+    \var{cols}~n~m = m
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  This follows by induction on $m$.
+\end{proof}
+
+\unbwtcorrect
+\begin{proof}
+\begin{align*}
+     \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l)
+  &= (\var{recreate}~(\var{length}~l))(\var{bwl}~l))_{(\var{bwn})~l}\\
+  &= (\var{cols}~(\var{length}~l)~(\var{lexsort}~(\var{rots}~l)))_{(\var{bwn})~l} \\
+  &= (\var{lexsort}~(\var{rots}~l))_{(\var{bwn})~l} \\
+  &= l \qedhere
+\end{align*}
+\end{proof}
+
 \section{\var{Sorted}, \var{Permutation}, and \var{StablePermutation}}
@@
 \var{Sorted} and \var{Permutation}, but no definition for stability.
-The Coq standard library also includes an implementation of mergesort.
-However, it uses Coq's module system to enable parametricity over
-ordered types, which is much less convenient to use than typeclasses.
-In this section, we define our own version of \var{Sorted} and
-\var{StablePerm} that are based on our \var{Preord} typeclass.
+It also includes an implementation of mergesort, but it uses Coq's
+module system to enable parametricity over ordered types, which is
+much less convenient to use than typeclasses. In this section, we
+define our own version of \var{Sorted} and \var{StablePerm} that are
+based on our \var{Preord} typeclass.
 
@@
 
-Although these two inductive definitions are intuitive, they are not
-obviously equivalent to the first definition, which is more clearly
-correct. We will prove all three equivalent.
+These two inductive definitions are not obviously equivalent to the
+first definition, which is more clearly correct. We will prove all
+three equivalent.
 
@@
 
+The Coq standard library comes with a definition of \var{Permutation}
+that holds when two lists contain exactly the same elements:
+\begin{gather*}
+  \infer[\var{perm\_nil}]{\var{Permutation}~[]~[]}{}
+  \\
+  \\
+  \infer[\var{perm\_skip}]{\var{Permutation}~(x :: l)~(x :: l')}{%
+    \var{Permutation}~l~l'
+  }
+  \\
+  \\
+  \infer[\var{perm\_swap}]{\var{Permutation}~(x :: y :: l)~(y :: x :: l)}{}
+  \\
+  \\
+  \infer[\var{perm\_trans}]{\var{Permutation}~l~l''}{
+    \var{Permutation}~l~l' & \var{Permutation}~l'~l''
+  }
+\end{gather*}
+Although we have no need to redefine it, we present here some
+alternative definitions. This is mainly because these definitions have
+analagous definitions for \var{StablePerm}, which we are definining
+ourselves.
+
 The word \textit{permutation} has two related meanings that are often
@@
 \lstinputlisting[
-  widthgobble=2, firstline=510, lastline=511,
+  widthgobble=2, firstline=521, lastline=522,
 ]{../theories/BWT/Sorting/PermFun.v}
 
-As with \var{Sorted}, we would also like an inductive definition of
-\var{Permutation}. The Coq standard library includes one:
-\begin{align*}
-  \infer[\var{perm\_nil}]{\var{Permutation}~[]~[]}{}
-  &&
-  \infer[\var{perm\_skip}]{\var{Permutation}~(x :: l)~(x :: l')}{%
-    \var{Permutation}~l~l'
-  }
-  \\
-  \\
-  \infer[\var{perm\_swap}]{\var{Permutation}~(x :: y :: l)~(y :: x :: l)}{}
-  &&
-  \infer[\var{perm\_trans}]{\var{Permutation}~l~l''}{
-    \var{Permutation}~l~l' & \var{Permutation}~l'~l''
-  }
-\end{align*}
-
-We know we will eventually need to reason inductively about
-permutation functions, so we would like to define the relationship
-between applying a permutation function $i :: p$ and applying $p$.
-Conceptually, the first element of $\var{apply}~(i::p)~l$ will be
-$l_i$. The rest will be the result of applying some new permutation
-$p'$, where everything greater than $i$ has been shifted down by one,
-to some new list $l'$, which has its $i$th element removed. So, we
-define \var{rem\_PermFun} and \var{rem\_nth} respectively, hoping to
-prove that
-\begin{equation*}
-  \var{apply}~(i::p)~l =
-  l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l)
-\end{equation*}
-\lstinputlisting[
-  widthgobble=4, firstline=229, lastline=229,
-]{../theories/BWT/Sorting/PermFun.v}
-\vspace{\parskip}
-\lstinputlisting[
-  widthgobble=2, firstline=599, lastline=607,
-]{../theories/BWT/Lib/List.v}
-
-These two theorems define how \var{rem\_nth} affects indexing.
-\begin{theorem}[nth\_lt\_rem\_nth, nth\_ge\_rem\_nth]
-  For all lists $l$ and indices $i$ and $j$,
-  \begin{equation*}
-    j < i \implies (\var{rem\_nth}~i~l)_j = l_j
-  \end{equation*}
-  and
-  \begin{equation*}
-    j \ge i \implies (\var{rem\_nth}~i~l)_j = l_{j+1}
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  Both theorems follow by induction on $i$.
-\end{proof}
-
-As we expect, removing the $n$th element and sticking it on the front
-preserves all elements.
-\begin{theorem}[rem\_nth\_Perm]
-  For all list $l$ and indices $i$
-  \begin{equation*}
-    \var{Permutation}~l~(l_i :: \var{rem\_nth}~i~l)
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  This follows from induction on $l$.
-\end{proof}
-
-\begin{theorem}[rem\_PermFun\_correct]
-  For all permutation functions $p$, lists $l$, and indices $i$.
-  \begin{equation*}
-    \var{apply}~(i::p)~l =
-    l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l)
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  Working backwards from our goal,
-  \begin{align*}
-    \var{apply}~(i::p)~l &=
-    l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l) \\
-    l_i :: \var{map}~(\lambda j.~l_j)~p &=
-    l_i :: \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_j)~(\var{rem\_PermFun}~i~p) \\
-    \var{map}~(\lambda j.~l_j)~p &=
-    \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_j)~(\var{map}~(\lambda j. \bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j)~p) \\
-    \var{map}~(\lambda j.~l_j)~p &=
-    \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_{\bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j})~p \\
-  \end{align*}
-  From here, we need to prove that the mapping functions agree on all
-  the elements of $p$, namely that for all $j \in p$
-  \begin{equation*}
-    l_j = (\var{rem\_nth}~i~l)_{\bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j}
-  \end{equation*}
-  We know that $i \neq j$, because $j \in p$ and $\var{NoDup}~(i :: p)$.
-  Therefore there are two cases, $i < j$ and $i > j$. When $i < j$, we have
-  \begin{equation*}
-    l_j = (\var{rem\_nth}~i~l)_{j-1}
-  \end{equation*}
-  by \var{nth\_ge\_rem\_nth}, and when $i > j$, we have
-  \begin{equation*}
-    l_j = (\var{rem\_nth}~i~l)_j
-  \end{equation*}
-  by \var{nth\_lt\_rem\_nth}.
-\end{proof}
-
-\begin{theorem}[apply\_correct]
-  For all permutation functions $p$ and lists $l$,
-  \begin{equation*}
-    \var{Permutation}~(\var{apply}~p~l)~l
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  First we let $n = \var{length}~l$, so that we can reason by
-  induction on $n$. The base case is uninteresting. In the inductive
-  case, we have the induction hypothesis that for $p :
-  \var{list}~\mathbb{N}$ and $l : \var{list}~A$
-  \begin{equation*}
-    \var{lenth}~l = n \implies \var{PermFun}~n~p \implies
-    \var{Permutation}~(\var{apply}~p~l)~l
-  \end{equation*}
-  Given $\var{PermFun}~(n+1)~(i::p')$, we must show that
-  \begin{equation*}
-    \var{Permutation}~(\var{apply}~(i :: p')~(a :: l))~(a :: l)
-  \end{equation*}
-
-  We reason as follows, using the notation $\eqperm$ to
-  chain together \var{Permutation}s:
-  \begin{align*}
-    &\var{apply}~(i :: p')~(a :: l) \\
-    \eqperm{}& \text{\{ by \var{rem\_PermFun\_correct} \}} \\
-    &(a :: l)_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a::l))
-    \\
-    \eqperm{}& \text{\{ by IH \}} \\
-    &(a :: l)_i :: \var{rem\_nth}~i~(a::l)
-    \\
-    \eqperm{}& \text{\{ \var{rem\_nth\_Perm} \}} \\
-    & a :: l \qedhere
-  \end{align*}
-\end{proof}
-
-\begin{theorem}[PermutationEx\_iff]
-  For all lists $l$ and $l'$,
-  \begin{equation*}
-    \var{Permutation}~l~l' \iff \var{PermutationEx}~l~l'
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  The $\impliedby$ direction follows from \var{apply\_correct}.
-
-  ($\implies$) We reason by induction on the evidence of
-  $\var{Permutation}~l~l'$. The empty case is simple.
-  \begin{itemize}
-  \item We have a permutation function $p$ from the induction
-    hypothesis such that $\var{apply}~p~l = l'$, and we need to
-    construct a permutation function $p'$ such that
-    $\var{apply}~p'~(x::l) = x :: l'$. We let $p' = 0 ::
-    \var{map}~(+1)~p$, shifting every index up by one and mapping
-    the first element of the list to itself.
-  \item We have to construct a permutation function $p$ such that
-    $\var{apply}~p~(y :: x :: l) = x :: y :: l$. We let $p = 1 :: 0 ::
-    \var{map}~(+2)~(\var{seq}~0~(\var{length}~l))$, leaving every
-    element in place except for the first 2.
-  \item We have permutation functions $p_1$ and $p_2$ from our two
-    induction hypotheses, such that
-    \begin{align*}
-      \var{apply}~p_1~l = l' && \var{apply}~p_2~l' = l''
-    \end{align*}
-    We need to construct a $p'$ such that $\var{apply}~p'~l=l''$, so
-    we let $p' = p_2 \comp p_1$. For the definition of composition of
-    permutations, see \cref{appendix:perm_comp}. Then,
-    \begin{equation*}
-      \var{apply}~(p_2 \comp p_1)~l =
-      \var{apply}~p_2~(\var{apply}~p_1~l) = l'' \qedhere
-    \end{equation*}
-  \end{itemize}
-\end{proof}
+We prove \var{PermutationEx} equivalent to \var{Permutation} in
+\Cref{thm:permutationex_iff}.
 
@@
 characters in that column, but it is crucial that it respect the
-initial ordering in cases of a tie.
+existing ordering in cases of a tie.
 
@@
 so we say $\var{StablePermFun}~l~p$ for some list $l$.
-\lstinputlisting[ widthgobble=2, firstline=557, lastline=561,
-]{../theories/BWT/Sorting/PermFun.v} As above, we can also define this
-property using the inverse permutation: \lstinputlisting[
-  widthgobble=2, firstline=563, lastline=567,
-]{../theories/BWT/Sorting/PermFun.v} As usual, we will prove these
+\lstinputlisting[
+  widthgobble=2, firstline=568, lastline=572,
+]{../theories/BWT/Sorting/PermFun.v}
+As above, we can also define this
+property using the inverse permutation:
+\lstinputlisting[
+  widthgobble=2, firstline=574, lastline=578,
+]{../theories/BWT/Sorting/PermFun.v}
+As usual, we will prove these
 definitions equivalent. First, though, we need to define \var{image}
-and \var{preimage}. \lstinputlisting[ widthgobble=2, firstline=85,
-  lastline=86, ]{../theories/BWT/Sorting/PermFun.v} We will not prove
+and \var{preimage}.
+\lstinputlisting[
+  widthgobble=2, firstline=81, lastline=82,
+]{../theories/BWT/Sorting/PermFun.v}
+We will not prove
 it here, but $\var{preimage}~p$ and $\var{image}~p$ are both of
@@
 \lstinputlisting[
-  widthgobble=2, firstline=607, lastline=608,
+  widthgobble=2, firstline=618, lastline=619,
 ]{../theories/BWT/Sorting/PermFun.v}
@@
 the same order, since the original lists are specifically stable
-permutations of one another.
+permutations of one another \cite{leroy_2018}.
 \lstinputlisting[
@@
 
-We will prove equivalence between these three definitions, but first
-we will mention some properties of \var{Stable}. \var{Stable} is an
-equivalence relation, meaning it is reflexive, transitive, and symmetric.
-
-\begin{lemma}[StablePerm\_app]
-  For all lists $l$, $l'$, $m$ and $m'$,
-  \begin{equation*}
-    \var{StablePerm}~l~l' \implies \var{StablePerm}~m~m' \implies
-    \var{StablePerm}~(l \dplus m)~(l' \dplus m')
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  This follows from the distributivity of \var{filter} over \var{app}.
-\end{proof}
-
-\begin{lemma}[StablePerm\_skip]
-  For all lists $l$ and $l'$, and elements $a$,
-  \begin{equation*}
-    \var{StablePerm}~l~l' \implies \var{StablePerm}~(a::l)~(a::l')
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  For any $x$, we have to show $\var{filter}~(\equiv x)~(a::l) =
-  \var{filter}~(\equiv x)~(a::l')$. If $x \equiv a$, $a$ is added to both lists,
-  and if $x \not\equiv a$ it is added to neither.
-\end{proof}
-
-\begin{lemma}[StablePerm\_swap]
-  For all lists $l$ and $l'$, and elements $a$ and $b$ where $a \not\equiv b$,
-  \begin{equation*}
-     \var{StablePerm}~(a::b::l)~(b::a::l)
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  For any $x$, we have to show $\var{filter}~(\equiv x)~(a::b::l) =
-  \var{filter}~(\equiv x)~(b::a::l)$. The proof follows from case analysis
-  on $x \equiv a$ and $x \equiv b$.
-\end{proof}
-
-\begin{lemma}[StablePerm\_cons\_app]
-  For all lists $l$, $l_1$ and $l_2$, and elements $a$ such that $\forall b
-  \in l_1,~a \not\equiv b$,
-  \begin{equation*}
-     \var{StablePerm}~l~(l_1 \dplus l_2) \implies \var{StablePerm}~(a
-     :: l)~(l_1 \dplus a :: l_2)
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  For any $x$, we have to show
-  \begin{equation*}
-    \var{filter}~(\equiv x)~(a::l) = \var{filter}~(\equiv x)~(l_1 \dplus a :: l_2)
-  \end{equation*}
-  If $x \not\equiv a$ the proof is easy, so assume $x \equiv a$. Then we have
-  \begin{align*}
-      \var{filter}~(\equiv x)~(l_1 \dplus a :: l_2)
-    &= \var{filter}~(\equiv x)~l_1 \dplus \var{filter}~(a :: l_2) \\
-    &= [] \dplus a::\var{filter}~l_2 \\
-    &= a::\var{filter}~l
-  \end{align*}
-  where we know that $\var{filter}~(\equiv x)~l_1 = []$ because $\forall b \in
-  l_1,~a \not\equiv b$, and the last step follows from
-  $\var{StablePerm}~l~(l_1 \dplus l_2)$.
-\end{proof}
-
-\begin{theorem}[StablePerm\_destr]
-  For all lists $l$ and $l'$, and elements $h$ and $h'$ such that $h
-  \not\equiv h'$,
-  \begin{gather*}
-    \var{StablePerm}~(h :: l)~(h' :: l') \implies \\
-    \exists~l_1~l_2, ~l = l_1 \dplus h' :: l_2 \land (\forall x \in l_1, h' \not\equiv x)
-  \end{gather*}
-\end{theorem}
-\begin{proof}
-  Let $l_1 = \var{take\_while}~(\not\equiv h')~l$ and $l_2 =
-  \var{drop\_while}~(\not\equiv h')~l$. Clearly $l = l_1 \dplus l_2$, and
-  because $\var{StablePerm}~(h :: l)~(h' :: l')$, we know the first
-  element of $l_2$ must be $h'$. Let $l_2'$ be the rest of $l_2$, and
-  we have two lists $l_1$ and $l_2'$ that satisfy the theorem.
-\end{proof}
-
-\begin{theorem}[StablePerm\_length]
-  For all lists $l$ and $l'$
-  \begin{equation*}
-    \var{StablePerm}~l~l' \implies \var{length}~l = \var{length}~l'
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  We reason by induction on $n = \var{length}~l$. In the inductive
-  case, we need to show that $\var{length}~(h::t) =
-  \var{length}~(h'::t')$ given that $\var{StablePerm}~(h::t)~(h'::t')$
-  and the induction hypothesis that for all lists of length $n$,
-  \var{StablePerm} implies their lengths are equal.
-
-  If $h \equiv h'$, then from $\var{StablePerm}~(h::t)~(h'::t')$ we know
-  that $h = h'$. We can then remove $h$ from the front of $h::t$ and
-  $h::t'$ and apply the induction hypothesis.
-
-  If $h \not\equiv h'$, we can destructure $t$ and $t'$ as $l_1 \dplus h'
-  :: l_2$ and $l_1' \dplus h :: l_2'$ respectively, using
-  \var{StablePerm\_destr}. Then,
-  \begin{equation*}
-      \var{length}~t
-    = \var{length}~(h' :: l_1  \dplus l_2)
-    = \var{length}~(h  :: l_1' \dplus l_2')
-    =\var{length}~t'
-  \end{equation*}
-  where the first and last steps follow from the induction hypothesis,
-  using the fact that $\var{StablePerm}~t~(h'::l_1 \dplus l_2)$ and
-  $\var{StablePerm}~t'~(h::l_1' \dplus l_2')$.
-\end{proof}
-
-\begin{theorem}[StablePerm\_Perm]
-  For all lists $l$ and $l'$,
-  \begin{equation*}
-    \var{StablePerm}~l~l' \implies \var{Permutation}~l~l'
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  We reason by strong induction on the length of both lists (we know
-  the lengths are equal by \var{StablePerm\_length}). In the inductive
-  case, we need to show $\var{Permutation}~(h :: t)~(h' :: t')$. If $h
-  \equiv h$, we know that $h = h'$, and the proof follows by
-  \var{perm\_skip}.
-
-  If $h \not\equiv$, we again destructure $t$ and $t'$ as $l_1 \dplus h' ::
-  l_2$ and $l_1' \dplus h :: l_2'$ respectively, using
-  \var{StablePerm\_destr}.
-  \begin{align*}
-    h::t ={}& h :: l_1 \dplus h' :: l_2 \\
-    \eqperm{}& h :: h' :: l_1 \dplus l_2 \\
-    \eqperm{}& h' :: h :: l_1 \dplus l_2 \\
-    \eqperm{}& h' :: h :: l_1' \dplus l_2' \\
-    \eqperm{}& h' :: l_1' \dplus h :: l_2' \\
-    ={}& h'::t
-  \end{align*}
-  The crucial step is $l_1 \dplus l_2 \eqperm l_1' \dplus l_2'$, which
-  follows from the induction hypothesis.
-\end{proof}
-
-\begin{theorem}[StablePermInd\_iff]
-  For all lists $l$ and $l'$
-  \begin{equation*}
-    \var{StablePerm}~l~l' \iff \var{StablePermInd}~l~l'
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  The ($\impliedby$) direction follows easily by induction on
-  $\var{StablePermInd}~l~l'$, using \var{StablePerm\_skip},
-  \var{StablePerm\_swap}, and \var{StablePerm\_trans}.
-
-  The ($\implies$) direction is proved very similarly to
-  \var{StablePerm\_Perm}, using \var{StablePerm\_destr} to split up
-  the lists.
-\end{proof}
-
-Now we prove equivalence with \var{StablePermEx}.
-
-\begin{theorem}[apply\_correct\_stable]
-  For all lists $l$ and permutation functions $p$,
-  \begin{equation*}
-    \var{StablePermFun}~l~p \implies \var{StablePerm}~(\var{apply}~p~l)~l
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  We reason by induction on the length of $l$. In the inductive case,
-  we need to show
-  \begin{equation*}
-    \var{StablePerm}~(\var{apply}~(i :: p)~(a :: l))~(a :: l)
-  \end{equation*}
-
-  Using $\eqstable$ to chain \var{StablePerm}s transitively,
-  \begin{align*}
-    \var{apply}~(i :: p)~(a :: l)
-    ={}& (a :: l)_i  ::
-    \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a :: l)) \\
-    \eqstable{}& (a :: l)_i :: \var{rem\_nth}~i~(a :: l) \\
-    \eqstable{}& (a :: l)_i
-  \end{align*}
-  where we rely on the induction hypothesis between the first and
-  second lines.
-\end{proof}
-
-\begin{theorem}[StablePermEx\_iff]
-  For all lists $l$ and $l'$
-  \begin{equation*}
-    \var{StablePerm}~l~l' \iff \var{StablePermEx}~l~l'
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  We use \var{StablePermInd\_iff} and prove $\var{StablePermInd}~l~l'
-  \iff \var{StablePermEx}~l~l'$.
-
-  The $\implies$ direction is very similar to the $\implies$ direction
-  of the \var{PermutationEx\_iff} proof: we reason by induction on
-  $\var{StablePermInd}~l~l'$, constructing the same permutation
-  functions in each case, but this time we have to prove that it is
-  stable as well. The $\impliedby$ direction follows from
-  \var{apply\_correct\_stable}.
-\end{proof}
+We prove equivalence between these three definitions in
+\cref{appendix:stableperm_def_eq}.
 
@@
 
+\stablesortunique
 \begin{proof}
@@
 \begin{corollary}[Sort\_StableSort\_Ord]
+  \label{thm:stablesort_ord}
   When sorting using an order,
@@
 
-\stablesortunique
-\begin{proof}
-  This follows from \var{Sort\_StableSort\_Ord} and \var{StableSort\_unique}.
+\begin{theorem}[Sort\_Ord\_unique]
+  \label{thm:sort_ord_unique}
+  When sorting using an order, given any two function $f, g:
+  \var{list}~A \to \var{list}~A$,
+  \begin{equation*}
+    \var{Sort}~f \implies \var{Sort}~g \implies
+    (\forall l,~f~l = g~l)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  This follows from \Cref{thm:stablesort_ord} and \Cref{thm:stablesort_unique}.
 \end{proof}
@@
 \lstinputlisting[
-  widthgobble=2, firstline=323, lastline=324,
+  widthgobble=2, firstline=326, lastline=327,
 ]{../theories/BWT/Sorting/Key.v}
@@
 offer an explanation of why these theorems makes sense, but these
-explanations do not mirror the Coq proofs we have for these theorems.
+explanations do not mirror the Coq proofs we have for these theorems,
+which are hard to follow.
 
@@
 \lstinputlisting[
-  widthgobble=2, firstline=26, lastline=27,
+  widthgobble=2, firstline=27, lastline=28,
 ]{../theories/BWT/Sorting/RadixSort.v}
-Here \var{hdsort} is just an insertion sort that uses a lexicographic
-ordering restricted to the first element of a list.
+Here \var{hdsort} is just an insertion sort using $\le_1$ ordering.
 
@@
 \cref{appendix:hdsort_sorted_S}.
-\begin{theorem}[hdsort\_sorted\_S]
+\begin{restatable*}[hdsort\_sorted\_S]{theorem}{hdsortsortedS}
   For all matrices $m$ and naturals $j$,
@@
 
+\radixsortcorrect
+\begin{proof}
+  This follows from \var{radixsort\_sorted} and \var{radixsort\_stable}.
+\end{proof}
+
 \section{Relationship to the standard BWT}
@@
 sorted column is the same. Therefore, we can calculate that
-permutation function once and store it in a variable $p$. The
-$i$th column of the recreated matrix is simply
-$(\var{apply}~p)^i~(\var{bwl}~l)$, so we can define
+permutation function once and store it in a variable $p$. The $i$th
+column of the recreated matrix is simply $(\var{apply}~p)^i~f$ where
+$f$ is the first column of the sorted rotation matrix, so we can
+define
 \begin{lstlisting}
@@
 \end{lstlisting}
-In the definition above, we are repeatedly applying \var{p} to the
+In the definition above, we are repeatedly applying $p$ to the
 first column of the rotation matrix, $\var{apply}~p~l$. Then we
@@
 Now, instead of indexing into each column at $i$, where the columns
-are all just the result of successively applying $p$ to $l$, we will
+are all the result of successively applying $p$ to $l$, we will
 successively take the image of $i$ under $p$ to generate the indices
@@
 \section{Conclusions}
-We have implemented and proved that the the inverse BWT actually
-inverts the forwards BWT, basing our work on Bird et al.'s program
-calculation derivation \cite{birdmu,pearls}. In the process, we
-defined the properties of stable sorts, and implemented and proved
-correct an LSD radixsort.
+We have implemented and proved that the the inverse BWT inverts the
+forwards BWT, basing our work on Bird et al.'s program calculation
+derivation \cite{birdmu,pearls}. In the process, we defined the
+properties of stable sorts, and implemented and proved correct an LSD
+radixsort.
 
@@
 arithmetic coding. As luck would have it, Bird has already analyzed
-arithmetic coding \cite{pearls}.
+arithmetic coding \cite{pearls}, and the other two phases are
+relatively simple.
 \clearpage
@@
 that it uses insertion sort as the inner sort, but
-\var{StableSort\_unique} allows us to extend that proof to a radixsort
-that uses any stable sort. We also use it as the implementation of
-\var{lexsort} because it is convenient, but similarly
-\var{Sort\_Ord\_unique} ensures that it can be replaced with any other
-sort.
+\Cref{thm:stablesort_unique} allows us to extend that proof to a
+radixsort that uses any stable sort. We also use it as the
+implementation of \var{lexsort} because it is convenient, but
+similarly \Cref{thm:sort_ord_unique} ensures that it can be replaced with
+any other sort.
 
@@
 
-In this section, we prove the following theorem:
-\begin{theorem}[hdsort\_sorted\_S]
-  For all matrices $m$ and naturals $j$,
-  \begin{equation*}
-    \var{PrefixSorted}~j~(\var{map}~\var{tl}~m) \implies
-    \var{PrefixSorted}~(j+1)~(\var{hdsort}~m)
-  \end{equation*}
-\end{theorem}
-
-First we prove this lemma:
 \begin{lemma}[insert\_Sorted\_S]
@@
 
+\hdsortsortedS
+\begin{proof}
+  This follows by induction from \var{insert\_Sorted\_S}
+\end{proof}
+
+\section{\var{Permutation} definitions equivalence}
+\label{appendix:perm_def_eq}
+
+We know we will eventually need to reason inductively about
+permutation functions, so we would like to define the relationship
+between applying a permutation function $i :: p$ and applying $p$.
+Conceptually, the first element of $\var{apply}~(i::p)~l$ will be
+$l_i$. The rest will be the result of applying some new permutation
+$p'$, where everything greater than $i$ has been shifted down by one,
+to some new list $l'$, which has its $i$th element removed. So, we
+define \var{rem\_PermFun} and \var{rem\_nth} respectively, hoping to
+prove that
+\begin{equation*}
+  \var{apply}~(i::p)~l =
+  l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l)
+\end{equation*}
+\lstinputlisting[
+  widthgobble=4, firstline=229, lastline=229,
+]{../theories/BWT/Sorting/PermFun.v}
+\vspace{\parskip}
+\lstinputlisting[
+  widthgobble=2, firstline=599, lastline=607,
+]{../theories/BWT/Lib/List.v}
+
+These two theorems define how \var{rem\_nth} affects indexing.
+\begin{theorem}[nth\_lt\_rem\_nth, nth\_ge\_rem\_nth]
+  For all lists $l$ and indices $i$ and $j$,
+  \begin{equation*}
+    j < i \implies (\var{rem\_nth}~i~l)_j = l_j
+  \end{equation*}
+  and
+  \begin{equation*}
+    j \ge i \implies (\var{rem\_nth}~i~l)_j = l_{j+1}
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  Both theorems follow by induction on $i$.
+\end{proof}
+
+As we expect, removing the $n$th element and sticking it on the front
+preserves all elements.
+\begin{theorem}[rem\_nth\_Perm]
+  For all list $l$ and indices $i$
+  \begin{equation*}
+    \var{Permutation}~l~(l_i :: \var{rem\_nth}~i~l)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  This follows from induction on $l$.
+\end{proof}
+
+\begin{theorem}[rem\_PermFun\_correct]
+  For all permutation functions $p$, lists $l$, and indices $i$.
+  \begin{equation*}
+    \var{apply}~(i::p)~l =
+    l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  Working backwards from our goal,
+  \begin{align*}
+    \var{apply}~(i::p)~l &=
+    l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l) \\
+    l_i :: \var{map}~(\lambda j.~l_j)~p &=
+    l_i :: \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_j)~(\var{rem\_PermFun}~i~p) \\
+    \var{map}~(\lambda j.~l_j)~p &=
+    \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_j)~(\var{map}~(\lambda j. \bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j)~p) \\
+    \var{map}~(\lambda j.~l_j)~p &=
+    \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_{\bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j})~p \\
+  \end{align*}
+  From here, we need to prove that the mapping functions agree on all
+  the elements of $p$, namely that for all $j \in p$
+  \begin{equation*}
+    l_j = (\var{rem\_nth}~i~l)_{\bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j}
+  \end{equation*}
+  We know that $i \neq j$, because $j \in p$ and $\var{NoDup}~(i :: p)$.
+  Therefore there are two cases, $i < j$ and $i > j$. When $i < j$, we have
+  \begin{equation*}
+    l_j = (\var{rem\_nth}~i~l)_{j-1}
+  \end{equation*}
+  by \var{nth\_ge\_rem\_nth}, and when $i > j$, we have
+  \begin{equation*}
+    l_j = (\var{rem\_nth}~i~l)_j
+  \end{equation*}
+  by \var{nth\_lt\_rem\_nth}.
+\end{proof}
+
+\begin{theorem}[apply\_correct]
+  For all permutation functions $p$ and lists $l$,
+  \begin{equation*}
+    \var{Permutation}~(\var{apply}~p~l)~l
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  First we let $n = \var{length}~l$, so that we can reason by
+  induction on $n$. The base case is uninteresting. In the inductive
+  case, we have the induction hypothesis that for $p :
+  \var{list}~\mathbb{N}$ and $l : \var{list}~A$
+  \begin{equation*}
+    \var{lenth}~l = n \implies \var{PermFun}~n~p \implies
+    \var{Permutation}~(\var{apply}~p~l)~l
+  \end{equation*}
+  Given $\var{PermFun}~(n+1)~(i::p')$, we must show that
+  \begin{equation*}
+    \var{Permutation}~(\var{apply}~(i :: p')~(a :: l))~(a :: l)
+  \end{equation*}
+
+  We reason as follows, using the notation $\eqperm$ to
+  chain together \var{Permutation}s:
+  \begin{align*}
+    &\var{apply}~(i :: p')~(a :: l) \\
+    \eqperm{}& \text{\{ by \var{rem\_PermFun\_correct} \}} \\
+    &(a :: l)_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a::l))
+    \\
+    \eqperm{}& \text{\{ by IH \}} \\
+    &(a :: l)_i :: \var{rem\_nth}~i~(a::l)
+    \\
+    \eqperm{}& \text{\{ \var{rem\_nth\_Perm} \}} \\
+    & a :: l \qedhere
+  \end{align*}
+\end{proof}
+
+\begin{theorem}[PermutationEx\_iff]
+  \label{thm:permutationex_iff}
+  For all lists $l$ and $l'$,
+  \begin{equation*}
+    \var{Permutation}~l~l' \iff \var{PermutationEx}~l~l'
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  The $\impliedby$ direction follows from \var{apply\_correct}.
+
+  ($\implies$) We reason by induction on the evidence of
+  $\var{Permutation}~l~l'$. The empty case is simple.
+  \begin{itemize}
+  \item We have a permutation function $p$ from the induction
+    hypothesis such that $\var{apply}~p~l = l'$, and we need to
+    construct a permutation function $p'$ such that
+    $\var{apply}~p'~(x::l) = x :: l'$. We let $p' = 0 ::
+    \var{map}~(+1)~p$, shifting every index up by one and mapping
+    the first element of the list to itself.
+  \item We have to construct a permutation function $p$ such that
+    $\var{apply}~p~(y :: x :: l) = x :: y :: l$. We let $p = 1 :: 0 ::
+    \var{map}~(+2)~(\var{seq}~0~(\var{length}~l))$, leaving every
+    element in place except for the first 2.
+  \item We have permutation functions $p_1$ and $p_2$ from our two
+    induction hypotheses, such that
+    \begin{align*}
+      \var{apply}~p_1~l = l' && \var{apply}~p_2~l' = l''
+    \end{align*}
+    We need to construct a $p'$ such that $\var{apply}~p'~l=l''$, so
+    we let $p' = p_2 \comp p_1$. For the definition of composition of
+    permutations, see \cref{appendix:perm_comp}. Then,
+    \begin{equation*}
+      \var{apply}~(p_2 \comp p_1)~l =
+      \var{apply}~p_2~(\var{apply}~p_1~l) = l'' \qedhere
+    \end{equation*}
+  \end{itemize}
+\end{proof}
+
+\section{\var{StablePerm} definitions equivalence}
+\label{appendix:stableperm_def_eq}
+
+\begin{lemma}[StablePerm\_app]
+  For all lists $l$, $l'$, $m$ and $m'$,
+  \begin{equation*}
+    \var{StablePerm}~l~l' \implies \var{StablePerm}~m~m' \implies
+    \var{StablePerm}~(l \dplus m)~(l' \dplus m')
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  This follows from the distributivity of \var{filter} over \var{app}.
+\end{proof}
+
+\begin{lemma}[StablePerm\_skip]
+  For all lists $l$ and $l'$, and elements $a$,
+  \begin{equation*}
+    \var{StablePerm}~l~l' \implies \var{StablePerm}~(a::l)~(a::l')
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  For any $x$, we have to show $\var{filter}~(\equiv x)~(a::l) =
+  \var{filter}~(\equiv x)~(a::l')$. If $x \equiv a$, $a$ is added to both lists,
+  and if $x \not\equiv a$ it is added to neither.
+\end{proof}
+
+\begin{lemma}[StablePerm\_swap]
+  For all lists $l$ and $l'$, and elements $a$ and $b$ where $a \not\equiv b$,
+  \begin{equation*}
+     \var{StablePerm}~(a::b::l)~(b::a::l)
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  For any $x$, we have to show $\var{filter}~(\equiv x)~(a::b::l) =
+  \var{filter}~(\equiv x)~(b::a::l)$. The proof follows from case analysis
+  on $x \equiv a$ and $x \equiv b$.
+\end{proof}
+
+\begin{lemma}[StablePerm\_cons\_app]
+  For all lists $l$, $l_1$ and $l_2$, and elements $a$ such that $\forall b
+  \in l_1,~a \not\equiv b$,
+  \begin{equation*}
+     \var{StablePerm}~l~(l_1 \dplus l_2) \implies \var{StablePerm}~(a
+     :: l)~(l_1 \dplus a :: l_2)
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  For any $x$, we have to show
+  \begin{equation*}
+    \var{filter}~(\equiv x)~(a::l) = \var{filter}~(\equiv x)~(l_1 \dplus a :: l_2)
+  \end{equation*}
+  If $x \not\equiv a$ the proof is easy, so assume $x \equiv a$. Then we have
+  \begin{align*}
+      \var{filter}~(\equiv x)~(l_1 \dplus a :: l_2)
+    &= \var{filter}~(\equiv x)~l_1 \dplus \var{filter}~(a :: l_2) \\
+    &= [] \dplus a::\var{filter}~l_2 \\
+    &= a::\var{filter}~l
+  \end{align*}
+  where we know that $\var{filter}~(\equiv x)~l_1 = []$ because $\forall b \in
+  l_1,~a \not\equiv b$, and the last step follows from
+  $\var{StablePerm}~l~(l_1 \dplus l_2)$.
+\end{proof}
+
+\begin{theorem}[StablePerm\_destr]
+  For all lists $l$ and $l'$, and elements $h$ and $h'$ such that $h
+  \not\equiv h'$,
+  \begin{gather*}
+    \var{StablePerm}~(h :: l)~(h' :: l') \implies \\
+    \exists~l_1~l_2, ~l = l_1 \dplus h' :: l_2 \land (\forall x \in l_1, h' \not\equiv x)
+  \end{gather*}
+\end{theorem}
+\begin{proof}
+  Let $l_1 = \var{take\_while}~(\not\equiv h')~l$ and $l_2 =
+  \var{drop\_while}~(\not\equiv h')~l$. Clearly $l = l_1 \dplus l_2$, and
+  because $\var{StablePerm}~(h :: l)~(h' :: l')$, we know the first
+  element of $l_2$ must be $h'$. Let $l_2'$ be the rest of $l_2$, and
+  we have two lists $l_1$ and $l_2'$ that satisfy the theorem.
+\end{proof}
+
+\begin{theorem}[StablePerm\_length]
+  For all lists $l$ and $l'$
+  \begin{equation*}
+    \var{StablePerm}~l~l' \implies \var{length}~l = \var{length}~l'
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We reason by induction on $n = \var{length}~l$. In the inductive
+  case, we need to show that $\var{length}~(h::t) =
+  \var{length}~(h'::t')$ given that $\var{StablePerm}~(h::t)~(h'::t')$
+  and the induction hypothesis that for all lists of length $n$,
+  \var{StablePerm} implies their lengths are equal.
+
+  If $h \equiv h'$, then from $\var{StablePerm}~(h::t)~(h'::t')$ we know
+  that $h = h'$. We can then remove $h$ from the front of $h::t$ and
+  $h::t'$ and apply the induction hypothesis.
+
+  If $h \not\equiv h'$, we can destructure $t$ and $t'$ as $l_1 \dplus h'
+  :: l_2$ and $l_1' \dplus h :: l_2'$ respectively, using
+  \var{StablePerm\_destr}. Then,
+  \begin{equation*}
+      \var{length}~t
+    = \var{length}~(h' :: l_1  \dplus l_2)
+    = \var{length}~(h  :: l_1' \dplus l_2')
+    =\var{length}~t'
+  \end{equation*}
+  where the first and last steps follow from the induction hypothesis,
+  using the fact that $\var{StablePerm}~t~(h'::l_1 \dplus l_2)$ and
+  $\var{StablePerm}~t'~(h::l_1' \dplus l_2')$.
+\end{proof}
+
+\begin{theorem}[StablePerm\_Perm]
+  For all lists $l$ and $l'$,
+  \begin{equation*}
+    \var{StablePerm}~l~l' \implies \var{Permutation}~l~l'
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We reason by strong induction on the length of both lists (we know
+  the lengths are equal by \var{StablePerm\_length}). In the inductive
+  case, we need to show $\var{Permutation}~(h :: t)~(h' :: t')$. If $h
+  \equiv h$, we know that $h = h'$, and the proof follows by
+  \var{perm\_skip}.
+
+  If $h \not\equiv$, we again destructure $t$ and $t'$ as $l_1 \dplus h' ::
+  l_2$ and $l_1' \dplus h :: l_2'$ respectively, using
+  \var{StablePerm\_destr}.
+  \begin{align*}
+    h::t ={}& h :: l_1 \dplus h' :: l_2 \\
+    \eqperm{}& h :: h' :: l_1 \dplus l_2 \\
+    \eqperm{}& h' :: h :: l_1 \dplus l_2 \\
+    \eqperm{}& h' :: h :: l_1' \dplus l_2' \\
+    \eqperm{}& h' :: l_1' \dplus h :: l_2' \\
+    ={}& h'::t
+  \end{align*}
+  The crucial step is $l_1 \dplus l_2 \eqperm l_1' \dplus l_2'$, which
+  follows from the induction hypothesis.
+\end{proof}
+
+\begin{theorem}[StablePermInd\_iff]
+  For all lists $l$ and $l'$
+  \begin{equation*}
+    \var{StablePerm}~l~l' \iff \var{StablePermInd}~l~l'
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  The ($\impliedby$) direction follows easily by induction on
+  $\var{StablePermInd}~l~l'$, using \var{StablePerm\_skip},
+  \var{StablePerm\_swap}, and \var{StablePerm\_trans}.
+
+  The ($\implies$) direction is proved very similarly to
+  \var{StablePerm\_Perm}, using \var{StablePerm\_destr} to split up
+  the lists.
+\end{proof}
+
+Now we prove equivalence with \var{StablePermEx}.
+
+\begin{theorem}[apply\_correct\_stable]
+  For all lists $l$ and permutation functions $p$,
+  \begin{equation*}
+    \var{StablePermFun}~l~p \implies \var{StablePerm}~(\var{apply}~p~l)~l
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We reason by induction on the length of $l$. In the inductive case,
+  we need to show
+  \begin{equation*}
+    \var{StablePerm}~(\var{apply}~(i :: p)~(a :: l))~(a :: l)
+  \end{equation*}
+
+  Using $\eqstable$ to chain \var{StablePerm}s transitively,
+  \begin{align*}
+    \var{apply}~(i :: p)~(a :: l)
+    ={}& (a :: l)_i  ::
+    \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a :: l)) \\
+    \eqstable{}& (a :: l)_i :: \var{rem\_nth}~i~(a :: l) \\
+    \eqstable{}& (a :: l)_i
+  \end{align*}
+  where we rely on the induction hypothesis between the first and
+  second lines.
+\end{proof}
+
+\begin{theorem}[StablePermEx\_iff]
+  For all lists $l$ and $l'$
+  \begin{equation*}
+    \var{StablePerm}~l~l' \iff \var{StablePermEx}~l~l'
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We use \var{StablePermInd\_iff} and prove $\var{StablePermInd}~l~l'
+  \iff \var{StablePermEx}~l~l'$.
+
+  The $\implies$ direction is very similar to the $\implies$ direction
+  of the \var{PermutationEx\_iff} proof: we reason by induction on
+  $\var{StablePermInd}~l~l'$, constructing the same permutation
+  functions in each case, but this time we have to prove that it is
+  stable as well. The $\impliedby$ direction follows from
+  \var{apply\_correct\_stable}.
+\end{proof}
+
 \section{Permutation composition}
@@
 
+\clearpage
+
 \nocite{*}

### 2019-10-06 draft_paper/draft_paper.tex
Fix link
Compile succesfully(ish) with acmart

@@
-\documentclass[11pt]{thesis}
-
-\title{Towards a Verified Bzip2: A Functional Specification for Burrows--Wheeler}
+\documentclass[sigplan,10pt,anonymous,review]{thesis}
+
+\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
+\acmYear{2020}
+\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
+\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\startPage{1}
+
+%% Copyright information
+%% Supplied to authors (based on authors' rights management selection;
+%% see authors.acm.org) by publisher for camera-ready submission;
+%% use 'none' for review submission.
+\setcopyright{none}
+%\setcopyright{acmcopyright}
+%\setcopyright{acmlicensed}
+%\setcopyright{rightsretained}
+%\copyrightyear{2018}           %% If different from \acmYear
+
+%% Bibliography style
+\bibliographystyle{ACM-Reference-Format}
+%% Citation style
+%\citestyle{acmauthoryear}  %% For author/year citations
+%\citestyle{acmnumeric}     %% For numeric citations
+%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
+                            %% sorting of references within a single citation;
+                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
+                            %% rendered as [14,5,2] rather than [2,5,14].
+%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
+                            %% compression of sequential references within a
+                            %% single citation;
+                            %% e.g., \cite{Baker12,Baker14,Baker16}
+                            %% rendered as [2,3,4] rather than [2-4].
+
+\begin{document}
+
+\title{Towards a Verified Bzip2}
+\subtitle{A Functional Specification for Burrows--Wheeler}
+
 \author{Jake Waksbaum}
-\date{May 6, 2019}
-
-\begin{document}
-
-\pagenumbering{gobble}
-
-\begin{titlepage}
-  \centering \Large{\textbf{Towards a Verified Bzip2: \\ A Functional Specification for Burrows--Wheeler
-  }} \vspace{10mm}
-
-  \Large\textit{Jake Waksbaum} \\
-  \vspace{25mm} \large{Junior Two-semester Independent Work \\ Department of
-    Computer Science \\ Princeton University \\ Fall 2018--Spring 2019\\}
-  \vspace{8mm}
-  \begin{figure}[!h]
-    \centering
-    \includegraphics[scale=0.1]{seal.png}
-  \end{figure}
-  \vspace{1mm}
-  \normalsize{Advised by Andrew W. Appel (COS)}\\
-  \normalsize{Second Reader: David Walker (COS)}
-
-  \vfill
-  \pledge{Jake Waksbaum}
-\end{titlepage}
-
-\clearpage
-
-\section*{Acknowledgements}
-
-Thanks to Professor Andrew Appel for being extremely generous with his
-time and his guidance. He was always available for questions, and
-allowed me to focus my work on the areas I found interesting.
-
-I'd also like to thank Matthew Weaver, Lennart Berringer, and Qinshi
-Wang for answering my questions throughout the year. Zachary Kincaid
-provided useful comments on a draft of this paper.
-
-I'd like to thank Benjamin Huang, Yael Marans, and Eli Waksbaum for
-their help and support.
-
-\clearpage
-
-\tableofcontents
-\clearpage
-
-\pagenumbering{arabic}
+\affiliation{
+  \department{Department of Computer Science}
+  \institution{Princeton University}
+  \city{Princeton}
+  \state{New Jersey}
+  \postcode{08540}
+  \country{USA}
+}
+\email{waksbaum@princeton.edu}
+
+\author{Andrew Appel}
+\affiliation{
+  \department{Department of Computer Science}
+  \institution{Princeton University}
+  \city{Princeton}
+  \state{New Jersey}
+  \postcode{08540}
+  \country{USA}
+}
+\email{appel@princeton.edu}
+
+\begin{abstract}
+Text of abstract \ldots.
+\end{abstract}
+
+%% 2012 ACM Computing Classification System (CSS) concepts
+%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
+\begin{CCSXML} <ccs2012>
+<concept>
+<concept_id>10003752.10003809.10010031.10002975</concept_id>
+<concept_desc>Theory of computation~Data compression</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10003809.10010031.10010033</concept_id>
+<concept_desc>Theory of computation~Sorting and searching</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10010940.10010992.10010998</concept_id>
+<concept_desc>Software and its engineering~Formal methods</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Theory of computation~Data compression}
+\ccsdesc[500]{Theory of computation~Sorting and searching}
+\ccsdesc[500]{Software and its engineering~Formal methods}
+%% End of generated code
+
+\keywords{Formal methods, Sorting and seraching, Data compression}
+
+\maketitle
 
@@
 definition or theorem in our Coq sources, which can be found at
-\url{https://github.com/jbaum98/verifed_bzip2/tree/2019-05-06}.
+\url{https://github.com/jbaum98/verified_bzip2/tree/2019-05-06}.
 Definitions are copied verbatim. Theorems are rephrased for clarity,
@@
   \begin{align*}
-       (\var{iter}~f~(n+1)~z)_{i+1} \\
+       (\var{iter}~f~(n+1)~z)_{i+1}
     &= (z :: \var{iter}~f~n~(f~z))_{i+1} \\
@@
 relatively simple.
-\clearpage
+
+\begin{acks}
+Thanks to Professor Andrew Appel for being extremely generous with his
+time and his guidance. He was always available for questions, and
+allowed me to focus my work on the areas I found interesting.
+
+I'd also like to thank Matthew Weaver, Lennart Berringer, and Qinshi
+Wang for answering my questions throughout the year. Zachary Kincaid
+provided useful comments on a draft of this paper.
+
+I'd like to thank Benjamin Huang, Yael Marans, and Eli Waksbaum for
+their help and support.
+\end{acks}
+
+\bibliography{draft_paper}
 
@@
 
-\clearpage
-
-\nocite{*}
-\printbibliography{}
-
 \end{document}

### 2019-10-07 draft_paper/draft_paper.tex
Add CPP conference
Change citation style to author/year
Small cleanups
Inline all the listings

@@
 
-\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
+\acmConference
+    [CPP'20]
+    {ACM SIGPLAN International Conference on Certified Programs and Proofs}
+    {January 20--21, 2020}{New Orleans, LA, USA}
 \acmYear{2020}
-\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM} \acmDOI{} %
+\acmDOI{10.1145/nnnnnnn.nnnnnnn}
 \startPage{1}
@@
 %% Citation style
-%\citestyle{acmauthoryear}  %% For author/year citations
+\citestyle{acmauthoryear}  %% For author/year citations
 %\citestyle{acmnumeric}     %% For numeric citations
@@
     \var{StableSort}~f \implies \var{StableSort}~g \implies
-    (\forall l,~f~l = g~l)
+    (\forall l,f~l = g~l)
   \end{equation*}
@@
 \var{Preord} typeclass in Coq:
-\lstinputlisting[firstline=9, lastline=14]{../theories/BWT/Sorting/Ord.v}
+\begin{lstlisting}
+Class Preord (A : Type) :=
+  { le : A -> A -> Prop;
+    le_trans : forall x y z, le x y -> le y z -> le x z;
+    le_total : forall x y, le x y \/ le y x;
+    le_dec : forall x y, {le x y} + {~le x y};
+  }.
+\end{lstlisting}
 We do not explicitly require reflexivity because it follows from
@@
 for this:
-\lstinputlisting[firstline=175, lastline=176, widthgobble=2]{../theories/BWT/Sorting/Ord.v}
+\begin{lstlisting}
+Class Ord (A : Type) `{Preord A} :=
+  { eqv_eq : forall x y, eqv x y -> x = y }.
+\end{lstlisting}
 For the BWT, we require an \var{Ord} instance for $A$, but
@@
 moves the first element of a list to the end.
-\lstinputlisting[widthgobble=2, firstline=24, lastline=28]{../theories/BWT/Rotation/Rotation.v}
+\begin{lstlisting}
+Definition lrot (l: list A) : list A :=
+  match l with
+  | [] => []
+  | hd :: tl => tl ++ [hd]
+  end.
+\end{lstlisting}
 \var{rrot} moves the last element of a list to the front, and
 relies on \var{init}, which drops the last element of a list.
-\lstinputlisting[widthgobble=2, firstline=18, lastline=22]{../theories/BWT/Lib/List.v}
-\lstinputlisting[widthgobble=2, firstline=17, lastline=21]{../theories/BWT/Rotation/Rotation.v}
+\begin{lstlisting}
+Fixpoint init l : list A :=
+  match l with
+  | [] | [_] => []
+  | hd :: tl => hd :: init tl
+  end.
+\end{lstlisting}
+\begin{lstlisting}
+Definition rrot (l: list A) : list A :=
+  match l with
+  | [] => []
+  | hd :: tl => last l hd :: init l
+  end.
+\end{lstlisting}
 Note that in order to ensure that $\var{last}$ is total, it takes
@@
 where $f^n$ denotes repeated function application.
-\lstinputlisting[
-  widthgobble=2,
-  firstline=14,
-  lastline=18
-]{../theories/BWT/Lib/Iterate.v}
+\begin{lstlisting}
+Fixpoint iter (f : A -> A) (n : nat) (z : A)
+  : list A :=
+  match n with
+  | O   => []
+  | S m => z :: iter f m (f z)
+  end.
+\end{lstlisting}
 
@@
 function \var{rep} for repeated function application.
-\lstinputlisting[
-  widthgobble=2, firstline=8, lastline=12,
-]{../theories/BWT/Lib/Repeat.v}
+\begin{lstlisting}
+Fixpoint rep (f : A -> A) (n : nat) (z : A) : A :=
+  match n with
+  | O => z
+  | S m => f (rep f m z)
+  end.
+\end{lstlisting}
 In proofs, we will still use the notation $f^i z$ for
@@
 Now we can define \var{rots} to produce the rotation matrix.
-\lstinputlisting[
-  widthgobble=2,
-  firstline=17,
-  lastline=18,
-]{../theories/BWT/Rotation/Rots.v}
+\begin{lstlisting}
+Definition rots (l : list A) : list (list A) :=
+  iter lrot (length l) l.
+\end{lstlisting}
 
@@
 using any sorting algorithm.
-\lstinputlisting[
-  widthgobble=2,
-  firstline=31,
-  lastline=35
-]{../theories/BWT/BurrowsWheeler.v}
+\begin{lstlisting}
+Definition bwl (l : list A) : list A :=
+  match l with
+  | [] => []
+  | hd :: _ =>
+    List.map (fun x => last x hd) (lexsort (rots l))
+  end.
+\end{lstlisting}
 
@@
 defined with respect to some decidable equivalence relation.
-\lstinputlisting[
-  widthgobble=2,
-  firstline=12,
-  lastline=19,
-]{../theories/BWT/Lib/FindIndex.v}
+\begin{lstlisting}
+Fixpoint findIndex (x : A) (ls : list A) : nat :=
+  match ls with
+  | [] => 0
+  | hd :: tl =>
+    if x == hd then 0
+    else S (findIndex x tl)
+  end.
+\end{lstlisting}
 
@@
 finds the row that is Leibniz equal to $l$.
-\lstinputlisting[
-  widthgobble=2,
-  firstline=75,
-  lastline=75,
-]{../theories/BWT/BurrowsWheeler.v}
+\begin{lstlisting}
+Definition bwn (l : list A) : nat :=
+  findIndex l (lexsort (rots l)).
+\end{lstlisting}
 
@@
 the last column of the sorted rotation matrix.
-\lstinputlisting[
-  widthgobble=2, firstline=156, lastline=160,
-]{../theories/BWT/BurrowsWheeler.v}
+\begin{lstlisting}
+Fixpoint recreate (j : nat) (l : list A)
+  : list (list A) :=
+  match j with
+  | O    => map (const []) l
+  | S j' => hdsort (prepend_col l (recreate j' l))
+  end.
+\end{lstlisting}
 This definition relies on \var{prepend\_col} and \var{zipWith}, defined below.
-\lstinputlisting[
-  widthgobble=2, firstline=47, lastline=48,
-]{../theories/BWT/Columns.v}
-\lstinputlisting[
-  widthgobble=2, firstline=16, lastline=20,
-]{../theories/BWT/Lib/ZipWith.v}
+\begin{lstlisting}
+Definition prepend_col
+  : list A -> list (list A) -> list (list A)
+  := zipWith cons.
+
+Fixpoint zipWith (a : list A) (b : list B)
+  : list C :=
+  match (a, b) with
+  | (ahd :: atl, bhd :: btl) =>
+    f ahd bhd :: zipWith atl btl
+  | _ => []
+  end.
+\end{lstlisting}
 
 Now we can redefine \var{unbwt} as
-\lstinputlisting[
-  widthgobble=2,
-  firstline=197,
-  lastline=198,
-]{../theories/BWT/BurrowsWheeler.v}
+\begin{lstlisting}
+Definition unbwt (i : nat) (l : list A) : list A :=
+  nth i (recreate (length l) l) l.
+\end{lstlisting}
 
@@
 invariant using $\var{cols}~j$, which takes the first $j$ columns of a matrix.
-\lstinputlisting[
-  widthgobble=2,
-  firstline=19,
-  lastline=20,
-]{../theories/BWT/Columns.v}
+\begin{lstlisting}
+Definition cols (j : nat)
+  : list (list A) -> list (list A) := map (firstn j).
+\end{lstlisting}
 
@@
 represents a permutation function on lists of length $n$:
-\lstinputlisting[
-  firstline=14, lastline=15,
-]{../theories/BWT/Sorting/PermFun.v}
+\begin{lstlisting}
+Definition PermFun (n : nat) (p : list nat)
+  : Prop := NoDup p /\ forall i, In i p <-> i < n.
+\end{lstlisting}
 Then, we can define \var{apply}, which permutes a list according to a
 given permutation:
-\lstinputlisting[
-  widthgobble=2, firstline=161, lastline=165,
-]{../theories/BWT/Sorting/PermFun.v}
+\begin{lstlisting}
+Definition apply p l : list A :=
+  match l with
+  | [] => []
+  | d :: _ => map (fun i => nth i l d) p
+  end.
+\end{lstlisting}
 And finally, we can define \var{PermutationEx}, which holds between
 two lists when they are permutations of each other.
-\lstinputlisting[
-  widthgobble=2, firstline=521, lastline=522,
-]{../theories/BWT/Sorting/PermFun.v}
-
+\begin{lstlisting}
+Definition PermutationEx l l' : Prop :=
+  exists p, PermFun (length l) p /\ apply p l = l'.
+\end{lstlisting}
 We prove \var{PermutationEx} equivalent to \var{Permutation} in
@@
 another when any element occurs an equal number of times in each:
-\lstinputlisting[
-  widthgobble=2, firstline=14, lastline=15,
-]{../theories/BWT/Sorting/PermutationCount.v}
+\begin{lstlisting}
+Definition PermutationCount l l' : Prop := forall x,
+  count_occ equiv_dec l x = count_occ equiv_dec l' x.
+\end{lstlisting}
 \var{count\_occ} counts the number of occurrences of a given variable
@@
 so we say $\var{StablePermFun}~l~p$ for some list $l$.
-\lstinputlisting[
-  widthgobble=2, firstline=568, lastline=572,
-]{../theories/BWT/Sorting/PermFun.v}
+\begin{lstlisting}
+Definition StablePermFun (l : list A) (p : list nat)
+  := PermFun (length l) p /\ forall i j d,
+    nth i (apply p l) d === nth j (apply p l) d ->
+    i < j < length l -> image p i < image p j.
+\end{lstlisting}
 As above, we can also define this
 property using the inverse permutation:
-\lstinputlisting[
-  widthgobble=2, firstline=574, lastline=578,
-]{../theories/BWT/Sorting/PermFun.v}
+\begin{lstlisting}
+Definition StablePermFun_preimage
+  (l : list A) (p : list nat) :=
+  PermFun (length l) p /\ forall i j d,
+    nth i l d === nth j l d ->
+    i < j < length l -> preimage p i < preimage p j.
+\end{lstlisting}
+
 As usual, we will prove these
@@
 and \var{preimage}.
-\lstinputlisting[
-  widthgobble=2, firstline=81, lastline=82,
-]{../theories/BWT/Sorting/PermFun.v}
-We will not prove
-it here, but $\var{preimage}~p$ and $\var{image}~p$ are both of
-bijections of type $\mathbb{N}_{<n} \to \mathbb{N}_{<n}$ and are
-inverses of each other. We also have the following two theorems:
+\begin{lstlisting}
+Definition image (p : list nat) i := nth i p 0.
+Definition preimage (p : list nat) i := findIndex i p.
+\end{lstlisting}
+We will not prove it here, but $\var{preimage}~p$ and $\var{image}~p$
+are both of bijections of type $\mathbb{N}_{<n} \to \mathbb{N}_{<n}$ and
+are inverses of each other. We also have the following two theorems:
 \begin{theorem}[nth\_preimage\_apply, nth\_image\_apply]
@@
 stable permutations of each other:
-\lstinputlisting[
-  widthgobble=2, firstline=618, lastline=619,
-]{../theories/BWT/Sorting/PermFun.v}
+\begin{lstlisting}
+Definition StablePermEx l l' :=
+  exists p, StablePermFun l p /\ apply p l = l'.
+\end{lstlisting}
 
@@
 we modify \var{Permutation} to create an inductive relation for stable
-lists? Intuitively, the problematic inference rule is
-\var{Permutation\_swap}: we can only swap two elements and preserve
-stability if they are not equivalent. With this fix, we have \var{StablePermInd}:
+lists? Intuitively, the problem is \var{Permutation\_swap}: we can
+only swap two elements and preserve stability if they are not
+equivalent. With this fix, we have \var{StablePermInd}:
 \begin{gather*}
@@
 permutations of one another \cite{leroy_2018}.
-\lstinputlisting[
-  widthgobble=2, firstline=21, lastline=22,
-]{../theories/BWT/Sorting/StablePerm.v}
-
+\begin{lstlisting}
+Definition StablePerm (l l': list A) : Prop := forall x,
+  filter (equiv_decb x) l = filter (equiv_decb x) l'.
+\end{lstlisting}
 We prove equivalence between these three definitions in
@@
 more elegantly.
-\lstinputlisting[
-  widthgobble=2, firstline=15, lastline=19,
-]{../theories/BWT/Sorting/Sort.v}
+\begin{lstlisting}
+Definition Sort f :=
+  forall l, Sorted (f l) /\ Permutation (f l) l.
+\end{lstlisting}
 
@@
 of lists is sorted if you only look at the first $n$ elements of each:
-\lstinputlisting[
-  widthgobble=2, firstline=326, lastline=327,
-]{../theories/BWT/Sorting/Key.v}
+\begin{lstlisting}
+Definition PrefixSorted (n : nat) : list (list A) -> Prop
+  := Sorted (keyOrd (firstn n)).
+\end{lstlisting}
 
@@
 Or, in Coq:
-\lstinputlisting[
-  widthgobble=2, firstline=27, lastline=28,
-]{../theories/BWT/Sorting/RadixSort.v}
+\begin{lstlisting}
+Definition radixsort m (n : nat) : list (list A)
+  := rep (hdsort ∘ map rrot) n m.
+\end{lstlisting}
 Here \var{hdsort} is just an insertion sort using $\le_1$ ordering.
@@
 
-\lstinputlisting[
-  widthgobble=2, firstline=15, lastline=26,
-]{../theories/BWT/Sorting/InsertionSort.v}
+\begin{lstlisting}
+  Fixpoint insert (x : A) (l : list A) :=
+    match l with
+    | [] => [x]
+    | h :: t =>
+      if le_dec x h then x :: h :: t else h :: insert x t
+    end.
+
+  Fixpoint sort (l : list A) : list A :=
+    match l with
+    | [] => []
+    | h :: t => insert h (sort t)
+    end.
+\end{lstlisting}
 
@@
 \end{equation*}
-\lstinputlisting[
-  widthgobble=4, firstline=229, lastline=229,
-]{../theories/BWT/Sorting/PermFun.v}
-\vspace{\parskip}
-\lstinputlisting[
-  widthgobble=2, firstline=599, lastline=607,
-]{../theories/BWT/Lib/List.v}
+\begin{lstlisting}
+Definition rem_PermFun i :=
+  map (fun j => if lt_dec i j then pred j else j).
+
+Fixpoint rem_nth i l :=
+  match l with
+  | [] => []
+  | h :: t =>
+    match i with
+    | 0 => t
+    | S i' => h :: rem_nth i' t
+    end
+  end.
+\end{lstlisting}
 
@@
 simply apply one permutation to the other permutation to compose them.
-\lstinputlisting[
-  widthgobble=2, firstline=409, lastline=409
-]{../theories/BWT/Sorting/PermFun.v}
+\begin{lstlisting}
+Definition compose p2 p1 := apply p2 p1.
+\end{lstlisting}
 This definition, while simple, does not make it obvious that
@@
 \begin{lstlisting}
-  Fixpoint combine (l : list A) (l' : list B) : list (A*B) :=
+Fixpoint combine (l : list A) (l' : list B)
+  : list (A*B) :=
   match l,l' with

### 2019-10-07 draft_paper/draft_paper.tex
Fix all egregious overflows

@@
   For all $j$ and matrices $m$ that has rows all of length less than $j$,
-  \begin{equation*}
-    \var{cols}~(j+1)~(\var{map}~\var{rrot}~m) =
+  \begin{gather*}
+    \var{cols}~(j+1)~(\var{map}~\var{rrot}~m) = \\
     \var{prepend\_col}~(\var{map}~\var{last}~m)~(\var{cols}~j~m)
-  \end{equation*}
+  \end{gather*}
 \end{theorem}
@@
   We reason by induction on $m$. In the inductive case we need to show that
-  \begin{equation*}
-    \var{cols}~(j+1)~(\var{map}~\var{rrot}~(r::m)) =
+  \begin{gather*}
+    \var{cols}~(j+1)~(\var{map}~\var{rrot}~(r::m)) = \\
     \var{prepend\_col}~(\var{map}~\var{last}~(r::m))~(\var{cols}~j~(r::m))
-  \end{equation*}
+  \end{gather*}
   Simplifying, we have that the tails are equal by the induction
@@
 \begin{proof}
-\begin{align*}
-  &\var{cols}~(j+1)~(\var{lexsort}~(\var{rots}~l)) \\
-={}& \var{cols}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
+  Letting $m = \var{lexsort}~(\var{rots}~l)$,
+  \begin{align*}
+    &\var{cols}~(j+1)~m \\
+    ={}& \var{cols}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~m)) \\
     ={}&
-\var{hdsort}~(\var{cols}~(j+1)~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
+    \var{hdsort}~(\var{cols}~(j+1)~(\var{map}~\var{rrot}~m)) \\
     ={}&
-\var{hdsort}~(\var{prepend\_col}~(\var{map}~\var{last}~(\var{lexsort}~(\var{rots}~l)))~(\var{cols}~j~(\var{lexsort}~(\var{rots}~l))) \\
+    \var{hdsort}~(\var{prepend\_col}~(\var{map}~\var{last}~m)~(\var{cols}~j~m)) \\
     ={}&
@@
 \begin{gather*}
-  \begin{align*}
   \infer[\var{SortedLocal\_nil}]{\var{SortedLocal}~[]}{}
-    &&
+  \\
+  \\
   \infer[\var{SortedLocal\_1}]{\var{SortedLocal}~[a]}{}
-  \end{align*}
   \\
@@
 
-  ($\impliedby$) We have $\var{SortedIx}~(a :: l')$, and we need to
-  show $\var{Sorted}~a :: l'$ given the induction hypothesis
-  $\var{SortedIx}~l' \implies \var{Sorted}~l'$. By
-  \var{SortedLocal\_iff}, we can just prove $\var{SortedLocal}~(a ::
-  l')$. If $l' = []$, we can apply $\var{SortedLocal\_1}$, so assume
-  $l' = b :: l''$. To apply \var{SortedLocal\_cons} we need to show
-  $\var{SortedLocal}~(b :: l'')$, which we have by the induction
-  hypothesis and $\var{SortedIx}~(a :: b :: l'')$, and that $a \le b$.
-  That follows because $a = (a :: b :: l'')_0$ and $b = (a :: b ::
-  l'')_1$, so by $\var{SortedIx}~(a :: b :: l'')$, we are done.
+  ($\impliedby$) We have $\var{SortedIx}\ \allowbreak(a :: l')$, and
+  we need to show $\var{Sorted}\ \allowbreak(a :: l')$ given the
+  induction hypothesis $\var{SortedIx}~l' \implies \var{Sorted}~l'$.
+  By \var{SortedLocal\_iff}, we can just prove
+  $\var{SortedLocal}\ \allowbreak(a :: l')$. If $l' = []$, we can
+  apply $\var{SortedLocal\_1}$, so assume $l' = b :: l''$. To apply
+  \var{SortedLocal\_cons} we need to show
+  $\var{SortedLocal}\ \allowbreak(b :: l'')$, which we have by the
+  induction hypothesis and $\var{SortedIx}\ \allowbreak(a :: b ::
+  l'')$, and that $a \le b$. That follows because $a = (a :: b ::
+  l'')_0$ and $b = (a :: b :: l'')_1$, so by
+  $\var{SortedIx}\ \allowbreak(a :: b :: l'')$, we are done.
 \end{proof}
@@
 $f : \mathbb{N}_{<n} \to A$ is represented by the $\var{list}~A$
-$[f(0),~f(1),\ldots,f(n-1)]$. For permutations, this is consistent with the
+$[f(0),~f(1),\allowbreak\ldots,f(n-1)]$. For permutations, this is consistent with the
 notation where we denote by $2~1~0$ the permutation $p(0) = 2$, $p(1)
@@
   We reason inductively on the proof of $\var{StablePerm}~l_1~l_2$
-  using \var{StablePermInd\_iff}. The important case is the
-  \var{StablePermInd\_swap} case, where we have to show
+  using \var{StablePermInd\_iff}. The important case is the swap case,
+  where we have to show
   \begin{equation*}
@@
   \begin{align*}
-    (\var{hdsort} \comp \var{map}~\var{rrot})^{j+1}~m ={}&
-    \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \comp
+     {}& (\var{hdsort} \comp \var{map}~\var{rrot})^{j+1}~m \\
+    ={}& \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \comp
          \var{map}~\var{rrot})^j~m) \\
@@
 \begin{proof}
-  We reason by induction on the evidence that $\var{StablePerm}~m~m'$
+  We reason by induction on the evidence that $m \eqperm m'$
   using \var{StablePermInd\_iff}. The important case is the swap case,
@@
   \begin{align*}
-    (\var{hdsort} \comp \var{map}~\var{rrot})^{j+1}~m ={}&
-    \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \comp
+    {}& (\var{hdsort} \comp \var{map}~\var{rrot})^{j+1}~m \\
+    ={}&\var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \comp
     \var{map}~\var{rrot})^j~m) \\
@@
 \begin{proof}
-  The proof is almost identical to the proof of \var{radixsort\_perm}.
+  The proof similar to the proof of \var{radixsort\_perm}.
 \end{proof}
@@
 \begin{theorem}[map\_tl\_prepend]
-  For all matrices $m : \var{list}~(\var{list}~A)$ and columns $c :
+  For all matrices $m : \var{list}\ \allowbreak(\var{list}~A)$ and columns $c :
   \var{list}~A$, where $c$ has at least as many elements as $m$ has columns,
@@
   \begin{align*}
-    \var{map}~\var{tl}~(\var{map}~\var{rrot}~m) &= \var{map}~\var{tl}~(\var{prepend\_col}~(\var{map}~\var{last}~m)~(\var{map}~\var{init}~m) \\
-    &= \var{map}~\var{init}~m \qedhere
+    {}& \var{map}~\var{tl}~(\var{map}~\var{rrot}~m) \\
+    ={}& \var{map}~\var{tl}~(\var{prepend\_col}~(\var{map}~\var{last}~m)~(\var{map}~\var{init}~m) \\
+    ={}& \var{map}~\var{init}~m \qedhere
   \end{align*}
@@
   \begin{align*}
-       &\var{PrefixSorted}~(j+1)~(\var{radixsort}~(r :: m)~(j+1)) \\
-    \iff{}&\var{PrefixSorted}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{radixsort}~(r :: m)~j))) \\
-    \iff{}&\var{PrefixSorted}~j~(\var{map}~\var{tl}~(\var{map}~\var{rrot}~(\var{radixsort}~(r :: m)~j))) \\
-    \iff{}&\var{PrefixSorted}~j~(\var{map}~\var{init}~(\var{radixsort}~(r :: m)~j)) \\
-    \iff{}&\var{Sorted}~(\var{map}~(\var{firstn}~j)~(\var{map}~\var{init}~(\var{radixsort}~(r :: m)~j))) \\
-    \iff{}&\var{Sorted}~(\var{map}~(\var{firstn}~j)~(\var{radixsort}~(r :: m)~j)) \\
-    \iff{}&\var{PrefixSorted}~j~(\var{radixsort}~(r :: m)~j)
+       &\var{PrefixSorted}~(j+1)~(\var{radixsort}~m~(j+1)) \\
+    \iff{}&\var{PrefixSorted}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{radixsort}~m~j))) \\
+    \iff{}&\var{PrefixSorted}~j~(\var{map}~\var{tl}~(\var{map}~\var{rrot}~(\var{radixsort}~m~j))) \\
+    \iff{}&\var{PrefixSorted}~j~(\var{map}~\var{init}~(\var{radixsort}~m~j)) \\
+    \iff{}&\var{Sorted}~(\var{map}~(\var{firstn}~j)~(\var{map}~\var{init}~(\var{radixsort}~m~j))) \\
+    \iff{}&\var{Sorted}~(\var{map}~(\var{firstn}~j)~(\var{radixsort}~m~j)) \\
+    \iff{}&\var{PrefixSorted}~j~(\var{radixsort}~m~j)
   \end{align*}
@@
 \begin{lstlisting}
-Definition recreate' (p : list nat) (l : list A) : list (list A) :=
+Definition recreate' (p : list nat) (l : list A)
+  : list (list A) :=
   transpose (iter (apply p) (length l) (apply p l))
@@
     let p := calc_sort_perm l in
-    let indices := tl (iter (image p) (S (length l)) i) in
-    map (fun j => nth l j d) indices
+    let indices :=
+      tl (iter (image p) (S (length l)) i)
+    in map (fun j => nth l j d) indices
 \end{lstlisting}
@@
     \var{map}~(\lambda j.~l_j)~p &=
-    \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_j)~(\var{map}~(\lambda j. \bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j)~p) \\
+    \var{map}~(\lambda j.~(\var{rem\_nth}~i~l)_j)\\
+      &\phantom{=} \quad (\var{map}~(\lambda j. \bvar{if}~i < j~\bvar{then}~j-1~\bvar{else}~j)~p) \\
     \var{map}~(\lambda j.~l_j)~p &=
@@
   For all lists $l$, $l'$, $m$ and $m'$,
-  \begin{equation*}
-    \var{StablePerm}~l~l' \implies \var{StablePerm}~m~m' \implies
+  \begin{gather*}
+    \var{StablePerm}~l~l' \implies \var{StablePerm}~m~m' \implies \\
     \var{StablePerm}~(l \dplus m)~(l' \dplus m')
-  \end{equation*}
+  \end{gather*}
 \end{lemma}
@@
 \begin{theorem}[apply\_correct\_stable]
+  \label{thm:apply_correct_stable}
   For all lists $l$ and permutation functions $p$,
@@
   \begin{align*}
-    \var{apply}~(i :: p)~(a :: l)
+    {}& \var{apply}~(i :: p)~(a :: l) \\
     ={}& (a :: l)_i  ::
@@
 \begin{proof}
-  We use \var{StablePermInd\_iff} and prove $\var{StablePermInd}~l~l'
-  \iff \var{StablePermEx}~l~l'$.
+  We prove $\var{StablePermInd}~l~l' \iff \var{StablePermEx}~l~l'$.
 
@@
   stable as well. The $\impliedby$ direction follows from
-  \var{apply\_correct\_stable}.
+  \Cref{thm:apply_correct_stable}.
 \end{proof}
@@
 \begin{theorem}[Permutation\_combine\_eq]
+  \label{thm:Permutation_combine_eq}
   For all lists $xs$, $ys_1$, and $ys_2$ with all the same length,
@@
   \begin{align*}
-       \var{compose}~(\var{seq}~0~(n+1))~(i::p)
-    &= \var{map}~(\lambda j.~(i::p)_j)~(\var{seq}~0~(n+1)) \\
-    &= \var{map}~(\lambda j.~(i::p)_j)~(0 :: \var{seq}~1~(n+1)) \\
-    &= i :: \var{map}~(\lambda j.~(i::p)_j)~(\var{seq}~1~(n+1)) \\
-    &= i :: \var{map}~(\lambda j.~(i::p)_j)~(\var{map}~(\lambda j.~j+1)~(\var{seq}~0~n)) \\
-    &= i :: \var{map}~(\lambda j.~(i::p)_{j+1})~(\var{seq}~0~n) \\
-    &= i :: \var{map}~(\lambda j.~p_j)~(\var{seq}~0~n) \\
-    &= i :: \var{compose}~(\var{seq}~0~n)~p \\
-    &= i :: p \qedhere
+     {}& \var{compose}~(\var{seq}~0~(n+1))~(i::p) \\
+    ={}& \var{map}~(\lambda j.~(i::p)_j)~(\var{seq}~0~(n+1)) \\
+    ={}& \var{map}~(\lambda j.~(i::p)_j)~(0 :: \var{seq}~1~(n+1)) \\
+    ={}& i :: \var{map}~(\lambda j.~(i::p)_j)~(\var{seq}~1~(n+1)) \\
+    ={}& i :: \var{map}~(\lambda j.~(i::p)_j)~(\var{map}~(\lambda j.~j+1)~(\var{seq}~0~n)) \\
+    ={}& i :: \var{map}~(\lambda j.~(i::p)_{j+1})~(\var{seq}~0~n) \\
+    ={}& i :: \var{map}~(\lambda j.~p_j)~(\var{seq}~0~n) \\
+    ={}& i :: \var{compose}~(\var{seq}~0~n)~p \\
+    ={}& i :: p \qedhere
   \end{align*}
@@
   and
-  \begin{equation*}
+  \begin{gather*}
     \var{combine}~\iota~l
     \eqperm
-    \var{apply}~p_1~(\var{combine}~\iota~l)
+    \var{apply}~p_1~(\var{combine}~\iota~l) \\
     \eqperm
     \var{apply}~p_2~(\var{apply}~p_1~(\var{combine}~\iota~l))
-  \end{equation*}
-  We combine these by transitivity, and then simplify, using \var{app}
-  as an abbreviation for \var{apply}.
-  \begin{align*}
-    \var{pap}~(p_2 \comp p_1)~(\var{combine}~\iota~l)
+  \end{gather*}
+  We combine these by transitivity, and then simplify. For space, we
+  denote \var{apply} by only juxtaposition, and \var{combine} by $\oplus$.
+  \begin{align*}
+    (p_2 \comp p_1)~(\iota \oplus l)
     &\eqperm
-    \var{app}~p_2~(\var{app}~p_1~(\var{combine}~\iota~l)) \\
-    \var{combine}~(\var{app}~(p_2 \comp p_1)~\iota)~(\var{app}~(p_2 \comp p_1)~l)
+    p_2~(p_1~(\iota \oplus l)) \\
+    ((p_2 \comp p_1)~\iota) \oplus ((p_2 \comp p_1)~l)
     &\eqperm
-    \var{app}~p_2~(\var{combine}~(\var{app}~p_1~\iota)~(\var{app}~p_1~l)) \\
-    \var{combine}~(p_2 \comp p_1)~(\var{app}~(p_2 \comp p_1)~l)
+    p_2~((p_1~\iota) \oplus (p_1~l)) \\
+    (p_2 \comp p_1) \oplus ((p_2 \comp p_1)~l)
     &\eqperm
-    \var{app}~p_2~(\var{combine}~p_1~(\var{app}~p_1~l)) \\
-    \var{combine}~(p_2 \comp p_1)~(\var{app}~(p_2 \comp p_1)~l)
+    p_2~(p_1 \oplus (p_1~l)) \\
+    (p_2 \comp p_1) \oplus ((p_2 \comp p_1)~l)
     &\eqperm
-    \var{combine}~(\var{app}~p_2~p_1)~(\var{app}~p_2~(\var{app}~p_1~l))
-  \end{align*}
-  Because $p_2 \comp p_1 = \var{app}~p_2~p_1$, we can apply
-  \var{Permutation\_combine\_eq}.
+    (p_2~p_1)\oplus(p_2~(p_1~l))
+  \end{align*}
+  Because $p_2 \comp p_1 = p_2~p_1$, we can apply
+  \Cref{thm:Permutation_combine_eq} to show $(p_2 \comp p_1)~l \eqperm
+  p_2~(p_1~l)$
 \end{proof}

### 2019-10-08 draft_paper/draft_paper.tex
Finish first draf of new intro and abstract

@@
 \begin{abstract}
-Text of abstract \ldots.
+Stable sorts are an important and fundamental type of sorting
+algorithm. Some algorithms, such as radix sorts, rely upon the
+stability of a sorting subroutine for their correctness. In other
+cases, stable sorts are useful for reasoning about algorithms, often
+because any two stable sorts are interchangeable. One such algorithm
+is the Burrows--Wheeler transform (BWT), which has applications in
+compression, full-text indexing and pattern searching.
+
+There has not been much work on formally verifying stable sorts, or
+rather, the stability of stable sorts. In this paper, we develop more
+fully the theory of stable sorts, and use this theory to prove the
+correctness of radixsort and the BWT.
 \end{abstract}
@@
 
-\keywords{Formal methods, Sorting and seraching, Data compression}
+\keywords{Formal methods, Sorting and searching, Data compression}
 
@@
 \label{sec:intro}
+
+Stable sorts are an important and fundamental type of sorting
+algorithm. A sort is stable if it preserves the relative order of
+equal elements: for example, if we are sorting a list of words by the
+first characters, every word that starts with \texttt{a} appears in
+the same order relative to the other words that start with \texttt{a}
+as in the original list. Stability can be directly useful in certain
+contexts, such as when sorting a table by a single column. However,
+stability is also crucial to correctness of certain general-purpose
+algorithms.
+
+One such algorith, or set of algorithms, are radix sorts. Radix sorts
+operate on lists with fixed-size elements, such as strings of a given
+length or integers with a given number of digits. A Least Significant
+Digit (LSD) radix sort uses some stable sort to sort each column,
+starting with the rightmost (or least significant) and ending with the
+leftmost (or most significant) digit. Radix sorts can be faster than
+other comparison-based sorts such as quicksort, achieving linear
+worst-case runtimes in the length of the list. This is because they
+are less general, in a sense: rather than relying only on a comparison
+function, they exploit more of the elements' structure. The cost of
+requiring that structure is that we are restricted to lists with
+fixed-size elements. In practice, though, many common types of data
+are of fixed size: integers, floating point numbers, and fixed-length
+strings, as well as structs and tuples of other fixed size data types.
+In addition, radix sorts can be extended to handle variable-length
+strings \cite{McIlroy93}.
+
+% https://probablydance.com/2016/12/02/investigating-radix-sort/
+
+Stable sorts are also important for reasoning about sorting algorithms
+more broadly. This is because any two stable sorting algorithms are
+equivalent, in the sense that they will produce the same outputs for a
+given input. Crucially, this allows us to reason using elegant
+algorithms and replace them with efficient algorithms later on. We
+make use of this technique in order to reason about LSD radix sort, as
+well as an algorithm called the Burrows-Wheeler transform (BWT).
 
@@
 sort the string, but that would be useless for compression because the
-original string would be unrecoverable.
+original string would be unrecoverable. Notably, the BWT is the core
+pass in the popular, open-source bzip2 compression software
+\cite{tsai_2016}. In addition, the BWT has been recently applied to
+full-text indexing and pattern searching, for example in genomics
+\cite{ferragina_index, dna}.
+
+There has not been much work on formally verifying stable sorts, or
+rather, the stability of stable sorts. Some work has been done on
+verifying the stability of mergesort
+\cite{leroy_2018,Sternagel2013,Leino2015,deGouw2014}, but we do not
+know of any formally verified implementation of a radix sort including
+stability. In addition, to our knowledge, no one has produced a
+formally verified implementation of the BWT, by which we mean an
+implementation of the forwards-transform, the inverse-transform, and a
+proof that they the inverse-transform inverts the forwards-transform.
+
+In this paper, we develop more fully the theory of stable sorts: we
+define stability in 3 different ways and prove equivalence between
+them, and we prove that all stable sorts are interchangeable. We use
+this to prove the correctness and stability of an LSD radix sort.
+Finally, we prove implement and prove correct the BWT, basing our
+proof on the excellent derivation of Bird
+et~al\cite{alg-of-prog,birdmu}. Bird's technique of proof by program
+calculation transferred very well to a machine-checked setting; the
+most difficult part was proving the lemmas the derivation relies upon,
+which include the correctness of LSD radix sort.
+
+\section{Main Ideas}
 

### 2019-10-11 draft_paper/draft_paper.tex
Change the title
stash

--- 2019-10-07 draft_paper/draft_paper.tex
@@
 
-\title{Towards a Verified Bzip2}
-\subtitle{A Functional Specification for Burrows--Wheeler}
+\title[Coq Formalization of Stable Sorts, Radix Sort, and BWT]{A Coq
+  Formalization of Stable Sorts, Radix Sort, and the Burrows-Wheeler
+  Transform}
 
@@
 \begin{abstract}
-Text of abstract \ldots.
+Stable sorts are an important and fundamental type of sorting
+algorithm. Some algorithms, such as radix sorts, rely upon the
+stability of a sorting subroutine for their correctness. In other
+cases, stable sorts are useful for reasoning about algorithms, often
+because any two stable sorts are interchangeable. One such algorithm
+is the Burrows--Wheeler transform (BWT), which has applications in
+compression, full-text indexing and pattern searching.
+
+There has not been much work on formally verifying stable sorts, or
+rather, the stability of stable sorts. While Bird
+et~al.\cite{birdmu,pearls} provide an elegant derivation and proof for
+the BWT, formalizing it in Coq reveals that it relies on a few lemmas
+that are not trivial to prove, including the correctness of
+least-significant digit (LSD) radix sort. In this paper, we develop
+more fully the theory of stable sorts, and use this theory to prove
+the correctness of LSD radix sort and thereby the BWT.
 \end{abstract}
@@
 
-The Burrows--Wheeler transform (BWT) is an invertible transformation
-that makes a string more amenable to compression by other methods
-\cite{bw}. It accomplishes this by creating runs of repeated letters
-in the produced string. Of course, a simple way to create runs is to
-sort the string, but that would be useless for compression because the
-original string would be unrecoverable.
-
-To apply the transformation to a string of length $n$:
+Stable sorts are an important and fundamental type of sorting
+algorithm. A sort is stable if it preserves the relative order of
+equal elements. For example, if we are stably sorting the list
+$[\var{banana},\ \allowbreak\var{armpit},\ \allowbreak\var{carrot},\ \allowbreak\var{apple}]$,
+we have to produce
+$[\var{armpit},\ \allowbreak\var{apple},\ \allowbreak\var{banana},\ \allowbreak\var{carrot}]$
+and not
+$[\var{apple},\ \allowbreak\var{armpit},\ \allowbreak\var{banana},\ \allowbreak\var{carrot}]$.
+Even though both are sorted on the first character, because
+\var{armpit} preceded \var{apple} in the original lists, it must also
+precede \var{apple} in the stably sorted list.
+
+Stability can be directly useful in certain contexts, such as when
+sorting a table by a single column. However, stability is also crucial
+to the correctness of certain general-purpose algorithms. One such
+algorithm, or set of algorithms, is radix sort. Radix sorts are used
+to sort data lexicographically, and can be more efficient than
+comparison-based sorting methods, both in theory and in practice.
+Radix sorts are especially well-suited to sorting strings
+\cite{McIlroy93}. Many radix sorts rely on a some stable sorting
+subroutine.
+
+Stable sorts are also important for reasoning about sorting algorithms
+more broadly. This is because any two stable sorting algorithms are
+equivalent, in the sense that they will produce the same outputs for a
+given input. Crucially, this allows us to reason using simple
+algorithms and replace them with more efficient algorithms later on.
+We make use of this technique in order to reason about LSD radix sort,
+as well as the Burrows-Wheeler transform (BWT).
+
+The Burrows--Wheeler transform is an invertible transformation that
+makes a string more amenable to compression by other methods
+\cite{bw}. Notably, the BWT is the core pass in the popular,
+open-source bzip2 compression software \cite{tsai_2016}. In addition,
+the BWT has been applied to full-text indexing and pattern searching,
+for example in genomics \cite{ferragina_index, dna}.
+
+There has not been much work on formally verifying stable sorts, or
+rather, the stability of stable sorts. Some work has been done on
+verifying the stability of mergesort
+\cite{leroy_2018,Sternagel2013,Leino2015,deGouw2014}, but we do not
+know of any formally verified implementation of a radix sort that
+includes stability.
+
+In addition, to our knowledge no one has produced a formally verified
+implementation of the BWT, by which we mean an implementation with a
+proof that the inverse-transform inverts the forwards-transform. Bird
+et~al.\cite{alg-of-prog,birdmu} provide an excellent derivation of the
+BWT and its inverse that we use as the basis of our implementation and
+proof. However, it leaves a few crucial lemmas unproven, including the
+correctness of LSD radix sort.
+
+In this paper, we develop more fully the theory of stable sorts: we
+give multiple definitions for permutations and stable permutations and
+prove them equivalent. This allows us to choose whichever definition
+makes the proof easiest when we are reasoning about stability. We also
+proved that any two sorts are interchangeable.
+\begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
+  \label{thm:stablesort_unique}
+  For all functions $f, g: \var{list}~A \to \var{list}~A$,
+  \begin{equation*}
+    \var{StableSort}~f \implies \var{StableSort}~g \implies
+    (\forall l,f~l = g~l)
+  \end{equation*}
+\end{restatable*}
+
+Using \cref{thm:stablesort_unique}, we prove the correctness of an LSD
+radix sort that relies on insertion sort to sort each column. This
+choice was crucial to the proof: parametrizing radix sort over any
+stable sort made it difficult to reason about, but reasoning about
+insertion sort using induction is very elegant.
+\cref{thm:stablesort_unique} allows us to recover the generality of
+parametrizing over all stable sorts, because insertion sort is
+equivalent to any other stable sort.
+
+Finally, we use the correctness of radix sort to reason about the BWT.
+Specifically, we implement the functions $\var{bwl} : \var{list}~A \to
+\var{list}~A$ and $\var{bwn} : \var{list}~A \to \mathbb{N}$ that produce
+the string and index part of the BWT respectively, and the function
+$\var{unbwt} : \mathbb{N} \to \var{list}~A \to \var{list}~A$ that reverses
+the transform. We then prove the following theorem:
+\begin{restatable*}[unbwt\_correct]{theorem}{unbwtcorrect}
+  \label{thm:unbwt_correct}
+  For all lists $l$,
+  \begin{equation*}
+    \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l) = l
+  \end{equation*}
+\end{restatable*}
+
+The implementation is useful in its own right as a functional
+specification for a potential verified implementation of a program
+that relies on Burrows--Wheeler, such as bzip2. But more generally,
+this provides evidence that Bird's approach of program calculation, in
+which programs are derived from their specifications via algebraic
+manipulations, is useful for proving the correctness of algorithms in
+a machine-checked setting. Bird has applied this approach to greedy
+algorithms, the Boyer-Moore algorithm, the Knuth-Morris-Pratt
+algorithm, Sudoku solvers, arithmetic coding, the Schorr-Waite
+algorithm, the Johnson-Trotter algorithm, and many more problems
+\cite{pearls}.
+
+Every definition and theorem in this paper corresponds directly to a
+definition or theorem in our Coq sources, which we provide as
+supplementary materials. Definitions are copied verbatim, modulo
+whitespace. Theorems are rephrased for clarity, with the name of the
+corresponding Coq theorem given in parentheses. The only exceptions
+are \cref{sec:opt} and places where we give hypothetical definitions
+to drive the explanation.
+
+\section{Main ideas}
+\label{sec:main_ideas}
+
+\subsection{Stable sorting}
+\label{subsec:stable_sorting}
+
+Whenver we sort a list, we do so with respect to some ordering on its
+elements. For example, we could sort a list of students by first name,
+last name, birthday, etc. More precisely, for sorting we need a total,
+decideable preorder: a \textit{preorder} is a relation $\le$ that is
+reflexive and transitive, it is \textit{total} if we can compare any
+two elements, and it is \textit{decidable} if we have a procedure that
+takes any two elements $x$ and $y$ and tells us if $x \le y$ or $x \ge y$.
+
+Given a total decideable preorder, we can specify what contract a
+function~$f$ must fulfill to be considered a sorting function.
+Intuitively, it must produce a list that is sorted, and that contains
+all of the same elements as the original list.
+\begin{definition}[Sort]
+  A function $f : \var{list}~A \to \var{list}~A$ is a \var{Sort} when
+  for all inputs $l$,
+  \begin{equation*}
+    \var{Sorted}~(f~l) \land \var{Permutation}~(f~l)~l
+  \end{equation*}
+\end{definition}
+
+Now we have to define \var{Sorted} and \var{Permutation}
+appropriately. For both properties, and for other properties we will
+discuss later, there are multiple reasonable definitions. We will
+state all the definitions we can think of, and prove equivalence
+between them. This increases our confidence that we have defined
+things correctly, and allows us to choose whichever definition we find
+most convenient for a given proof.
+
+We can define \var{Sorted} in two main ways: %TODO cite coq book and knuth
+\begin{enumerate}
+\item A list $l$ is sorted when for all
+  indices $i$ and $j$,
+\begin{equation*}
+  i \le j \implies l_i \le l_j
+\end{equation*}
+\item We can inductively define when a list $l$ is sorted, by saying
+  that the empty list is sorted, and that given a sorted lists $t$, we
+  may conclude that $h :: t$ is also sorted if $h$ is less~than or
+  equal~to every element in $t$.
+\end{enumerate}
+
+Similarly, we can define \var{Permutation} in three main ways:
+\begin{enumerate}
+\item Two lists $l$ and $l'$ are permutations of one another if every
+  $x$ occurs an equal number of times in $l$ and $l'$.
+\item Two lists $l$ and $l'$ with lengths $n$ are permutations of one
+  another when there exists a permutation function~$p$ that maps the
+  indices of $l$ to indices of $l'$, so that for every index~$i$
+  \begin{equation*}
+      l_i = l'_{p(i)}
+  \end{equation*}
+  $p$ must be a bijection so that every index of $l$ is mapped to by
+  exactly one index of $l'$.
+\item We can define inductively when two lists are permutations of one another:
+  \begin{itemize}
+  \item As a base case, two empty lists are permutations of one
+    another.
+  \item If two lists $l$ and $l'$ are permutations of one another,
+    then we can conclude that $x :: l$ and $x :: l'$ are permutations
+    of one another.
+  \item As a second base case, we can conclude that for any $x$, $y$,
+    and $l$, that $x :: y :: l$ is a permutation of $y :: x :: l$.
+  \item If $l_1$ is a permutation of $l_2$ and $l_2$ is a permutation
+    of $l_3$, then $l_1$ is a permutation of $l_3$.
+  \end{itemize}
+\end{enumerate}
+
+Now that we've defined what a sort is, we can start to consider stable
+sorts. Stable sorts must preserve the relative orders of
+``equivalent'' elements. This equivalence comes from the ordering we
+are using: any total decidable preorder also give us a decideable
+equivalence relation $\equiv$, where $x \equiv y$ when $x \le y$ and $x \ge y$.
+\begin{definition}[StableSort]
+  A function $f : \var{list}~A \to \var{list}~A$ is a \var{StableSort} when
+  for all inputs $l$,
+  \begin{equation*}
+    \var{Sorted}~(f~l) \land \var{StablePerm}~(f~l)~l
+  \end{equation*}
+\end{definition}
+
+Stable sorts still must produce sorted outputs, but the output cannot
+be just any permutation of the input. So we need a stricter notion of
+a permutation that respects equivalence, which we will call a stable
+permutation.
+
+To define stable permutations, we can tweak the each
+definitions of permutation mentioned above:
+\begin{enumerate}
+\item Two lists $l$ and $l'$ are stable permutations of one another if
+  for every $x$,
+  \begin{equation*}
+    \var{filter}~(\lambda y. y \equiv x)~l = \var{filter}~(\lambda y. y \equiv x)~l'
+  \end{equation*}
+  This is a stricter version of the first definition of permutation
+  above, because that can be rephrased as $\var{filter}~(\lambda y. y = x)~l
+  = \var{filter}~(\lambda y. y = x)~l'$.
+\item For two lists $l$ and $l'$ of length $n$ to be stable
+  permutations of one another, there must exists a \textit{stable}
+  permutation function~$p$ that relates them. That means that $p$
+  satisfies an additional stability property: for all indices $i < j$
+  where $l_{i} \equiv l_{j}$, we must have $p(i) < p(j)$.
+\item We can modify the third inference rule, which allows swapping
+  $x$ and $y$, so that you can only apply it when $x \nequiv y$: the new
+  rule says that for any $x$, $y$, and $l$ such that $x \nequiv y$ , you can
+  conclude that $x :: y :: l$ is a permutation of $y :: x :: l$.
+\end{enumerate}
+
+From these definitions, we can see that stable sorts are restricted in
+which permutation they produce. In fact, there is only one stable
+permutation that is also sorted, so any two stable sorts must produce
+the same results for every input, and are thus interchangeable.
+\stablesortunique
+
+This can be generalized a little further: even if a sorting algorithm
+itself is not a stable sorting algorithm, if the ordering we are using
+doesn't allow us to distinguish between equivalent elements, then we
+will not be able to tell if equivalent elements are being swapped.
+More precisely, notice that it isn't always true that $x \equiv y$ implies
+$x = y$. For example, if we are sorting strings by their first
+character, then $\text{\texttt{apple}} \equiv \text{\texttt{armpit}}$ under
+this ordering, but clearly $\text{\texttt{apple}} \neq
+\text{\texttt{armpit}}$. When we do have that the equivalence ($\equiv$)
+implies Leibniz equality ($=$), we say that we have an \textit{order},
+as we have when we are sorting strings using all of their characters.
+Under an order, as opposed to a preorder, we can prove that every
+permutation is a stable permutation and every sort is a stable sort,
+which implies that every sort is equivalent to every other sort.
+\begin{theorem}[all\_perm\_stable]
+  If we have an equivalence that implies Leibniz equality, then for
+  all lists $l$ and $l'$
+  \begin{equation*}
+      \var{Permutation}~l~l' \implies \var{StablePerm}~l~l'
+  \end{equation*}
+\end{theorem}
+\begin{corollary}[Sort\_Ord\_unique]
+  \label{thm:sort_ord_unique}
+  When sorting using an order, given any two function $f, g:
+  \var{list}~A \to \var{list}~A$,
+  \begin{equation*}
+    \var{Sort}~f \implies \var{Sort}~g \implies
+    (\forall l, f~l = g~l)
+  \end{equation*}
+\end{corollary}
+
+\subsection{LSD radix sort}
+\label{sec:radix_sort}
+
+LSD radix sort is a simple radix sort that is used to
+lexicographically sort a list of strings of equal length. LSD radix
+sort operates by stably sorting on each column, from the end of the
+string to the front (\cref{fig:lsd_radixsort}). The algorithm works
+because lexicographic order says that if two strings are equal up to a
+column, then the tie is broken according to the rest of the string.
+When we stably sort on a column, ties are broken in favor or the
+existing order, which was established by sorted on the following
+columns.
+
+\begin{figure}[!hb]
+  \centering
+  \begin{tt}
+    \begin{tabular}{c}
+      d a \textbf{b} \\
+      a d \textbf{d} \\
+      c a \textbf{b} \\
+      f a \textbf{d} \\
+      f e \textbf{e} \\
+      b a \textbf{d} \\
+      d a \textbf{d} \\
+      b e \textbf{e} \\
+      f e \textbf{d} \\
+      b e \textbf{d} \\
+      e b \textbf{b} \\
+      a c \textbf{e}
+    \end{tabular}
+    $\rightarrow$
+    \begin{tabular}{c}
+      d \textbf{a} b \\
+      c \textbf{a} b \\
+      e \textbf{b} b \\
+      a \textbf{d} d \\
+      f \textbf{a} d \\
+      b \textbf{a} d \\
+      d \textbf{a} d \\
+      f \textbf{e} d \\
+      b \textbf{e} d \\
+      f \textbf{e} e \\
+      b \textbf{e} e \\
+      a \textbf{c} e
+    \end{tabular}
+    $\rightarrow$
+    \begin{tabular}{c}
+      \textbf{d} a b \\
+      \textbf{c} a b \\
+      \textbf{f} a d \\
+      \textbf{b} a d \\
+      \textbf{d} a d \\
+      \textbf{e} b b \\
+      \textbf{a} c e \\
+      \textbf{a} d d \\
+      \textbf{f} e d \\
+      \textbf{b} e d \\
+      \textbf{f} e e \\
+      \textbf{b} e e
+    \end{tabular}
+    $\rightarrow$
+    \begin{tabular}{c}
+      a c e \\
+      a d d \\
+      b a d \\
+      b e d \\
+      b e e \\
+      c a b \\
+      d a b \\
+      d a d \\
+      e b b \\
+      f a d \\
+      f e d \\
+      f e e
+    \end{tabular}
+  \end{tt}
+  \caption{LSD Radixsort}
+  \label{fig:lsd_radixsort}
+\end{figure}
+
+\subsection{The Burrows-Wheeler transform}
+\label{subsec:bwt}
+
+The Burrows--Wheeler transform was originally introduced as a pass in a
+compression algorithm \cite{bw}, and continues to be used as a part of
+the popular, open-source bzip2 compression software \cite{tsai_2016}.
+Here we describe a standard compression algorithm built around BWT.
+\begin{enumerate}
+\item Apply the BWT, which tends to group create runs of identical characters.
+\item Apply Move-to-Front (MTF) encoding, in which a character is
+  represented by the number of different characters encountered since
+  the last occurrence of this character. This creates many runs of
+  zeroes from the runs in the output of BWT. On English texts, often
+  50-60\% of the values produced by the MTF pass are zeroes
+  \cite{fenwick2007}.
+\item Apply Run-Length Encoding (RLE) to encode only the runs of
+  zeroes with their lengths, using a bijective base-2 encoding
+  \cite{bw-analysis, tsai_2016}.
+\item Compress with some zeroth-order coder like a Huffman coder.
+\end{enumerate}
+
+The BWT is also at the heart of algorithms that exploit the
+relationship between the rotation matrix in the BWT and the suffix
+array data structure in order to search for patterns in a compressed
+text \cite{ferragina_index}. These techniques have been applied to
+enable fast matching of DNA subsequences \cite{dna}.
+
+To apply the BWT to a string of length $n$:
 \begin{enumerate}
 \item Form the \(n \times n\) matrix of all cyclic shifts of the
-  original string (also referred to as the rotation matrix). Starting
+  original string (referred to as the rotation matrix). Starting
   with the original string, each row is formed from the previous one
@@
 \item Sort the rows lexicographically
-\item Take the last column of the matrix. That string is the result of the transformation.
+\item Take the last column of the matrix. That string is the result of
+  the transformation.
 \item Note the index where the original string appears in the sorted
@@
 \end{enumerate}
-As shown in \Cref{fig:bw_ex}, applying the algorithm to the string
+As shown in \cref{fig:bw_ex}, applying the algorithm to the string
 \var{abracadabra!} produces the new string \var{ard!rcaaabb} and the
-index \var{3}. The transformation tends to produce runs like the three
-\var{a}s or the two \var{b}s in \var{ard!rcaaabb}, and it is this
-feature that makes the resulting string easier to compress. Although
-it is not obvious, the transformed string, together with the index,
-provide enough information to recover the the original string.
+index \var{3}.
 
@@
 
-Bird and Mu \cite{birdmu,pearls} analyze the Burrows--Wheeler algorithm
-in a functional setting. They express the algorithm as the composition
-of many smaller functions, and then derive an expression for the
-inverse transform from the specification that it inverts the forwards
-transform. In addition to providing an intuitive derivation of the
-inverse transform, this provides the outline of a proof of for the
-inverse BWT.
-
-Our main contribution is a formalization in the Coq proof assistant of
-the BWT, together with a machine-checked proof that our implementation
-is correct. Specifically, we implement the functions $\var{bwl} :
-\var{list}~A \to \var{list}~A$ and $\var{bwn} : \var{list}~A \to
-\mathbb{N}$ that produce the string and index part of the BWT
-respectively, and the function $\var{unbwt} : \mathbb{N} \to
-\var{list}~A \to \var{list}~A$ that reverses the transform. We prove the
-following theorem:
-\begin{restatable*}[unbwt\_correct]{theorem}{unbwtcorrect}
-  \label{thm:unbwt_correct}
-  For all lists $l$,
-  \begin{equation*}
-    \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l) = l
-  \end{equation*}
-\end{restatable*}
-Our implementation and proof are based on Bird et~al.'s derivations.
-
-The implementation is useful in its own right as a functional
-specification for a potential verified implementation of a program
-that relies on Burrows--Wheeler, such as bzip2\cite{appel-func-spec}.
-But more generally, this provides evidence that Bird's approach of
-program calculation, in which programs are derived from their
-specifications via algebraic manipulations, is useful for proving the
-correctness of algorithms in a machine-checked setting. Bird has
-applied this approach to greedy algorithms, the Boyer-Moore algorithm,
-the Knuth-Morris-Pratt algorithm, Sudoku solvers, arithmetic coding,
-the Schorr-Waite algorithm, the Johnson-Trotter algorithm, and many
-more problems \cite{pearls}.
-
-As a part of the proof of \Cref{thm:unbwt_correct}, we proved two
-other noteworthy results that have not, to our knowledge, been proved
-before in Coq. We implemented a least-significant-digit (LSD)
-radixsort and proved its correctness,
-\begin{restatable*}[radixsort\_correct]{theorem}{radixsortcorrect}
-  \label{thm:radixsort_correct}%
-  For all matrices $m$ with $n$ columns,
-  \begin{equation*}
-    \var{Sorted}~(\var{radixsort}~m~n) \land \var{StablePerm}~(\var{radixsort}~m~n)~m
-  \end{equation*}
-\end{restatable*}
-and we proved that any two stable sorts are interchangeable,
-\begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
-  \label{thm:stablesort_unique}
-  For all functions $f, g: \var{list}~A \to \var{list}~A$,
-  \begin{equation*}
-    \var{StableSort}~f \implies \var{StableSort}~g \implies
-    (\forall l,f~l = g~l)
-  \end{equation*}
-\end{restatable*}
-This in turn required developing suitable definitions for
-\var{Sorted}, \var{Permutation}, and \var{StablePerm}.
-
-The rest of this paper is organized as follows. \Cref{sec:bwt} details
-the history of the BWT and its applications. \Cref{sec:ord} defines
-what we need to know about a type $A$ to sort a $\var{list}~A$ as a
-part of BWT. \Cref{sec:forwards_BWT} implements the forwards BWT in
-Coq. \Cref{sec:inverse_BWT} implements the inverse BWT and proves
-\Cref{thm:unbwt_correct}. \Cref{sec:sorted_perm_stable} provides
-multiple definitions of \var{Sorted}, \var{Permutation}, and
-\var{StablePermutation}, and proves equivalence between them.
-\Cref{sec:unique} proves \Cref{thm:stablesort_unique}. \Cref{sec:ord}
-defines lexicographic orderings, need to sort the rotation matrix.
-\Cref{sec:sort} radixsort and proves \Cref{thm:radixsort_correct}.
-\var{sec:opt} explains the optimization that is most commonly applied
-to BWT that relates our implementation to the standard version of the
-algorithm.
-
-Every definition and theorem in this paper corresponds directly to a
-definition or theorem in our Coq sources, which can be found at
-\url{https://github.com/jbaum98/verified_bzip2/tree/2019-05-06}.
-Definitions are copied verbatim. Theorems are rephrased for clarity,
-with the name of the corresponding Coq theorem given in parentheses.
-The only exceptions are \var{sec:opt} and places where we give
-hypothetical definitions to provide an explanation.
-
-\section{History and applications of the BWT}
-\label{sec:bwt}
-
-The Burrows--Wheeler transform was originally introduced as a pass in a
-compression algorithm, and continues to be used as a part of the
-popular, open-source bzip2 compression software \cite{tsai_2016}.
-
-Here we describe a standard compression algorithm built around BWT.
-First the input is transformed with BWT, which tends to group together
-characters. Intuitively, the last column of the rotation matrix
-contains the characters that directly preced the characters in the
-first column of the rotation matrix. Sorting lexicographically, which
-completely sorts the first column. If there is a correlation between a
-character and the character preceding it, bringing together identical
-characters in the first column will tend to bring together identical
-characters in the last column.
-
-The next pass is a Move-to-Front (MTF) encoding, in which a character
-is represented by the number of different characters encountered since
-the last occurrence of this character. This creates many runs of zeroes
-from the runs in the output of BWT. On English texts, often 50-60\% of
-the values produced by the MTF pass are zeroes \cite{fenwick2007}.
-Often a Run-Length Encoding (RLE) pass is then used to encode only the
-runs of zeroes with their lengths, using a bijective base-2 encoding
-\cite{bw-analysis, tsai_2016}. Finally, some sort of zeroth-order
-coder like a Huffman coder is used.
-
-BWT is also at the heart of algorithms that exploit the relationship
-between the rotation matrix in BWT and the suffix array data structure
-in order to search for patterns in a compressed text
-\cite{ferragina_index}. These techniques have been applied to enable
-fast matching of DNA subsequences \cite{dna}.
-
-\section{Orders and preorders}
-\label{sec:ord}
-
-Although it is commonly demonstrated on strings, the BWT can be
-applied to any list whose elements are comparable. To make that
-precise, if we have a type $A$ and a relation on that type $\le$, we say
-that this forms a preorder on $A$ when $\le$ is reflexive and transitive, that is
-$x \le x$ for all $x:A$, and
-\begin{equation*}
-  x \le y \implies y \le z \implies x \le z
-\end{equation*}
-for all $x,y,z:A$.
-
-We say that we have a \textit{total} preorder if for any two elements
-$x$ and $y$, $x \le y \lor y \le x$.
-
-We say that we have a \textit{decidable} preorder if we have a
-decision procedure for determining whether $x \le y$ or $x \not\le y$.
-
-We represent a total, decidable preorder on type with the
-\var{Preord} typeclass in Coq:
-\begin{lstlisting}
-Class Preord (A : Type) :=
-  { le : A -> A -> Prop;
-    le_trans : forall x y z, le x y -> le y z -> le x z;
-    le_total : forall x y, le x y \/ le y x;
-    le_dec : forall x y, {le x y} + {~le x y};
-  }.
-\end{lstlisting}
-We do not explicitly require reflexivity because it follows from
-totality. Given a \var{Preord} instance for a type $A$, we will be
-able to sort $\var{list}~A$ as shown in \cref{sec:sort}.
-
-From $\le$, we can define
-\begin{align*}
-x \ge y \coloneqq y \le x && x < y \coloneqq \lnot (y \le x) && x > y \coloneqq \lnot (x \le y)
-\end{align*}
-
-We can also define
-\begin{equation*}
-  x \equiv y \coloneqq x \le y \land y \le x
-\end{equation*}
-$\equiv$ is a
-decidable equivalence relation, which means it is reflexive,
-transitive, and symmetric, and we have a decision procedure for it.
-
-The builtin class \var{EqDec} is used to describe types with decidable
-equivalence relations, so for any \var{Preord} instance we have the
-\var{EqDec} instance \var{Preord\_EqDec}. In Coq, the notation
-\var{=\hspace{0pt}=\hspace{0pt}=} is used for the equivalence relation of an \var{EqDec}
-instance and \var{=\hspace{0pt}=} for its decision procedure.
-
-We say that we have a total, decidable \textit{order} when we have
-that for all $x$ and $y$,
-\begin{equation*}
-  x \equiv y \implies x = y
-\end{equation*}
-that is, the induced equivalence relation of the preorder ($\equiv$)
-implies Leibniz equality ($=$). We use the \var{Ord} class in Coq
-for this:
-\begin{lstlisting}
-Class Ord (A : Type) `{Preord A} :=
-  { eqv_eq : forall x y, eqv x y -> x = y }.
-\end{lstlisting}
-For the BWT, we require an \var{Ord} instance for $A$, but
-elsewhere, such as in sorting, \var{Preord} will suffice.
-
-\section{The forwards BWT}
-\label{sec:forwards_BWT}
-
-First, we define the left and right rotations of a list. \var{lrot}
-moves the first element of a list to the end.
-\begin{lstlisting}
-Definition lrot (l: list A) : list A :=
-  match l with
-  | [] => []
-  | hd :: tl => tl ++ [hd]
-  end.
-\end{lstlisting}
-\var{rrot} moves the last element of a list to the front, and
-relies on \var{init}, which drops the last element of a list.
-\begin{lstlisting}
-Fixpoint init l : list A :=
-  match l with
-  | [] | [_] => []
-  | hd :: tl => hd :: init tl
-  end.
-\end{lstlisting}
-\begin{lstlisting}
-Definition rrot (l: list A) : list A :=
-  match l with
-  | [] => []
-  | hd :: tl => last l hd :: init l
-  end.
-\end{lstlisting}
-Note that in order to ensure that $\var{last}$ is total, it takes
-an extra argument to return in the case of an empty list.
-
-Now we produce the rotation matrix by repeatedly applying \var{lrot}.
-We will use \var{iter}, which takes a function $f : A \to A$, a number
-$n$, and an initial value $z$, and returns the list of intermediate
-values obtained by applying $f$ to $z$ $n$ times. Or,
-\begin{equation}
-  \var{iter}~f~n~z = [z,~f~z,~f~(f~z),~\ldots,~f^{n-1} z]
-  \label{eq:iter}
-\end{equation}
-where $f^n$ denotes repeated function application.
-\begin{lstlisting}
-Fixpoint iter (f : A -> A) (n : nat) (z : A)
-  : list A :=
-  match n with
-  | O   => []
-  | S m => z :: iter f m (f z)
-  end.
-\end{lstlisting}
-
-We can state \Cref{eq:iter} as a theorem by defining the
-function \var{rep} for repeated function application.
-\begin{lstlisting}
-Fixpoint rep (f : A -> A) (n : nat) (z : A) : A :=
-  match n with
-  | O => z
-  | S m => f (rep f m z)
-  end.
-\end{lstlisting}
-In proofs, we will still use the notation $f^i z$ for
-$\var{rep}~f~i~z$. We will also use that notation $l_i$ for the $i$th
-element of a list. In Coq, that is given by $\var{nth}~i~l~d$, where
-$d:A$ is a default value in case $i \ge \var{length}~l$.
-
-\begin{theorem}[iter\_nth]
-  For all $f$, $n$, $z$, and $i < n$,
-  \begin{equation*}
-    (\var{iter}~f~n~z)_i = f^i~z
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  The theorem follows by induction on $i$. The inductive case is
-  \begin{align*}
-       (\var{iter}~f~(n+1)~z)_{i+1}
-    &= (z :: \var{iter}~f~n~(f~z))_{i+1} \\
-    &= (\var{iter}~f~n~(f~z))_i \\
-    &= f^i~(f~z) \\
-    &= f^{i+1}~z \qedhere
-  \end{align*}
-\end{proof}
-
-Now we can define \var{rots} to produce the rotation matrix.
-\begin{lstlisting}
-Definition rots (l : list A) : list (list A) :=
-  iter lrot (length l) l.
-\end{lstlisting}
-
-\var{bwl} produces the transformed string by taking the last column of
-the sorted rotation matrix. \var{lexsort} sorts the rotation matrix
-using insertion sort (\cref{appendix:insertion_sort}) and the induced
-lexicographic ordering(\cref{sec:lex_ord}). By
-\Cref{thm:stablesort_ord} though, \var{lexsort} can be implemented
-using any sorting algorithm.
-\begin{lstlisting}
-Definition bwl (l : list A) : list A :=
-  match l with
-  | [] => []
-  | hd :: _ =>
-    List.map (fun x => last x hd) (lexsort (rots l))
-  end.
-\end{lstlisting}
-
-\var{bwn} returns the index in the sorted rotation matrix where the
-original string appears. It relies on \var{findIndex}, which is
-defined with respect to some decidable equivalence relation.
-\begin{lstlisting}
-Fixpoint findIndex (x : A) (ls : list A) : nat :=
-  match ls with
-  | [] => 0
-  | hd :: tl =>
-    if x == hd then 0
-    else S (findIndex x tl)
-  end.
-\end{lstlisting}
-
-When we use \var{findIndex} in \var{bwn}, we are using the
-induced equivalence relation from the \var{Ord} instance, so it
-finds the row that is Leibniz equal to $l$.
-\begin{lstlisting}
-Definition bwn (l : list A) : nat :=
-  findIndex l (lexsort (rots l)).
-\end{lstlisting}
-
-Given these definitions, we can quickly prove that \var{bwn}
-satisfies its specification:
-\begin{restatable*}[bwn\_correct]{theorem}{bwncorrect}
-  For every non-empty list $xs$,
-  \begin{equation*}
-    (\var{lexsort}~(\var{rots}~xs))_{(\var{bwn}~xs)} = xs
-  \end{equation*}
-\end{restatable*}
-
-First we prove a few lemmas:
-\begin{lemma}[findIndex\_correct]
-  For all elements $x$ and lists $l$,
-  \begin{equation*}
-    (\exists y \in l,~x \equiv y) \implies l_{(\var{findIndex}~x~xs)} \equiv x
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  This follows by induction on $l$: At every step, either the
-  current element is equivalent to $x$ and we return $0$, or there
-  must be such a $y$ in the rest of the list.
-\end{proof}
-
-\begin{lemma}[orig\_in\_sorted\_rots]
-  For every non-empty list $l$,
-  \begin{equation*}
-    \exists y \in \var{lexsort}~(\var{rots}~l),~x = y
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  Clearly $l$ is in $\var{rots}~l$ at index 0, and because
-  \var{lexsort} produces a permutation of its input, $l$ must be
-  in $\var{lexsort}~(\var{rots}~l)$ as well.
-\end{proof}
-
-\bwncorrect
-\begin{proof}
-  This follows from \var{findIndex\_correct} if we can show that
-  $\exists y \in xs,~x = y$, which we have by \var{orig\_in\_sorted\_rots}.
-\end{proof}
-
-\section{The inverse BWT}
-\label{sec:inverse_BWT}
-
-Now we want to construct a function \var{unbwt} such that for all $l$,
-\begin{equation}
-  \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l) = l
-  \label{eq:unbwt}
-\end{equation}
-If we could recreate the entire sorted rotation matrix from
-$\var{bwl}~l$, it would be simple to recover $l$ by indexing into
-the matrix at $\var{bwn}~l$.
-
-More precisely, suppose we had a function function \var{recreate}
-such that
-\begin{equation}
-  \label{eq:recreate}
-  \var{recreate}~(\var{bwl}~l) = \var{lexsort}~(\var{rots}~l)
-\end{equation}
-Then, we could define \var{unbwt} as
-\begin{lstlisting}
-Definition unbwt (i : nat) (l : list A) : list A :=
-  nth i (recreate l) l.
-\end{lstlisting}
-and \Cref{eq:unbwt} would follow directly from the \var{bwn\_correct}:
-\begin{align*}
-     \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l)
-  &= (\var{recreate}~(\var{bwl}~l))_{(\var{bwn})~l}\\
-  &= (\var{lexsort}~(\var{rots}~l))_{(\var{bwn})~l} \\
-  &= l
-\end{align*}
-
-How, then, do we implement \var{recreate} such that \Cref{eq:recreate}
-holds? A key insight is that because any two methods of sorting the
-rotation matrix must produce the same output, we can pretend that
-\var{lexsort} is implemented using an sorting algorithm we find
-convenient (\Cref{thm:sort_ord_unique}). Specifically, if we examine
-how an LSD radixsort sorts the rotation matrix
-(\Cref{fig:radixsort_rots}) we notice an interesting pattern. Each
-column of the rotation matrix is a permutation of the original string,
-so when we sort a on particular column, we always produce the same
-result in that column. As we move right-to-left, the column containing
-the sorted input string moves with us until it reaches the left edge
-of the matrix.
+The BWT tends to produce runs like the three \var{a}s or the two
+\var{b}s in \var{ard!rcaaabb}, and it is this feature that makes the
+resulting string easier to compress. It creates those runs by
+exploiting the correlation between adjacent characters, which means
+that by knowing one character, one can predict the character that
+follows it with some success. Sorting the rotation matrix
+lexicographically brings together rows that start with the same
+characters. But, the beginning and end of each row were adjacent in
+the original string, so this will also tend to bring together rows
+that end with the same characters.
+
+\subsection{Inverting the BWT}
+\label{subsec:bwt}
+
+Although it is not obvious, the transformed string and the index
+provide enough information to recover the the original string.
+Conceptually, we can find the original string by using the transformed
+string to reconstruct the entire sorted rotation matrix, and then
+using the index to find the row containing the original string. This,
+though, relies on our ability to reconstruct the sorted rotation
+matrix from just its last column.
+
+A key insight is that because of \cref{thm:sort_ord_unique}, we can
+pretend that during the forwards transform, the rotation matrix is
+being sorted using any sorting algorithm we find convenient.
+Specifically, if we examine how an LSD radixsort sorts the rotation
+matrix (\cref{fig:radixsort_rots}) we notice an interesting pattern.
+Each column of the rotation matrix is a permutation of the original
+string, so when we sort a on particular column, we always produce the
+same result. As we move right-to-left, the column containing the
+sorted input string moves with us until it reaches the left edge of
+the matrix.
 
@@
 
-This suggests that we can, as we had hoped, recreate the matrix from a
-single column (\Cref{fig:recreate}). But why specifically the last
-column?
+This suggests that we can recreate the matrix from the last column by
+sorting it to produce the first column, then repeatedly prepending the
+last column to the growing matrix and stably sorting by it
+(\cref{fig:recreate}. As the matrix grows from right to left, we are
+mimicing the process of LSD radix sort in \cref{fig:radixsort_rots},
+in that the sorted portions to the right of the shaded column are
+identical.
 
@@
 
-The reason is that the last column is the solution to the fixpoint equation
+The choice of the last column is crucial, because as the ``original
+string'' in the stable column sort, it acts as the tie-breaker. The
+last column is the solution to the fixpoint equation
 \begin{equation*}
@@
 
+\section{Implementing the forwards BWT}
+\label{sec:forwards_BWT}
+
+First, we define the left and right rotations of a list. \var{lrot}
+moves the first element of a list to the end.
+\begin{lstlisting}
+Definition lrot (l: list A) : list A :=
+  match l with
+  | [] => []
+  | hd :: tl => tl ++ [hd]
+  end.
+\end{lstlisting}
+\var{rrot} moves the last element of a list to the front, and
+relies on \var{init}, which drops the last element of a list.
+\begin{lstlisting}
+Definition rrot (l: list A) : list A :=
+  match l with
+  | [] => []
+  | hd :: tl => last l hd :: init l
+  end.
+\end{lstlisting}
+Note that in order to ensure that $\var{last}$ is total, it takes
+an extra argument to return in the case of an empty list.
+
+Now we produce the rotation matrix by repeatedly applying \var{lrot}.
+We will use \var{iter}, which takes a function $f : A \to A$, a number
+$n$, and an initial value $z$, and returns the list of intermediate
+values obtained by applying $f$ to $z$, $n$ times. Or,
+\begin{equation*}
+  \var{iter}~f~n~z = [z,~f~z,~f~(f~z),~\ldots,~f^{n-1} z]
+  \label{eq:iter}
+\end{equation*}
+We use the notation $f^i z$ for $\var{rep}~f~i~z$, which implements
+repeated function application. We will also use the notation $l_i$ for
+the $i$th element of a list. In Coq, that is given by
+$\var{nth}~i~l~d$, where $d:A$ is a default value in case $i \ge
+\var{length}~l$.
+
+Now we can define \var{rots} to produce the rotation matrix.
+\begin{lstlisting}
+Definition rots (l : list A) : list (list A) :=
+  iter lrot (length l) l.
+\end{lstlisting}
+
+\var{bwl} produces the transformed string by taking the last column of
+the sorted rotation matrix. \var{lexsort} sorts the rotation matrix
+using insertion sort (\cref{appendix:insertion_sort}) according to the
+induced lexicographic ordering(\cref{sec:lex_ord}).
+\begin{lstlisting}
+Definition bwl (l : list A) : list A :=
+  match l with
+  | [] => []
+  | hd :: _ =>
+    List.map (fun x => last x hd) (lexsort (rots l))
+  end.
+\end{lstlisting}
+
+\var{bwn} returns the index in the sorted rotation matrix where the
+original string appears. It relies on \var{findIndex}, which searches
+a list for an element $y$ that is equal to the given element $x$, and
+returns the index of $y$.
+\begin{lstlisting}
+Definition bwn (l : list A) : nat :=
+  findIndex l (lexsort (rots l)).
+\end{lstlisting}
+
+Given these definitions, we can quickly prove that \var{bwn}
+satisfies its specification using a few lemmas.
+\begin{lemma}[findIndex\_correct]
+  For all elements $x$ and lists $l$,
+  \begin{equation*}
+    (\exists y \in l,~x \equiv y) \implies l_{(\var{findIndex}~x~xs)} \equiv x
+  \end{equation*}
+\end{lemma}
+\begin{lemma}[orig\_in\_sorted\_rots]
+  For every non-empty list $l$,
+  \begin{equation*}
+    \exists y \in \var{lexsort}~(\var{rots}~l),~x = y
+  \end{equation*}
+\end{lemma}
+\begin{restatable}[bwn\_correct]{theorem}{bwncorrect}
+  \label{thm:bwn_correct}
+  For every non-empty list $xs$,
+  \begin{equation*}
+    (\var{lexsort}~(\var{rots}~xs))_{(\var{bwn}~xs)} = xs
+  \end{equation*}
+\end{restatable}
+
+\section{Impelementing the inverse BWT}
+\label{sec:inverse_BWT}
+
+Now we want to construct a function \var{unbwt} such that for all $l$,
+\begin{equation}
+  \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l) = l
+  \label{eq:unbwt}
+\end{equation}
+If we could recreate the entire sorted rotation matrix from
+$\var{bwl}~l$, it would be simple to recover $l$ by indexing into
+the matrix at $\var{bwn}~l$.
+
+More precisely, suppose we had a function function \var{recreate}
+such that
+\begin{equation}
+  \label{eq:recreate}
+  \var{recreate}~(\var{bwl}~l) = \var{lexsort}~(\var{rots}~l)
+\end{equation}
+Then, we could define \var{unbwt} as
+\begin{lstlisting}
+Definition unbwt (i : nat) (l : list A) : list A :=
+  nth i (recreate l) l.
+\end{lstlisting}
+and \cref{eq:unbwt} would follow directly from \cref{thm:bwn_correct}.
+
 We implement \var{recreate} as follows, introducing a parameter $j$ so
@@
 \end{lstlisting}
-This definition relies on \var{prepend\_col} and \var{zipWith}, defined below.
-\begin{lstlisting}
-Definition prepend_col
-  : list A -> list (list A) -> list (list A)
-  := zipWith cons.
-
-Fixpoint zipWith (a : list A) (b : list B)
-  : list C :=
-  match (a, b) with
-  | (ahd :: atl, bhd :: btl) =>
-    f ahd bhd :: zipWith atl btl
-  | _ => []
-  end.
-\end{lstlisting}
+This definition relies on \var{prepend\_col}, which prepends a column
+to a matrix.
 
@@
 Since we recreate the matrix column by column, we can define an
-invariant using $\var{cols}~j$, which takes the first $j$ columns of a matrix.
-\begin{lstlisting}
-Definition cols (j : nat)
-  : list (list A) -> list (list A) := map (firstn j).
-\end{lstlisting}
-
+invariant based on \cref{eq:recreate} that uses $\var{cols}~j$, which
+takes the first $j$ columns of a matrix.
 \begin{restatable*}[recreate\_correct\_inv]{theorem}{recreatecorrectinv}
   For all lists $l$ and $j \le \var{length}~l$,
-  \begin{equation*}
-    \var{recreate}~j~(\var{bwl}~l) =
-    \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
-  \end{equation*}
+  $\var{recreate}~j~(\var{bwl}~l) =
+  \var{cols}~j~(\var{lexsort}~(\var{rots}~l))$
 \end{restatable*}
 
-Assuming we can prove this theorem, we will be able to prove that our
-implementation of \var{unbwt} is correct along the same lines as before.
-
-To prove this invariant, we need to prove a few other theorems that
+If we can prove this theorem, we will be able to prove that our
+implementation of \var{unbwt} is correct along the same lines as
+before. To do so, we need to prove a few other theorems that
 have already come up when explaining the algorithm conceptually.
+
 First, we note that moving the last column of the rotation matrix to the front
@@
 matrix is symmetric.
-
 \begin{theorem}[map\_rrot\_rots]
@@
 \begin{proof}
-  By \Cref{thm:sort_ord_unique}, any two sorts are equivalent, and we
+  By \cref{thm:sort_ord_unique}, any two sorts are equivalent, and we
   can replace \var{lexsort} with an LSD radixsort implemented as
@@
 \end{proof}
-
 \begin{theorem}[cols\_map\_rrot]
@@
 
-When defining these properties, it is very important that they mean
-what we would like them to mean. If we prove our implementations
-satisfy our specifications, but our specifications are incorrect, then
-the proof is worthless. In addition, it would be nice if the
-properties are defined in a way that is elegant to reason about. So,
-we define each property in multiple ways and prove equivalence between
-them: some definitions are more obviously correct, while others might
-be easier to reason about it different situations. Even if a
-definition is never used, proving equivalence gives us confidence
-about the correctness of the definitions.
+In this section, we restate the definition outlined in
+\cref{sec:main_ideas} more formally, and prove the appropriate
+equivalences.
 
@@
 
-For a sorting function to be correct, its output must be sorted with
-respect to the ordering defined by the \var{Preord} instance. There a
-a few ways we might define what it means for a list to be sorted. In
-mathematics, we might say that a list $l$ is sorted if for all indices
-$i$ and $j$,
+\begin{definition}{SortedIx}
+  A list $l$ is sorted if for all indices $i$ and $j$,
   \begin{equation*}
@@
   \end{equation*}
-We will call this property \var{SortedIx}.
-
-However, it is often easier to reason about inductively defined
-properties, so we would like to define \var{Sorted} inductively. We
-can say that as a base case, the empty list is sorted, and given a
-sorted list, we can add an element that is smaller than all the
-existing elements and the result will also be sorted:
+\end{definition}
+\begin{definition}{Sorted}
   \begin{align*}
@@
   \end{align*}
-
-By transitivity it suffices to check that the new element is
-smaller than just the first element of the sorted list. That leads to a third definition of sorted:
+\end{definition}
+
+By transitivity it suffices to check that the new element is smaller
+than just the first element of the sorted list. That leads to a third
+definition of sorted:
+\begin{definition}{SortedLocal}
   \begin{gather*}
@@
   \end{gather*}
-
-These two inductive definitions are not obviously equivalent to the
-first definition, which is more clearly correct. We will prove all
-three equivalent.
+\end{definition}
 
@@
 \end{theorem}
-\begin{proof}
-  We reason by induction on $l$. The base cases are easy, so we
-  address the inductive case, $l = a :: l'$
-
-  ($\implies$) We have to prove that $(a :: l')_i \le (a :: l')_j$ given
-  $i \le j$ and $\var{Sorted}~l$. We reason by cases on $i$ and $j$:
-  \begin{itemize}
-  \item $i = j = 0$. By reflexivity, $a \le a$.
-  \item $i = 0,~j=j' + 1$. We need to show $a \le l'_{j'}$. We know that
-    $a \le x$ for all $x \in l'$ by $\var{Sorted}~(a::l')$, so we are
-    done.
-  \item $i = i' +1,~j = j' + 1$. We need to show $l'_{i'} \le l'_{j'}$,
-    which follows from the induction hypothesis.
-  \end{itemize}
-
-  ($\impliedby$) We have $\var{SortedIx}\ \allowbreak(a :: l')$, and
-  we need to show $\var{Sorted}\ \allowbreak(a :: l')$ given the
-  induction hypothesis $\var{SortedIx}~l' \implies \var{Sorted}~l'$.
-  By \var{SortedLocal\_iff}, we can just prove
-  $\var{SortedLocal}\ \allowbreak(a :: l')$. If $l' = []$, we can
-  apply $\var{SortedLocal\_1}$, so assume $l' = b :: l''$. To apply
-  \var{SortedLocal\_cons} we need to show
-  $\var{SortedLocal}\ \allowbreak(b :: l'')$, which we have by the
-  induction hypothesis and $\var{SortedIx}\ \allowbreak(a :: b ::
-  l'')$, and that $a \le b$. That follows because $a = (a :: b ::
-  l'')_0$ and $b = (a :: b :: l'')_1$, so by
-  $\var{SortedIx}\ \allowbreak(a :: b :: l'')$, we are done.
-\end{proof}
 
@@
 
-The Coq standard library comes with a definition of \var{Permutation}
-that holds when two lists contain exactly the same elements:
-\begin{gather*}
-  \infer[\var{perm\_nil}]{\var{Permutation}~[]~[]}{}
-  \\
-  \\
-  \infer[\var{perm\_skip}]{\var{Permutation}~(x :: l)~(x :: l')}{%
-    \var{Permutation}~l~l'
-  }
-  \\
-  \\
-  \infer[\var{perm\_swap}]{\var{Permutation}~(x :: y :: l)~(y :: x :: l)}{}
-  \\
-  \\
-  \infer[\var{perm\_trans}]{\var{Permutation}~l~l''}{
-    \var{Permutation}~l~l' & \var{Permutation}~l'~l''
-  }
-\end{gather*}
-Although we have no need to redefine it, we present here some
-alternative definitions. This is mainly because these definitions have
-analagous definitions for \var{StablePerm}, which we are definining
-ourselves.
-
 The word \textit{permutation} has two related meanings that are often
-used interchangeably. We sometimes say that two lists are permutations of each
-other if they contain all of the same elements, just appearing in a
-different order. However, in mathematics we often say that a
+used interchangeably. We sometimes say that two lists are permutations
+of each other if they contain all of the same elements, just appearing
+in a different order. However, in mathematics we often say that a
 permutation is actually the bijection on indices that defines the
@@
 each other, and that $p$ is the \textit{permutation function} that
-relates $l$ to $l'$. We say that two lists $l$ and $l'$ are
-permutations of each other is precisely when
+relates $l$ to $l'$.
+
+So, we can also say that two lists $l$ and $l'$ are permutations of
+each other is precisely when
 \begin{itemize}
@@
 \item there exists a permutation function $p : \mathbb{N}_{<n} \to
-    \mathbb{N}_{<n}$, where $p$ is a bijection, such that for all indices $i$,
-    \begin{equation*}
-      l_{p(i)} = l'_i
+  \mathbb{N}_{<n}$, where $p$ is a bijection, such that for all
+  indices $i$,
+  \begin{equation*}
+    l_i = l'_{p(i)}
   \end{equation*}
 \end{itemize}
+
+\begin{definition}{PermutationCount}
+  Two lists are permutations of one another when for any $x$,
+  \begin{equation*}
+    \var{count\_occ}~l~x = \var{count\_occ}~l'~x
+  \end{equation*}
+  where $\var{count\_occ}~l~x$ counts the number of elements in $l$
+  that are equal to $x$ under some equivalence relation.
+\end{definition}
+\var{count\_occ} counts the number of occurrences of a given variable
+using a decision procedure for Leibniz equality. Here that decision
+procedure is \texttt{equiv\_dec} from an \var{EqDec} instance where we
+require the equivalence relation itself by Leibniz equality.
 
@@
 $f : \mathbb{N}_{<n} \to A$ is represented by the $\var{list}~A$
-$[f(0),~f(1),\allowbreak\ldots,f(n-1)]$. For permutations, this is consistent with the
-notation where we denote by $2~1~0$ the permutation $p(0) = 2$, $p(1)
-= 1$, $p(2) = 0$. So, $p$ here would be represented by $[2,~1,~0]$.
-
-To restrict the range to $\mathbb{N}_{<n}$ we say that $\forall i,~i \in p
-\implies ~i<n$; to make the function surjective, we strengthen that to
-$\forall i,~i \in p \iff ~i<n$; and to make it injective we require that $p$ not
-contain any duplicate entries. The inductive property \var{NoDup} from
-the Coq standard library captures that requirement:
-\begin{align*}
-  \infer[\var{NoDup\_nil}]{\var{NoDup}~[]}{}
-  &&
-  \infer[\var{NoDup\_cons}]{\var{NoDup}~(x :: l)}{%
-    x \notin l & \var{NoDup}~l
-  }
-\end{align*}
-
-This leads to the following definition of the property
-$\var{PermFun}~n~p$, which holds when $p : \var{list}~\mathbb{N}$
-represents a permutation function on lists of length $n$:
-\begin{lstlisting}
-Definition PermFun (n : nat) (p : list nat)
-  : Prop := NoDup p /\ forall i, In i p <-> i < n.
-\end{lstlisting}
+$[f(0),~f(1),\allowbreak\ldots,f(n-1)]$. For permutations, this is
+consistent with the notation where we denote by $2~1~0$ the
+permutation $p(0) = 2$, $p(1) = 1$, $p(2) = 0$. So, $p$ here would be
+represented by $[2,~1,~0]$. To restrict the range to $\mathbb{N}_{<n}$
+we say that $\forall i,~i \in p \implies ~i<n$; to make the function
+surjective, we strengthen that to $\forall i,~i \in p \iff ~i<n$; and to make it
+injective we require that $p$ not contain any duplicate entries. The
+inductive property \var{NoDup} from the Coq standard library captures
+that requirement. This leads to the following definition:
+\begin{definition}{PermFun}
+  A list of natural numbers $p$ is a permutation function on $n$
+  elements if it contains no duplicate entries, and if for all natural
+  numbers $i$, $i \in p \iff i < n$.
+\end{definition}
+
 Then, we can define \var{apply}, which permutes a list according to a
@@
 \end{lstlisting}
-And finally, we can define \var{PermutationEx}, which holds between
-two lists when they are permutations of each other.
-\begin{lstlisting}
-Definition PermutationEx l l' : Prop :=
-  exists p, PermFun (length l) p /\ apply p l = l'.
-\end{lstlisting}
-We prove \var{PermutationEx} equivalent to \var{Permutation} in
-\Cref{thm:permutationex_iff}.
-
-For completeness, we will briefly mention a third definition of a
-permutation relation that says that two lists are permutations of one
-another when any element occurs an equal number of times in each:
-\begin{lstlisting}
-Definition PermutationCount l l' : Prop := forall x,
-  count_occ equiv_dec l x = count_occ equiv_dec l' x.
-\end{lstlisting}
-\var{count\_occ} counts the number of occurrences of a given variable
-using a decision procedure for Leibniz equality. Here that decision
-procedure is \texttt{equiv\_dec} from an \var{EqDec} instance where we
-require the equivalence relation itself by Leibniz equality.
+We will use $p(l)$ as notation for $\var{apply}~p~l$ when $l$ is a
+list.
+
+And finally, we can define a notion of when two lists are permutations
+of one another based on permutation functions.
+\begin{definition}{PermutationEx}
+  Two lists $l$ and $l'$ are permutations of one another when there
+  exists a permutation function $p$ on $\var{length}~l$ elements such
+  that $p(l) = l'$.
+\end{definition}
 
@@
 \end{theorem}
-\begin{proof}
-  The ($\implies$) direction follows from induction in the evidence of
-  $\var{Permutation}~l~l'$.
-
-  In the ($\impliedby$) direction, we reason by induction on $l$. In
-  the inductive case, we have to show $\var{Permutation}~(a::l)~l'$.
-  We know that $a \in l'$ because $\var{count\_occ}~a~l' =
-  \var{count\_occ}~a~(a::l) > 0$. That means we can deconstruct $l'$
-  as $L \dplus a :: R$, and
-  \begin{equation*}
-    L \dplus a :: R \eqperm a :: L \dplus R \eqperm a :: l
-  \end{equation*}
-  where the last step follows from the induction hypothesis.
-\end{proof}
 
@@
 
-A stable sort is a sort that doesn't swap any two elements considered
-equivalent by the ordering. This is relevant when the induced
-equivalence relation does not imply equality, so that we can
-distinguish between elements that the sort considers identical. A good
-example of this is in radixsort: when we sort on a single column, the
-sort can't tell the difference between strings that contain equal
-characters in that column, but it is crucial that it respect the
-existing ordering in cases of a tie.
-
-For stable sorts, we need a stronger property than \var{Permutation}.
-We can extend our mathematical definition. We say that two lists $l$
-and $l'$ are stable permutations of each other when we have a
-permutation $p$ that relates them, and that satisfies an additional
-stability property: for all indices $i$ and $j$ where
-$(\var{apply}~p~l)_i \equiv (\var{apply}~p~l)_j$
-\begin{gather*}
-  i < j \implies p(i) < p(j)
-\end{gather*}
-Equivalently, we can say that for all indices $i$ and $j$ where $l_j \equiv
-l_j$ we must have
-\begin{gather*}
-  i < j \implies p^{-1}(i) < p^{-1}(j)
-\end{gather*}
-
-Based on this, we define a stronger version of the \var{PermFun}
-property, \var{StablePermFun}. However, while $\var{PermFun}~n~p$
-means that $p$ represents a permutation function on lists of length
-$n$, we need to specify the list with respect to which $p$ is stable,
-so we say $\var{StablePermFun}~l~p$ for some list $l$.
-\begin{lstlisting}
-Definition StablePermFun (l : list A) (p : list nat)
-  := PermFun (length l) p /\ forall i j d,
-    nth i (apply p l) d === nth j (apply p l) d ->
-    i < j < length l -> image p i < image p j.
-\end{lstlisting}
-As above, we can also define this
-property using the inverse permutation:
-\begin{lstlisting}
-Definition StablePermFun_preimage
-  (l : list A) (p : list nat) :=
-  PermFun (length l) p /\ forall i j d,
-    nth i l d === nth j l d ->
-    i < j < length l -> preimage p i < preimage p j.
-\end{lstlisting}
-
-As usual, we will prove these
-definitions equivalent. First, though, we need to define \var{image}
-and \var{preimage}.
-\begin{lstlisting}
-Definition image (p : list nat) i := nth i p 0.
-Definition preimage (p : list nat) i := findIndex i p.
-\end{lstlisting}
-We will not prove it here, but $\var{preimage}~p$ and $\var{image}~p$
-are both of bijections of type $\mathbb{N}_{<n} \to \mathbb{N}_{<n}$ and
-are inverses of each other. We also have the following two theorems:
-\begin{theorem}[nth\_preimage\_apply, nth\_image\_apply]
-  For all lists $l$, permutations $p$, and indices $i$,
-  \begin{align*}
-    (\var{apply}~p~l)_{p^{-1}(i)} = l_i && l_{p(i)} = (\var{apply}~p~l)_i
-  \end{align*}
-\end{theorem}
-\begin{proof}
-  The first equation (\var{nth\_preimage\_apply}) holds because:
-  \begin{align*}
-       (\var{apply}~p~l)_{p^{-1}(i)}
-    &= (\var{map}~(\lambda j.~l_j)~p)_{\var{findIndex}~i~p} \\
-    &= l_{p_{(\var{findIndex}~i~p)}} \\
-    &= l_i
-  \end{align*}
-  and the second follows from this:
-  \begin{equation*}
-      (\var{apply}~p~l)_i
-    = (\var{apply}~p~l)_{p^{-1}(p(i))}
-    = l_{p(i)}
-    \qedhere
-  \end{equation*}
-\end{proof}
-
-\begin{theorem}[StablePermFun\_iff]
-  For all lists $l$ and permutation $p$,
-  \begin{equation*}
-    \var{StablePermFun}~l~p \iff \var{StablePermFun\_preimage}~l~p
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  We will only prove the $\implies$ direction; the other direction is
-  very similar. We have that $l_i \equiv l_j$, and that $i < j$, and we
-  have to show that $p^{-1}(i) < p^{-1}(j)$. We will show this by
-  proving that $p^{-1}(i) \ge p^{-1}(j)$ leads to a contradiction.
-
-  Suppose $p^{-1}(i) \ge p^{-1}(j)$. We actually know that $p^{-1}(i) >
-  p^{-1}(j)$ because $i \neq j$. From $l_i \equiv l_j$ and
-  $\var{StablePermFun}~l~p$, along with $p^{-1}(j) < p^{-1}(i)$, we
-  know that $p(p^{-1}(j)) < p(p^{-1}(i))$. However, this implies $j <
-  i$, which is a contradiction. Therefore, $p^{-1}(i) < p^{-1}(j)$.
-\end{proof}
-
-Now, we can define \var{StablePermEx} to describe when two lists are
-stable permutations of each other:
-\begin{lstlisting}
-Definition StablePermEx l l' :=
-  exists p, StablePermFun l p /\ apply p l = l'.
-\end{lstlisting}
-
-While \var{StablePermEx} corresponds to \var{PermutationEx}, how can
-we modify \var{Permutation} to create an inductive relation for stable
-lists? Intuitively, the problem is \var{Permutation\_swap}: we can
-only swap two elements and preserve stability if they are not
-equivalent. With this fix, we have \var{StablePermInd}:
+\begin{definition}{StablePerm}
+  Two lists are stable permutations of one aonther if for any $x$,
+  \begin{equation*}
+    \var{filter}~(\lambda y. y \equiv x)~l = \var{filter}~(\lambda y. y \equiv x)~l'
+  \end{equation*}
+\end{definition}
+
+\begin{definition}{StablePermFun\_preimage}
+  A permutation function $p$ is a stable permutation function on some
+  $l : \var{list}~A$ when for all $i < j$ such that $l_i \equiv l_j$, we
+  also have $p(i) < p(j)$.
+\end{definition}
+\begin{definition}{PermutationEx}
+  Two lists $l$ and $l'$ are stable permutations of one another when
+  there exists a stable permutation function~$p$ on $l$ such that
+  $p(l) = l'$.
+\end{definition}
+
+\begin{definition}{StablePermInd}
   \begin{gather*}
@@
   \end{gather*}
-
-Unfortunately, neither of these two definitions will be particularly
-useful for proving the theorems we would like to prove with them. The
-best definition for stability does not come by analogy with
-\var{Permutation}. The intuition is that for any two lists that are
-stable permutations of each other, if we filter each list by
-equivalence with any element, the resulting lists should be identical.
-This is because they must have the same elements, since the original
-lists are permutations of one another, and those elements must be in
-the same order, since the original lists are specifically stable
-permutations of one another \cite{leroy_2018}.
-\begin{lstlisting}
-Definition StablePerm (l l': list A) : Prop := forall x,
-  filter (equiv_decb x) l = filter (equiv_decb x) l'.
-\end{lstlisting}
-We prove equivalence between these three definitions in
-\cref{appendix:stableperm_def_eq}.
+\end{definition}
 
@@
   given $\var{Sorted}~(h :: t)$, $\var{Sorted}~(h' :: t')$, and
-  $\var{StablePerm}~(h :: t)~(h' :: t')$.
-
-  First, we can say that $h \equiv h'$: we know that $h \in h' :: t'$ and $h'
-  \in h :: t$. If $h = h'$, then we are done, so assuming $h \neq h'$ we
-  know $h \in t'$ and $h' \in t$. Because both $h::t$ and $h'::t'$ are
-  sorted, we know that $h \le h'$ and $h' \le h$, implying $h \equiv h'$.
-
-  This means that $h = h'$: from $\var{StablePerm}~(h :: t)~(h' ::
-  t')$ we have
+  $\var{StablePerm}~(h :: t)~(h' :: t')$, as well as $t = t'$ from the
+  induction hypothesis. So it suffices to show $h = h'$.
+
+  First, we can prove that $h \equiv h'$. We know that $h \in h' :: t'$ and
+  $h' \in h :: t$ because they are permutations of one another. If $h =
+  h'$, then we are done, so assuming $h \neq h'$ we know $h \in t'$ and $h'
+  \in t$. Because both $h::t$ and $h'::t'$ are sorted, we know that $h \le
+  h'$ and $h' \le h$, implying $h \equiv h'$.
+
+  From this we can prove that $h = h'$. From $\var{StablePerm}~(h ::
+  t)~(h' :: t')$ we have
   \begin{align*}
@@
   \end{align*}
-  We also have $t = t'$ by the induction hypothesis, so we are done.
+  We also have $t = t'$ by the induction hypothesis, so we can
+  conclude that $h = h'$.
 \end{proof}
@@
 \begin{theorem}[all\_perm\_stable]
-  If we have that equivalence implies Leibniz equality, then for all
-  lists $l$ and $l'$
+  If we have an equivalence that implies Leibniz equality, then for
+  all lists $l$ and $l'$
   \begin{equation*}
@@
 \begin{proof}
-  This follows from \Cref{thm:stablesort_ord} and \Cref{thm:stablesort_unique}.
+  This follows from \cref{thm:stablesort_ord} and \cref{thm:stablesort_unique}.
 \end{proof}
@@
 Given a preorder on a type $A$, we can define a lexicographic preorder
-on $\var{list}~A$. Intuitively, we want to say that to compare two
-lists, we compare their elements from left to right. As soon as we
-encounter two elements that are not equivalent, their relationship
-defines the relationship between the two lists. If we make it all the
-way through one of the lists and all the elements are equivalent, we
-say that the shorter list is smaller. If they are the same length and
-all the elements are equivalent, then the lists are equivalent. To
-capture this, we define the inductive relation \var{lex\_le}, which we
-will denote by $\lexle$ in proofs.
+on $\var{list}~A$. To capture this, we define the inductive relation
+\var{lex\_le}, which we will denote by $\lexle$ in proofs.
 \begin{gather*}
@@
 \end{proof}
-
-\section{Radixsort}
-\label{sec:sort}
-
-In a least-significant-digit radix sort, we lexicographically sort a
-list of strings of equal length by stably sorting on each column, from
-the end of the string to the front (\Cref{fig:lsd_radixsort}). This
-works because the lexicographic order says that if two strings are
-equal up to a column then the tie is broken in according to the rest
-of the string. This is reflected in the fact that when we stably sort
-on a column, ties are broken in favor or the existing order, which was
-established by sorted on the following columns.
-
-\begin{figure}[!hb]
-  \centering
-  \begin{tt}
-    \begin{tabular}{c}
-      d a \textbf{b} \\
-      a d \textbf{d} \\
-      c a \textbf{b} \\
-      f a \textbf{d} \\
-      f e \textbf{e} \\
-      b a \textbf{d} \\
-      d a \textbf{d} \\
-      b e \textbf{e} \\
-      f e \textbf{d} \\
-      b e \textbf{d} \\
-      e b \textbf{b} \\
-      a c \textbf{e}
-    \end{tabular}
-    $\rightarrow$
-    \begin{tabular}{c}
-      d \textbf{a} b \\
-      c \textbf{a} b \\
-      e \textbf{b} b \\
-      a \textbf{d} d \\
-      f \textbf{a} d \\
-      b \textbf{a} d \\
-      d \textbf{a} d \\
-      f \textbf{e} d \\
-      b \textbf{e} d \\
-      f \textbf{e} e \\
-      b \textbf{e} e \\
-      a \textbf{c} e
-    \end{tabular}
-    $\rightarrow$
-    \begin{tabular}{c}
-      \textbf{d} a b \\
-      \textbf{c} a b \\
-      \textbf{f} a d \\
-      \textbf{b} a d \\
-      \textbf{d} a d \\
-      \textbf{e} b b \\
-      \textbf{a} c e \\
-      \textbf{a} d d \\
-      \textbf{f} e d \\
-      \textbf{b} e d \\
-      \textbf{f} e e \\
-      \textbf{b} e e
-    \end{tabular}
-    $\rightarrow$
-    \begin{tabular}{c}
-      a c e \\
-      a d d \\
-      b a d \\
-      b e d \\
-      b e e \\
-      c a b \\
-      d a b \\
-      d a d \\
-      e b b \\
-      f a d \\
-      f e d \\
-      f e e
-    \end{tabular}
-  \end{tt}
-  \caption{LSD Radixsort}
-  \label{fig:lsd_radixsort}
-\end{figure}
 
@@
 
-\radixsortcorrect
+\begin{theorem}[radixsort\_correct]
+  \label{thm:radixsort_correct}%
+  For all matrices $m$ with $n$ columns,
+  \begin{equation*}
+    \var{Sorted}~(\var{radixsort}~m~n) \land \var{StablePerm}~(\var{radixsort}~m~n)~m
+  \end{equation*}
+\end{theorem}
 \begin{proof}
@@
 
-It is important to note that standard implementations of the BWT do
-not actually reconstruct the entire rotation matrix, as this would be
-inefficient. Instead, they avoid this using the optimization described
-in this section. Although, we did not prove or implement this
-optimization in Coq, it would be straightforward to do so.
+Standard implementations of the BWT do not actually reconstruct the
+entire rotation matrix, as this would be inefficient. Instead, they
+avoid this using the optimization described in this section. Although,
+we did not prove or implement this optimization in Coq, it would be
+straightforward to do so.
 
@@
 We have implemented and proved that the the inverse BWT inverts the
-forwards BWT, basing our work on Bird et al.'s program calculation
+forwards BWT, basing our work on Bird et~al.'s program calculation
 derivation \cite{birdmu,pearls}. In the process, we defined the
@@
 that it uses insertion sort as the inner sort, but
-\Cref{thm:stablesort_unique} allows us to extend that proof to a
+\cref{thm:stablesort_unique} allows us to extend that proof to a
 radixsort that uses any stable sort. We also use it as the
 implementation of \var{lexsort} because it is convenient, but
-similarly \Cref{thm:sort_ord_unique} ensures that it can be replaced with
+similarly \cref{thm:sort_ord_unique} ensures that it can be replaced with
 any other sort.
@@
   stable as well. The $\impliedby$ direction follows from
-  \Cref{thm:apply_correct_stable}.
+  \cref{thm:apply_correct_stable}.
 \end{proof}
@@
 we can think of it as permuting the original indices $\iota =
-[0,~1,~\ldots,~n-1]$. If we rephrase \Cref{eq:compose_apply} the effect on
+[0,~1,~\ldots,~n-1]$. If we rephrase \cref{eq:compose_apply} the effect on
 on $I$, we get
@@
   Because $p_2 \comp p_1 = p_2~p_1$, we can apply
-  \Cref{thm:Permutation_combine_eq} to show $(p_2 \comp p_1)~l \eqperm
+  \cref{thm:Permutation_combine_eq} to show $(p_2 \comp p_1)~l \eqperm
   p_2~(p_1~l)$

### 2019-10-16 draft_paper/draft_paper.tex
stash

@@
 is the Burrows--Wheeler transform (BWT), which has applications in
-compression, full-text indexing and pattern searching.
+compression, full-text indexing, and pattern searching.
 
 There has not been much work on formally verifying stable sorts, or
-rather, the stability of stable sorts. While Bird
-et~al.\cite{birdmu,pearls} provide an elegant derivation and proof for
-the BWT, formalizing it in Coq reveals that it relies on a few lemmas
-that are not trivial to prove, including the correctness of
-least-significant digit (LSD) radix sort. In this paper, we develop
-more fully the theory of stable sorts, and use this theory to prove
-the correctness of LSD radix sort and thereby the BWT.
+rather, the stability of stable sorts. While Bird et~al.
+\cite{birdmu,pearls} provide an elegant derivation and proof for the
+BWT, formalizing it in Coq reveals that they rely on a few lemmas that
+are not trivial to prove, including the correctness of
+least-significant digit (LSD) radix sort. We develop the theory of
+stable sorts, and use this theory to prove the correctness of LSD
+radix sort and thereby the BWT.
 \end{abstract}
@@
 to sort data lexicographically, and can be more efficient than
-comparison-based sorting methods, both in theory and in practice.
-Radix sorts are especially well-suited to sorting strings
-\cite{McIlroy93}. Many radix sorts rely on a some stable sorting
-subroutine.
+comparison-based sorting methods, both in theory and in practice
+\cite{McIlroy93,Bentley:1997:FAS:314161.314321,10.1007/978-3-540-89097-3_3}.
+Many radix sorts rely on a some stable sorting subroutine.
 
@@
 et~al.\cite{alg-of-prog,birdmu} provide an excellent derivation of the
-BWT and its inverse that we use as the basis of our implementation and
-proof. However, it leaves a few crucial lemmas unproven, including the
-correctness of LSD radix sort.
+BWT and its inverse based on the technique program calculation, in
+which programs are calculated from their specifications using
+equational reasoning. However, their derivation leaves a few crucial
+lemmas unproven, including the correctness of LSD radix sort.
 
@@
 proved that any two sorts are interchangeable.
-\begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
+\begin{restatable*}[StableSort\_unique]{corollary}{stablesortunique}
   \label{thm:stablesort_unique}
-  For all functions $f, g: \var{list}~A \to \var{list}~A$,
-  \begin{equation*}
-    \var{StableSort}~f \implies \var{StableSort}~g \implies
+  For all stable sorting functions $f, g: \var{list}~A \to
+  \var{list}~A$,
+  \begin{equation*}
     (\forall l,f~l = g~l)
@@
 stable sort made it difficult to reason about, but reasoning about
-insertion sort using induction is very elegant.
-\cref{thm:stablesort_unique} allows us to recover the generality of
-parametrizing over all stable sorts, because insertion sort is
-equivalent to any other stable sort.
+insertion sort is very elegant. \cref{thm:stablesort_unique} allows us
+to recover the generality of parametrizing over all stable sorts,
+because it proves that insertion sort is equivalent to any other
+stable sort.
 
@@
 Specifically, we implement the functions $\var{bwl} : \var{list}~A \to
-\var{list}~A$ and $\var{bwn} : \var{list}~A \to \mathbb{N}$ that produce
+\var{list}~A$ and $\var{bwn} : \var{list}~A \to \nat$ that produce
 the string and index part of the BWT respectively, and the function
-$\var{unbwt} : \mathbb{N} \to \var{list}~A \to \var{list}~A$ that reverses
+$\var{unbwt} : \nat \to \var{list}~A \to \var{list}~A$ that reverses
 the transform. We then prove the following theorem:
@@
 
-\section{Main ideas}
-\label{sec:main_ideas}
-
-\subsection{Stable sorting}
-\label{subsec:stable_sorting}
-
-Whenver we sort a list, we do so with respect to some ordering on its
+\section{Sorting}
+\label{sec:sorting}
+
+Whenever we sort a list, we do so with respect to some ordering on its
 elements. For example, we could sort a list of students by first name,
@@
 
-Given a total decideable preorder, we can specify what contract a
-function~$f$ must fulfill to be considered a sorting function.
-Intuitively, it must produce a list that is sorted, and that contains
-all of the same elements as the original list.
+Given a total decideable preorder, we can define when a list is in
+ascending order with respect to that preorder. In this case, as will
+be the case with other properties we will discuss, there are multiple
+reasonable definitions. Our approach will be to state all the
+definitions we can think of and to prove equivalence between them.
+This increases our confidence that we have defined things correctly,
+and allows us to choose whichever definition we find most convenient
+for a given proof.
+
+We can define \var{Sorted} in two main ways: an index-based definition
+(\var{SortedIx}), and an inductive definition (\var{Sorted}).
+\begin{definition}{SortedIx}
+  A list $l$ is sorted if for all indices $i$ and $j$,
+  \begin{equation*}
+    i \le j \implies l_i \le l_j
+  \end{equation*}
+\end{definition}
+
+\begin{definition}{Sorted}
+  \begin{align*}
+    \infer[\var{Sorted\_nil}]{\var{Sorted}~[]}{}
+    &&
+    \infer[\var{Sorted\_cons}]{\var{Sorted}~(a :: l)}{%
+      (\forall~x \in l,~a < x) & \var{Sorted}~l
+    }
+  \end{align*}
+\end{definition}
+
+\begin{restatable}[SortedIx\_iff]{theorem}{sortedixiff}
+  For all lists $l$,
+  \begin{equation*}
+    \var{Sorted}~l \iff \var{SortedIx}~l
+  \end{equation*}
+\end{restatable}
+\begin{proof}
+  See \cref{appendix:proofs}.
+\end{proof}
+
+However, a sorting function cannot produce just any sorted list as its
+output; the sorted list must also contain all the same elements as the
+original list. To capture this, we could use the \var{Permutation}
+relation defined in the Coq standard library. This relation holds on
+two lists when they contain all the same elements, and is defined
+inductively:
+\begin{definition}[Permutation]
+  \label{def:permutation}
+  \begin{gather*}
+    \infer[\var{perm\_nil}]{\var{Permutation}~[]~[]}{}
+    \\
+    \allowbreak
+    \\
+    \infer[\var{perm\_skip}]{\var{Permutation}~(x :: l)~(x :: l')}{%
+      \var{Permutation}~l~l'
+    }
+    \\
+    \allowbreak
+    \\
+    \infer[\var{perm\_swap}]{\var{Permutation}~(x :: y :: l)~(y :: x :: l)}{}
+    \\
+    \allowbreak
+    \\
+    \infer[\var{perm\_trans}]{\var{Permutation}~l~l''}{
+      \var{Permutation}~l~l' & \var{Permutation}~l'~l''
+    }
+  \end{gather*}
+\end{definition}
+
+Before we go further, we should clarify what we mean by the word
+\textit{permutation}, as it has two related meanings that are often
+used interchangeably. We sometimes say that two lists are permutations
+of one other if they contain all of the same elements, perhaps in a
+different order, like the lists $l =
+[\var{zebra},\ \allowbreak\var{appliance},\ \allowbreak\var{kite}]$
+and $l' =
+[\var{appliance},\ \allowbreak\var{kite},\ \allowbreak\var{zebra}]$
+However, in mathematics we sometimes say that a permutation is
+actually the bijection on the indices that defines the correspondence
+between the two lists. In this case, we could map indices in $l$ to
+indices in $l'$ by mapping $0 \mapsto 2$, mapping $1 \mapsto 0$, and mapping $2 \mapsto
+0$. Formally, we say that $p$ is the permutation that relates $l$ to
+$l'$ if for all indices $i$ and $j$,
+\begin{equation*}
+  l_i = l'_{p(i)}
+\end{equation*}
+For consistency with Coq's built-in \var{Permuation} relation, in a
+situation like this we will say that $l$ and $l'$ are permutations of
+one other, and that $p$ is the \textit{permutation function} that
+relates $l$ to $l'$.
+
+This suggests an alternative definition for \var{Permutation}:
+\begin{definition}{PermutationEx}
+  \label{def:permutation_ex}
+  Two lists $l$ and $l'$ are permutations of one another when, letting
+  $n = \var{length}~l$, there exists a permutation function $p$ on $n$
+  elements such that $p(l) = l'$.
+\end{definition}
+Here we are overloading the notation $p(l)$ to denote the result of
+shuffling a list~$l$ according to a permutation function~$p$. To
+define this precisely, we have to first define how we will represent
+permutation functions in Coq.
+
+A permutation function on $n$ elements maps indices to indices, and so
+has type $\nat_{<n} \to \nat_{<n}$. In Coq, we can represent
+a finite function as a list of its outputs: $f : \nat_{<n} \to A$
+is represented by the list $[f(0),~f(1),\allowbreak\ldots,~f(n-1)]$.
+Applied to permutations, this means that the permutation~$p$ mentioned
+above would be represented by $[2,~1,~0]$. This is consistent with
+``one-line'' notation for permutations, in which a permutation on $n$
+elements is represented by its result when applied to the
+list~$[0,~2,\ldots,n-1]$. Now we can make precise the notation
+$p(l)$\footnote{In Coq, this corresponds the the \var{apply} function,
+  whose definition is slightly more complex due to the extra default
+  parameter for \var{nth}.}
+\begin{equation*}
+  p(l) \coloneq \var{map}~(\lambda i. l_i)~p
+\end{equation*}
+
+However, not any list of natural numbers represents a valid
+permutation function, because a permutation must be a bijection. We
+can define the property $\var{PermFun}~n~p$, which holds when $p :
+\var{list}~\nat$ represents a permutation function on lists of
+length $n$:
+\begin{definition}[PermFun]
+  A list~$p$ of natural numbers is a permutation function on $n$
+  elements if for all natural numbers $i$, $i \in p \iff i < n$, and if
+  $p$ contains no duplicate elements\footnote{For this we can use the
+    \var{NoDup} predicate from Coq's standard library.}.
+\end{definition}
+
+We should also mention a third definition for permutations that can be
+used on types with decideable equality:
+\begin{definition}[PermutationCount]
+  \label{def:permutation_count}
+  Two lists are permutations of one another when for any $x$,
+  \begin{equation*}
+    \var{count\_occ}~l~x = \var{count\_occ}~l'~x
+  \end{equation*}
+  where $\var{count\_occ}~l~x$ counts the number of elements in $l$
+  that are equal to $x$.
+\end{definition}
+
+With this we can define the specification for sorting functions:
 \begin{definition}[Sort]
-  A function $f : \var{list}~A \to \var{list}~A$ is a \var{Sort} when
-  for all inputs $l$,
+  A function $f : \var{list}~A \to \var{list}~A$ is a sorting function
+  when for all inputs $l$,
   \begin{equation*}
@@
 
-Now we have to define \var{Sorted} and \var{Permutation}
-appropriately. For both properties, and for other properties we will
-discuss later, there are multiple reasonable definitions. We will
-state all the definitions we can think of, and prove equivalence
-between them. This increases our confidence that we have defined
-things correctly, and allows us to choose whichever definition we find
-most convenient for a given proof.
-
-We can define \var{Sorted} in two main ways: %TODO cite coq book and knuth
-\begin{enumerate}
-\item A list $l$ is sorted when for all
-  indices $i$ and $j$,
-\begin{equation*}
-  i \le j \implies l_i \le l_j
-\end{equation*}
-\item We can inductively define when a list $l$ is sorted, by saying
-  that the empty list is sorted, and that given a sorted lists $t$, we
-  may conclude that $h :: t$ is also sorted if $h$ is less~than or
-  equal~to every element in $t$.
-\end{enumerate}
-
-Similarly, we can define \var{Permutation} in three main ways:
-\begin{enumerate}
-\item Two lists $l$ and $l'$ are permutations of one another if every
-  $x$ occurs an equal number of times in $l$ and $l'$.
-\item Two lists $l$ and $l'$ with lengths $n$ are permutations of one
-  another when there exists a permutation function~$p$ that maps the
-  indices of $l$ to indices of $l'$, so that for every index~$i$
-  \begin{equation*}
-      l_i = l'_{p(i)}
-  \end{equation*}
-  $p$ must be a bijection so that every index of $l$ is mapped to by
-  exactly one index of $l'$.
-\item We can define inductively when two lists are permutations of one another:
-  \begin{itemize}
-  \item As a base case, two empty lists are permutations of one
-    another.
-  \item If two lists $l$ and $l'$ are permutations of one another,
-    then we can conclude that $x :: l$ and $x :: l'$ are permutations
-    of one another.
-  \item As a second base case, we can conclude that for any $x$, $y$,
-    and $l$, that $x :: y :: l$ is a permutation of $y :: x :: l$.
-  \item If $l_1$ is a permutation of $l_2$ and $l_2$ is a permutation
-    of $l_3$, then $l_1$ is a permutation of $l_3$.
-  \end{itemize}
-\end{enumerate}
-
-Now that we've defined what a sort is, we can start to consider stable
-sorts. Stable sorts must preserve the relative orders of
-``equivalent'' elements. This equivalence comes from the ordering we
-are using: any total decidable preorder also give us a decideable
-equivalence relation $\equiv$, where $x \equiv y$ when $x \le y$ and $x \ge y$.
-\begin{definition}[StableSort]
-  A function $f : \var{list}~A \to \var{list}~A$ is a \var{StableSort} when
-  for all inputs $l$,
-  \begin{equation*}
-    \var{Sorted}~(f~l) \land \var{StablePerm}~(f~l)~l
-  \end{equation*}
-\end{definition}
+\section{Stable sorting}
+\label{sec:stable_sorting}
+
+Now that we've defined sorts, we can start to consider stable sorts.
+Stable sorts must preserve the relative orders of ``equivalent''
+elements. This equivalence comes from the ordering we are using: any
+total decidable preorder also give us a decideable equivalence
+relation $\equiv$, where $x \equiv y$ when $x \le y$ and $x \ge y$.
 
 Stable sorts still must produce sorted outputs, but the output cannot
-be just any permutation of the input. So we need a stricter notion of
+be just any permutation of the input. We need a stricter notion of
 a permutation that respects equivalence, which we will call a stable
 permutation.
-
-To define stable permutations, we can tweak the each
-definitions of permutation mentioned above:
-\begin{enumerate}
+\begin{definition}[StableSort]
+  A function $f : \var{list}~A \to \var{list}~A$ is a stable sorting
+  function when for all inputs $l$,
+  \begin{equation*}
+    \var{Sorted}~(f~l) \land \var{StablePerm}~(f~l)~l
+  \end{equation*}
+\end{definition}
+
+To define stable permutations, we can tweak each of the definitions of
+permutation mentioned above. For example, we can modify our
+function-based definition.
+
+\begin{definition}[StablePermEx]
+  Two lists $l$ and $l'$ are stable permutations of one another when
+  there exists a \textit{stable} permutation function~$p$ on $l$ such
+  that $p(l) = l'$.
+\end{definition}
+
+\begin{definition}[StablePermFun]
+  A permutation function $p$ is a \textit{stable} permutation function
+  on some list $l$ when for all indices $i < j$ where $l_i \equiv l_j$, we
+  also have $p(i) < p(j)$.
+\end{definition}
+
+We can also modify the inductive definition of permutations to respect
+equivalence. The problematic rule is \var{perm\_swap} because we could
+porentially swap two equivalent elements. We will restrict it, so that
+you can only swap $x$ and $y$ when $x \nequiv y$.
+
+\begin{definition}[StablePermInd]
+  \begin{gather*}
+    \infer[\var{StablePermInd\_nil}]{\var{StablePermInd}~[]~[]}{}
+    \\ \\
+    \infer[\var{StablePermInd\_skip}]{\var{StablePermInd}~(x :: l)~(x :: l')}{%
+      \var{StablePermInd}~l~l'
+    }
+    \\ \\
+    \infer[\var{StablePermInd\_swap}]
+          {\var{StablePermInd}~(x :: y :: l)~(y :: x :: l)}
+          {x \not\equiv y}
+    \\ \\
+    \infer[\var{StablePermInd\_trans}]{\var{StablePermInd}~l~l''}{
+      \var{StablePermInd}~l~l' & \var{StablePermInd}~l'~l''
+    }
+  \end{gather*}
+\end{definition}
+
+Finally, we can say that if two lists are in a stable permutation, and
+we examine only the the elements equivalent to $x$ for some $x$, then
+those elements should appear in the same order in both lists.
+\begin{definition}[StablePerm]
 \item Two lists $l$ and $l'$ are stable permutations of one another if
@@
   \end{equation*}
-  This is a stricter version of the first definition of permutation
-  above, because that can be rephrased as $\var{filter}~(\lambda y. y = x)~l
-  = \var{filter}~(\lambda y. y = x)~l'$.
-\item For two lists $l$ and $l'$ of length $n$ to be stable
-  permutations of one another, there must exists a \textit{stable}
-  permutation function~$p$ that relates them. That means that $p$
-  satisfies an additional stability property: for all indices $i < j$
-  where $l_{i} \equiv l_{j}$, we must have $p(i) < p(j)$.
-\item We can modify the third inference rule, which allows swapping
-  $x$ and $y$, so that you can only apply it when $x \nequiv y$: the new
-  rule says that for any $x$, $y$, and $l$ such that $x \nequiv y$ , you can
-  conclude that $x :: y :: l$ is a permutation of $y :: x :: l$.
-\end{enumerate}
+\end{definition}
+In a sense, this is a stricter version of
+\cref{def:permutation_count}, because that can be rephrased as
+$\var{filter}~(\lambda y. y = x)~l = \var{filter}~(\lambda y. y = x)~l'$.
 
@@
 which permutation they produce. In fact, there is only one stable
-permutation that is also sorted, so any two stable sorts must produce
-the same results for every input, and are thus interchangeable.
-\stablesortunique
+permutation that is also sorted.
+\begin{theorem}[StablePerm\_Sorted\_eq]
+  For all lists $l$,
+  \begin{equation*}
+    \var{Sorted}~l \implies \var{Sorted}~l' \implies
+    \var{StablePerm}~l~l' \implies l = l'.
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We reason by induction on $l$. In the inductive case, we have to
+  show that
+  \begin{equation*}
+    h :: t = h' :: t'
+  \end{equation*}
+  given $\var{Sorted}~(h :: t)$, $\var{Sorted}~(h' :: t')$, and
+  $\var{StablePerm}~(h :: t)~(h' :: t')$, as well as $t = t'$ from the
+  induction hypothesis. So it suffices to show $h = h'$.
+
+  First, we can prove that $h \equiv h'$. We know that $h \in h' :: t'$ and
+  $h' \in h :: t$ because they are permutations of one another. If $h =
+  h'$, then we are done, so assuming $h \neq h'$ we know $h \in t'$ and $h'
+  \in t$. Because both $h::t$ and $h'::t'$ are sorted, we know that $h \le
+  h'$ and $h' \le h$, implying $h \equiv h'$.
+
+  From this we can prove that $h = h'$. From $\var{StablePerm}~(h ::
+  t)~(h' :: t')$ we have
+  \begin{align*}
+    \var{filter}~(\equiv h)~(h :: t) &= \var{filter}~(\equiv h)~(h' :: t') \\
+    h :: \var{filter}~(\equiv h)~t &= h' :: \var{filter}~(\equiv h)~t'
+  \end{align*}
+  We also have $t = t'$ by the induction hypothesis, so we can
+  conclude that $h = h'$.
+\end{proof}
+
+This means that any two stable sorts must produce the same results for
+every input, and are thus interchangeable.
+\stablesortunique*
 
@@
 which implies that every sort is equivalent to every other sort.
+
 \begin{theorem}[all\_perm\_stable]
@@
 \end{theorem}
+\begin{proof}
+  We reason by induction on the evidence of $\var{Permutation}~l~l'$.
+  Every case other than the \var{perm\_swap} case follows easily.
+  In the \var{perm\_swap} case, we must show that
+  \begin{equation*}
+    \var{StablePerm}~(y :: x :: l)~(x :: y :: l)
+  \end{equation*}
+  for all $x$ and $y$. If $x \not\equiv y$, we can just apply
+  \var{StablePerm\_swap}. If $x \equiv y$, then we know that $x = y$, and
+  we can show
+  \begin{equation*}
+    \var{StablePerm}~(x :: x :: l)~(x :: x :: l)
+  \end{equation*}
+  by applying \var{perm\_skip} twice.
+\end{proof}
+
+\begin{corollary}[Sort\_StableSort\_Ord]
+  \label{thm:stablesort_ord}
+  When sorting using an order,
+  \begin{equation*}
+    \var{Sort}~f \iff \var{StableSort}~f
+  \end{equation*}
+  for any $f$.
+\end{corollary}
+
 \begin{corollary}[Sort\_Ord\_unique]
   \label{thm:sort_ord_unique}
-  When sorting using an order, given any two function $f, g:
+  When sorting using an order, for any two sorting function $f, g:
   \var{list}~A \to \var{list}~A$,
   \begin{equation*}
-    \var{Sort}~f \implies \var{Sort}~g \implies
     (\forall l, f~l = g~l)
@@
 
-\subsection{LSD radix sort}
+\section{LSD radix sort}
 \label{sec:radix_sort}
@@
 LSD radix sort is a simple radix sort that is used to
-lexicographically sort a list of strings of equal length. LSD radix
-sort operates by stably sorting on each column, from the end of the
-string to the front (\cref{fig:lsd_radixsort}). The algorithm works
-because lexicographic order says that if two strings are equal up to a
-column, then the tie is broken according to the rest of the string.
-When we stably sort on a column, ties are broken in favor or the
-existing order, which was established by sorted on the following
-columns.
-
-\begin{figure}[!hb]
+lexicographically sort strings of equal length. We can define what it
+means to sort lexicographically:
+
+\begin{definition}[lex\_le]
+  Given a preorder~$\le$ on a type $A$, we can define another preorder,
+  $\lexle$ on $\var{list}~A$.
+  \begin{gather*}
+    \infer[\var{lex\_le\_nil}]{[] \lexle ys}{}
+    \\
+    \\
+    \infer[\var{lex\_le\_cons\_lt}]{(x :: xs) \lexle (y :: ys)}{%
+      x < y
+    }
+    \\
+    \\
+    \infer[\var{lex\_le\_cons\_eq}]{(x :: xs) \lexle (y :: ys)}{%
+      x \equiv y & xs \lexle ys
+    }
+  \end{gather*}
+\end{definition}
+
+LSD radix sort operates by conceptually placing all the strings in a
+matrix, and by stably sorting on each column, from the end of the
+string to the front (\cref{fig:lsd_radixsort}). This works because if
+two strings match on a column, the stable sort ``defers'' to the
+preexisting ordering, and that preexisting ordering was determined by
+sorting by the tail of the strings. This mirrors the way a
+lexicographic ordering ``defers'' to the rest of a string when two
+strings match on earlier characters.
+
+\begin{figure}
   \centering
   \begin{tt}
+    \setlength{\tabcolsep}{0pt}
     \begin{tabular}{c}
-      d a \textbf{b} \\
-      a d \textbf{d} \\
-      c a \textbf{b} \\
-      f a \textbf{d} \\
-      f e \textbf{e} \\
-      b a \textbf{d} \\
-      d a \textbf{d} \\
-      b e \textbf{e} \\
-      f e \textbf{d} \\
-      b e \textbf{d} \\
-      e b \textbf{b} \\
-      a c \textbf{e}
+    fit \\
+    rib \\
+    fly \\
+    vat \\
+    age \\
+    mix \\
+    row \\
+    fog
+    \end{tabular}
+    $\rightarrow$
+    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    ri&b& \\
+    ag&e& \\
+    fo&g& \\
+    fi&t& \\
+    va&t& \\
+    mi&x& \\
+    fl&y& \\
+    ro&w&
+    \end{tabular}
+    $\rightarrow$
+    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    v&a&t \\
+    a&g&e \\
+    r&i&b \\
+    f&i&t \\
+    m&i&x \\
+    f&l&y \\
+    f&o&g \\
+    r&o&w
+    \end{tabular}
+    $\rightarrow$
+    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    &a&ge \\
+    &f&it \\
+    &f&ly \\
+    &f&og \\
+    &m&ix \\
+    &r&ib \\
+    &r&ow \\
+    &v&at
+    \end{tabular}
+  \end{tt}
+  \caption{LSD radix in-place}
+  \label{fig:lsd_radixsort}
+\end{figure}
+
+In a functional setting, we can define an LSD radixsort elegantly by
+repeatedly moving the last column to the front of the matrix and
+sorting on the first column (\cref{fig:lsd_radixsort_func}). We will
+take as a parameter the number of columns $n$.
+\begin{equation*}
+  \var{radixsort}~m~n = (\var{hdsort} \comp \var{map}~\var{rrot})^n~m
+\end{equation*}
+
+\begin{figure}
+  \centering
+  \begin{tt}
+    \setlength{\tabcolsep}{0pt}
+    \begin{tabular}{c}
+    fit \\
+    rib \\
+    fly \\
+    vat \\
+    age \\
+    mix \\
+    row \\
+    fog
     \end{tabular}
@@
     \begin{tabular}{c}
-      d \textbf{a} b \\
-      c \textbf{a} b \\
-      e \textbf{b} b \\
-      a \textbf{d} d \\
-      f \textbf{a} d \\
-      b \textbf{a} d \\
-      d \textbf{a} d \\
-      f \textbf{e} d \\
-      b \textbf{e} d \\
-      f \textbf{e} e \\
-      b \textbf{e} e \\
-      a \textbf{c} e
+    tfi \\
+    bri \\
+    yfl \\
+    tva \\
+    eag \\
+    xmi \\
+    wro \\
+    gfo
     \end{tabular}
     $\rightarrow$
-    \begin{tabular}{c}
-      \textbf{d} a b \\
-      \textbf{c} a b \\
-      \textbf{f} a d \\
-      \textbf{b} a d \\
-      \textbf{d} a d \\
-      \textbf{e} b b \\
-      \textbf{a} c e \\
-      \textbf{a} d d \\
-      \textbf{f} e d \\
-      \textbf{b} e d \\
-      \textbf{f} e e \\
-      \textbf{b} e e
+    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    &b&ri \\
+    &e&ag \\
+    &g&fo \\
+    &t&fi \\
+    &t&va \\
+    &x&mi \\
+    &y&fl \\
+    &w&ro
     \end{tabular}
     $\rightarrow$
-    \begin{tabular}{c}
-      a c e \\
-      a d d \\
-      b a d \\
-      b e d \\
-      b e e \\
-      c a b \\
-      d a b \\
-      d a d \\
-      e b b \\
-      f a d \\
-      f e d \\
-      f e e
+    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    ibr \\
+    gea \\
+    ogf \\
+    itf \\
+    atv \\
+    ixm \\
+    lyf \\
+    owr
+    \end{tabular}
+    $\rightarrow$
+    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    &a&tv \\
+    &g&ea \\
+    &i&br \\
+    &i&tf \\
+    &i&xm \\
+    &l&yf \\
+    &o&gf \\
+    &o&wr
+    \end{tabular}
+    $\rightarrow$
+    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    vat \\
+    age \\
+    rib \\
+    fit \\
+    mix \\
+    fly \\
+    fog \\
+    row
+    \end{tabular}
+    $\rightarrow$
+    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    &a&ge \\
+    &f&it \\
+    &f&ly \\
+    &f&og \\
+    &m&ix \\
+    &r&ib \\
+    &r&ow \\
+    &v&at
     \end{tabular}
   \end{tt}
-  \caption{LSD Radixsort}
-  \label{fig:lsd_radixsort}
+  \caption{LSD radix using rotations}
+  \label{fig:lsd_radixsort_func}
 \end{figure}
 
-\subsection{The Burrows-Wheeler transform}
-\label{subsec:bwt}
+Here \var{hdsort} is an insertion sort that only examines the first
+character of each string. We can implement this by defining orderings
+that compare elements using some key function:
+
+\begin{definition}[keyOrd]
+  Given a preorder on a type $K$ and a key function $\var{key} : A \to
+  K$, we can define a preorder on $A$ that compares all elements using
+  the key function. That is, we can define the relation
+  $\le_{\var{key}}$ on $A$ such that for all $x, y : A$
+  \begin{equation*}
+    x \le_{\var{key}} y \coloneqq (\var{key}~x) \le (\var{key}~y)
+  \end{equation*}
+\end{definition}
+
+Now we can define prefix lexicographic orderings on lists, where the
+lexicographic ordering is only allowed to compare up to the first $n$
+elements of two lists. We define the relation $\le_n$ on $\var{list}~A$
+such that for all $xs, ys : \var{list}~A$,
+\begin{equation*}
+  xs \le_n ys \coloneqq (\var{firstn}~n~xs) \lexle (\var{firstn}~n~ys)
+\end{equation*}
+We also define $\var{PrefixSorted}~n$ as \var{Sorted} specialized to
+the $\le_n$ ordering. \var{hdsort} is insertion sort specialized to the
+$\le_1$ ordering.
+
+We would like to prove that radix sort is a stable sort.
+
+\begin{corollary}[radixsort\_correct]
+  \label{thm:radixsort_correct}%
+  For all matrices $m$ with $n$ columns,
+  \begin{equation*}
+    \var{Sorted}~(\var{radixsort}~m~n) \land \var{StablePerm}~(\var{radixsort}~m~n)~m
+  \end{equation*}
+\end{corollary}
+
+For both properties, our approach will be to prove some invariant that
+holds after $j$ iterations. For \var{StablePerm}, we can say that
+after $j$ iterations of rotating and sorting, the matrix should be a
+stable permutation of the original matrix had we rotated $j$ times.
+
+\begin{restatable}[radixsort\_stable\_inv]{theorem}{radixsortstableinv}
+  \label{thm:radixsort_stable_inv}
+  For all matrices $m$ and $j : \nat$,
+  \begin{equation*}
+    \var{StablePerm}~(\var{radixsort}~m~j)~((\var{map}~\var{rrot})^j~m)
+  \end{equation*}
+\end{restatable}
+
+For \var{Sorted}, the invariant will be that after $j$ iterations, the
+matrix is sorted on the first $j$ columns.
+
+\begin{restatable}[radixsort\_sorted\_inv]{theorem}{radixsortsortedinv}
+  \label{thm:radixsort_sorted_inv}
+  For all matrices $m$ with rows of length $n$, and $j \le n$,
+  \begin{equation*}
+    \var{PrefixSorted}~j~(\var{radixsort}~m~j)
+  \end{equation*}
+\end{restatable}
+
+To prove both of these invariants, we will have to be able to reason
+about the relationship between the various orderings we are using: we
+use $\le_1$ in \var{hdsort}, $\le_j$ in $\var{PrefixSorted}~j$, and
+$\lexle \approx \le_n$ everywhere else.
+
+Let's generalize, and consider two orders $\le_j$ and $\le_k$ such that $j
+\le k$. If in the first $j$ elements we can conclude that $x <_j y$,
+then examining more elements can't change that.
+
+\begin{theorem}[key\_lt\_firstn\_ge]
+  \label{thm:key_lt_firstn_ge}
+  For all $j$ and $k$ such that $j \le k$,
+  \begin{equation*}
+    x <_j y \implies x <_k y
+  \end{equation*}
+\end{theorem}
+
+We can rephrase this in terms of $\le$ and $\equiv$.
+
+\begin{corollary}[key\_le\_firstn\_ge]
+  For all $j$ and $k$ such that $j \le k$,
+  \begin{equation*}
+    x \le_k y \implies x \le_j y
+  \end{equation*}
+\end{corollary}
+\begin{proof}
+  This is the contrapositive of \cref{thm:key_lt_firstn_ge}.
+\end{proof}
+
+\begin{corollary}
+  For all $j$ and $k$ such that $j \le k$,
+  \begin{equation*}
+    x \equiv_k y \implies x \equiv_j y
+  \end{equation*}
+\end{corollary}
+
+These theorems show that $\equiv_j$ is weaker than $\equiv_k$ in that it can
+distinguish between fewer elements. If we consider the implications
+this has for stability, stability with respect to a weaker order is
+actually a stronger property, because the weaker order, considering
+more elements equivalent, is more restrictive about which swaps it
+allows.
+
+\begin{theorem}[StablePerm\_weaken]
+  \label{thm:stableperm_weaken}
+  Suppose we have a two preorders $\le$ and $\le'$ on $A$, and two lists
+  $l_1,l_2 : \var{list}~A$, such that for all $x$ and $y$ in $l_1$,
+  \begin{equation*}
+    x \le' y \implies x \le y.
+  \end{equation*}
+  That is, $\le'$ is stronger than $\le$.
+
+  Then, denoting by \var{StablePerm} stability relative to $\le$ and and
+  by $\var{StablePerm}'$ stability relative to $\le'$,
+  \begin{equation*}
+    \var{StablePerm}~l_1~l_2 \implies \var{StablePerm}'~l_1~l_2
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  We reason inductively on the proof of $\var{StablePerm}~l_1~l_2$.
+  The important case is the swap case, where we have to show
+  \begin{equation*}
+    \var{StablePerm}'~(y :: x :: l)~(x :: y :: l)
+  \end{equation*}
+  given that $x \not\equiv y$.
+
+  We have that $x \equiv' y \implies x \equiv y$, so $x \not\equiv y \implies x
+  \not\equiv' y$.
+\end{proof}
+
+Now we can prove the stability invariant:
+\radixsortstableinv*
+\begin{proof}
+  We reason by induction on $j$. In the inductive case, we have to prove
+  \begin{equation*}
+    \var{StablePerm}~(\var{radixsort}~m~(j+1))~((\var{map}~\var{rrot})^{j+1}~m)
+  \end{equation*}
+  Expanding $\var{radixsort}~m~(j+1)$:
+  \begin{align*}
+     \var{radixsort}~m~(j+1)
+    ={}& (\var{hdsort} \comp \var{map}~\var{rrot})^{j+1}~m \\
+    ={}& \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \comp \var{map}~\var{rrot})^j~m)
+  \end{align*}
+  By \cref{thm:stableperm_weaken}, the stability relative to $\le_1$
+  that is guaranteed by \var{hdsort} implies stability relative to
+  $\lexle$, so we have only to show.
+  \begin{align*}
+    \var{StablePerm}~&(\var{map}~\var{rrot}~(\var{hdsort} \comp \var{map}~\var{rrot})^j~m) \\
+                     &((\var{map}~\var{rrot})^{j+1}~m)
+  \end{align*}
+  By \cref{thm:stableperm_map_rrot} (proved below), we can take off
+  the outer $\var{map}~\var{rrot}$ from both sides.
+  \begin{equation*}
+    \var{StablePerm}~
+    ((\var{hdsort} \comp \var{map}~\var{rrot})^j~m)~((\var{map}~\var{rrot})^j~m)
+  \end{equation*}
+  This is exactly the induction hypothesis.
+\end{proof}
+
+\begin{lemma}[StablePerm\_map\_rrot]
+  \label{thm:stableperm_map_rrot}
+  For all matrices $m$ and $m'$,
+  \begin{equation*}
+    \var{StablePerm}~m~m' \implies
+    \var{StablePerm}~(\var{map}~\var{rrot}~m)~(\var{map}~\var{rrot}~m').
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  We reason by induction on $\var{StablePerm}~m~m'$. The important
+  case is the swap case, where we have to show
+  \begin{equation*}
+    \var{StablePerm}~(\var{map}~\var{rrot}~(y :: x :: l))~(\var{map}~\var{rrot}~(x :: y :: l))
+  \end{equation*}
+  given that $x \not\equivle y$.
+
+  Here we can apply the swap ruel if we can show that $\var{rrot}~x
+  \not\equivle \var{rrot}~y$. This follows from the fact that
+  \begin{equation*}
+    x \equivle y \iff \var{rrot}~x \equivle \var{rrot}~y \qedhere
+  \end{equation*}
+\end{proof}
+
+We would also like to prove \cref{thm:radixsort_sorted_inv}. To do so,
+we need to be able to reason about the process of adding a column to
+the front of a sorted matrix and sorting on that column. Intuitively,
+we need to know that if we have a matrix that is sorted on the first
+$j$ columns, and we add a new column to the front and sort on it, the
+result list will be sorted on the first $j+1$ columns.
+\begin{restatable}[hdsort\_sorted\_S]{theorem}{hdsortsortedS}
+  \label{thm:hdsortsortedS}
+  For all matrices $m$ and naturals $j$,
+  \begin{equation*}
+    \var{PrefixSorted}~j~(\var{map}~\var{tl}~m) \implies
+    \var{PrefixSorted}~(j+1)~(\var{hdsort}~m)
+  \end{equation*}
+\end{restatable}
+\begin{proof}
+  See \cref{appendix:hdsort_sorted_S}.
+\end{proof}
+
+\radixsortsortedinv*
+\begin{proof}
+  We reason by induction on $j$. In the inductive case, we must prove.
+  \begin{equation*}
+    \var{PrefixSorted}~(j+1)~(\var{radixsort}~m~(j+1))
+  \end{equation*}
+  We can reason backwards from our goal. Unfolding \var{radixsort},
+  \begin{equation*}
+    \var{PrefixSorted}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{radixsort}~m~j)))
+  \end{equation*}
+  By \cref{thm:hdsortsortedS},
+  \begin{equation*}
+    \var{PrefixSorted}~j~(\var{map}~\var{tl}~(\var{map}~\var{rrot}~(\var{radixsort}~m~j)))
+  \end{equation*}
+  Because $\var{tl} \comp \var{rrot} = \var{init}$,
+  \begin{equation*}
+    \var{PrefixSorted}~j~(\var{map}~\var{init}~(\var{radixsort}~m~j))
+  \end{equation*}
+  Unfolding $\var{PrefixSorted}~j$,
+  \begin{equation*}
+    \var{Sorted}~(\var{map}~(\var{firstn}~j)~(\var{map}~\var{init}~(\var{radixsort}~m~j)))
+  \end{equation*}
+  Here, because $j$ is less than the length of all the rows,
+  $\var{firstn}~j \comp \var{init} = \var{firstn}~j$.
+  \begin{equation*}
+    \var{Sorted}~(\var{map}~(\var{firstn}~j)~(\var{radixsort}~m~j))
+  \end{equation*}
+  Folding back up $\var{PrefixSorted}~j$, we have the induction hypothesis
+  \begin{equation*}
+    \var{PrefixSorted}~j~(\var{radixsort}~m~j) \qedhere
+  \end{equation*}
+\end{proof}
+
+\section{The Burrows-Wheeler transform}
+\label{sec:bwt}
 
@@
 \begin{enumerate}
-\item Apply the BWT, which tends to group create runs of identical characters.
+\item Apply the BWT, which tends to create runs of identical
+  characters.
 \item Apply Move-to-Front (MTF) encoding, in which a character is
@@
 
-\begin{figure}[!hb]
+\begin{figure}
   \centering
@@
 resulting string easier to compress. It creates those runs by
-exploiting the correlation between adjacent characters, which means
+exploiting the correlation between adjacent characters. In many
+real-world inputs, adjacent characters are correlated, which means
 that by knowing one character, one can predict the character that
@@
 
-\subsection{Inverting the BWT}
-\label{subsec:bwt}
+\section{Inverting the BWT}
+\label{sec:invert_bwt}
 
@@
 
-\begin{figure}[!hb]
+\begin{figure*}
   \centering
@@
   \label{fig:radixsort_rots}
-\end{figure}
+\end{figure*}
 
@@
   \label{fig:recreate}
-\end{figure}
+\end{figure*}
 
@@
 We will use \var{iter}, which takes a function $f : A \to A$, a number
-$n$, and an initial value $z$, and returns the list of intermediate
-values obtained by applying $f$ to $z$, $n$ times. Or,
+$n$, and an initial value $z$, and returns a list of length $n$ formed
+by repeatedly applying $f$ to $z$:
 \begin{equation*}
@@
 repeated function application. We will also use the notation $l_i$ for
-the $i$th element of a list. In Coq, that is given by
+the $i$th element of a list, though in Coq, that is given by
 $\var{nth}~i~l~d$, where $d:A$ is a default value in case $i \ge
@@
 
-\section{\var{Sorted}, \var{Permutation}, and \var{StablePermutation}}
-\label{sec:sorted_perm_stable}
-
-The previous section's proofs relied on the correctness and stability
-of various sorts. The Coq standard library includes definitions for
-\var{Sorted} and \var{Permutation}, but no definition for stability.
-It also includes an implementation of mergesort, but it uses Coq's
-module system to enable parametricity over ordered types, which is
-much less convenient to use than typeclasses. In this section, we
-define our own version of \var{Sorted} and \var{StablePerm} that are
-based on our \var{Preord} typeclass.
-
-In this section, we restate the definition outlined in
-\cref{sec:main_ideas} more formally, and prove the appropriate
-equivalences.
-
-\subsection{\var{Sorted}}
-\label{subsec:sorted}
-
-\begin{definition}{SortedIx}
-  A list $l$ is sorted if for all indices $i$ and $j$,
-  \begin{equation*}
-    i \le j \implies l_i \le l_j
-  \end{equation*}
-\end{definition}
-\begin{definition}{Sorted}
-  \begin{align*}
-    \infer[\var{Sorted\_nil}]{\var{Sorted}~[]}{}
-    &&
-    \infer[\var{Sorted\_cons}]{\var{Sorted}~(a :: l)}{%
-      (\forall~x \in l,~a < x) & \var{Sorted}~l
-    }
-  \end{align*}
-\end{definition}
-
-By transitivity it suffices to check that the new element is smaller
-than just the first element of the sorted list. That leads to a third
-definition of sorted:
-\begin{definition}{SortedLocal}
-  \begin{gather*}
-    \infer[\var{SortedLocal\_nil}]{\var{SortedLocal}~[]}{}
-    \\
-    \\
-    \infer[\var{SortedLocal\_1}]{\var{SortedLocal}~[a]}{}
-    \\
-    \\
-    \infer[\var{SortedLocal\_cons}]{\var{SortedLocal}~(a :: b :: l)}{%
-      a < b & \var{SortedLocal}~(b :: l)
-    }
-  \end{gather*}
-\end{definition}
-
-\begin{theorem}[SortedLocal\_iff]
-  For all lists $l$,
-  \begin{equation*}
-    \var{Sorted}~l \iff \var{SortedLocal}~l
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  Both directions follow by induction on $l$, making use of the
-  transitivity of $\le$.
-\end{proof}
-
-\begin{theorem}[SortedIx\_iff]
-  For all lists $l$,
-  \begin{equation*}
-    \var{Sorted}~l \iff \var{SortedIx}~l
-  \end{equation*}
-\end{theorem}
-
-\subsection{\var{Permutation}}
-\label{subsec:perm}
-
-The word \textit{permutation} has two related meanings that are often
-used interchangeably. We sometimes say that two lists are permutations
-of each other if they contain all of the same elements, just appearing
-in a different order. However, in mathematics we often say that a
-permutation is actually the bijection on indices that defines the
-correspondence between the two lists. For example, we say that $p$ is
-the permutation that relates $l$ to $l'$ if for all indices $i$ and
-$j$,
-\begin{equation*}
-  l_{p(i)} = l'_i
-\end{equation*}
-For consistency with Coq's built-in \var{Permuation} relation, in a
-situation like this we will say that $l$ and $l'$ are permutations of
-each other, and that $p$ is the \textit{permutation function} that
-relates $l$ to $l'$.
-
-So, we can also say that two lists $l$ and $l'$ are permutations of
-each other is precisely when
-\begin{itemize}
-\item $\var{length}~l = \var{length}~l' = n$
-\item there exists a permutation function $p : \mathbb{N}_{<n} \to
-  \mathbb{N}_{<n}$, where $p$ is a bijection, such that for all
-  indices $i$,
-  \begin{equation*}
-    l_i = l'_{p(i)}
-  \end{equation*}
-\end{itemize}
-
-\begin{definition}{PermutationCount}
-  Two lists are permutations of one another when for any $x$,
-  \begin{equation*}
-    \var{count\_occ}~l~x = \var{count\_occ}~l'~x
-  \end{equation*}
-  where $\var{count\_occ}~l~x$ counts the number of elements in $l$
-  that are equal to $x$ under some equivalence relation.
-\end{definition}
-\var{count\_occ} counts the number of occurrences of a given variable
-using a decision procedure for Leibniz equality. Here that decision
-procedure is \texttt{equiv\_dec} from an \var{EqDec} instance where we
-require the equivalence relation itself by Leibniz equality.
-
-In Coq, we can represent a finite function as a list of its outputs:
-$f : \mathbb{N}_{<n} \to A$ is represented by the $\var{list}~A$
-$[f(0),~f(1),\allowbreak\ldots,f(n-1)]$. For permutations, this is
-consistent with the notation where we denote by $2~1~0$ the
-permutation $p(0) = 2$, $p(1) = 1$, $p(2) = 0$. So, $p$ here would be
-represented by $[2,~1,~0]$. To restrict the range to $\mathbb{N}_{<n}$
-we say that $\forall i,~i \in p \implies ~i<n$; to make the function
-surjective, we strengthen that to $\forall i,~i \in p \iff ~i<n$; and to make it
-injective we require that $p$ not contain any duplicate entries. The
-inductive property \var{NoDup} from the Coq standard library captures
-that requirement. This leads to the following definition:
-\begin{definition}{PermFun}
-  A list of natural numbers $p$ is a permutation function on $n$
-  elements if it contains no duplicate entries, and if for all natural
-  numbers $i$, $i \in p \iff i < n$.
-\end{definition}
-
-Then, we can define \var{apply}, which permutes a list according to a
-given permutation:
-\begin{lstlisting}
-  Definition apply p l : list A :=
-  match l with
-  | [] => []
-  | d :: _ => map (fun i => nth i l d) p
-  end.
-\end{lstlisting}
-We will use $p(l)$ as notation for $\var{apply}~p~l$ when $l$ is a
-list.
-
-And finally, we can define a notion of when two lists are permutations
-of one another based on permutation functions.
-\begin{definition}{PermutationEx}
-  Two lists $l$ and $l'$ are permutations of one another when there
-  exists a permutation function $p$ on $\var{length}~l$ elements such
-  that $p(l) = l'$.
-\end{definition}
-
-\begin{theorem}[PermutationCount\_iff]
-  For all lists $l$ and $l'$,
-  \begin{equation*}
-    \var{Permutation}~l~l' \iff \var{PermutationCount}~l~l'
-  \end{equation*}
-\end{theorem}
-
-\subsection{\var{StablePerm}}
-\label{subsec:stable}
-
-\begin{definition}{StablePerm}
-  Two lists are stable permutations of one aonther if for any $x$,
-  \begin{equation*}
-    \var{filter}~(\lambda y. y \equiv x)~l = \var{filter}~(\lambda y. y \equiv x)~l'
-  \end{equation*}
-\end{definition}
-
-\begin{definition}{StablePermFun\_preimage}
-  A permutation function $p$ is a stable permutation function on some
-  $l : \var{list}~A$ when for all $i < j$ such that $l_i \equiv l_j$, we
-  also have $p(i) < p(j)$.
-\end{definition}
-\begin{definition}{PermutationEx}
-  Two lists $l$ and $l'$ are stable permutations of one another when
-  there exists a stable permutation function~$p$ on $l$ such that
-  $p(l) = l'$.
-\end{definition}
-
-\begin{definition}{StablePermInd}
-  \begin{gather*}
-    \infer[\var{StablePermInd\_nil}]{\var{StablePermInd}~[]~[]}{}
-    \\ \\
-    \infer[\var{StablePermInd\_skip}]{\var{StablePermInd}~(x :: l)~(x :: l')}{%
-      \var{StablePermInd}~l~l'
-    }
-    \\ \\
-    \infer[\var{StablePermInd\_swap}]
-          {\var{StablePermInd}~(x :: y :: l)~(y :: x :: l)}
-          {x \not\equiv y}
-          \\ \\
-          \infer[\var{StablePermInd\_trans}]{\var{StablePermInd}~l~l''}{
-            \var{StablePermInd}~l~l' & \var{StablePermInd}~l'~l''
-          }
-  \end{gather*}
-\end{definition}
-
-\section{Uniqueness of stable sorts}
-\label{sec:unique}
-
-Using our definitions of \var{StablePerm} and \var{Sorted}, we can
-prove the theorem that allowed us to replace \var{lexsort} with
-radixsort.
-
-\begin{theorem}[StablePerm\_Sorted\_eq]
-  For all lists $l$,
-  \begin{equation*}
-    \var{Sorted}~l \implies \var{Sorted}~l' \implies
-    \var{StablePerm}~l~l' \implies l = l'.
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  We reason by induction on $l$. In the inductive case, we have to
-  show that
-  \begin{equation*}
-    h :: t = h' :: t'
-  \end{equation*}
-  given $\var{Sorted}~(h :: t)$, $\var{Sorted}~(h' :: t')$, and
-  $\var{StablePerm}~(h :: t)~(h' :: t')$, as well as $t = t'$ from the
-  induction hypothesis. So it suffices to show $h = h'$.
-
-  First, we can prove that $h \equiv h'$. We know that $h \in h' :: t'$ and
-  $h' \in h :: t$ because they are permutations of one another. If $h =
-  h'$, then we are done, so assuming $h \neq h'$ we know $h \in t'$ and $h'
-  \in t$. Because both $h::t$ and $h'::t'$ are sorted, we know that $h \le
-  h'$ and $h' \le h$, implying $h \equiv h'$.
-
-  From this we can prove that $h = h'$. From $\var{StablePerm}~(h ::
-  t)~(h' :: t')$ we have
-  \begin{align*}
-    \var{filter}~(\equiv h)~(h :: t) &= \var{filter}~(\equiv h)~(h' :: t') \\
-    h :: \var{filter}~(\equiv h)~t &= h' :: \var{filter}~(\equiv h)~t'
-  \end{align*}
-  We also have $t = t'$ by the induction hypothesis, so we can
-  conclude that $h = h'$.
-\end{proof}
-
-If we define \var{Sort} and \var{StableSort} as we did
-\cref{sec:sorted_perm_stable}, we can restate the theorem slightly
-more elegantly.
-\begin{lstlisting}
-  Definition Sort f :=
-  forall l, Sorted (f l) /\ Permutation (f l) l.
-\end{lstlisting}
-
-\stablesortunique
-\begin{proof}
-  This follows from \var{StablePerm\_Sorted\_eq}.
-\end{proof}
-
-However, we do not want to require that \var{lexsort} be implemented
-using a stable sorting algorithm. Luckily, because we require that $A$
-be equipped with an order and not just a preorder, we know that
-equivalent elements are actually indistinguishable. This means that
-all sorting algorithms can be considered to be stable: it is
-impossible to tell if any two equivalent elements were swapped because
-you cannot distinguish between equivalent elements.
-
-\begin{theorem}[all\_perm\_stable]
-  If we have an equivalence that implies Leibniz equality, then for
-  all lists $l$ and $l'$
-  \begin{equation*}
-    \var{Permutation}~l~l' \implies \var{StablePerm}~l~l'
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  We reason by induction on the evidence of $\var{Permutation}~l~l'$.
-  Every case other than the \var{perm\_swap} case follows easily.
-  In the \var{perm\_swap} case, we must show that
-  \begin{equation*}
-    \var{StablePerm}~(y :: x :: l)~(x :: y :: l)
-  \end{equation*}
-  for all $x$ and $y$. If $x \not\equiv y$, we can just apply
-  \var{StablePerm\_swap}. If $x \equiv y$, then we know that $x = y$, and
-  we can show
-  \begin{equation*}
-    \var{StablePerm}~(x :: x :: l)~(x :: x :: l)
-  \end{equation*}
-  by applying \var{perm\_skip} twice.
-\end{proof}
-
-\begin{corollary}[Sort\_StableSort\_Ord]
-  \label{thm:stablesort_ord}
-  When sorting using an order,
-  \begin{equation*}
-    \var{Sort}~f \iff \var{StableSort}~f
-  \end{equation*}
-  for any $f$.
-\end{corollary}
-\begin{proof}
-  This follows from \var{all\_perm\_stable}.
-\end{proof}
-
-\begin{theorem}[Sort\_Ord\_unique]
-  \label{thm:sort_ord_unique}
-  When sorting using an order, given any two function $f, g:
-  \var{list}~A \to \var{list}~A$,
-  \begin{equation*}
-    \var{Sort}~f \implies \var{Sort}~g \implies
-    (\forall l,~f~l = g~l)
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  This follows from \cref{thm:stablesort_ord} and \cref{thm:stablesort_unique}.
-\end{proof}
-
-\section{Lexicographic orders and keys}
-\label{sec:lex_ord}
-
-Given a preorder on a type $A$, we can define a lexicographic preorder
-on $\var{list}~A$. To capture this, we define the inductive relation
-\var{lex\_le}, which we will denote by $\lexle$ in proofs.
-\begin{gather*}
-  \infer[\var{lex\_le\_nil}]{[] \lexle ys}{}
-  \\
-  \\
-  \infer[\var{lex\_le\_cons\_lt}]{(x :: xs) \lexle (y :: ys)}{%
-    x < y
-  }
-  \\
-  \\
-  \infer[\var{lex\_le\_cons\_eq}]{(x :: xs) \lexle (y :: ys)}{%
-    x \equiv y & xs \lexle ys
-  }
-\end{gather*}
-
-Given a preorder on a type $K$ and a key function $\var{key} : A \to K$,
-we can define a preorder on $A$ that compares all elements using the
-key function. That is, we can define the relation $\le_{\var{key}}$ on
-$A$ such that for all $x, y : A$
-\begin{equation*}
-  x \le_{\var{key}} y \coloneqq (\var{key}~x) \le (\var{key}~y)
-\end{equation*}
-
-\begin{theorem}[key\_sort\_inv]
-  Given a key function $\var{key} : A \to K$ and a function $f : A \to A$
-  such that for all $x : A$,
-  \begin{equation*}
-    \var{key}~(f~x) = \var{key}~x
-  \end{equation*}
-  we have that for all lists $l$
-  \begin{equation*}
-    \var{sort}~(\var{map}~\var{f}~l) = \var{map}~\var{f}~(\var{sort}~l).
-  \end{equation*}
-  where \var{sort} is insertion sort.\footnote{This theorem actually
-    holds for any function with type $\forall X, (X \to X \to \mathbb{B}) \to
-    \var{list}~X \to \var{list}~X$ due to parametricity \cite{free}.}
-\end{theorem}
-\begin{proof}
-  The proof follows by induction, unfolding the definition of
-  \var{sort}. All decisions are made by calling \var{key}, so both
-  sides are equal.
-\end{proof}
-
-Taken together, we can define prefix lexicographic orderings on lists,
-where the lexicographic ordering is only allowed to compare up to the
-first $n$ elements of two lists. We define the relation $\le_n$ on
-$\var{list}~A$ such that for all $xs, ys : \var{list}~A$,
-\begin{equation*}
-  xs \le_n ys \coloneqq (\var{firstn}~n~xs) \lexle (\var{firstn}~n~ys)
-\end{equation*}
-For convenience, we define $\var{PrefixSorted}~n$ to hold when a list
-of lists is sorted if you only look at the first $n$ elements of each:
-\begin{lstlisting}
-  Definition PrefixSorted (n : nat) : list (list A) -> Prop
-  := Sorted (keyOrd (firstn n)).
-\end{lstlisting}
-
-We can describe the relationship between orders that can look at fewer
-or more characters with the following two theorems. We will briefly
-offer an explanation of why these theorems makes sense, but these
-explanations do not mirror the Coq proofs we have for these theorems,
-which are hard to follow.
-
-\begin{theorem}[key\_lt\_firstn\_ge]
-  For all $j$ and $k$ such that $j \le k$,
-  \begin{equation*}
-    x <_j y \implies x <_k y
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  If $x <_j y$, then there must be some index $i < j$ such that $x_i <
-  y_i$. Then $i < k$ as well, so $x <_k y$.
-\end{proof}
-
-\begin{theorem}[key\_le\_firstn\_ge]
-  For all $j$ and $k$ such that $k \ge j$,
-  \begin{equation*}
-    x \le_k y \implies x \le_j y
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  If $x \le_k y$, then either the first $k$ characters are all
-  equivalent, in which case the first $j$ characters are all
-  equivalent as well, or there was some $i < k$ such that $x_i < y_i$,
-  but at every index before $i$ $x$ and $y$ are equivalent. If $j <
-  i$, then the first $j$ characters are still equivalent, and we have
-  $x \equiv y \implies x \le_j y$. If $j \ge i$, then the deciding character is
-  among the first $j$ and $x \le_j k$.
-\end{proof}
-
-If we consider the implications this has for stability, a weaker order
-can distinguish between fewer elements, which means that it considers
-more elements equivalent. This means that stability with respect to a
-weaker order is actually a stronger property, because the weaker
-order, considering more elements equivalent, is more restrictive about
-which swaps it allows.
-
-\begin{theorem}[StablePerm\_weaken]
-  Suppose we have a two preorders $O$ and $O'$ on $A$, and two lists
-  $l_1$ and $l_2$, with elements of type $A$.
-
-  We also know that on the elements of $l_1$, $O'$ is stronger than
-  $O$. If we denote by $\le$ and $\le'$ the relations for $O$ and $O'$
-  respectively, this means that for all $x$ and $y$ in $l_1$,
-  \begin{equation*}
-    x \le' y \implies x \le y.
-  \end{equation*}
-  Then, denoting by \var{StablePerm} stability relative to the
-  equivalence relation of $O$, and by $\var{StablePerm}'$ stability
-  relative to the equivalence relation of $O'$, we have
-  \begin{equation*}
-    \var{StablePerm}~l_1~l_2 \implies \var{StablePerm}'~l_1~l_2
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  We reason inductively on the proof of $\var{StablePerm}~l_1~l_2$
-  using \var{StablePermInd\_iff}. The important case is the swap case,
-  where we have to show
-  \begin{equation*}
-    \var{StablePerm}'~(y :: x :: l)~(x :: y :: l)
-  \end{equation*}
-  given that $x \not\equiv y$. To show this, we need to know that $x \not\equiv'
-  y$, which follows from the fact that $O'$ is stronger than $O$.
-\end{proof}
-
-In a functional setting, we can define an LSD radixsort elegantly by
-repeatedly moving the last column to the front of the matrix and
-sorting on the first column. For a matrix with $n$ columns,
-\begin{equation*}
-  \var{radixsort} = (\var{hdsort} \comp \var{map}~\var{rrot})^n
-\end{equation*}
-Or, in Coq:
-\begin{lstlisting}
-  Definition radixsort m (n : nat) : list (list A)
-  := rep (hdsort ∘ map rrot) n m.
-\end{lstlisting}
-Here \var{hdsort} is just an insertion sort using $\le_1$ ordering.
-
-First we want to prove that \var{radixsort} permutes the rows of the
-matrix. We will start by proving the invariant that if we rotate the
-last column to the front and sort $j$ times, the resulting matrix is a
-permutation of simply rotating the last column to the front $j$
-times.
-\begin{lemma}[radixsort\_perm\_inv]
-  For all matrices $m$ and $j : \mathbb{N}$,
-  \begin{equation*}
-    \var{Permutation}~((\var{map}~\var{rrot})^j~m)~
-    ((\var{hdsort} \comp \var{map}~\var{rrot})^j~m)
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  We reason by induction on $j$. In the inductive case,
-  \begin{align*}
-    {}& (\var{hdsort} \comp \var{map}~\var{rrot})^{j+1}~m \\
-    ={}& \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \comp
-    \var{map}~\var{rrot})^j~m) \\
-    \eqperm{}& \var{map}~\var{rrot}~((\var{hdsort} \comp
-    \var{map}~\var{rrot})^j~m) \\
-    \eqperm{}& \var{map}~\var{rrot}~((\var{map}~\var{rrot})^j~m) \\
-    ={}& (\var{map}~\var{rrot})^{j+1}~m \qedhere
-  \end{align*}
-\end{proof}
-
-\begin{theorem}[radixsort\_perm]
-  For all matrices $m$ with rows of length $n$,
-  \begin{equation*}
-    \var{Permutation}~m~(\var{radixsort}~m~n)
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  \begin{align*}
-    m ={}&
-    \var{map}~\var{rrot}^n~m \\
-    ={}& (\var{map}~\var{rrot})^n~m \\
-    \eqperm{}& (\var{hdsort} \comp \var{map}~\var{rrot})^n~m \\
-    ={}& \var{radixsort}~m~n \qedhere
-  \end{align*}
-\end{proof}
-
-We will prove the stability of \var{radixsort} in a similar way, but
-first we need a lemma.
-\begin{lemma}[StablePerm\_map\_rrot]
-  For all matrices $m$ and $m'$,
-  \begin{equation*}
-    \var{StablePerm}~m~m' \implies
-    \var{StablePerm}~(\var{map}~\var{rrot}~m)~(\var{map}~\var{rrot}~m').
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  We reason by induction on the evidence that $m \eqperm m'$
-  using \var{StablePermInd\_iff}. The important case is the swap case,
-  where we have to show
-  \begin{equation*}
-    \var{StablePerm}~(\var{map}~(y :: x :: l))~(\var{map}~(x :: y :: l))
-  \end{equation*}
-  given that $x \not\equivle y$. For this it suffices to show that
-  $\var{rrot}~x \not\equivle \var{rrot}~y$. Since lexicographic
-  equivalence means that all the pairs of elements at corresponding
-  indices are equivalent, when \var{rrot} moves the first elements of
-  both lists to the end of both lists, it preserves lexicographic
-  equivalence. Therefore
-  \begin{equation*}
-    x \equivle y \iff \var{rrot}~x \equivle \var{rrot}~y \qedhere
-  \end{equation*}
-\end{proof}
-
-\begin{lemma}[radixsort\_stable\_inv]
-  For all matrices $m$ and $j : \mathbb{N}$,
-  \begin{equation*}
-    \var{StablePerm}~((\var{map}~\var{rrot})^j~m)~
-    ((\var{hdsort} \comp \var{map}~\var{rrot})^j~m)
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  We reason by induction on $j$. In the inductive case,
-  \begin{align*}
-    {}& (\var{hdsort} \comp \var{map}~\var{rrot})^{j+1}~m \\
-    ={}&\var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \comp
-    \var{map}~\var{rrot})^j~m) \\
-    \eqstable{}& \var{map}~\var{rrot}~((\var{hdsort} \comp
-    \var{map}~\var{rrot})^j~m) \\
-    \eqstable{}& \var{map}~\var{rrot}~((\var{map}~\var{rrot})^j~m) \\
-    ={}& (\var{map}~\var{rrot})^{j+1}~m
-  \end{align*}
-  Going from the first line to the second line requires using
-  \var{StablePerm\_weaken} to show that the stability we are
-  guaranteed by \var{hdsort}, namely stability with respect to an
-  ordering that examines at most one element, implies stability with
-  respect to a regular lexicographic ordering.
-\end{proof}
-
-\begin{theorem}[radixsort\_stable]
-  For all matrices $m$ with rows of length $n$,
-  \begin{equation*}
-    \var{StablePerm}~m~(\var{radixsort}~m~n)
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  The proof similar to the proof of \var{radixsort\_perm}.
-\end{proof}
-
-We would also like to establish an invariant to prove that
-\var{radixsort} produces a sorted list. To do so, we need to be able
-to reason about the process of adding a column to the front of a
-sorted matrix and sorting on that column. Intuitively, we need to know
-that if we have a matrix that is sorted on the first $j$ columns, and
-we add a new column to the front and sort on it, the result list will
-be sorted on the first $j+1$ columns. For a proof, see
-\cref{appendix:hdsort_sorted_S}.
-\begin{restatable*}[hdsort\_sorted\_S]{theorem}{hdsortsortedS}
-  For all matrices $m$ and naturals $j$,
-  \begin{equation*}
-    \var{PrefixSorted}~j~(\var{map}~\var{tl}~m) \implies
-    \var{PrefixSorted}~(j+1)~(\var{hdsort}~m)
-  \end{equation*}
-\end{restatable*}
-
-We also need the following property about \var{prepend\_col}, that
-states that prepending a column and then removing the first column is
-a no-op.
-\begin{theorem}[map\_tl\_prepend]
-  For all matrices $m : \var{list}\ \allowbreak(\var{list}~A)$ and columns $c :
-  \var{list}~A$, where $c$ has at least as many elements as $m$ has columns,
-  \begin{equation*}
-    \var{map}~\var{tl}~(\var{prepend\_col}~c~m) = m
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  We reason by induction on $m$. In the inductive case, we have to
-  prove that
-  \begin{align*}
-    &\var{map}~\var{tl}~(\var{prepend\_col}~(h_c :: t_c)~(r :: m)) \\
-    ={}&\var{map}~\var{tl}~((h_c :: r) :: \var{prepend\_col}~t_c~m) \\
-    ={}&r :: \var{map}~\var{tl}~(\var{prepend\_col}~t_c~m) \\
-    ={}&r :: m
-  \end{align*}
-  where the last line follows from the induction hypothesis.
-\end{proof}
-
-\begin{corollary}
-  For all matrices $m$ and rows $r$,
-  \begin{equation*}
-    \var{map}~\var{tl}~(\var{map}~\var{rrot}~m) = \var{map}~\var{init}~m
-  \end{equation*}
-\end{corollary}
-\begin{proof}
-  \begin{align*}
-    {}& \var{map}~\var{tl}~(\var{map}~\var{rrot}~m) \\
-    ={}& \var{map}~\var{tl}~(\var{prepend\_col}~(\var{map}~\var{last}~m)~(\var{map}~\var{init}~m) \\
-    ={}& \var{map}~\var{init}~m \qedhere
-  \end{align*}
-\end{proof}
-
-Now we can establish our invariant:
-
-\begin{lemma}[radixsort\_sorted\_inv]
-  For all matrices $m$ with rows of length $n$, and $j \le n$,
-  \begin{equation*}
-    \var{PrefixSorted}~j~(\var{radixsort}~m~j)
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  We reason by induction on $j$. The base case is true because every
-  list is sorted under ordering where you examine no elements of the
-  list. In the inductive case, we reason backwards from our goal:
-  \begin{align*}
-    &\var{PrefixSorted}~(j+1)~(\var{radixsort}~m~(j+1)) \\
-    \iff{}&\var{PrefixSorted}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{radixsort}~m~j))) \\
-    \iff{}&\var{PrefixSorted}~j~(\var{map}~\var{tl}~(\var{map}~\var{rrot}~(\var{radixsort}~m~j))) \\
-    \iff{}&\var{PrefixSorted}~j~(\var{map}~\var{init}~(\var{radixsort}~m~j)) \\
-    \iff{}&\var{Sorted}~(\var{map}~(\var{firstn}~j)~(\var{map}~\var{init}~(\var{radixsort}~m~j))) \\
-    \iff{}&\var{Sorted}~(\var{map}~(\var{firstn}~j)~(\var{radixsort}~m~j)) \\
-    \iff{}&\var{PrefixSorted}~j~(\var{radixsort}~m~j)
-  \end{align*}
-  The last line follows from the induction hypothesis.
-\end{proof}
-
-\begin{theorem}[radixsort\_sorted]
-  For all matrices $m$ with rows of length $n$,
-  \begin{equation*}
-    \var{Sorted}~(\var{radixsort}~m~n)
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  \begin{align*}
-    &\var{Sorted}~(\var{radixsort}~m~n) \\
-    \iff{}&\var{Sorted}~(\var{map}~(\var{firstn}~n)~(\var{radixsort}~m~n)) \\
-    \iff{}&\var{PrefixSorted}~n~(\var{radixsort}~m~n))
-  \end{align*}
-  The last line follows from \var{radixsort\_sorted\_inv}.
-\end{proof}
-
-\begin{theorem}[radixsort\_correct]
-  \label{thm:radixsort_correct}%
-  For all matrices $m$ with $n$ columns,
-  \begin{equation*}
-    \var{Sorted}~(\var{radixsort}~m~n) \land \var{StablePerm}~(\var{radixsort}~m~n)~m
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  This follows from \var{radixsort\_sorted} and \var{radixsort\_stable}.
-\end{proof}
-
 \section{Relationship to the standard BWT}
@@
 So assuming we have some function $\var{calc\_sort\_perm} :
-\var{list}~A \to \var{list}~\mathbb{N}$ to calculate the permutation
+\var{list}~A \to \var{list}~\nat$ to calculate the permutation
 function that sorts $l$, we redefine \var{unbwt} as
@@
 \appendix
+
+\section{Additional Proofs}
+\label{appendix:proofs}
+
+\sortedixiff*
+\begin{proof}
+  We reason by induction on $l$. The base cases are easy, so we
+  address the inductive case, $l = a :: l'$
+
+  ($\implies$) We have to prove that $(a :: l')_i \le (a :: l')_j$ given
+  $i \le j$ and $\var{Sorted}~l$. We reason by cases on $i$ and $j$:
+  \begin{itemize}
+  \item $i = j = 0$. By reflexivity, $a \le a$.
+  \item $i = 0,~j=j' + 1$. We need to show $a \le l'_{j'}$. We know that
+    $a \le x$ for all $x \in l'$ by $\var{Sorted}~(a::l')$, so we are
+    done.
+  \item $i = i' +1,~j = j' + 1$. We need to show $l'_{i'} \le l'_{j'}$,
+    which follows from the induction hypothesis.
+  \end{itemize}
+
+  ($\impliedby$) We have $\var{SortedIx}\ \allowbreak(a :: l')$, and
+  we need to show $\var{Sorted}\ \allowbreak(a :: l')$ given the
+  induction hypothesis $\var{SortedIx}~l' \implies \var{Sorted}~l'$.
+  By \var{SortedLocal\_iff}, we can just prove
+  $\var{SortedLocal}\ \allowbreak(a :: l')$. If $l' = []$, we can
+  apply $\var{SortedLocal\_1}$, so assume $l' = b :: l''$. To apply
+  \var{SortedLocal\_cons} we need to show
+  $\var{SortedLocal}\ \allowbreak(b :: l'')$, which we have by the
+  induction hypothesis and $\var{SortedIx}\ \allowbreak(a :: b ::
+  l'')$, and that $a \le b$. That follows because $a = (a :: b ::
+  l'')_0$ and $b = (a :: b :: l'')_1$, so by
+  $\var{SortedIx}\ \allowbreak(a :: b :: l'')$, we are done.
+\end{proof}
 
@@
 
-\hdsortsortedS
+\hdsortsortedS*
 \begin{proof}
@@
   case, we have the induction hypothesis that for $p :
-  \var{list}~\mathbb{N}$ and $l : \var{list}~A$
+  \var{list}~\nat$ and $l : \var{list}~A$
   \begin{equation*}

### 2019-10-17 draft_paper/draft_paper.tex
Finish content

@@
 
-\author{Andrew Appel}
+\author{Andrew W. Appel}
 \affiliation{
@@
 There has not been much work on formally verifying stable sorts, or
-rather, the stability of stable sorts. While Bird et~al.
-\cite{birdmu,pearls} provide an elegant derivation and proof for the
-BWT, formalizing it in Coq reveals that they rely on a few lemmas that
-are not trivial to prove, including the correctness of
+rather, the stability of stable sorts. While Bird
+et~al.\ \cite{birdmu,pearls} provide an elegant derivation and proof
+for the BWT, formalizing it in Coq reveals that they rely on a few
+lemmas that are not trivial to prove, including the correctness of
 least-significant digit (LSD) radix sort. We develop the theory of
@@
 equal elements. For example, if we are stably sorting the list
-$[\var{banana},\ \allowbreak\var{armpit},\ \allowbreak\var{carrot},\ \allowbreak\var{apple}]$,
-we have to produce
-$[\var{armpit},\ \allowbreak\var{apple},\ \allowbreak\var{banana},\ \allowbreak\var{carrot}]$
+$[\var{mix},\ \allowbreak\var{fog},\ \allowbreak\var{age},\ \allowbreak\var{fit}]$
+by the first character of each word, we have to produce
+$[\var{age},\ \allowbreak\var{fog},\ \allowbreak\var{fit},\ \allowbreak\var{mix}]$
 and not
-$[\var{apple},\ \allowbreak\var{armpit},\ \allowbreak\var{banana},\ \allowbreak\var{carrot}]$.
-Even though both are sorted on the first character, because
-\var{armpit} preceded \var{apple} in the original lists, it must also
-precede \var{apple} in the stably sorted list.
+$[\var{age},\ \allowbreak\var{fit},\ \allowbreak\var{fog},\ \allowbreak\var{mix}]$.
+Even though both are sorted on the first character, because \var{fog}
+preceded \var{fit} in the original list, it must also precede
+\var{fit} in the stably sorted list.
 
@@
 \cite{McIlroy93,Bentley:1997:FAS:314161.314321,10.1007/978-3-540-89097-3_3}.
-Many radix sorts rely on a some stable sorting subroutine.
+Most radix sorts rely on a some stable sorting subroutine.
 
@@
 In addition, to our knowledge no one has produced a formally verified
-implementation of the BWT, by which we mean an implementation with a
-proof that the inverse-transform inverts the forwards-transform. Bird
-et~al.\cite{alg-of-prog,birdmu} provide an excellent derivation of the
-BWT and its inverse based on the technique program calculation, in
-which programs are calculated from their specifications using
-equational reasoning. However, their derivation leaves a few crucial
-lemmas unproven, including the correctness of LSD radix sort.
+implementation of the BWT. Bird et~al.\ \cite{alg-of-prog,birdmu}
+provide an excellent derivation of the BWT and its inverse based on
+the technique of program calculation, in which programs are calculated
+from their specifications using algebraic transformations. However,
+their derivation leaves a few crucial lemmas unproven, including the
+correctness of LSD radix sort.
 
@@
 prove them equivalent. This allows us to choose whichever definition
-makes the proof easiest when we are reasoning about stability. We also
-proved that any two sorts are interchangeable.
+makes the proof easiest. We also prove that any two sorts are
+interchangeable.
 \begin{restatable*}[StableSort\_unique]{corollary}{stablesortunique}
@@
 
-Using \cref{thm:stablesort_unique}, we prove the correctness of an LSD
-radix sort that relies on insertion sort to sort each column. This
-choice was crucial to the proof: parametrizing radix sort over any
-stable sort made it difficult to reason about, but reasoning about
-insertion sort is very elegant. \cref{thm:stablesort_unique} allows us
-to recover the generality of parametrizing over all stable sorts,
-because it proves that insertion sort is equivalent to any other
-stable sort.
+Throughout our proofs about LSD radix sort and the BWT, we reason
+about sorting algorithms whose implementation is irrelevent as long as
+they are correct. We could, in principle, rely only on the fact that
+they fulfill the specification of a sorting algorithm
+(\cref{def:sort}). However, in practice this is difficult, because the
+inductive definition of permutations can produce unhelpful induction
+hypotheses, especially in the transitive case. Insetad, we reason
+about specifically insertion sort. Therefore
+\cref{thm:stablesort_unique} is quite useful, because we can reason
+using insertion sort, and later on replace it with any other
+(presumably faster) stable sort.
 
@@
 that relies on Burrows--Wheeler, such as bzip2. But more generally,
-this provides evidence that Bird's approach of program calculation, in
-which programs are derived from their specifications via algebraic
-manipulations, is useful for proving the correctness of algorithms in
-a machine-checked setting. Bird has applied this approach to greedy
-algorithms, the Boyer-Moore algorithm, the Knuth-Morris-Pratt
-algorithm, Sudoku solvers, arithmetic coding, the Schorr-Waite
-algorithm, the Johnson-Trotter algorithm, and many more problems
-\cite{pearls}.
+this provides evidence that Bird's approach of program calculation is
+useful for proving the correctness of algorithms in a machine-checked
+setting. Bird has applied this approach to greedy algorithms, the
+Boyer-Moore algorithm, the Knuth-Morris-Pratt algorithm, Sudoku
+solvers, arithmetic coding, the Schorr-Waite algorithm, the
+Johnson-Trotter algorithm, and many more problems \cite{pearls}.
 
@@
 supplementary materials. Definitions are copied verbatim, modulo
-whitespace. Theorems are rephrased for clarity, with the name of the
-corresponding Coq theorem given in parentheses. The only exceptions
-are \cref{sec:opt} and places where we give hypothetical definitions
-to drive the explanation.
+whitespace. Theorems and definitions are rephrased for clarity, with
+the name of the corresponding Coq theorem given in parentheses. The
+only exceptions are \cref{sec:opt} and places where we give
+hypothetical definitions to drive the explanation.
 
@@
 (\var{SortedIx}), and an inductive definition (\var{Sorted}).
-\begin{definition}{SortedIx}
+
+\begin{definition}[SortedIx]
   A list $l$ is sorted if for all indices $i$ and $j$,
@@
 
-\begin{definition}{Sorted}
-  \begin{align*}
+We use the notation $l_i$ for the $i$th element of a list, though in
+Coq, that is given by $\var{nth}~i~l~d$, where $d:A$ is a default
+value in case $i \ge \var{length}~l$.
+
+\begin{definition}[Sorted]
+  \label{def:sorted}
+  \begin{gather*}
     \infer[\var{Sorted\_nil}]{\var{Sorted}~[]}{}
-    &&
+    \\
     \infer[\var{Sorted\_cons}]{\var{Sorted}~(a :: l)}{%
@@
     }
-  \end{align*}
+  \end{gather*}
 \end{definition}
@@
 \begin{proof}
-  See \cref{appendix:proofs}.
-\end{proof}
-
-However, a sorting function cannot produce just any sorted list as its
-output; the sorted list must also contain all the same elements as the
-original list. To capture this, we could use the \var{Permutation}
-relation defined in the Coq standard library. This relation holds on
-two lists when they contain all the same elements, and is defined
-inductively:
+  See \cref{appendix:sorted}.
+\end{proof}
+
+However, a sorting function cannot produce just any sorted list; the
+output must also contain the same elements as the original list. To
+capture this, we could use the \var{Permutation} relation defined in
+the Coq standard library. This relation holds on two lists when they
+contain all the same elements, and is defined inductively:
 \begin{definition}[Permutation]
@@
 \end{definition}
+While this definition is useful, proofs that do induction directly on
+the evidence of \var{Permutation} can sometimes get stuck on the
+transitive case, hence the remark in the introduction about the
+usefulness of \cref{thm:stablesort_unique}.
 
@@
 indices in $l'$ by mapping $0 \mapsto 2$, mapping $1 \mapsto 0$, and mapping $2 \mapsto
-0$. Formally, we say that $p$ is the permutation that relates $l$ to
-$l'$ if for all indices $i$ and $j$,
+0$. $p$ is the permutation that relates $l$ to $l'$ if for all indices
+$i$ and $j$,
 \begin{equation*}
@@
 This suggests an alternative definition for \var{Permutation}:
-\begin{definition}{PermutationEx}
+\begin{definition}[PermutationEx]
   \label{def:permutation_ex}
-  Two lists $l$ and $l'$ are permutations of one another when, letting
-  $n = \var{length}~l$, there exists a permutation function $p$ on $n$
-  elements such that $p(l) = l'$.
+  Two lists $l$ and $l'$ are permutations of one another when there
+  exists a permutation function $p$ on $n = \var{length}~l$ elements
+  such that $p(l) = l'$.
 \end{definition}
@@
 A permutation function on $n$ elements maps indices to indices, and so
-has type $\nat_{<n} \to \nat_{<n}$. In Coq, we can represent
-a finite function as a list of its outputs: $f : \nat_{<n} \to A$
-is represented by the list $[f(0),~f(1),\allowbreak\ldots,~f(n-1)]$.
-Applied to permutations, this means that the permutation~$p$ mentioned
-above would be represented by $[2,~1,~0]$. This is consistent with
+has type $\nat_{<n} \to \nat_{<n}$. In Coq, we can represent a finite
+function as a list of its outputs: $f : \nat_{<n} \to A$ is represented
+by the list $[f(0),~f(1),\allowbreak\ldots,~f(n-1)]$. Applied to
+permutations, this means that the permutation~$p$ mentioned above
+would be represented by $[2,~1,~0]$. This is consistent with
 ``one-line'' notation for permutations, in which a permutation on $n$
 elements is represented by its result when applied to the
-list~$[0,~2,\ldots,n-1]$. Now we can make precise the notation
-$p(l)$\footnote{In Coq, this corresponds the the \var{apply} function,
-  whose definition is slightly more complex due to the extra default
-  parameter for \var{nth}.}
+list~$[0,~2,\ldots,n-1]$. Now we can define the action of a permutation on
+a list.
 \begin{equation*}
@@
 \end{equation*}
+We implement this in Coq using an \var{apply} function. The definition
+is slightly complicated by the fact that \var{nth} takes an extra
+parameter to return in case $i$ is out of bounds.
+\begin{lstlisting}
+Definition apply p l : list A :=
+  match l with
+  | [] => []
+  | d :: _ => map (fun i => nth i l d) p
+  end.
+\end{lstlisting}
 
@@
 can define the property $\var{PermFun}~n~p$, which holds when $p :
-\var{list}~\nat$ represents a permutation function on lists of
-length $n$:
+\var{list}~\nat$ represents a permutation function on $n$ elements.
 \begin{definition}[PermFun]
@@
   elements if for all natural numbers $i$, $i \in p \iff i < n$, and if
-  $p$ contains no duplicate elements\footnote{For this we can use the
-    \var{NoDup} predicate from Coq's standard library.}.
+  $p$ contains no duplicate elements.
 \end{definition}
+
+\begin{restatable}[PermutationEx\_iff]{theorem}{permutationexiff}
+  \label{thm:permutationex_iff}
+  For all lists $l$ and $l'$,
+  \begin{equation*}
+    \var{Permutation}~l~l' \iff \var{PermutationEx}~l~l'
+  \end{equation*}
+\end{restatable}
+\begin{proof}
+  See \cref{appendix:perm}.
+\end{proof}
 
@@
 
+\begin{restatable}[PermutationCount\_iff]{theorem}{permutationcountiff}
+  \label{thm:permutationcount_iff}
+  For all lists $l$ and $l'$,
+  \begin{equation*}
+    \var{Permutation}~l~l' \iff \var{PermutationCount}~l~l'
+  \end{equation*}
+\end{restatable}
+\begin{proof}
+  See \cref{appendix:perm}.
+\end{proof}
+
 With this we can define the specification for sorting functions:
 \begin{definition}[Sort]
+  \label{def:sort}
   A function $f : \var{list}~A \to \var{list}~A$ is a sorting function
@@
 
-Now that we've defined sorts, we can start to consider stable sorts.
-Stable sorts must preserve the relative orders of ``equivalent''
-elements. This equivalence comes from the ordering we are using: any
-total decidable preorder also give us a decideable equivalence
-relation $\equiv$, where $x \equiv y$ when $x \le y$ and $x \ge y$.
+Now that we've defined sorts, we can consider stable sorts. Stable
+sorts must preserve the relative orders of ``equivalent'' elements.
+This equivalence comes from the ordering we are using: any total
+decidable preorder also give us a decideable equivalence relation $\equiv$,
+where $x \equiv y$ when $x \le y$ and $x \ge y$.
 
@@
 To define stable permutations, we can tweak each of the definitions of
-permutation mentioned above. For example, we can modify our
+permutations mentioned above. For example, we can modify our
 function-based definition.
@@
 equivalence. The problematic rule is \var{perm\_swap} because we could
-porentially swap two equivalent elements. We will restrict it, so that
+potentially swap two equivalent elements. We will restrict it, so that
 you can only swap $x$ and $y$ when $x \nequiv y$.
@@
 Finally, we can say that if two lists are in a stable permutation, and
-we examine only the the elements equivalent to $x$ for some $x$, then
+for some $x$ we examine only the the elements equivalent to $x$, then
 those elements should appear in the same order in both lists.
@@
 $\var{filter}~(\lambda y. y = x)~l = \var{filter}~(\lambda y. y = x)~l'$.
+
+\begin{restatable}[StablePermEx\_iff]{theorem}{stablepermexiff}
+  For all lists $l$ and $l'$
+  \begin{equation*}
+    \var{StablePerm}~l~l' \iff \var{StablePermEx}~l~l'
+  \end{equation*}
+\end{restatable}
+
+\begin{restatable}[StablePermInd\_iff]{theorem}{stablepermindiff}
+  For all lists $l$ and $l'$
+  \begin{equation*}
+    \var{StablePerm}~l~l' \iff \var{StablePermInd}~l~l'
+  \end{equation*}
+\end{restatable}
 
@@
 every input, and are thus interchangeable.
-\stablesortunique*
-
-This can be generalized a little further: even if a sorting algorithm
-itself is not a stable sorting algorithm, if the ordering we are using
-doesn't allow us to distinguish between equivalent elements, then we
-will not be able to tell if equivalent elements are being swapped.
-More precisely, notice that it isn't always true that $x \equiv y$ implies
-$x = y$. For example, if we are sorting strings by their first
-character, then $\text{\texttt{apple}} \equiv \text{\texttt{armpit}}$ under
-this ordering, but clearly $\text{\texttt{apple}} \neq
-\text{\texttt{armpit}}$. When we do have that the equivalence ($\equiv$)
-implies Leibniz equality ($=$), we say that we have an \textit{order},
-as we have when we are sorting strings using all of their characters.
-Under an order, as opposed to a preorder, we can prove that every
-permutation is a stable permutation and every sort is a stable sort,
-which implies that every sort is equivalent to every other sort.
+\stablesortunique
+
+This can often be extended even to unstable sorting algorithms.
+Remember, it isn't always true that $x \equiv y$ implies $x = y$. For
+example, if we are sorting strings by their first character, then
+$\text{\texttt{fit}} \equiv \text{\texttt{fog}}$, but clearly
+$\text{\texttt{fit}} \neq \text{\texttt{fog}}$. When we do have that the
+equivalence ($\equiv$) implies Leibniz equality ($=$), we say that we have
+an \textit{order}, as we have when we are sorting strings using all of
+their characters. Under an order, as opposed to a preorder, equivalent
+elements are actually indistinguishable, so you can't tell if they
+have been swapped.
 
@@
 
-LSD radix sort is a simple radix sort that is used to
-lexicographically sort strings of equal length. We can define what it
-means to sort lexicographically:
+Least-significant digit (LSD) radix sort is a simple radix sort that
+is used to lexicographically sort strings of equal length. We can
+define what it means to sort lexicographically:
 
 \begin{definition}[lex\_le]
-  Given a preorder~$\le$ on a type $A$, we can define another preorder,
-  $\lexle$ on $\var{list}~A$.
+  Given a preorder~$\le$ on a type $A$, we define the preorder $\lexle$
+  on $\var{list}~A$.
   \begin{gather*}
@@
 LSD radix sort operates by conceptually placing all the strings in a
-matrix, and by stably sorting on each column, from the end of the
+matrix, and then stably sorting on each column, from the end of the
 string to the front (\cref{fig:lsd_radixsort}). This works because if
@@
   \end{tt}
-  \caption{LSD radix in-place}
+  \caption{LSD radix sort in-place}
   \label{fig:lsd_radixsort}
@@
 
-In a functional setting, we can define an LSD radixsort elegantly by
+In a functional setting, we can define an LSD radix sort elegantly by
 repeatedly moving the last column to the front of the matrix and
-sorting on the first column (\cref{fig:lsd_radixsort_func}). We will
-take as a parameter the number of columns $n$.
-\begin{equation*}
-  \var{radixsort}~m~n = (\var{hdsort} \comp \var{map}~\var{rrot})^n~m
-\end{equation*}
+sorting on the first column (\cref{fig:lsd_radixsort_func}). First, we
+define \var{rrot}, which moves the last element of a list to the
+front, and relies on \var{init}, which drops the last element of a
+list. Note that in order to ensure that $\var{last}$ is total, it
+takes an extra argument to return in the case of an empty list.
+\begin{lstlisting}
+Definition rrot (l: list A) : list A :=
+  match l with
+  | [] => []
+  | hd :: tl => last l hd :: init l
+  end.
+
+Fixpoint init l : list A :=
+  match l with
+  | [] | [_] => []
+  | hd :: tl => hd :: init tl
+  end.
+\end{lstlisting}
+
+Now we can define \var{radixsort}, taking the number of columns $n$ as
+a parameter.
+\begin{equation*}
+  \var{radixsort}~m~n \coloneq (\var{hdsort} \comp \var{map}~\var{rrot})^n~m
+\end{equation*}
+We use the notation $f^i z$ for $\var{rep}~f~i~z$, which implements
+repeated function application.
 
@@
     \end{tabular}
-    $\rightarrow$
+    $\overset{\var{rot}}{\rightarrow}$
     \begin{tabular}{c}
@@
     \end{tabular}
-    $\rightarrow$
+    $\overset{\var{sort}}{\rightarrow}$
     \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
@@
     \end{tabular}
-    $\rightarrow$
+    $\overset{\var{rot}}{\rightarrow}$
     \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
@@
   \end{tt}
-  \caption{LSD radix using rotations}
+  \caption{LSD radix sort using rotations}
   \label{fig:lsd_radixsort_func}
@@
 
-Here \var{hdsort} is an insertion sort that only examines the first
+Here \var{hdsort} is an insertion sort
+(\cref{appendix:insertion_sort}) that only examines the first
 character of each string. We can implement this by defining orderings
@@
 lexicographic ordering is only allowed to compare up to the first $n$
-elements of two lists. We define the relation $\le_n$ on $\var{list}~A$
+elements of two lists.
+
+\begin{definition}[lexle\_firstn]
+  We define the relation $\le_n$ on $\var{list}~A$
   such that for all $xs, ys : \var{list}~A$,
-\begin{equation*}
-  xs \le_n ys \coloneqq (\var{firstn}~n~xs) \lexle (\var{firstn}~n~ys)
-\end{equation*}
+  \begin{align*}
+    xs \le_n ys &\coloneqq xs \le_{\var{firstn}~n} ys \\
+              &= (\var{firstn}~n~xs) \lexle (\var{firstn}~n~ys)
+  \end{align*}
+\end{definition}
 We also define $\var{PrefixSorted}~n$ as \var{Sorted} specialized to
@@
 after $j$ iterations of rotating and sorting, the matrix should be a
-stable permutation of the original matrix had we rotated $j$ times.
-
-\begin{restatable}[radixsort\_stable\_inv]{theorem}{radixsortstableinv}
+stable permutation of the original matrix had we only rotated $j$
+times without sorting.
+
+\begin{restatable*}[radixsort\_stable\_inv]{theorem}{radixsortstableinv}
   \label{thm:radixsort_stable_inv}
@@
   \end{equation*}
-\end{restatable}
+\end{restatable*}
 
@@
 
-\begin{restatable}[radixsort\_sorted\_inv]{theorem}{radixsortsortedinv}
+\begin{restatable*}[radixsort\_sorted\_inv]{theorem}{radixsortsortedinv}
   \label{thm:radixsort_sorted_inv}
@@
 
-\begin{theorem}[key\_lt\_firstn\_ge]
+\begin{lemma}[key\_lt\_firstn\_ge]
   \label{thm:key_lt_firstn_ge}
@@
   \end{equation*}
-\end{theorem}
+\end{lemma}
 
@@
 
-These theorems show that $\equiv_j$ is weaker than $\equiv_k$ in that it can
-distinguish between fewer elements. If we consider the implications
-this has for stability, stability with respect to a weaker order is
-actually a stronger property, because the weaker order, considering
-more elements equivalent, is more restrictive about which swaps it
-allows.
-
-\begin{theorem}[StablePerm\_weaken]
+These theorems show that $\equiv_j$ is weaker than $\equiv_k$ in the sense that
+it can distinguish between fewer elements. If we consider the
+implications this has for stability, stability with respect to a
+weaker order is actually a stronger property, because the weaker
+order, considering more elements equivalent, is more restrictive about
+which swaps it allows.
+
+\begin{lemma}[StablePerm\_weaken]
   \label{thm:stableperm_weaken}
@@
 
-  Then, denoting by \var{StablePerm} stability relative to $\le$ and and
+  Then, denoting by \var{StablePerm} stability relative to $\le$ and
   by $\var{StablePerm}'$ stability relative to $\le'$,
@@
   \end{equation*}
-\end{theorem}
+\end{lemma}
 \begin{proof}
@@
 Now we can prove the stability invariant:
-\radixsortstableinv*
+\radixsortstableinv
 \begin{proof}
@@
 
-We would also like to prove \cref{thm:radixsort_sorted_inv}. To do so,
-we need to be able to reason about the process of adding a column to
-the front of a sorted matrix and sorting on that column. Intuitively,
-we need to know that if we have a matrix that is sorted on the first
-$j$ columns, and we add a new column to the front and sort on it, the
-result list will be sorted on the first $j+1$ columns.
-\begin{restatable}[hdsort\_sorted\_S]{theorem}{hdsortsortedS}
+We would also like to prove the sorted invariant
+(\cref{thm:radixsort_sorted_inv}). To do so, we need to be able to
+reason about the process of adding a column to the front of a sorted
+matrix and sorting on that column. Intuitively, we need to know that
+if we have a matrix that is sorted on the first $j$ columns, and we
+add a new column to the front and sort on it, the result will be
+sorted on the first $j+1$ columns.
+\begin{restatable}[hdsort\_sorted\_S]{lemma}{hdsortsortedS}
   \label{thm:hdsortsortedS}
@@
 
-\radixsortsortedinv*
+\radixsortsortedinv
 \begin{proof}
@@
   \end{equation*}
-  By \cref{thm:hdsortsortedS},
+  By \cref{thm:hdsortsortedS}, this is true if the tail of the matrix
+  was already sorted on the first $j$ columns.
   \begin{equation*}
@@
   by removing the first character and sticking it on the end.
-\item Sort the rows lexicographically
+\item Sort the rows lexicographically.
 \item Take the last column of the matrix. That string is the result of
@@
 being sorted using any sorting algorithm we find convenient.
-Specifically, if we examine how an LSD radixsort sorts the rotation
+Specifically, if we examine how an LSD radix sort sorts the rotation
 matrix (\cref{fig:radixsort_rots}) we notice an interesting pattern.
@@
   \end{tt}
-  \caption{Sorting the rotation matrix with LSD radixsort. The
+  \caption{Sorting the rotation matrix with LSD radix sort. The
     highlighted column contains the sorted original string.}
@@
 We also can notice that the columns to the right of the sorted column
-move with it for similar reasons. The column directly to its right is
-the result of sorting the input string, and then stably sorting it
-again by the character that precedes it in the original string. The
-next column is the result of sorting by each character, sorting by the
-preceding character, and then sorting by the character 2 spaces in
-front.
+move with it for similar reasons. The column directly to the right of
+the shaded column is the result of sorting the input string, and then
+stably sorting it again by the character that precedes it in the
+original string. The next column is the result of sorting by each
+character, sorting by the preceding character, and then sorting by the
+character 2 spaces in front.
 
@@
 last column to the growing matrix and stably sorting by it
-(\cref{fig:recreate}. As the matrix grows from right to left, we are
+(\cref{fig:recreate}). As the matrix grows from right to left, we are
 mimicing the process of LSD radix sort in \cref{fig:radixsort_rots},
@@
 \end{equation*}
-where \var{hdsort} sorts on the first column of a matrix,
-$\var{prefix\_shift}~c$ adds $c$ as the first column of a matrix,
-dropping the last column, and $m$ is the rotation matrix. This means
-as we repeatedly prefix $l$ and sort, we preserve the ``rest of the
-matrix'', even the parts that we have yet to reconstruct.
+where $\var{prefix\_shift}~c$ adds $c$ as the first column of a
+matrix, dropping the last column, and $m$ is the rotation matrix. This
+means as we repeatedly prefix $l$ and sort, we preserve the ``rest of
+the matrix'', even the parts that we have yet to reconstruct.
 
@@
 
-First, we define the left and right rotations of a list. \var{lrot}
-moves the first element of a list to the end.
+First, we define the left rotation of a list. \var{lrot} moves the
+first element of a list to the end.
 \begin{lstlisting}
@@
 \end{lstlisting}
-\var{rrot} moves the last element of a list to the front, and
-relies on \var{init}, which drops the last element of a list.
-\begin{lstlisting}
-Definition rrot (l: list A) : list A :=
-  match l with
-  | [] => []
-  | hd :: tl => last l hd :: init l
-  end.
-\end{lstlisting}
-Note that in order to ensure that $\var{last}$ is total, it takes
-an extra argument to return in the case of an empty list.
-
 Now we produce the rotation matrix by repeatedly applying \var{lrot}.
@@
 \end{equation*}
-We use the notation $f^i z$ for $\var{rep}~f~i~z$, which implements
-repeated function application. We will also use the notation $l_i$ for
-the $i$th element of a list, though in Coq, that is given by
-$\var{nth}~i~l~d$, where $d:A$ is a default value in case $i \ge
-\var{length}~l$.
-
+\begin{lstlisting}
+Fixpoint iter (f : A -> A) (n : nat) (z : A)
+  : list A :=
+  match n with
+  | O   => []
+  | S m => z :: iter f m (f z)
+  end.
+\end{lstlisting}
 Now we can define \var{rots} to produce the rotation matrix.
@@
 \var{bwl} produces the transformed string by taking the last column of
-the sorted rotation matrix. \var{lexsort} sorts the rotation matrix
-using insertion sort (\cref{appendix:insertion_sort}) according to the
-induced lexicographic ordering(\cref{sec:lex_ord}).
+the sorted rotation matrix. \var{lexsort} lexicographically sorts the
+rotation matrix using insertion sort (\cref{appendix:insertion_sort}).
 \begin{lstlisting}
@@
 \begin{lstlisting}
+Fixpoint findIndex (x : A) (ls : list A) : nat :=
+  match ls with
+  | [] => 0
+  | hd :: tl =>
+    if x == hd then 0
+    else S (findIndex x tl)
+  end.
+
 Definition bwn (l : list A) : nat :=
@@
 
-Given these definitions, we can quickly prove that \var{bwn}
-satisfies its specification using a few lemmas.
+Given these definitions, we can quickly prove that \var{bwn} satisfies
+its specification.
+
 \begin{lemma}[findIndex\_correct]
@@
 \end{lemma}
+\begin{proof}
+  This follows by induction on $l$: At every step, either the
+  current element is equivalent to $x$ and we return $0$, or there
+  must be such a $y$ in the rest of the list.
+\end{proof}
+
 \begin{lemma}[orig\_in\_sorted\_rots]
@@
 \end{lemma}
+\begin{proof}
+  Clearly $l$ is in $\var{rots}~l$ at index 0, and because
+  \var{lexsort} produces a permutation of its input, $l$ must be
+  in $\var{lexsort}~(\var{rots}~l)$ as well.
+\end{proof}
+
 \begin{restatable}[bwn\_correct]{theorem}{bwncorrect}
@@
 \end{restatable}
-
-\section{Impelementing the inverse BWT}
+\begin{proof}
+  This follows from \var{findIndex\_correct} if we can show that
+  $\exists y \in xs,~x = y$, which we have by \var{orig\_in\_sorted\_rots}.
+\end{proof}
+
+\section{Implementing the inverse BWT}
 \label{sec:inverse_BWT}
@@
 
-More precisely, suppose we had a function function \var{recreate}
+More precisely, suppose we had a function \var{recreate}
 such that
@@
   end.
+
+Definition prepend_col
+  : list A -> list (list A) -> list (list A)
+  := zipWith cons.
+
+Fixpoint zipWith (a : list A) (b : list B)
+  : list C :=
+  match (a, b) with
+  | (ahd :: atl, bhd :: btl) =>
+    f ahd bhd :: zipWith atl btl
+  | _ => []
+  end.
 \end{lstlisting}
-This definition relies on \var{prepend\_col}, which prepends a column
-to a matrix.
 
@@
   For all lists $l$ and $j \le \var{length}~l$,
-  $\var{recreate}~j~(\var{bwl}~l) =
-  \var{cols}~j~(\var{lexsort}~(\var{rots}~l))$
+  \begin{equation*}
+  \var{recreate}~j~(\var{bwl}~l) = \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
+  \end{equation*}
 \end{restatable*}
@@
 implementation of \var{unbwt} is correct along the same lines as
-before. To do so, we need to prove a few other theorems that
-have already come up when explaining the algorithm conceptually.
-
-First, we note that moving the last column of the rotation matrix to the front
-is the same as moving the last row to the top, because the rotation
-matrix is symmetric.
+before. To do so, we need to prove a few other theorems.
+
+First, we note that moving the last column of the rotation matrix to
+the front is the same as moving the last row to the top, because the
+rotation matrix is symmetric.
 \begin{theorem}[map\_rrot\_rots]
@@
 Now we can prove that the last column is indeed a fixpoint.
-\begin{theorem}[lexsort\_rots\_hdsort]
+\begin{lemma}[lexsort\_rots\_hdsort]
   For all lists $l$,
@@
   \end{equation*}
-\end{theorem}
-\begin{proof}
-  By \cref{thm:sort_ord_unique}, any two sorts are equivalent, and we
-  can replace \var{lexsort} with an LSD radixsort implemented as
-  \begin{equation*}
-    \var{radixsort}~m \coloneqq (\var{hdsort} \comp \var{map}~\var{rrot})^n~m
-  \end{equation*}
-  where $m$ is a matrix with $n$ columns. In this case, we are sorting
-  the rotation matrix and we let $n = \var{length}~l$. Then we have
+\end{lemma}
+\begin{proof}
+  By \cref{thm:sort_ord_unique}, any two sorts on an order are
+  equivalent, and we can replace \var{lexsort} with \var{radixsort}.
   \begin{align*}
@@
   The last step follows because the function $\var{hdsort} \comp
-  \var{rrot}$ permutes the rotation matrix.
+  \var{rrot}$ permutes the rotation matrix right before it is sorted
+  by \var{lexsrot}, so it doesn't change the final result.
 \end{proof}
@@
 
-\begin{theorem}[cols\_hdsort\_comm]
+\begin{lemma}[cols\_hdsort\_comm]
   For all matrices $m$ and naturals $j$,
@@
   and $\var{cols}~(j+1)$ doesn't alter the first elements of the rows
-  of $m$.\footnote{This argument is made precise by
-    \var{key\_sort\_inv} in \cref{sec:lex_ord}.}
-\end{proof}
-\begin{theorem}[cols\_map\_rrot]
-  For all $j$ and matrices $m$ that has rows all of length less than $j$,
+  of $m$.
+\end{proof}
+
+\begin{lemma}[cols\_map\_rrot]
+  For all matrices $m$ and $j$ such that for all rows $r$,
+  $\var{length}~r > j$,
   \begin{gather*}
@@
   \end{gather*}
-\end{theorem}
+\end{lemma}
 \begin{proof}
@@
 \begin{proof}
-  Letting $m = \var{lexsort}~(\var{rots}~l)$,
+  We reason by induction on $j$. Letting $m =
+  \var{lexsort}~(\var{rots}~l)$,
   \begin{align*}
@@
     \var{hdsort}~(\var{prepend\_col}~(\var{bwl}~l)~(\var{recreate}~j~(\var{bwl}~l))) \\
-    ={}& \var{recreate}~j~(\var{bwl}~l) \qedhere
+    ={}& \var{recreate}~(j + 1)~(\var{bwl}~l) \qedhere
   \end{align*}
@@
 Standard implementations of the BWT do not actually reconstruct the
-entire rotation matrix, as this would be inefficient. Instead, they
-avoid this using the optimization described in this section. Although,
-we did not prove or implement this optimization in Coq, it would be
+entire rotation matrix, as this would be inefficient. They avoid this
+using the optimization described in this section. Although, we did not
+prove or implement this optimization in Coq, it would be
 straightforward to do so.
@@
   : list (list A) :=
-  transpose (iter (apply p) (length l) (apply p l))
+  let f := apply p l in
+  transpose (iter (apply p) (length l) f)
 \end{lstlisting}
-In the definition above, we are repeatedly applying $p$ to the
-first column of the rotation matrix, $\var{apply}~p~l$. Then we
-take the transpose to create the matrix from the columns.
 
@@
 successively take the image of $i$ under $p$ to generate the indices
-into $l$ where those characters appear. This is justified via the
+into $l$ where those characters appear.
+\begin{lstlisting}
+Definition image (p : list nat) i := nth i p 0.
+\end{lstlisting}
+This is justified via the
 following theorem:
@@
 forwards BWT, basing our work on Bird et~al.'s program calculation
-derivation \cite{birdmu,pearls}. In the process, we defined the
-properties of stable sorts, and implemented and proved correct an LSD
-radixsort.
+derivation \cite{birdmu,pearls}. In the process, we defined stable
+sorts, proved that all stable sorts are equivalent, and implemented
+and proved correct an LSD radix sort.
 
@@
 step. Most of the time and effort went into proving the correctness of
-radixsort. If we exclude the parts related to sorting or to other
+LSD radix sort. If we exclude the parts related to sorting or to other
 general-purpose functions not specific to the BWT, the entire verified
@@
 \cref{sec:opt}, can be used as a functional specification to for a
-verified implementation of bzip2. Of course, it would also be
-necessary to have functional specifications for the other phases of
-bzip2, such as Run-Length Encoding, Move-to-Front Encoding, and
-arithmetic coding. As luck would have it, Bird has already analyzed
-arithmetic coding \cite{pearls}, and the other two phases are
+verified implementation of bzip2 \cite{tsai_2016}. Of course, it would
+also be necessary to have functional specifications for the other
+phases of bzip2, such as Run-Length Encoding, Move-to-Front Encoding,
+and arithmetic coding. As luck would have it, Bird has already
+analyzed arithmetic coding \cite{pearls}, and the other two phases are
 relatively simple.
@@
 
+\nocite{*}
 \bibliography{draft_paper}
 
+\clearpage
+
 \appendix
 
-\section{Additional Proofs}
-\label{appendix:proofs}
+\section{Sorted}
+\label{appendix:sorted}
+
+There is a third definition for \var{Sorted} that is very similar to
+\cref{def:sorted}, but uses transitivity to avoid parametrizing over
+every element in the list.
+\begin{definition}[SortedLocal]
+  \begin{gather*}
+    \infer[\var{SortedLocal\_nil}]{\var{SortedLocal}~[]}{}
+    \\
+    \\
+    \infer[\var{SortedLocal\_1}]{\var{SortedLocal}~[a]}{}
+    \\
+    \\
+    \infer[\var{SortedLocal\_cons}]{\var{SortedLocal}~(a :: b :: l)}{%
+      a < b & \var{SortedLocal}~(b :: l)
+    }
+  \end{gather*}
+\end{definition}
 
@@
 
-\section{Insertion sort}
-\label{appendix:insertion_sort}
-
-Insertion sort is a simple and elegant stable sort that is easy to
-reason about. The theorems proved in \cref{sec:unique} allow us to
-prove things using insertion sort that apply to any stable sort. For
-example, our proof of correctness for radixsort relies on the fact
-that it uses insertion sort as the inner sort, but
-\cref{thm:stablesort_unique} allows us to extend that proof to a
-radixsort that uses any stable sort. We also use it as the
-implementation of \var{lexsort} because it is convenient, but
-similarly \cref{thm:sort_ord_unique} ensures that it can be replaced with
-any other sort.
-
+\section{Permutation}
+\label{appendix:perm}
+
+\permutationcountiff*
+\begin{proof}
+  The ($\implies$) direction follows from induction in the evidence of
+  $\var{Permutation}~l~l'$.
+
+  In the ($\impliedby$) direction, we reason by induction on $l$. In
+  the inductive case, we have to show $\var{Permutation}~(a::l)~l'$.
+  We know that $a \in l'$ because $\var{count\_occ}~a~l' =
+  \var{count\_occ}~a~(a::l) > 0$. That means we can deconstruct $l'$
+  as $L \dplus a :: R$, and
+  \begin{equation*}
+    L \dplus a :: R \eqperm a :: L \dplus R \eqperm a :: l
+  \end{equation*}
+  where the last step follows from the induction hypothesis.
+\end{proof}
+
+We can define the preimage of an index under a permutation function.
 \begin{lstlisting}
-  Fixpoint insert (x : A) (l : list A) :=
-  match l with
-  | [] => [x]
-  | h :: t =>
-  if le_dec x h then x :: h :: t else h :: insert x t
-  end.
-
-  Fixpoint sort (l : list A) : list A :=
-  match l with
-  | [] => []
-  | h :: t => insert h (sort t)
-  end.
+Definition preimage (p : list nat) i := findIndex i p.
 \end{lstlisting}
 
-\begin{lemma}[insert\_perm]
-  For all lists $l$ and elements $x$,
-  \begin{equation*}
-    \var{Permutation}~(x::l)~(\var{insert}~x~l)
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  We reason by induction on $l$. In the inductive case, we need to show
-  \begin{equation*}
-    \var{Permutation}~(x::h::t)~(\var{insert}~x~(h::t))
-  \end{equation*}
-  given the induction hypothesis that
-  \begin{equation*}
-    \var{Permutation}~(x::t)~(\var{insert}~x~t)
-  \end{equation*}
-  If $x \le h$, then $\var{insert}~x~(h::t) = x :: h :: t$ and we are
-  done. Otherwise, $\var{insert}~x~(h::t) = h :: \var{insert}~x~t$ and
-  we can apply the induction hypothesis.
-\end{proof}
-
-\begin{theorem}[sort\_perm]
-  For all lists $l$,
-  \begin{equation*}
-    \var{Permutation~l}~(\var{sort}~l)
-  \end{equation*}
+We will not prove it here, but $\var{preimage}~p$ and $\var{image}~p$
+are both of bijections of type $\mathbb{N}_{<n} \to \mathbb{N}_{<n}$ and
+are inverses of each other. We also have the following two theorems:
+\begin{theorem}[nth\_preimage\_apply, nth\_image\_apply]
+  For all lists $l$, permutations $p$, and indices $i$,
+  \begin{align*}
+    (\var{apply}~p~l)_{p^{-1}(i)} = l_i && l_{p(i)} = (\var{apply}~p~l)_i
+  \end{align*}
 \end{theorem}
 \begin{proof}
-  The proof follows by induction on $l$. The inductive case is
-  \begin{equation*}
-    h :: t \eqperm h :: \var{sort}~t \eqperm
-    \var{insert}~h~(\var{sort}~t) = \var{sort}~(h :: t)
-  \end{equation*}
-  where the first step follows from the induction hypothesis, and the
-  second from \var{insert\_perm}.
-\end{proof}
-
-\begin{lemma}[insert\_sorted]
-  For all lists $l$ and elements $x$,
-  \begin{equation*}
-    \var{Sorted}~l \implies \var{Sorted}~(\var{insert}~x~l)
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  We reason by induction on the proof that $\var{Sorted}~l$. In the
-  inductive case we need prove that
-  \begin{equation*}
-    \var{Sorted}~(\var{insert}~x~(h::t))
-  \end{equation*}
-  If $x \le h$, then $\var{insert}~x~(h::t) = x :: h :: t$, and by
-  \var{SortedLocal\_cons} we are done.
-
-  If $x > h$, $\var{insert}~x~(h::t) = h :: \var{insert}~x~t$, and it
-  suffices to show that $\forall y \in \var{insert}~x~t,~h \le y$. By
-  \var{insert\_perm}, this is equivalent to $\forall y \in x::t,~h \le y$ From
-  the the fact that $\var{Sorted}~(h::t)$, we know that $\forall y \in t,~h \le
-  y$, and we have that $h < x$, so we are done.
-\end{proof}
-
-\begin{theorem}[sort\_sorted]
-  For all lists $l$,
-  \begin{equation*}
-    \var{Sorted}~(\var{sort}~l)
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  The proof follows by induction on $l$ and \var{insert\_sorted}.
-\end{proof}
-
-\begin{lemma}[insert\_stable]
-  For all lists $l$ and elements $x$,
-  \begin{equation*}
-    \var{StablePerm}~(x::l)~(\var{insert}~x~l)
-  \end{equation*}
-\end{lemma}
-\begin{proof}
-  We reason by induction on $l$. In the inductive case, we need to show
-  \begin{equation*}
-    \var{StablePerm}~(x::h::t)~(\var{insert}~x~(h::t))
-  \end{equation*}
-  given the induction hypothesis that
-  \begin{equation*}
-    \var{StablePerm}~(x::t)~(\var{insert}~x~t)
-  \end{equation*}
-  If $x \le h$, then $\var{insert}~x~(h::t) = x :: h :: t$ and we are
-  done. Otherwise, $x > h$ and $\var{insert}~x~(h::t) = h ::
-  \var{insert}~x~t$.
-  \begin{equation*}
-    x :: h :: t \eqstable h :: x :: t \eqstable h :: \var{insert}~x~t
-  \end{equation*}
-  where the first step follows from the fact that $x \not\equiv h$ and the
-  second from the induction hypothesis.
-\end{proof}
-
-\begin{theorem}[sort\_stable]
-  For all lists $l$,
-  \begin{equation*}
-    \var{StablePerm}~l~(\var{sort}~l)
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  The proof is almost identical to the proof for \var{sort\_perm}.
-\end{proof}
-
-\begin{theorem}[sort\_StableSort]
-  \begin{equation*}
-    \var{StableSort}~\var{sort}
-  \end{equation*}
-\end{theorem}
-\begin{proof}
-  This follows from \var{sort\_sorted} and \var{sort\_stable}.
-\end{proof}
-
-\section{\var{hdsorted\_sorted\_S}}
-\label{appendix:hdsort_sorted_S}
-
-\begin{lemma}[insert\_Sorted\_S]
-  For all matrices $m$, rows $r$, and naturals $j$,
-  \begin{gather*}
-    \var{PrefixSorted}~(j+1)~(\var{hdsort}~m) \implies \\
-    \var{PrefixSorted}~j~(\var{tl}~r :: \var{map}~{tl}~m) \implies \\
-    \var{PrefixSorted}~(j+1)~(\var{insert}_1~r~(\var{hdsort}~m))
-  \end{gather*}
-  where $\var{insert}_1$ is \var{insert} that uses the 1-prefix
-  lexicographic ordering used by \var{hdsort}.
-\end{lemma}
-\begin{proof}
-  We can destruct $\var{insert}_1~r~(\var{hdsort}~m)$ as $m_1 \dplus r
-  :: m_2$, where $m_1 \dplus m_2 = \var{hdsort}~m$. Then by
-  \var{Sorted\_app\_cons}, it suffices to show
-  \begin{gather*}
-    \var{PrefixSorted}~(j+1)~m_1 \land \\
-    \var{PrefixSorted}~(j+1)~(r :: m_2) \land \\
-    (\forall x \in m_1,~ r >_{j+1} x)
-  \end{gather*}
-  \begin{itemize}
-  \item $\var{PrefixSorted}~(j+1)~m_1$ follows from the fact that
-    $\var{hdsort}~m = m_1 \dplus m_2$ and
-    $\var{PrefixSorted}~(j+1)~(\var{hdsort}~m)$.
-  \item By similar reasoning, we know that
-    $\var{PrefixSorted}~(j+1)~m_2$, so in order to show
-    $\var{PrefixSorted}~(j+1)~(r :: m_2)$ all we need to prove is that
-    for every row $x$ in $m_2$,
-    \begin{equation*}
-      r \le_{j+1} x
-    \end{equation*}
-    We know from the fact that $\var{insert}_1~r~(\var{hdsort}~m) =
-    m_1 \dplus r :: m_2$ that $r \le_1 x$. Then by \var{key\_firstn\_S},
-    we only need to show that
-    \begin{equation*}
-      \var{tl}~r \le_j \var{tl}~x
-    \end{equation*}
-    This follows from $\var{PrefixSorted}~j~(\var{tl}~r :: \var{map}~{tl}~m)$.
-  \item We have that for any $x \in m_1$, $r >_1 x$ by the fact that
-    $\var{insert}_1~r~(\var{hdsort}~m) = m_1 \dplus r :: m_2$. This
-    gives us $r >_{j+1} x$ by \var{key\_lt\_firstn\_ge}. \qedhere
-  \end{itemize}
-\end{proof}
-
-\hdsortsortedS*
-\begin{proof}
-  This follows by induction from \var{insert\_Sorted\_S}
-\end{proof}
-
-\section{\var{Permutation} definitions equivalence}
-\label{appendix:perm_def_eq}
+  The first equation (\var{nth\_preimage\_apply}) holds because:
+  \begin{align*}
+       (\var{apply}~p~l)_{p^{-1}(i)}
+    &= (\var{map}~(\lambda j.~l_j)~p)_{\var{findIndex}~i~p} \\
+    &= l_{p_{(\var{findIndex}~i~p)}} \\
+    &= l_i
+  \end{align*}
+  and the second follows from this:
+  \begin{equation*}
+      (\var{apply}~p~l)_i
+    = (\var{apply}~p~l)_{p^{-1}(p(i))}
+    = l_{p(i)}
+    \qedhere
+  \end{equation*}
+\end{proof}
 
@@
 
-\begin{theorem}[PermutationEx\_iff]
-  \label{thm:permutationex_iff}
-  For all lists $l$ and $l'$,
-  \begin{equation*}
-    \var{Permutation}~l~l' \iff \var{PermutationEx}~l~l'
-  \end{equation*}
-\end{theorem}
+\permutationexiff*
 \begin{proof}
@@
 
-\section{\var{StablePerm} definitions equivalence}
-\label{appendix:stableperm_def_eq}
+\section{Stability}
+\label{appendix:stability}
 
@@
 
-\begin{theorem}[StablePermInd\_iff]
-  For all lists $l$ and $l'$
-  \begin{equation*}
-    \var{StablePerm}~l~l' \iff \var{StablePermInd}~l~l'
-  \end{equation*}
-\end{theorem}
+\stablepermindiff*
 \begin{proof}
@@
 
-\begin{theorem}[StablePermEx\_iff]
-  For all lists $l$ and $l'$
-  \begin{equation*}
-    \var{StablePerm}~l~l' \iff \var{StablePermEx}~l~l'
-  \end{equation*}
-\end{theorem}
+\stablepermexiff*
 \begin{proof}
@@
   \cref{thm:apply_correct_stable}.
+\end{proof}
+
+\section{Insertion sort}
+\label{appendix:insertion_sort}
+
+\begin{lstlisting}
+Fixpoint insert (x : A) (l : list A) :=
+  match l with
+  | [] => [x]
+  | h :: t =>
+    if le_dec x h then x :: h :: t else h :: insert x t
+  end.
+
+Fixpoint sort (l : list A) : list A :=
+  match l with
+  | [] => []
+  | h :: t => insert h (sort t)
+  end.
+\end{lstlisting}
+
+\begin{lemma}[insert\_perm]
+  For all lists $l$ and elements $x$,
+  \begin{equation*}
+    \var{Permutation}~(x::l)~(\var{insert}~x~l)
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  We reason by induction on $l$. In the inductive case, we need to show
+  \begin{equation*}
+    \var{Permutation}~(x::h::t)~(\var{insert}~x~(h::t))
+  \end{equation*}
+  given the induction hypothesis that
+  \begin{equation*}
+    \var{Permutation}~(x::t)~(\var{insert}~x~t)
+  \end{equation*}
+  If $x \le h$, then $\var{insert}~x~(h::t) = x :: h :: t$ and we are
+  done. Otherwise, $\var{insert}~x~(h::t) = h :: \var{insert}~x~t$ and
+  we can apply the induction hypothesis.
+\end{proof}
+
+\begin{theorem}[sort\_perm]
+  For all lists $l$,
+  \begin{equation*}
+    \var{Permutation~l}~(\var{sort}~l)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  The proof follows by induction on $l$. The inductive case is
+  \begin{equation*}
+    h :: t \eqperm h :: \var{sort}~t \eqperm
+    \var{insert}~h~(\var{sort}~t) = \var{sort}~(h :: t)
+  \end{equation*}
+  where the first step follows from the induction hypothesis, and the
+  second from \var{insert\_perm}.
+\end{proof}
+
+\begin{lemma}[insert\_sorted]
+  For all lists $l$ and elements $x$,
+  \begin{equation*}
+    \var{Sorted}~l \implies \var{Sorted}~(\var{insert}~x~l)
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  We reason by induction on the proof that $\var{Sorted}~l$. In the
+  inductive case we need prove that
+  \begin{equation*}
+    \var{Sorted}~(\var{insert}~x~(h::t))
+  \end{equation*}
+  If $x \le h$, then $\var{insert}~x~(h::t) = x :: h :: t$, and by
+  \var{SortedLocal\_cons} we are done.
+
+  If $x > h$, $\var{insert}~x~(h::t) = h :: \var{insert}~x~t$, and it
+  suffices to show that $\forall y \in \var{insert}~x~t,~h \le y$. By
+  \var{insert\_perm}, this is equivalent to $\forall y \in x::t,~h \le y$ From
+  the the fact that $\var{Sorted}~(h::t)$, we know that $\forall y \in t,~h \le
+  y$, and we have that $h < x$, so we are done.
+\end{proof}
+
+\begin{theorem}[sort\_sorted]
+  For all lists $l$,
+  \begin{equation*}
+    \var{Sorted}~(\var{sort}~l)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  The proof follows by induction on $l$ and \var{insert\_sorted}.
+\end{proof}
+
+\begin{lemma}[insert\_stable]
+  For all lists $l$ and elements $x$,
+  \begin{equation*}
+    \var{StablePerm}~(x::l)~(\var{insert}~x~l)
+  \end{equation*}
+\end{lemma}
+\begin{proof}
+  We reason by induction on $l$. In the inductive case, we need to show
+  \begin{equation*}
+    \var{StablePerm}~(x::h::t)~(\var{insert}~x~(h::t))
+  \end{equation*}
+  given the induction hypothesis that
+  \begin{equation*}
+    \var{StablePerm}~(x::t)~(\var{insert}~x~t)
+  \end{equation*}
+  If $x \le h$, then $\var{insert}~x~(h::t) = x :: h :: t$ and we are
+  done. Otherwise, $x > h$ and $\var{insert}~x~(h::t) = h ::
+  \var{insert}~x~t$.
+  \begin{equation*}
+    x :: h :: t \eqstable h :: x :: t \eqstable h :: \var{insert}~x~t
+  \end{equation*}
+  where the first step follows from the fact that $x \not\equiv h$ and the
+  second from the induction hypothesis.
+\end{proof}
+
+\begin{theorem}[sort\_stable]
+  For all lists $l$,
+  \begin{equation*}
+    \var{StablePerm}~l~(\var{sort}~l)
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  The proof is almost identical to the proof for \var{sort\_perm}.
+\end{proof}
+
+\begin{theorem}[sort\_StableSort]
+  \begin{equation*}
+    \var{StableSort}~\var{sort}
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  This follows from \var{sort\_sorted} and \var{sort\_stable}.
+\end{proof}
+
+\section{\var{hdsorted\_sorted\_S}}
+\label{appendix:hdsort_sorted_S}
+
+\begin{lemma}[insert\_Sorted\_S]
+  For all matrices $m$, rows $r$, and naturals $j$,
+  \begin{gather*}
+    \var{PrefixSorted}~(j+1)~(\var{hdsort}~m) \implies \\
+    \var{PrefixSorted}~j~(\var{tl}~r :: \var{map}~{tl}~m) \implies \\
+    \var{PrefixSorted}~(j+1)~(\var{insert}_1~r~(\var{hdsort}~m))
+  \end{gather*}
+  where $\var{insert}_1$ is \var{insert} that uses the 1-prefix
+  lexicographic ordering used by \var{hdsort}.
+\end{lemma}
+\begin{proof}
+  We can destruct $\var{insert}_1~r~(\var{hdsort}~m)$ as $m_1 \dplus r
+  :: m_2$, where $m_1 \dplus m_2 = \var{hdsort}~m$. Then by
+  \var{Sorted\_app\_cons}, it suffices to show
+  \begin{gather*}
+    \var{PrefixSorted}~(j+1)~m_1 \land \\
+    \var{PrefixSorted}~(j+1)~(r :: m_2) \land \\
+    (\forall x \in m_1,~ r >_{j+1} x)
+  \end{gather*}
+  \begin{itemize}
+  \item $\var{PrefixSorted}~(j+1)~m_1$ follows from the fact that
+    $\var{hdsort}~m = m_1 \dplus m_2$ and
+    $\var{PrefixSorted}~(j+1)~(\var{hdsort}~m)$.
+  \item By similar reasoning, we know that
+    $\var{PrefixSorted}~(j+1)~m_2$, so in order to show
+    $\var{PrefixSorted}~(j+1)~(r :: m_2)$ all we need to prove is that
+    for every row $x$ in $m_2$,
+    \begin{equation*}
+      r \le_{j+1} x
+    \end{equation*}
+    We know from the fact that $\var{insert}_1~r~(\var{hdsort}~m) =
+    m_1 \dplus r :: m_2$ that $r \le_1 x$. Then by \var{key\_firstn\_S},
+    we only need to show that
+    \begin{equation*}
+      \var{tl}~r \le_j \var{tl}~x
+    \end{equation*}
+    This follows from $\var{PrefixSorted}~j~(\var{tl}~r :: \var{map}~{tl}~m)$.
+  \item We have that for any $x \in m_1$, $r >_1 x$ by the fact that
+    $\var{insert}_1~r~(\var{hdsort}~m) = m_1 \dplus r :: m_2$. This
+    gives us $r >_{j+1} x$ by \var{key\_lt\_firstn\_ge}. \qedhere
+  \end{itemize}
+\end{proof}
+
+\hdsortsortedS*
+\begin{proof}
+  This follows by induction from \var{insert\_Sorted\_S}
 \end{proof}

### 2019-10-18 draft_paper/draft_paper.tex
Finish revising appendices

@@
 algorithm. A sort is stable if it preserves the relative order of
-equal elements. For example, if we are stably sorting the list
+equivalent elements. For example, if we are stably sorting the list
 $[\var{mix},\ \allowbreak\var{fog},\ \allowbreak\var{age},\ \allowbreak\var{fit}]$
@@
 $[\var{age},\ \allowbreak\var{fit},\ \allowbreak\var{fog},\ \allowbreak\var{mix}]$.
-Even though both are sorted on the first character, because \var{fog}
+Even though both are sorted on the first characters, because \var{fog}
 preceded \var{fit} in the original list, it must also precede
@@
 \cite{McIlroy93,Bentley:1997:FAS:314161.314321,10.1007/978-3-540-89097-3_3}.
-Most radix sorts rely on a some stable sorting subroutine.
+Most radix sorts rely on a some stable sorting subroutine, as does
+least-significant-digit (LSD) radix sort (\cref{sec:radix_sort}).
 
@@
 Throughout our proofs about LSD radix sort and the BWT, we reason
-about sorting algorithms whose implementation is irrelevent as long as
+about sorting algorithms whose implementation is irrelevant as long as
 they are correct. We could, in principle, rely only on the fact that
@@
 inductive definition of permutations can produce unhelpful induction
-hypotheses, especially in the transitive case. Insetad, we reason
-about specifically insertion sort. Therefore
-\cref{thm:stablesort_unique} is quite useful, because we can reason
-using insertion sort, and later on replace it with any other
+hypotheses, especially in the transitive case. Instead, we reason
+about specifically insertion sort. By \cref{thm:stablesort_unique} we
+can always replace insertion sort (which is slow) with any other
 (presumably faster) stable sort.
@@
 last name, birthday, etc. More precisely, for sorting we need a total,
-decideable preorder: a \textit{preorder} is a relation $\le$ that is
+decidable preorder: a \textit{preorder} is a relation $\le$ that is
 reflexive and transitive, it is \textit{total} if we can compare any
@@
 
-Given a total decideable preorder, we can define when a list is in
-ascending order with respect to that preorder. In this case, as will
-be the case with other properties we will discuss, there are multiple
-reasonable definitions. Our approach will be to state all the
-definitions we can think of and to prove equivalence between them.
-This increases our confidence that we have defined things correctly,
-and allows us to choose whichever definition we find most convenient
-for a given proof.
+Given a preorder, we can define when a list is in ascending order with
+respect to that preorder. In this case, as will be the case with other
+properties we will discuss, there are multiple reasonable definitions.
+Our approach will be to state all the definitions we can think of and
+to prove equivalence between them. This increases our confidence that
+we have defined things correctly, and allows us to choose whichever
+definition we find most convenient for a given proof.
 
@@
 the Coq standard library. This relation holds on two lists when they
-contain all the same elements, and is defined inductively:
+contain all the same elements, and is defined inductively.
 \begin{definition}[Permutation]
@@
 \end{equation*}
-For consistency with Coq's built-in \var{Permuation} relation, in a
+For consistency with Coq's built-in \var{Permutation} relation, in a
 situation like this we will say that $l$ and $l'$ are permutations of
 one other, and that $p$ is the \textit{permutation function} that
-relates $l$ to $l'$.
-
-This suggests an alternative definition for \var{Permutation}:
+relates $l$ to $l'$. This suggests an alternative definition for
+\var{Permutation}.
+
 \begin{definition}[PermutationEx]
@@
 
-However, not any list of natural numbers represents a valid
-permutation function, because a permutation must be a bijection. We
-can define the property $\var{PermFun}~n~p$, which holds when $p :
-\var{list}~\nat$ represents a permutation function on $n$ elements.
+Not any list of natural numbers represents a valid permutation
+function, because a permutation must be a bijection. We can define the
+property $\var{PermFun}~n~p$, which holds when $p : \var{list}~\nat$
+represents a permutation function on $n$ elements.
 \begin{definition}[PermFun]
@@
 We should also mention a third definition for permutations that can be
-used on types with decideable equality:
+used on types with decidable equality.
 \begin{definition}[PermutationCount]
@@
 
-With this we can define the specification for sorting functions:
+With this we can define the specification for sorting functions.
 \begin{definition}[Sort]
@@
 Now that we've defined sorts, we can consider stable sorts. Stable
-sorts must preserve the relative orders of ``equivalent'' elements.
-This equivalence comes from the ordering we are using: any total
-decidable preorder also give us a decideable equivalence relation $\equiv$,
-where $x \equiv y$ when $x \le y$ and $x \ge y$.
+sorts must preserve the relative orders of equivalent elements. This
+equivalence comes from the ordering we are using: any total decidable
+preorder also gives us a decidable equivalence relation $\equiv$, where $x
+\equiv y$ when $x \le y$ and $x \ge y$.
 
@@
 \end{restatable}
+\begin{proof}
+  See \cref{appendix:stability}.
+\end{proof}
 
@@
 
-This can often be extended even to unstable sorting algorithms.
-Remember, it isn't always true that $x \equiv y$ implies $x = y$. For
-example, if we are sorting strings by their first character, then
-$\text{\texttt{fit}} \equiv \text{\texttt{fog}}$, but clearly
-$\text{\texttt{fit}} \neq \text{\texttt{fog}}$. When we do have that the
-equivalence ($\equiv$) implies Leibniz equality ($=$), we say that we have
-an \textit{order}, as we have when we are sorting strings using all of
-their characters. Under an order, as opposed to a preorder, equivalent
-elements are actually indistinguishable, so you can't tell if they
-have been swapped.
+This can be extended even to unstable sorting algorithms when $x \equiv y
+\implies x = y$. This isn't always true: for example, if we are
+sorting strings by their first character, then $\text{\texttt{fit}} \equiv
+\text{\texttt{fog}}$, but clearly $\text{\texttt{fit}} \neq
+\text{\texttt{fog}}$. When we do have that the equivalence ($\equiv$)
+implies Leibniz equality ($=$), we say that we have an \textit{order},
+as we have when we are sorting strings using all of their characters.
+Under an order, as opposed to a preorder, equivalent elements are
+actually indistinguishable, so you can't tell if they have been
+swapped.
 
@@
 is used to lexicographically sort strings of equal length. We can
-define what it means to sort lexicographically:
+define what it means to sort lexicographically.
 
@@
 sorting on the first column (\cref{fig:lsd_radixsort_func}). First, we
-define \var{rrot}, which moves the last element of a list to the
-front, and relies on \var{init}, which drops the last element of a
-list. Note that in order to ensure that $\var{last}$ is total, it
-takes an extra argument to return in the case of an empty list.
+define the right rotation of a list using \var{rrot}, which moves the
+last element of a list to the front. Our definition relies on
+\var{init}, which drops the last element of a list. Note that in order
+to ensure that $\var{last}$ is total, it takes an extra argument to
+return in the case of an empty list.
 \begin{lstlisting}
@@
 character of each string. We can implement this by defining orderings
-that compare elements using some key function:
+that compare elements using some key function.
 
 \begin{definition}[keyOrd]
-  Given a preorder on a type $K$ and a key function $\var{key} : A \to
-  K$, we can define a preorder on $A$ that compares all elements using
-  the key function. That is, we can define the relation
-  $\le_{\var{key}}$ on $A$ such that for all $x, y : A$
+  Given a preorder on a type~$K$ and a key function~$\var{key} : A \to
+  K$, we define a preorder~$\le_{\var{key}}$ on $A$ that compares all elements using
+  the key function.
   \begin{equation*}
@@
   \begin{align*}
-    xs \le_n ys &\coloneqq xs \le_{\var{firstn}~n} ys \\
+    xs \le_n ys &\coloneqq xs \le_{(\var{firstn}~n)} ys \\
               &= (\var{firstn}~n~xs) \lexle (\var{firstn}~n~ys)
@@
 after $j$ iterations of rotating and sorting, the matrix should be a
-stable permutation of the original matrix had we only rotated $j$
-times without sorting.
+stable permutation of the original matrix had we rotated $j$ times
+without sorting.
 
@@
 
-Let's generalize, and consider two orders $\le_j$ and $\le_k$ such that $j
-\le k$. If in the first $j$ elements we can conclude that $x <_j y$,
-then examining more elements can't change that.
+In general, let's consider the relationship between two orders $\le_j$
+and $\le_k$ such that $j \le k$. If when comparing two strings $x$ and
+$y$, in the first $j$ elements we can conclude that $x <_j y$, then
+examining more elements can't change that.
 
@@
 
-Now we can prove the stability invariant:
+Now we can prove the stability invariant.
 \radixsortstableinv
@@
 
-  Here we can apply the swap ruel if we can show that $\var{rrot}~x
+  Here we can apply the swap rule if we can show that $\var{rrot}~x
   \not\equivle \var{rrot}~y$. This follows from the fact that
@@
 enable fast matching of DNA subsequences \cite{dna}.
-
-To apply the BWT to a string of length $n$:
-\begin{enumerate}
-\item Form the \(n \times n\) matrix of all cyclic shifts of the
-  original string (referred to as the rotation matrix). Starting
-  with the original string, each row is formed from the previous one
-  by removing the first character and sticking it on the end.
-\item Sort the rows lexicographically.
-\item Take the last column of the matrix. That string is the result of
-  the transformation.
-\item Note the index where the original string appears in the sorted
-  rotation matrix. This piece of information is necessary to reverse
-  the transformation.
-\end{enumerate}
-As shown in \cref{fig:bw_ex}, applying the algorithm to the string
-\var{abracadabra!} produces the new string \var{ard!rcaaabb} and the
-index \var{3}.
 
@@
 
+To apply the BWT to a string of length $n$:
+\begin{enumerate}
+\item Form the \(n \times n\) matrix of all cyclic shifts of the
+  original string (referred to as the rotation matrix). Starting
+  with the original string, each row is formed from the previous one
+  by removing the first character and sticking it on the end.
+\item Sort the rows lexicographically.
+\item Take the last column of the matrix. That string is the result of
+  the transformation.
+\item Note the index where the original string appears in the sorted
+  rotation matrix. This piece of information is necessary to reverse
+  the transformation.
+\end{enumerate}
+As shown in \cref{fig:bw_ex}, applying the algorithm to the string
+\var{abracadabra!} produces the new string \var{ard!rcaaabb} and the
+index \var{3}.
+
 The BWT tends to produce runs like the three \var{a}s or the two
@@
 Although it is not obvious, the transformed string and the index
-provide enough information to recover the the original string.
-Conceptually, we can find the original string by using the transformed
-string to reconstruct the entire sorted rotation matrix, and then
-using the index to find the row containing the original string. This,
-though, relies on our ability to reconstruct the sorted rotation
-matrix from just its last column.
-
-A key insight is that because of \cref{thm:sort_ord_unique}, we can
-pretend that during the forwards transform, the rotation matrix is
-being sorted using any sorting algorithm we find convenient.
-Specifically, if we examine how an LSD radix sort sorts the rotation
-matrix (\cref{fig:radixsort_rots}) we notice an interesting pattern.
-Each column of the rotation matrix is a permutation of the original
-string, so when we sort a on particular column, we always produce the
-same result. As we move right-to-left, the column containing the
-sorted input string moves with us until it reaches the left edge of
-the matrix.
+provide enough information to recover the the original string. The
+crucial insight is that from the transformed string, which is the last
+column of the sorted rotation matrix, we can reconstruct the entire
+sorted rotation matrix. From there, we can find the row that contains
+the original string using the index.
+
+\begin{figure}
+  $
+  {\var{abra\textbf{c}}}%
+  {\underbrace{\var{adabra\textcolor{gray}{!abrac}}}}%
+  {\var{\textcolor{gray}{adabra!}}}
+  $
+  \caption{The last column of the sorted rotation matrix contains each
+    character of the original string, sorted by the $n$ characters
+    that follow it in the original string. This shows the $n$
+    characters that follow \var{c} are \var{adabra!abrac}, hence that
+    is ``\var{c}'s rotation''.}
+  \label{fig:last_col_key}
+\end{figure}
+
+How, though, does the last column encapsulate all the information in
+the sorted rotation matrix? First, we can notice that each column is a
+permutation of the original string. In particular, the last column
+contains each character of the original string, sorted by ``its
+rotation'': the $n$ characters that follow it in the original string
+(\cref{fig:last_col_key}).
+
+We can gain more insight if we examine how the last column is created
+when sorting the rotation matrix with an LSD radix sort as in
+\cref{fig:radixsort_rots} (this is justified by
+\cref{thm:sort_ord_unique}). After $j$ iterations, the column directly
+to the left of the shaded column (the ``next column'') contains all
+the characters in the original string sorted by the $j$ characters
+that follow it in the original string.
 
@@
   \end{tabular}
+  $\rightarrow$
+  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    acadabra&!&abr \\
+    racadabr&a&!ab \\
+    cadabra!&a&bra \\
+    !abracad&a&bra \\
+    abra!abr&a&cad \\
+    ra!abrac&a&dab \\
+    abracada&b&ra! \\
+    adabra!a&b&rac \\
+    bra!abra&c&ada \\
+    a!abraca&d&abr \\
+    bracadab&r&a!a \\
+    dabra!ab&r&aca
+  \end{tabular}
   $\rightarrow \cdots \rightarrow$
@@
   \caption{Sorting the rotation matrix with LSD radix sort. The
-    highlighted column contains the sorted original string.}
+    highlighted column is the last column that was sorted on.}
   \label{fig:radixsort_rots}
@@
 
-We also can notice that the columns to the right of the sorted column
-move with it for similar reasons. The column directly to the right of
-the shaded column is the result of sorting the input string, and then
-stably sorting it again by the character that precedes it in the
-original string. The next column is the result of sorting by each
-character, sorting by the preceding character, and then sorting by the
-character 2 spaces in front.
-
-This suggests that we can recreate the matrix from the last column by
-sorting it to produce the first column, then repeatedly prepending the
-last column to the growing matrix and stably sorting by it
-(\cref{fig:recreate}). As the matrix grows from right to left, we are
-mimicing the process of LSD radix sort in \cref{fig:radixsort_rots},
-in that the sorted portions to the right of the shaded column are
-identical.
+The ``next-column'' converges to the last column as the matrix is
+sorted, because on each iteration it can look one character further in
+the string. The last column has full knowledge because it can look at
+all $n$ characters. For example, in \cref{fig:radixsort_rots} we can
+see that the ``next column'' after three iterations is
+\var{ar!drcaaaabb}, which is very similar to the last column at the
+end of the process, \var{ard!rcaaaabb}. The only difference is the
+\var{!} and \var{d} are swapped, because the three characters that
+follow them are the same (\var{abr}), so we can't tell yet which
+should go first. By the end, we can see that \var{d}, whose rotation
+is \var{abra!abracad}, should go before \var{!}, whose rotation is
+\var{abracadabra!}.
+
+From this perspective, the last column is the best ``next column'' for
+the LSD radix sort, and we can use it at every iteration
+(\cref{fig:recreate}). This gives us a way to recover the entire
+sorted rotation matrix from the last column~$s$: starting with the
+matrix as just a single column $s$, sort the matrix so far by the
+first column, then prepend $s$ and repeat. As the matrix grows from
+right to left, we are mimicking the process of LSD radix sort in
+\cref{fig:radixsort_rots} in that the sorted portions to the right of
+the shaded column are identical.
 
@@
   \end{tabular}
+  $\rightarrow$
+  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
+    \textcolor{lightgray}{aaaaaaaa}&!&abr \\
+    \textcolor{lightgray}{rrrrrrrr}&a&!ab \\
+    \textcolor{lightgray}{dddddddd}&a&bra \\
+    \textcolor{lightgray}{!!!!!!!!}&a&bra \\
+    \textcolor{lightgray}{rrrrrrrr}&a&cad \\
+    \textcolor{lightgray}{cccccccc}&a&dab \\
+    \textcolor{lightgray}{aaaaaaaa}&b&ra! \\
+    \textcolor{lightgray}{aaaaaaaa}&b&rac \\
+    \textcolor{lightgray}{aaaaaaaa}&c&ada \\
+    \textcolor{lightgray}{aaaaaaaa}&d&abr \\
+    \textcolor{lightgray}{bbbbbbbb}&r&a!a \\
+    \textcolor{lightgray}{bbbbbbbb}&r&aca
+  \end{tabular}
   $\rightarrow \cdots \rightarrow$
@@
   \end{tt}
-  \caption{Recreating the rotation matrix from the last column.}
+  \caption{Recreating the rotation matrix from the last column: at
+    every iteration, we prepend the result of the BWT and sort on it.}
   \label{fig:recreate}
@@
 
-The choice of the last column is crucial, because as the ``original
-string'' in the stable column sort, it acts as the tie-breaker. The
-last column is the solution to the fixpoint equation
-\begin{equation*}
-  \var{hdsort}~(\var{prefix\_shift}~c~m) = m
-\end{equation*}
-where $\var{prefix\_shift}~c$ adds $c$ as the first column of a
-matrix, dropping the last column, and $m$ is the rotation matrix. This
-means as we repeatedly prefix $l$ and sort, we preserve the ``rest of
-the matrix'', even the parts that we have yet to reconstruct.
-
-We can think of the process of sorting the matrix as computing the
-solution to the fixpoint equation, subject to the constraints of the
-lexicographic ordering. For example, we know that the collection of
-all characters preceding an \texttt{a} in the sorted column must be
-some permutation of
-$[\texttt{c},~\texttt{d},~\texttt{r},~\texttt{r},~\texttt{!}]$,
-because those are the characters that precede \texttt{a}s in the
-original string. However, the order of these characters in the final
-string
-($[\texttt{r},~\texttt{d},~\texttt{!},~\texttt{r},~\texttt{c}]$) is
-important. An \texttt{r} must be first, because in the original string
-there is an \texttt{a} following an \texttt{r} which is itself
-followed by a \texttt{!}, the smallest character. Similarly, although
-the \texttt{d} and \texttt{!} both precede \texttt{a}s that are
-followed by \texttt{bra}s, the next character for the \texttt{d} is a
-\texttt{!}, and the next character for the \texttt{!} is a \texttt{c}.
-This requires that the \texttt{d} precede the \texttt{!} in the last
-column.
-
-Generally, there are constraints on the relative order of the letters
-in the column before the sorted column, and in each step we are
-ensuring that these constraints are not violated one more character
-deep in the string. By the time we have reached the last column, we
-have satisfied all the constraints.
+This intuition is related to the following lemma.
+\begin{restatable*}[lexsort\_rots\_hdsort]{lemma}{lexsrotrotshdsort}
+  For all lists $l$,
+  \begin{equation*}
+    (\var{hdsort} \comp
+    \var{map}~\var{rrot})~(\var{lexsort}~(\var{rots}~l)) =
+    \var{lexsort}~(\var{rots}~l)
+  \end{equation*}
+\end{restatable*}
+This lemma says that if we take the last column of the sorted rotation
+matrix and move it to the front, and then sort on it, we produce the
+same sorted rotation matrix that we started with. This lemma means
+that when we repeatedly prepend $s$ and sort by it, we are preserving
+the rest of the sorted rotation matrix, even the parts we have yet to
+reconstruct.
 
@@
   \var{lexsort} produces a permutation of its input, $l$ must be
-  in $\var{lexsort}~(\var{rots}~l)$ as well.
+  in $\var{lexsort}\ \allowbreak(\var{rots}~l)$ as well.
 \end{proof}
@@
 
-Now we want to construct a function \var{unbwt} such that for all $l$,
-\begin{equation}
-  \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l) = l
-  \label{eq:unbwt}
-\end{equation}
-If we could recreate the entire sorted rotation matrix from
-$\var{bwl}~l$, it would be simple to recover $l$ by indexing into
-the matrix at $\var{bwn}~l$.
-
-More precisely, suppose we had a function \var{recreate}
-such that
-\begin{equation}
-  \label{eq:recreate}
-  \var{recreate}~(\var{bwl}~l) = \var{lexsort}~(\var{rots}~l)
-\end{equation}
-Then, we could define \var{unbwt} as
-\begin{lstlisting}
-Definition unbwt (i : nat) (l : list A) : list A :=
-  nth i (recreate l) l.
-\end{lstlisting}
-and \cref{eq:unbwt} would follow directly from \cref{thm:bwn_correct}.
-
-We implement \var{recreate} as follows, introducing a parameter $j$ so
-that we can recursively recreate the first $j$ columns of the matrix,
-one column at a time. Here $l$ is meant to be the output of the BWT,
-the last column of the sorted rotation matrix.
+As outlined in \cref{sec:invert_bwt}, we recreate the sorted rotation
+matrix from the last column~$l$ using \var{recreate}. It takes a
+parameter $j : \nat$ so that we can recursively recreate the first $j$
+columns of the matrix, one column at a time.
 \begin{lstlisting}
@@
 
-Now we can redefine \var{unbwt} as
+The specificiation for \var{recreate} is that it recreates the first
+$j$ columns of the sorted rotation matrix. For this we need to define
+\var{cols}.
 \begin{lstlisting}
-Definition unbwt (i : nat) (l : list A) : list A :=
-  nth i (recreate (length l) l) l.
+Definition cols (j : nat)
+  : list (list A) -> list (list A)
+  := map (firstn j).
 \end{lstlisting}
 
-Since we recreate the matrix column by column, we can define an
-invariant based on \cref{eq:recreate} that uses $\var{cols}~j$, which
-takes the first $j$ columns of a matrix.
 \begin{restatable*}[recreate\_correct\_inv]{theorem}{recreatecorrectinv}
@@
 \end{restatable*}
-
-If we can prove this theorem, we will be able to prove that our
-implementation of \var{unbwt} is correct along the same lines as
-before. To do so, we need to prove a few other theorems.
 
@@
 rotation matrix is symmetric.
-\begin{theorem}[map\_rrot\_rots]
+\begin{lemma}[map\_rrot\_rots]
   For all lists $l$,
@@
 
-Now we can prove that the last column is indeed a fixpoint.
-\begin{lemma}[lexsort\_rots\_hdsort]
-  For all lists $l$,
-  \begin{equation*}
-    (\var{hdsort} \comp
-    \var{map}~\var{rrot})~(\var{lexsort}~(\var{rots}~l)) =
-    \var{lexsort}~(\var{rots}~l)
-  \end{equation*}
-\end{lemma}
+\lexsrotrotshdsort
 \begin{proof}
@@
 \end{proof}
-
-We need two more lemmas that describe the interaction between
-\var{cols} and $\var{map}~\var{rrot}$.
 
@@
 
-Now we are ready to prove the invariant:
+Now we are ready to prove the invariant.
 \recreatecorrectinv
@@
 \end{proof}
+
+Once we have the sorted rotation matrix, it is easy to recover the
+original string by indexing into the matrix at the correct row.
+\begin{lstlisting}
+Definition unbwt (i : nat) (l : list A) : list A :=
+  nth i (recreate (length l) l) l.
+\end{lstlisting}
 
@@
 Since in $\var{unbwt}~i~l$, we want the $i$th row of the matrix, we
-can eliminate the transpose by replacing the $\var{nth}~i$ with a
+can eliminate the transpose by replacing the $\var{nth}~i$ with
 $\var{map}~\var{nth}~i$ Assuming we already have calculated $p$,
@@
 \end{lstlisting}
-This is justified via the
-following theorem:
+This is justified via the following theorem.
 \begin{theorem}
@@
 
+\begin{theorem}[SortedLocal\_iff]
+  \label{thm:sortedlocal_iff}
+  For all lists $l$,
+  \begin{equation*}
+    \var{Sorted}~l \iff \var{SortedLocal}~l
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  Both directions follow by induction on $l$, making use of the
+  transitivity of $\le$.
+\end{proof}
+
 \sortedixiff*
@@
   ($\implies$) We have to prove that $(a :: l')_i \le (a :: l')_j$ given
-  $i \le j$ and $\var{Sorted}~l$. We reason by cases on $i$ and $j$:
+  $i \le j$ and $\var{Sorted}~(a::l)$. We reason by cases on $i$ and $j$:
   \begin{itemize}
@@
   ($\impliedby$) We have $\var{SortedIx}\ \allowbreak(a :: l')$, and
-  we need to show $\var{Sorted}\ \allowbreak(a :: l')$ given the
-  induction hypothesis $\var{SortedIx}~l' \implies \var{Sorted}~l'$.
-  By \var{SortedLocal\_iff}, we can just prove
+  we need to show $\var{Sorted}\ \allowbreak(a :: l')$. By
+  \cref{thm:sortedlocal_iff}, we can just prove
   $\var{SortedLocal}\ \allowbreak(a :: l')$. If $l' = []$, we can
   apply $\var{SortedLocal\_1}$, so assume $l' = b :: l''$. To apply
-  \var{SortedLocal\_cons} we need to show
-  $\var{SortedLocal}\ \allowbreak(b :: l'')$, which we have by the
-  induction hypothesis and $\var{SortedIx}\ \allowbreak(a :: b ::
-  l'')$, and that $a \le b$. That follows because $a = (a :: b ::
-  l'')_0$ and $b = (a :: b :: l'')_1$, so by
+  \var{SortedLocal\_cons} we need to show that $a \le b$. That follows
+  because $a = (a :: b :: l'')_0$ and $b = (a :: b :: l'')_1$, so by
   $\var{SortedIx}\ \allowbreak(a :: b :: l'')$, we are done.
@@
 \label{appendix:perm}
+
+In this section we sometimes use the notation $\eqperm$ to chain
+\var{Permutation} relationships transitively.
 
@@
 
-We can define the preimage of an index under a permutation function.
+We can define the iamge and preimage of an index under a permutation
+function.
 \begin{lstlisting}
+Definition image (p : list nat) i := nth i p 0.
 Definition preimage (p : list nat) i := findIndex i p.
 \end{lstlisting}
+We will use the notation $p(i)$ for $\var{image}~i$ and $p^{-1}(i)$
+for $\var{preimage}~i$. It should be clear from context whether $p(x)$
+denotes $\var{image}~x$ or $\var{apply}~x$, and when it is not we will
+use \var{image} and \var{apply} explicitlity.
 
 We will not prove it here, but $\var{preimage}~p$ and $\var{image}~p$
-are both of bijections of type $\mathbb{N}_{<n} \to \mathbb{N}_{<n}$ and
-are inverses of each other. We also have the following two theorems:
+are both of bijections of type $\nat_{<n} \to \nat_{<n}$ and are
+inverses of each other. We also have the following two theorems:
 \begin{theorem}[nth\_preimage\_apply, nth\_image\_apply]
@@
 \begin{proof}
-  The first equation (\var{nth\_preimage\_apply}) holds because:
+  The first equation (\var{nth\_preimage\_apply}) holds because
   \begin{align*}
@@
   \end{align*}
-  and the second follows from this:
+  and the second follows from the first.
   \begin{equation*}
@@
 
-  We reason as follows, using the notation $\eqperm$ to
-  chain together \var{Permutation}s:
   \begin{align*}
@@
   | h :: t =>
-    if le_dec x h then x :: h :: t else h :: insert x t
+    if le_dec x h
+    then x :: h :: t
+    else h :: insert x t
   end.
@@
 we can think of it as permuting the original indices $\iota =
-[0,~1,~\ldots,~n-1]$. If we rephrase \cref{eq:compose_apply} the effect on
-on $I$, we get
+[0,~1,~\ldots,~n-1]$. We can specialize \cref{eq:compose_apply} in terms of
+$\iota$.
 \begin{equation*}
@@
 
-We also need to show that \var{apply} distributes over \var{combine}:
+We also need to show that \var{apply} distributes over \var{combine}.
 \begin{theorem}[apply\_combine]
@@
   We can use \var{rem\_PermFun\_correct} to simplify both sides of the
-  equation. We have that the heads of both lists must are equal:
+  equation. We have that the heads of both lists must are equal.
   \begin{equation*}
@@
 
-Now we can prove the main theorem:
+Now we can prove the main theorem.
 \begin{theorem}[compose\_apply]

### 2019-10-19 draft_paper/draft_paper.tex
Fix preimage/image issue

@@
 
-A permutation function on $n$ elements maps indices to indices, and so
-has type $\nat_{<n} \to \nat_{<n}$. In Coq, we can represent a finite
-function as a list of its outputs: $f : \nat_{<n} \to A$ is represented
-by the list $[f(0),~f(1),\allowbreak\ldots,~f(n-1)]$. Applied to
-permutations, this means that the permutation~$p$ mentioned above
-would be represented by $[2,~1,~0]$. This is consistent with
-``one-line'' notation for permutations, in which a permutation on $n$
-elements is represented by its result when applied to the
-list~$[0,~2,\ldots,n-1]$. Now we can define the action of a permutation on
-a list.
+One natural representation for a permutation function~$p$ is a list of
+natural numbers, where each element~$i$ of $p$ stands in for the $i$th
+element of the original list. For example, the representation of the
+permutation mapping $l =
+[\var{zebra},\ \allowbreak\var{appliance},\ \allowbreak\var{kite}]$ to
+$l' =
+[\var{appliance},\ \allowbreak\var{kite},\ \allowbreak\var{zebra}]$
+would be $[1, 2, 0]$, because $\var{appliance} = l_1$, $\var{kite} =
+l_2$ and $\var{zebra} = l_0$. So, the mathematical permutation
+function~$p$ on $n$ elements is represented by the list
+$[p^{-1}(0),\ \allowbreak p^{-1}(1),\ \allowbreak,\ldots,\ \allowbreak
+  p^{-1}(n-1)]$. This gives us a very elegant definition for the
+action of a permutation on lists.
 \begin{equation*}
@@
 \end{equation*}
-We implement this in Coq using an \var{apply} function. The definition
-is slightly complicated by the fact that \var{nth} takes an extra
-parameter to return in case $i$ is out of bounds.
+We implement this in Coq with \var{apply}. The definition is slightly
+complicated by the fact that \var{nth} takes an extra parameter to
+return in case $i$ is out of bounds.
 \begin{lstlisting}
@@
 \begin{definition}[StablePermFun]
+  \label{def:stablepermfun}
   A permutation function $p$ is a \textit{stable} permutation function
@@
 \begin{lstlisting}
-Definition image (p : list nat) i := nth i p 0.
+Definition image (p : list nat) i := findIndex i p.
 \end{lstlisting}
@@
 
-We can define the iamge and preimage of an index under a permutation
+We can define the image and preimage of an index under a permutation
 function.
 \begin{lstlisting}
-Definition image (p : list nat) i := nth i p 0.
-Definition preimage (p : list nat) i := findIndex i p.
+Definition image (p : list nat) i := findIndex i p.
+Definition preimage (p : list nat) i := nth i p 0.
 \end{lstlisting}
@@
   \begin{align*}
-    (\var{apply}~p~l)_{p^{-1}(i)} = l_i && l_{p(i)} = (\var{apply}~p~l)_i
+    (\var{apply}~p~l)_{p(i)} = l_i && l_{p^{-1}(i)} = (\var{apply}~p~l)_i
   \end{align*}
@@
   \begin{align*}
-       (\var{apply}~p~l)_{p^{-1}(i)}
+       (\var{apply}~p~l)_{p(i)}
     &= (\var{map}~(\lambda j.~l_j)~p)_{\var{findIndex}~i~p} \\
@@
       (\var{apply}~p~l)_i
-    = (\var{apply}~p~l)_{p^{-1}(p(i))}
-    = l_{p(i)}
+    = (\var{apply}~p~l)_{p(p^{-1}(i))}
+    = l_{p^{-1}(i)}
     \qedhere
@@
 \label{appendix:stability}
+
+We defined \var{StablePermFun} in terms of the indices of the original
+list~$l$ (\cref{def:stablepermfun}), but we just as easily could have
+defined it in terms of the indices of the permuted list~$l'$.
+
+\begin{definition}[StablePermFun\_preimage]
+  A permutation function $p$ is a \textit{stable} permutation function
+  on some list $l$ when for all indices $i < j$ where $p(l)_i \equiv p(l)_j$, we
+  also have $p^{-1}(i) < p^{-1}(j)$.
+\end{definition}
+
+\begin{theorem}[StablePermFun\_iff]
+  For all lists~$l$ and $p : \var{list}~\nat$,
+  \begin{equation*}
+    \var{StablePermFun}~l~p \iff \var{StablePermFun\_preimage}~l~p
+  \end{equation*}
+\end{theorem}
+\begin{proof}
+  Both directions are very similar, we will prove just the $\implies$
+  direction here.
+
+  We have that $\var{StablePermFun}~l~p$, and we have an $i$ and $j$
+  such that $i < j$ and $p(l)_i \equiv p(l)_j$. We must show that
+  $p^{-1}(i) < p^{-1}(j)$. For the sake of contradiction, suppose
+  $p^{-1}(i) \ge p^{-1}(j)$. If $p^{-1}(i) = p^{-1}(j)$, then $i = j$,
+  contradicting $i < j$, so suppose $p^{-1}(i) > p^{-1}(j)$. Then,
+  since $p(l)_i \equiv p(l)_j$ implies $l_{p^{-1}(i)} \equiv l_{p^{-1}(j)}$,
+  from $\var{StablePermFun}~l~p$ we know that $p(p^{-1}(i)) >
+  p(p^{-1}(i))$, and therefore that $i > j$. This is also a
+  contradiction, so it must be that $p^{-1}(i) < p^{-1}(j)$.
+\end{proof}
 

### 2019-10-20 draft_paper/draft_paper.tex
Fix a few spelling errors

@@
 This intuition is related to the following lemma.
-\begin{restatable*}[lexsort\_rots\_hdsort]{lemma}{lexsrotrotshdsort}
+\begin{restatable*}[lexsort\_rots\_hdsort]{lemma}{lexsortrotshdsort}
   For all lists $l$,
@@
 
-The specificiation for \var{recreate} is that it recreates the first
+The specification for \var{recreate} is that it recreates the first
 $j$ columns of the sorted rotation matrix. For this we need to define
@@
 
-\lexsrotrotshdsort
+\lexsortrotshdsort
 \begin{proof}
@@
   \var{rrot}$ permutes the rotation matrix right before it is sorted
-  by \var{lexsrot}, so it doesn't change the final result.
+  by \var{lexsort}, so it doesn't change the final result.
 \end{proof}
@@
 denotes $\var{image}~x$ or $\var{apply}~x$, and when it is not we will
-use \var{image} and \var{apply} explicitlity.
+use \var{image} and \var{apply} explicitly.
 

### 2019-10-20 draft_paper/draft_paper.tex
Rephrase intro

@@
 
-There has not been much work on formally verifying stable sorts, or
-rather, the stability of stable sorts. Some work has been done on
-verifying the stability of mergesort
-\cite{leroy_2018,Sternagel2013,Leino2015,deGouw2014}, but we do not
-know of any formally verified implementation of a radix sort that
-includes stability.
-
-In addition, to our knowledge no one has produced a formally verified
-implementation of the BWT. Bird et~al.\ \cite{alg-of-prog,birdmu}
-provide an excellent derivation of the BWT and its inverse based on
-the technique of program calculation, in which programs are calculated
-from their specifications using algebraic transformations. However,
-their derivation leaves a few crucial lemmas unproven, including the
+Bird et~al.\ \cite{alg-of-prog,birdmu} provide an excellent derivation
+of the BWT and its inverse based on the technique of program
+calculation, in which programs are calculated from their
+specifications using algebraic transformations. However, their
+derivation leaves a few crucial lemmas unproven, including the
 correctness of LSD radix sort.
 
-In this paper, we develop more fully the theory of stable sorts: we
-give multiple definitions for permutations and stable permutations and
-prove them equivalent. This allows us to choose whichever definition
-makes the proof easiest. We also prove that any two sorts are
-interchangeable.
+Although there is a line of work on mergesort stability proofs
+\cite{leroy_2018,Sternagel2013,Leino2015,deGouw2014}, these proofs are
+not directly applicable to radix sort. In fact, the correctness of LSD
+radix sort remains unproven and more generally, nobody has contributed
+a general theory for reasoning about stable sorts.
+
+In this paper, we develop such a theory: we give multiple definitions
+for permutations and stable permutations and prove them equivalent.
+This allows us to choose whichever definition makes the proof easiest.
+We also prove that any two sorts are interchangeable.
 \begin{restatable*}[StableSort\_unique]{corollary}{stablesortunique}
